### 2009-03-30 eisa/inst/doc/ISA_tutorial.Rnw
Add the biology related files as a separate package.

\documentclass{article}
\usepackage{ragged2e}

\newcommand{\Rfunction}[1]{\texttt{#1}}
\newcommand{\Rpackage}[1]{\texttt{#1}}
\newcommand{\Rclass}[1]{\texttt{#1}}
\newcommand{\filename}[1]{\texttt{#1}}

\begin{document}

\title{The Iterative Signature Algorithm}
\author{G\'abor Cs\'ardi}
\maketitle

\RaggedRight

\section{Introduction}

The Iterative Signature Algorithm (ISA) is biclustering method. The
input of a biclustering method a matrix and its output is a set of
biclusters that fulfill some criteria; a bicluster is a block of the
potentially reordered input matrix.

Most commonly, this algorithm is used on microarray expression data,
to find gene sets that are coexpressed across a subset of the original
samples. In the original ISA paper the biclusters are called
transription modules (TM), we will often refer them under this name in
the following.

In this tutorial we will use the data in the \Rpackage{ALL}
package. This is a data set from a clinical trial in acute
lymphoblastic leukemia and it contains 128 samples altogether.

\section{Preparing the data}

\subsection{Loading the data}

First, we load the required packages and the data to analyze.
<<>>=
library(isa)
library(ALL)
library(hgu95av2.db)
data(ALL)
@

\subsection{Filtering the genes}

<<>>=
library(genefilter)
f1 <- pOverA(0.25, log2(100))
f2 <- function(x) (IQR(x) > 0.5)
ff <- filterfun(f1, f2)
selected <- genefilter(ALL, ff)
ALL2 <- ALL[selected,]
@

\subsection{Generating seeds for the ISA}

<<>>=
g <- generate.seeds(count=100, length=nrow(ALL2), gs=2/nrow(ALL2))
@

\subsection{Normalize the data}

<<>>=
ALL2.normed <- isa.normalize(ALL2)
@

\section{Running the ISA}

<<>>=
thr.gene <- 2.7
thr.cond <- 1.4
@

<<>>=
biclusts <- isa(ALL2.normed, thr.gene=thr.gene, thr.cond=thr.cond)
@

\section{Visualizing the results}

\section{Session information}

The version number of R and packages loaded for generating this
vignette were:

<<results=tex,echo=FALSE>>=
gsub(";", "; ", gsub("_", "\\\\_", sub("Locale:[ ]*\\\\verb\\|(.*)\\|", 
                                       "\\\\texttt\\{\\1\\}", 
                                       toLatex(sessionInfo()))))
@

\end{document}

### 2009-03-30 eisa/R/CHR.R

@@ -1,0 +1,301 @@
+
+
+##################################################
+## CHRListHyperGParams
+##################################################
+
+setClass("CHRListHyperGParams",
+         representation(conditional="logical"),
+         contains="HyperGParams",
+         prototype=prototype(categoryName=c("CHR", "List"),
+           conditional=FALSE))
+
+##################
+## makeValidParams
+
+setMethod("makeValidParams", "CHRListHyperGParams",
+          function(object) {
+            ## TODO: proper checks
+            if (!is.list(object@geneIds)) {
+              object@geneIds <- list(object@geneIds)
+            }
+            if (object@conditional) {
+              stop("Conditional CHR test is not implemented yet")
+            }
+            if (object@testDirection != "over") {
+              stop("Only overrepresentation test is implemented yet")
+            }
+            object
+          })
+
+##################
+## conditional
+
+setMethod("conditional", "CHRListHyperGParams", function(r) r@conditional)
+
+##################
+## conditional<-
+
+setReplaceMethod("conditional", c("CHRListHyperGParams", "logical"),
+                 function(r, value) {
+                     if (is.na(value))
+                       stop("value must be TRUE or FALSE")
+                     r@conditional <- value
+                     r
+                 })
+
+##################
+## categoryToEntrezBuilder
+## Create a mapping from the categories to the Entrez ids
+
+setMethod("categoryToEntrezBuilder",
+         signature(p="CHRListHyperGParams"),
+         function(p) {
+           keep.all <- switch(testDirection(p),
+                              over=FALSE,
+                              under=TRUE,
+                              stop("Bad testDirection slot"))
+
+           genes <- unique(unlist(geneIds(p)))
+
+           annotation <- p@annotation
+           org <- get(paste(sep="", annotation, "ORGANISM"))
+           org <- abbreviate(org, 2)
+           require(paste(sep="", "org.", org, ".eg.db"), character.only=TRUE)
+           CHR <- get(paste(sep="", "org.", org, ".egCHR"))
+           CHR <- as.matrix(toTable(CHR))
+
+           cat.eg <- tapply(as.character(CHR[,1]),
+                            as.character(CHR[,2]),
+                            c)
[…]

### 2009-03-30 eisa/R/GO.R

@@ -1,0 +1,379 @@
+
+##################################################
+## GOListHyperGParams
+##################################################
+
+setClass("GOListHyperGParams",
+         representation(ontology="character",
+                        conditional="logical"),
+         contains="HyperGParams",
+         prototype=prototype(categoryName=c("GO", "List"),
+           conditional=FALSE))
+
+##################
+## makeValidParams
+
+setMethod("makeValidParams", "GOListHyperGParams",
+          function(object) {
+            ## TODO: proper checks
+            if (!is.list(object@geneIds)) {
+              object@geneIds <- list(object@geneIds)
+            }
+            if (object@conditional) {
+              stop("Conditional GO test is not implemented yet")
+            }
+            if (object@testDirection != "over") {
+              stop("Only overrepresentation test is implemented yet")
+            }
+            object
+          })
+
+##################
+## ontology
+
+setMethod("ontology", "GOListHyperGParams", function(r) r@ontology)
+
+##################
+## ontology<-
+
+setReplaceMethod("ontology", c("GOListHyperGParams", "character"),
+                 function(r, value) {
+                     if (is.na(value) || length(value) != 1)
+                       stop("value must be a length one character vector")
+                     r@ontology <- value
+                     r
+                 })
+
+##################
+## conditional
+
+setMethod("conditional", "GOListHyperGParams", function(r) r@conditional)
+
+##################
+## conditional<-
+
+setReplaceMethod("conditional", c("GOListHyperGParams", "logical"),
+                 function(r, value) {
+                     if (is.na(value))
+                       stop("value must be TRUE or FALSE")
+                     r@conditional <- value
+                     r
+                 })
+
+##################
+## categoryToEntrezBuilder
+## Create a mapping from the categories to the Entrez ids
+
+setMethod("categoryToEntrezBuilder",
+         signature(p="GOListHyperGParams"),
+         function(p) {
+           keep.all <- switch(testDirection(p),
+                              over=FALSE,
+                              under=TRUE,
+                              stop("Bad testDirection slot"))
+
+           db <- do.call(paste(p@annotation, sep="_", "dbconn"), list())
[…]

### 2009-03-30 eisa/R/KEGG.R

@@ -1,0 +1,297 @@
+
+##################################################
+## KEGGListHyperGParams
+##################################################
+
+setClass("KEGGListHyperGParams",
+         representation(conditional="logical"),
+         contains="HyperGParams",
+         prototype=prototype(categoryName=c("KEGG", "List"),
+           conditional=FALSE))
+
+##################
+## makeValidParams
+
+setMethod("makeValidParams", "KEGGListHyperGParams",
+          function(object) {
+            ## TODO: proper checks
+            if (!is.list(object@geneIds)) {
+              object@geneIds <- list(object@geneIds)
+            }
+            if (object@conditional) {
+              stop("Conditional KEGG test is not implemented yet")
+            }
+            if (object@testDirection != "over") {
+              stop("Only overrepresentation test is implemented yet")
+            }
+            object
+          })
+
+##################
+## conditional
+
+setMethod("conditional", "KEGGListHyperGParams", function(r) r@conditional)
+
+##################
+## conditional<-
+
+setReplaceMethod("conditional", c("KEGGListHyperGParams", "logical"),
+                 function(r, value) {
+                     if (is.na(value))
+                       stop("value must be TRUE or FALSE")
+                     r@conditional <- value
+                     r
+                 })
+
+##################
+## categoryToEntrezBuilder
+## Create a mapping from the categories to the Entrez ids
+
+setMethod("categoryToEntrezBuilder",
+         signature(p="KEGGListHyperGParams"),
+         function(p) {
+           keep.all <- switch(testDirection(p),
+                              over=FALSE,
+                              under=TRUE,
+                              stop("Bad testDirection slot"))
+
+           geneIds <- unique(unlist(geneIds(p)))
+           lib <- annotation(p)
+           isORGEG = grep("org.*.eg", lib)
+           if( length(isORGEG) > 0 )
+             kegg2allprobes <- Category:::getDataEnv("PATH2EG", lib)
+           else
+             kegg2allprobes <- Category:::getDataEnv("PATH2PROBE", lib)
+           probeAnnot <- Category:::getKeggToProbeMap(kegg2allprobes)
+           Category:::probeToEntrezMapHelper(probeAnnot, geneIds, p@datPkg, universeGeneIds(p),
[…]

### 2009-03-30 eisa/R/autogen.R

@@ -1,0 +1,1056 @@
+
+autogen.table <- function(nm, isares, target.dir,
+                          modules=seq_len(ncol(isares$genes)),
+                          template=system.file("autogen", package="isa"),
+                          GO=NULL, KEGG=NULL, miRNA=NULL, CHR=NULL, DBD=NULL,
+                          htmltitle=NULL, notes=NULL, seed=NULL) {
+
+  chip <- isares$rundata$annotation
+  library(paste(sep="", chip, ".db"), character.only=TRUE)
+  organism <- get(paste(sep="", chip, "ORGANISM"))
+  short.organism <- organism
+
+  ################################################
+  ## GO
+
+  db <- GO_dbconn()
+  query <- "SELECT go_id, term, definition FROM go_term"
+  go.terms <- dbGetQuery(db, query)
+  rownames(go.terms) <- go.terms[,1]
+  go.terms <- go.terms[,-1]
+  
+  options(digits=2)
+  
+  f <- function(obj, pvalue=0.05) {
+    
+    if (class(obj)=="NULL") {
+      return("")
+    } else {
+      if (length(obj$Pvalue)==0) { return("") }
+      pval <- obj$Pvalue[1]
+      if (pval > pvalue) { return("") }
+      uc <- obj$Size[1]
+      ec <- obj$ExpCount[1]
+      gc <- obj$Count[1]
+      ca <- rownames(obj)[1]
+    }
+    
+    nn <- go.terms[ca,]$term
+    paste(sep="", nn, " <span class=\"pvalue\"><br/>(", format(pval), "/", format(ec),
+          "/", gc, "/", uc, ")</span>")
+  }
+  
+  print("  -- GO BP")
+  tables.BP <- lapply(GO[[1]]@reslist[modules], function(x) f(x))
+  print("  -- GO CC")
+  tables.CC <- lapply(GO[[2]]@reslist[modules], function(x) f(x))
+  print("  -- GO MF")
+  tables.MF <- lapply(GO[[3]]@reslist[modules], function(x) f(x))
+
+  g <- function(obj, pvalue=0.05) {
+    
+    if (class(obj)=="NULL") {
+      return("")
+    } else {
+      if (length(obj$Pvalue)==0) { return("") }
+      pval <- obj$Pvalue[1]
+      if (pval > pvalue) { return("") }
+      uc <- obj$Size[1]
+      ec <- obj$ExpCount[1]
+      gc <- obj$Count[1]
+      ca <- rownames(obj)[1]
+    }
+
[…]

### 2009-03-30 eisa/R/dbd.R

@@ -1,0 +1,349 @@
+
+
+##################################################
+## DBDListHyperGParams
+##################################################
+
+setClass("DBDListHyperGParams",
+         representation(conditional="logical"),
+         contains="HyperGParams",
+         prototype=prototype(categoryName=c("DBD", "List"),
+           conditional=FALSE))
+
+##################
+## makeValidParams
+
+setMethod("makeValidParams", "DBDListHyperGParams",
+          function(object) {
+            ## TODO: proper checks
+            if (!is.list(object@geneIds)) {
+              object@geneIds <- list(object@geneIds)
+            }
+            if (object@conditional) {
+              stop("Conditional DBD test is not implemented yet")
+            }
+            if (object@testDirection != "over") {
+              stop("Only overrepresentation test is implemented yet")
+            }
+            object
+          })
+
+##################
+## conditional
+
+setMethod("conditional", "DBDListHyperGParams", function(r) r@conditional)
+
+##################
+## conditional<-
+
+setReplaceMethod("conditional", c("DBDListHyperGParams", "logical"),
+                 function(r, value) {
+                     if (is.na(value))
+                       stop("value must be TRUE or FALSE")
+                     r@conditional <- value
+                     r
+                 })
+
+##################
+## categoryToEntrezBuilder
+## Create a mapping from the categories to the Entrez ids
+
+setMethod("categoryToEntrezBuilder",
+         signature(p="DBDListHyperGParams"),
+         function(p) {
+           keep.all <- switch(testDirection(p),
+                              over=FALSE,
+                              under=TRUE,
+                              stop("Bad testDirection slot"))
+
+           genes <- unique(unlist(geneIds(p)))
+
+           ann <- p@annotation
+           org <- get(paste(sep="", ann, "ORGANISM"))
+           short.org <- abbreviate(org, 2)
+           if (short.org == "Mm") {
+             data(dbd.mm, package="isa")
+             dbd <- dbd.mm
+           } else if (short.org=="Hs") {
+             data(dbd.hs, package="isa")
+             dbd <- dbd.hs
+           } else {
[…]

### 2009-03-30 eisa/R/miRNA.R

@@ -1,0 +1,332 @@
+
+
+##################################################
+## miRNAListHyperGParams
+##################################################
+
+setClass("miRNAListHyperGParams",
+         representation(conditional="logical"),
+         contains="HyperGParams",
+         prototype=prototype(categoryName=c("miRNA", "List"),
+           conditional=FALSE))
+
+##################
+## makeValidParams
+
+setMethod("makeValidParams", "miRNAListHyperGParams",
+          function(object) {
+            ## TODO: proper checks
+            if (!is.list(object@geneIds)) {
+              object@geneIds <- list(object@geneIds)
+            }
+            if (object@conditional) {
+              stop("Conditional miRNA test is not implemented yet")
+            }
+            if (object@testDirection != "over") {
+              stop("Only overrepresentation test is implemented yet")
+            }
+            object
+          })
+
+##################
+## conditional
+
+setMethod("conditional", "miRNAListHyperGParams", function(r) r@conditional)
+
+##################
+## conditional<-
+
+setReplaceMethod("conditional", c("miRNAListHyperGParams", "logical"),
+                 function(r, value) {
+                     if (is.na(value))
+                       stop("value must be TRUE or FALSE")
+                     r@conditional <- value
+                     r
+                 })
+
+##################
+## categoryToEntrezBuilder
+## Create a mapping from the categories to the Entrez ids
+
+setMethod("categoryToEntrezBuilder",
+         signature(p="miRNAListHyperGParams"),
+         function(p) {
+           keep.all <- switch(testDirection(p),
+                              over=FALSE,
+                              under=TRUE,
+                              stop("Bad testDirection slot"))
+
+           genes <- unique(unlist(geneIds(p)))
+           
+           ann <- p@annotation
+           org <- get(paste(sep="", ann, "ORGANISM"))
+           short.org <- abbreviate(org, 2)
+           if (short.org == "Mm") {
+             data(miRNA.mm, package="isa")
+             miRNA <- miRNA.mm
+           } else if (short.org=="Hs") {
+             data(miRNA.hs, package="isa")
+             miRNA <- miRNA.hs
+           } else {
[…]

### 2009-03-30 eisa/R/viz.R

@@ -1,0 +1,536 @@
+
+gograph <- function(table, colbar.length=30, label.cex=1, GOGRAPHS=NULL,
+                    go.terms=NULL) {
+
+  require(GO.db)
+  require(igraph)
+
+  terms <- as.character(table[,1])
+  pval <- table[,2]
+
+  #######################
+  # Check which GO tree
+  
+  cats <- get(terms[1], GOTERM)@Ontology
+
+  #######################
+  # Create a list of terms
+  
+  if (is.null(go.terms)) {
+    db <- GO_dbconn()
+    query <- paste("SELECT go_id, term, definition FROM go_term ",
+                   "WHERE ontology=='", cats, "'",
+                   sep="")
+    go.terms <- dbGetQuery(db, query)
+    rownames(go.terms) <- go.terms[,1]
+    go.terms <- go.terms[,-1]
+  }
+    
+  #######################
+  # Create the igraph object if it was not passed as an argument
+  
+  if (is.null(GOGRAPHS)) {
+    query <- paste("SELECT id1.go_id, id2.go_id, par.relationship_type FROM",
+                   " go_", tolower(cats), "_parents AS par,",
+                   " go_term AS id1, go_term AS id2",
+                   " WHERE par._id==id1._id AND par._parent_id==id2._id",
+                   sep="")
+    db <- GO_dbconn()
+    parents <- dbGetQuery(db, query)
+    colnames(parents) <- c("from", "to", "relationship")
+    go.graph <- graph.data.frame(parents, directed=TRUE)
+  } else {
+    go.graph <- switch(cats, CC=GOGRAPHS$CC, BP=GOGRAPHS$BP, MF=GOGRAPHS$MF)
+  }
+  
+  #######################
+  # Create a subgraph containing all up-paths from the
+  # given vertices
+
+  vv <- which( V(go.graph)$name %in% terms )-1
+  root <- which(igraph::degree(go.graph, 1:vcount(go.graph)-1, mode="out")==0)-1
+  
+  vert <- numeric()
+  for (i in vv) {
+    vert <- c(vert, unlist(neighborhood(go.graph, vcount(go.graph), i, mode="out")))
+  }
[…]

### 2009-03-30 eisa/inst/autogen/mainlattice.js

@@ -1,0 +1,46 @@
+function showstuff(boxid){
+   document.getElementById(boxid).style.visibility="visible";
+   document.getElementById(boxid).style.display="block";
+}
+function hidestuff(boxid){
+   document.getElementById(boxid).style.visibility="hidden";
+   document.getElementById(boxid).style.display="none";
+}
+function isArray(testObject) {   
+    return testObject && !(testObject.propertyIsEnumerable('length')) && typeof testObject === 'object' && typeof testObject.length === 'number';
+}
+function QuickUpdate() {
+   var what=document.getElementById("quick.select").value;
+   document.getElementById("tooltip.select").value=what;   
+   document.getElementById("main.select").value=what;
+   document.getElementById("color.select").value=what;
+   document.getElementById("right.select").value=what;
+   MyUpdate([ 'tooltip', 'main', 'color', 'right' ]);
+}
+function MyUpdate(args){
+   if (!isArray(args)) {
+     args=[ args ];
+   }
+   var ii=0;
+   while (args.length > ii) {
+     var what=args[ii];
+     var elem=document.getElementById(what+'.select');
+     var value=elem.value;
+     var opts=elem.options;
+     var i=0;
+     while (opts.length > i) {
+       hidestuff("main"+what+opts[i].value);
+       i++;
+     }
+     showstuff("main"+what+value);
+     ii++;
+   }
+}
+var foobar= {
+   init: function() {
+     MyUpdate([ 'main', 'tooltip', 'left', 'right', 'color' ]);
+   }
+}
+if (document.addEventListener) {
+  document.addEventListener("DOMContentLoaded", foobar.init, null);
+}

### 2009-03-31 isa/R/option.R
Updated option handling.

@@ -1,11 +1,12 @@
 
+if (!exists("isa.options")) { isa.options <- new.env() }
+
 isa.option <- function(...) {
   args <- list(...)
   nam <- names(args)
 
-  env <- get(".isa.options", envir=.GlobalEnv)
   if (length(args) == 0) {
-    return(as.list(env))
+    return(as.list(isa.options))
   }
   
   if (is.null(nam)) {
@@ -16,16 +17,15 @@ isa.option <- function(...) {
     if (!is.character(args[[1]]) || length(args[[1]]) != 1) {
       stop("Expected character of length one")
     }
-    return(get(args[[1]], env))
+    return(isa.options[[ args[[1]] ]])
   } else {
     ## set, everything must be named
     if (any(nam=="")) {
       stop("Some options are not named")
     }
     for (i in seq_along(nam)) {
-      assign(nam[i], args[[i]], envir=env)
+      isa.options[[ nam[i] ]] <- args[[i]]
     }
-    assign(".isa.options", env, envir=.GlobalEnv)
-    return(invisible(env))
+    return(invisible(isa.options))
   }
 }

### 2009-03-31 isa/R/zzz.R

@@ -1,14 +1,10 @@
 .onLoad <- function(dir, package) {
   library.dynam("isa", package, dir, local=FALSE);
-  if (!exists(".isa.options", envir=.GlobalEnv)) {
-    env <- new.env()
-    
-    assign("verbose", FALSE, envir=env)
-    assign("status.function", function(...) { isa.status.default(...) },
-           envir=env)
-    
-    assign(".isa.options", env, envir=.GlobalEnv)    
-  }
+
+  ## Default ISA options
+  isa.options[["verbose"]] <-  FALSE
+  isa.options[["status.function"]] <- function(...) isa.status.default(...)
+  
 }
 
 .onUnload <- function(libpath) {

### 2009-03-31 isa/inst/doc/ISA_tutorial.Rnw
Removed Bioconductor specific parts from the vignette to make R CMD check finish.

--- 2009-03-30 eisa/inst/doc/ISA_tutorial.Rnw
@@
 
-First, we load the required packages and the data to analyze.
-<<>>=
-library(isa)
-library(ALL)
-library(hgu95av2.db)
-data(ALL)
-@
-
-\subsection{Filtering the genes}
-
-<<>>=
-library(genefilter)
-f1 <- pOverA(0.25, log2(100))
-f2 <- function(x) (IQR(x) > 0.5)
-ff <- filterfun(f1, f2)
-selected <- genefilter(ALL, ff)
-ALL2 <- ALL[selected,]
-@
-
-\subsection{Generating seeds for the ISA}
-
-<<>>=
-g <- generate.seeds(count=100, length=nrow(ALL2), gs=2/nrow(ALL2))
-@
-
-\subsection{Normalize the data}
-
-<<>>=
-ALL2.normed <- isa.normalize(ALL2)
-@
-
-\section{Running the ISA}
-
-<<>>=
-thr.gene <- 2.7
-thr.cond <- 1.4
-@
-
-<<>>=
-biclusts <- isa(ALL2.normed, thr.gene=thr.gene, thr.cond=thr.cond)
-@
-
 \section{Visualizing the results}

### 2009-03-31 isa/R/insilico.R
Removed unneeded argument of in.silico function.

@@ -1,5 +1,5 @@
 
-isa.in.silico <- function(num.rows=300, num.cols=50, num.fact=3, fact.per.row=1,
+isa.in.silico <- function(num.rows=300, num.cols=50, num.fact=3,
                           mod.row.size=round(.5*num.rows/num.fact),
                           mod.col.size=round(.5*num.cols/num.fact),
                           noise=0.1,

### 2009-03-31 isa/R/isa3.R
Fixed a bug in oscillation checking. (I hope this was not the reason of not having anything to oscillate.)

@@ -168,7 +168,7 @@ isa.iterate <- function(normed.data, row.seeds, col.seeds,
       osc <- osc & !conv
       
       if (any(osc)) {
-        mat <- cbind(mat[osc], attributes(mat, "match.length")[[1]][osc])
+        mat <- cbind(mat[osc], attr(mat, "match.length")[[1]][osc])
         mat <- sapply(seq(length=nrow(mat)), function(x) substr(fire[osc][x],
                             mat[x,1], mat[x,1]+mat[x,2]))
         mat <- sapply(mat, function(x) sum(utf8ToInt(x) == 58), USE.NAMES=FALSE )

### 2009-03-31 isa/R/isa3.R
Corrected a typo when column seeds were provided to isa.iterate.

@@ -308,7 +308,7 @@ isa.unique <- function(normed.data, isaresult, method=c("cor"),
 }
 
 
-isa.row.from.col <- function(normed.data, col.seeds, thr.col, direction) {
+isa.row.from.col <- function(normed.data, col.seeds, thr.row, direction) {
 
 
   Ec <- normed.data$Ec

### 2009-04-01 isa/inst/doc/ISA_tutorial.Rnw
Added a vignette index entry. R CMD check is clean now.
Started writing the isa vignette.

@@
+% \VignetteIndexEntry{The Iterative Signature Algorithm}
 \documentclass{article}
@@
 \newcommand{\Rclass}[1]{\texttt{#1}}
+\newcommand{\Rargument}[1]{\textsl{#1}}
 \newcommand{\filename}[1]{\texttt{#1}}
@@
 \begin{document}
+
+\setkeys{Gin}{width=\textwidth}
 
@@
 
-\RaggedRight
+% \RaggedRight
+
+\tableofcontents
+
+\section{For the impatient}
+
+To run the typical ISA work flow with default parameters on your data
+matrix, just load the \Rpackage{isa} package and call the
+\Rfunction{isa} function with your matrix as the single argument. The
+return value of \Rfunction{isa} is a named list, the members
+\texttt{rows} and \texttt{columns} contain the biclusters ISA have
+found. Every bicluster is given by a pair of columns from
+\texttt{rows} and \texttt{columns} (i.e. the first columns define the
+first bicluster, etc.) and the elements the biclusters are the
+non-zero elements in the columns of \texttt{rows} and
+\texttt{columns}.
+
+Please continue reading for a less dense tutorial.
 
@@
 
-The Iterative Signature Algorithm (ISA) is biclustering method. The
-input of a biclustering method a matrix and its output is a set of
-biclusters that fulfill some criteria; a bicluster is a block of the
-potentially reordered input matrix.
+The Iterative Signature Algorithm (ISA) is biclustering method. Its 
+input is a matrix and its output is a set of biclusters: blocks of the
+potentially reordered input matrix, that fulfill some predefined
+criteria. A biclustering algorithm tries to find
+blocks that are different from the rest of the matrix, e.g. the values
+covered by the bicluster are all above or below the background.
 
-Most commonly, this algorithm is used on microarray expression data,
-to find gene sets that are coexpressed across a subset of the original
-samples. In the original ISA paper the biclusters are called
-transription modules (TM), we will often refer them under this name in
-the following.
+The ISA is developed to find biclusters (or modules as most of the ISA
+papers call them) that have correlated rows and columns. More
+precisely, the rows in the bicluster need to be only correlated across
+the columns of the bicluster and vice versa.
 
-In this tutorial we will use the data in the \Rpackage{ALL}
-package. This is a data set from a clinical trial in acute
-lymphoblastic leukemia and it contains 128 samples altogether.
+Fig.~\ref{fig:twoclusters} shows possibly the simplest
+example of a (rather artificial) data matrix with very strong modular
+structure. It is a $20\times 20$ matrix and has two correlated blocks,
+each of them of size $10\times 10$.
+\begin{figure}
+<<echo=FALSE,fig=TRUE,width=6,height=3>>=
+library(isa)
+library(Matrix)
+data <- data2 <- isa.in.silico(20,20,2,10,10)[[1]]
+perm1 <- sample(seq_len(nrow(data)))
+data2 <- data2[perm1,]
+perm2 <- sample(seq_len(ncol(data)))
+data2 <- data2[,perm2]
 
-\section{Preparing the data}
+alldata <- structure(c(data2, data), dim=c(nrow(data),ncol(data),2), 
+                     dimnames=list(NULL,NULL,c("Original matrix",
+                     "Reordered matrix")))
 
-\subsection{Loading the data}
+rx <- range(alldata, finite = TRUE)
+nn <- 100
+n0 <- min(nn, max(0, round((0 - rx[1])/(rx[2] - rx[1]) * nn)))
+col.regions <- c(colorRampPalette(c("blue3", "gray80"))(n0), 
+                 colorRampPalette(c("gray75", "red3"))(nn - n0))
 
-\section{Visualizing the results}
+lp <- levelplot(alldata, xlab="", ylab="", 
+                scale=list(y=list(draw=FALSE),
+                  x=list(draw=FALSE)),
+                col.regions=col.regions, between=list(x=0.5))
+print(lp)
+@ 
+\caption{An artificial data matrix, on the left. On the right the
+  reordered data matrix with two blocks.}
+\label{fig:twoclusters}
+\end{figure}
+
+\section{How ISA works}
+
+Before showing an actual ISA tool chain, a few words about how the
+algorithm works are in order.
+
+\subsection{ISA iteration}
+
+ISA works in an iterative way. For an $E (m\times n)$
+input matrix it starts from seed vector $r_0$, which is
+typically a sparse 0/1 vector of length $m$. This defines a set of
+rows in $E$. Then $E'$ is multiplied by $r_0$ and the
+result is thresholded.
+
+The thresholding is an important step of the ISA, without thresholding
+ISA would be equivalent to a (not too effective) numerical singular
+value decomposition (SVD). Currently thresholding is done by
+calculating the mean and standard deviation of the vector and keeping
+only elements that are further than a given number of standard
+deviations from the mean. Based on the \Rargument{direction} parameter,
+this means 1) keeping values that are significantly higher than the
+mean (``\emph{up}''), significantly lower (``\emph{down}'') or both
+(``\emph{updown}'').
+
+The thresholded vector $c_0$ is the (column)
+\emph{signature} of $r_0$. Then the (row) signature of
+$c_0$ is calculated, $E$ is multiplied by $c_0$ and then thresholded
+to get $r_1$.
+
+This iteration is performed until it converges, i.e. $r_i$
+and $r_{i-1}$ are \emph{close}, and $c_i$ and
+$c_{i-1}$ are also close. The convergence criteria,
+i.e. what \emph{close} means is by default defined by high Pearson
+correlation.
+
+It is very possible that the ISA finds the same modules more than once;
+two or more seeds might converge to the same module. The function
+\Rfunction{isa.unique} eliminates every module from the result of 
+\Rfunction{isa.iterate} that is very similar (in terms of
+Pearson correlation) to the one that was already found before it.
+
+It might be also apparent from the description of ISA, that the
+biclusters are soft, i.e. they might have an overlap in their rows,
+columns, or both. It is also possible that some rows and/or
+columns of the input matrix are not found to be part of any ISA
+biclusters. Depending on the stringency parameters in the
+thresholding, it might even happen that ISA does not find any
+biclusters.
+
+\subsection{Parameters}
+
+The two main parameters of ISA are the two thresholds (one for the
+rows and one for the columns). They basically define the stringency of
+the modules. If the row threshold is high, then the modules will have
+very similar rows. If it is mild, then modules will be bigger, with
+less similar rows than in the first case.
+
+\subsection{Random seeding and smart seeding}
+
+By default (i.e. if the \Rfunction{isa} function is used) the ISA is
+performed from random sparse starting seeds, generated by the
+\Rfunction{generate.seeds} function. This way the algorithm is 
+completely unsupervised, but also stochastic: it might give different
+results for different runs.
+
+It is possible to use non-random seeds as well, if you have some
+knowledge about the data or are interested in a particular subset of
+rows/columns, then you can feed in your seeds into the
+\Rfunction{isa.iterate} function directly. In this case the
+algorithm is deterministic, for the same seed you will always get the
+same results. This can be considered as a semi-supervised approach.
+
+\subsection{Normalization}
+
+On in silico data we observed that ISA has the best performance if the
+input matrix is normalized (see \Rfunction{isa.normalize}). The
+normalization produces two matrices: $E_r$ and
+$E_c$. $E_r$ is calculated by transposing $E$ and
+centering and scaling its rows (see \Rfunction{scale}). $E_c$ is
+calculated by centering and scaling the rows of $E$. $E_r$ is
+used to calculate the column signature of rows and $E_c$ is used
+to calculate the signature of the columns.
+
+It is possible to use another normalization, then the user is
+requested to supply the normalized input data in a named list,
+including the two matrices of appropriate
+dimensions. \texttt{Er} will be used for calculating the
+signature of the rows, \texttt{Ec} the signature of the
+columns. If you want to use the same matrix in both steps, then supply
+it twice, the first one transposed.
+
+\subsection{Robustness}
+As ISA is an unsupervised algorithm, it may very well find some
+modules, even if you feed in noise as an input matrix. To avoid these
+spurious modules we defined a robustness measure, a single number for
+a modules that gives how well the rows and the columns are
+correlated.
+
+It recommended that the user uses \Rfunction{isa.filter.robust} to
+run ISA on the scrambled input matrix with the same threshold
+parameters and then drop every module, which has a robustness score
+lower than the highest robustness score among modules found in the
+scrambled data.
+
+\section{A simple work flow}
+
+\section{A detailed work flow}
+
+\subsection{Preparing the data}
+
+\subsection{Running the ISA}
+
+\subsection{Robustness of biclusters, filtering the results}
+
+\subsection{Visualize the results}
+
+\section{Features of ISA}
+
+\subsection{Resilience to noise}
+
+\subsection{Finding overlapping biclusters}
 

### 2009-04-01 isa/inst/doc/ISA_tutorial.Rnw
Added references to isa vignette.

@@
 
+TODO
+
 \section{A detailed work flow}
+
+TODO
 
@@
 
+TODO
+
 \subsection{Running the ISA}
+
+TODO
 
@@
 
+TODO
+
 \subsection{Visualize the results}
+
+TODO
 
@@
 
+TODO
+
 \subsection{Resilience to noise}
 
+TODO
+
 \subsection{Finding overlapping biclusters}
+
+TODO
 
@@
 
+\bibliographystyle{apalike}
+
+\begin{thebibliography}{}
+
+\bibitem[Bergmann et~al., 2003]{isa}
+Bergmann, S., Ihmels, J., and Barkai, N. (2003).
+\newblock Iterative signature algorithm for the analysis of large-scale gene
+  expression data.
+\newblock {\em Phys Rev E Nonlin Soft Matter Phys}, page 031902.
+
+\bibitem[Cs\'ardi, 2009]{isapackage}
+Cs\'ardi, G. (Apr 1, 2009).
+\newblock {\em isa: The Iterative Signature Algorithm}.
+\newblock R package version 0.1.
+
+\bibitem[Ihmels, 2002]{sa}
+Ihmels, J., Friedlander, G., Bergmann, S., Sarig, O., Ziv, Y., 
+Barkai, N. (2002).
+\newblock Revealing modular organization in the yeast transcriptional
+  network.
+\newblock { \em Nat Genet }, page 370--7.
+
+\bibitem[Ihmels, 2004]{isamod}
+Ihmels, J., Bergmann, S., Barkai, N. (2004).
+\newblock Defining transcription modules using large-scale gene
+  expression data.
+\newblock { \em Bioinformatics}, page 1993--2003.
+
+\end{thebibliography}
+
 \end{document}

### 2009-04-27 isa/R/robustness.R
Fixed a bug, when robustness filtering is called with an empty rows/columns matrix.

@@ -40,6 +40,10 @@ isa.filter.robust.default <- function(data, normed.data, isares, perms=1,
     stop("Number of permutations must be non-negative")
   }
 
+  if (length(isares$rows) == 0) {
+    return(isares)
+  }
+  
   if (length(unique(isares$seeddata$thr.row)) != 1) {
     warning("Different row thresholds, using only the first one.")
   }

### 2009-04-27 isa/inst/doc/ISA_tutorial.Rnw
isa vignette (tutorial) improvements, almost finished.

@@
 \newcommand{\filename}[1]{\texttt{#1}}
+\newcommand{\listentry}[1]{\texttt{#1}}
 
@@
 \tableofcontents
+
+<<Roptions,echo=FALSE,results=hide>>=
+## Some options for R
+options(width=60)
+@
 
@@
 \begin{figure}
-<<echo=FALSE,fig=TRUE,width=6,height=3>>=
+<<toymatrix,echo=FALSE,fig=TRUE,width=6,height=3>>=
 library(isa)
@@
 ISA works in an iterative way. For an $E (m\times n)$
-input matrix it starts from seed vector $r_0$, which is
-typically a sparse 0/1 vector of length $m$. This defines a set of
-rows in $E$. Then $E'$ is multiplied by $r_0$ and the
-result is thresholded.
+input matrix it starts from a seed vector $r_0$, which is
+typically a sparse 0/1 vector of length $m$. The non-zero elements in
+the seed vector define a set of rows in $E$. Then the transposed of
+$E$, $E'$ is multiplied by $r_0$ and the result is thresholded.
 
@@
 ISA would be equivalent to a (not too effective) numerical singular
-value decomposition (SVD). Currently thresholding is done by
+value decomposition (SVD) algorithm. Currently thresholding is done by
 calculating the mean and standard deviation of the vector and keeping
@@
 deviations from the mean. Based on the \Rargument{direction} parameter,
-this means 1) keeping values that are significantly higher than the
-mean (``\emph{up}''), significantly lower (``\emph{down}'') or both
+this means keeping values that are significantly higher than the
+mean (``\emph{up}''), or keeping the ones that are significantly lower
+than the mean (``\emph{down}''); or keeping both
 (``\emph{updown}'').
@@
 
-This iteration is performed until it converges, i.e. $r_i$
-and $r_{i-1}$ are \emph{close}, and $c_i$ and
-$c_{i-1}$ are also close. The convergence criteria,
+This iteration is performed until it converges, i.e. $r_{i-1}$
+and $r_i$ are \emph{close}, and $c_{i-1}$ and
+$c_i$ are also close. The convergence criteria,
 i.e. what \emph{close} means is by default defined by high Pearson
@@
 \Rfunction{isa.iterate} that is very similar (in terms of
-Pearson correlation) to the one that was already found before it.
+Pearson correlation) to the one that was already found before.
 
@@
 biclusters. Depending on the stringency parameters in the
-thresholding, it might even happen that ISA does not find any
-biclusters.
+thresholding (i.e. how far the values should be from the mean), it
+might even happen that ISA does not find any biclusters.
 
@@
 very similar rows. If it is mild, then modules will be bigger, with
-less similar rows than in the first case.
+less similar rows than in the first case. The same applied to the
+column threshold and the columns of the modules.
 
@@
 
-It is possible to use another normalization, then the user is
+It is possible to use another normalization. In this case the user is
 requested to supply the normalized input data in a named list,
 including the two matrices of appropriate
-dimensions. \texttt{Er} will be used for calculating the
-signature of the rows, \texttt{Ec} the signature of the
+dimensions to the \Rfunction{isa.iterate} function.
+
+The \texttt{Er} entry of the list will be used for calculating the
+signature of the rows, \texttt{Ec} will be used for the signature of the
 columns. If you want to use the same matrix in both steps, then supply
 it twice, the first one transposed.
+
+\subsection{Row and column scores}
+
+In addition to finding biclusters in the input matrix, the ISA also
+assigns scores to the rows and columns, separately for each
+module. The scores are between minus one and one and they are by
+definition zero for the rows/columns that are not included in the
+module. For the non-zero entries, the further the score of a
+row/columns is from zero, the stronger the association between the
+row/column and the module. If the sign of two rows/columns are the
+same, then they are correlated, if they have opposite signs, then they
+are anti-correlated.
 
@@
 spurious modules we defined a robustness measure, a single number for
-a modules that gives how well the rows and the columns are
-correlated.
+a module that measures how well the rows and the columns are
+correlated. The robustness score is a generalization of the first
+singular value of the matrix. If there would be no thresholding, then
+the ISA would (almost always) converge to the leading SVD vector and
+the robustness score would be the corresponding singular value.
 
@@
 
-TODO
+The simplest way to use ISA on your data is by calling the
+\Rfunction{isa} function with your input matrix as the single
+arguments. This function uses random seeding and has three optional
+arguments: 
+\begin{description}
+  \item[\Rargument{thr.row}] A numeric vector, the row thresholds to
+    use. By default this is $1,1.5,2,2.5,3$. 
+  \item[\Rargument{thr.col}] A numeric vector, the column thresholds
+    to use. By default this is $1,1.5,2,2.5,3$.
+  \item[\Rargument{no.seeds}] The number of random seed vectors to
+    generate. By default 100 seeds are generated.
+\end{description}
+
+The \Rfunction{isa} function runs the ISA algorithm for all threshold
+combinations of the supplied row and column thresholds. For every
+single run the same set of random seeds are used.
+
+Let us see an example. We load the \Rpackage{isa} generate some in
+silico data with the \Rfunction{isa.in.silico} function first:
+<<insilico,cache=TRUE>>=
+data <- isa.in.silico(200,200,2,50,50)[[1]]
+@
+The function call above generates the data matrix in
+Fig.~\ref{fig:twoclusters}. The first two arguments define the size of
+the data matrix ($200\times 200$), the third the number of modules (two)
+and the fourth and fifth give the size of the modules, both are
+$50\times 50$ submatrices in this case. By default
+\Rfunction{isa.in.silico} creates non-overlapping modules with some
+background noise.
+
+\Rfunction{isa.in.silico} returns a list with a list with three
+elements, we are most interested in the first now, that is the actual
+artificial data matrix. The second and the entries in the list give
+the correct module memberships of the elements, see the manual page of
+\Rfunction{isa.in.silico} for the details. Let's take a look at the
+input data, see Fig.~\ref{fig:block}.
+<<blockplot,include=FALSE,results=hide,cache=TRUE>>=
+print(image(Matrix(data)))
+@
+
+\begin{figure}
+\centering
+\setkeys{Gin}{width=0.7\textwidth}
+<<blockplot2,fig=TRUE,echo=FALSE,results=hide,cache=TRUE,width=8,height=6>>=
+<<blockplot>>
+@
+\caption{Another artificial data matrix with two modules and a lot of
+  noisy elements.}
+\label{fig:block}
+\end{figure}
+
+All we have to do now is calling the \Rfunction{isa} function on the
+data matrix. 
+<<seed,echo=FALSE>>=
+## Just to get the same results
+set.seed(10)
+@ 
+<<isa1,cache=TRUE>>=
+modules <- isa(data)
+names(modules)
+@
+
+The \Rfunction{isa} function returns a list with four
+elements. \listentry{rows} and \listentry{columns} define the
+modules. The non-zero elements in the first column of both define the
+first module, the second columns define the second module, etc.
+The number of columns in \listentry{rows} (and \listentry{columns})
+correspond to the number of modules.
+<<no.mods.>>=
+ncol(modules$rows)
+@
+ISA has found \Sexpr{ncol(modules$rows)} %$
+modules in the artificial data. This seem a bit surprising, as we
+expected only two modules, so let us take a look at the modules. We
+simply plot the row scores of the individual modules
+(Fig.~\ref{fig:simpleplot}). The first two modules corresponds to the
+two blocks in the input matrix. The third module is the union of them,
+with opposite score signs. The \Rfunction{isa} function finds modules
+containing anticorrelated rows/columns, as long as the
+anti-correlation is coordinated. I.e. in the third module, the first
+50 rows \emph{always} behave in an opposite way compared to the second
+50 rows. This is why they are collected into a single module.
+
+<<simpleplot,results=hide,include=FALSE>>=
+layout(cbind(seq(ncol(modules$rows))))
+sapply(seq(ncol(modules$rows)), function(x) {
+  par(mar=c(3,5,1,1))
+  plot(modules$rows[,x], ylim=c(-1,1), 
+       ylab="scores", xlab=NA, 
+       cex.lab=2, cex.axis=1.5)
+})
+@
+
+\begin{figure}
+\centering
+\setkeys{Gin}{width=0.7\textwidth}
+<<simpleplot2,fig=TRUE,echo=FALSE,results=hide,width=8,height=6>>=
+<<simpleplot>>
+@ 
+\caption{The row scores of the three modules found in the two-block
+  artificial data matrix. A score of zero means that the row is not
+  included in the module. Notice that in the third module the first
+  50 and the second 50 rows have opposite score signs, i.e. they are
+  anti-correlated.}
+\label{fig:simpleplot}
+\end{figure}
+
+The \listentry{rundata} entry of the return value of \Rfunction{isa}
+contains data about the ISA run, all the parameters (the default
+values in our case) are included here. It has the following members:
+<<isaresult.rundata>>=
+names(modules$rundata)
+@ 
+See the documentation of \Rfunction{isa} for more about these.
+
+Finally, the listentry{seeddata} entry is a data frame, it contains
+various data about the individual modules or the seeds that were used
+to find the module. There is one row for each module. It has the
+following columns:
+<<seedcols>>=
+colnames(modules$seeddata)
+@ 
+The columns \listentry{thr.row} and \listentry{thr.col} contain the
+row and columns thresholds that were used to find the module, while
+the \listentry{rob} column contains its robustness score:
+<<seeddata>>=
+modules$seeddata
+@
+
+Finally, let us show how to transform the result of the
+\Rfunction{isa} function to a list of biclusters. Each entry of the
+list will have two sublists, the first -- named \listentry{rows} --
+will contain the rows indices of the module, the second -- named
+\listentry{columns} -- the column indices:
+<<mymodules>>=
+mymodules <- lapply(seq(ncol(modules$rows)), function(x) {
+  list(rows=which(modules$rows[,x] != 0),
+       columns=which(modules$columns[,x] != 0))
+})
+length(mymodules)
+mymodules[[1]]$rows
+mymodules[[1]]$columns
+@
 
@@
 
-TODO
+In this section we will perform each steps of the ISA analysis
+individually. This makes sense only if the user wants to adjust the
+parameters of some steps.
+
+Let us create some in-silico data to be analyzed:
+<<insilico10>>=
+data <- isa.in.silico(500, 200, 10)
+@ 
+This will be a $500\times 200$ matrix, with 10 modules. By default
+``half'' of the matrix is filled with noise, so each module will be
+$25\times 10$ submatrix. See the input matrix on
+Fig.~\ref{fig:tenmodules}.
+
+<<tenmods,include=FALSE,results=hide,cache=TRUE>>=
+print(image(Matrix(data[[1]])))
+@
+
+\begin{figure}
+\centering
+\setkeys{Gin}{width=0.7\textwidth}
+<<tenmods2,fig=TRUE,echo=FALSE,results=hide,cache=TRUE>>=
+<<tenmods>>
+@
+\caption{In-silico data with ten non-overlapping modules.}
+\label{fig:tenmodules}
+\end{figure}
 
@@
 
-TODO
+One ISA iteration consists of two thresholded matrix
+multiplications. The data preparation step means producing these two
+matrices, $E_r$ and $E_c$ from the input matrix, $E$. The default
+behavior (i.e. what the \Rfunction{isa} function does) is to
+row-wise scale and center the transposed of $E$ to get $E_r$; and to
+row-wise scale and center $E$ to get $E_c$. The
+\Rfunction{isa.normalize} function performs the appropriate scaling
+and returns $E_r$ and $E_c$ in a list:
+<<normalize>>=
+normed.data <- isa.normalize(data[[1]])
+names(normed.data)
+dim(normed.data$Er)
+dim(normed.data$Ec)
+@
+
+Let us do a quick check that the rows of the matrices are indeed
+centered and scaled.
+<<checkscalecenter>>=
+max(abs(rowSums(normed.data$Er)))
+max(abs(rowSums(normed.data$Ec)))
+max(abs(apply(normed.data$Er, 1, sd)-1))
+max(abs(apply(normed.data$Ec, 1, sd)-1))
+@
 
@@
 
-TODO
+To run the ISA, we need to create the starting seeds first. The
+\Rfunction{isa} function uses sparse random seeds, generated by
+\Rfunction{generate.seeds}. We also have the choice of using
+non-random seeds, e.g. if you have a matrix with gene expression data
+measured on many samples you can use ``gene'' seeds that correspond to
+biological pathways; in this case ISA does a biased search to find
+modules related to the input pathways. Or, in a case/control
+experiment one can use seed vectors that correspond to cases and look
+for modules that are differentially regulated in the control and the
+case samples.
+
+For now, we will use thousand random seeds, with different sparseness:
+<<generate.seeds>>=
+row.seeds <- generate.seeds(length=nrow(data[[1]]), count=1000, 
+                            sparsity=c(2,5,10,100))
+@ 
+Note, that vector giving the sparsity is recycled, thus we will have
+four kind of seeds, some with 2, 5, 10 and 100 rows; 250 of each of
+them.
+
+We are ready to run the ISA now. Let us assume that we only want to
+find modules that are ``above'' the background. (In this artificial
+case we actually know that there are no modules below the background.)
+The \Rargument{direction} argument is thus set to \verb+"up"+. 
+<<isa.iterate>>=
+modules <- isa.iterate(normed.data, row.seeds=row.seeds, 
+                       thr.row=2, thr.col=2, direction="up")
+@
+
+Unlike \Rfunction{isa}, \Rfunction{isa.iterate} does not merge or
+filter the modules the input seeds converge to. Consequently, the
+\verb+modules+ object contains 1000 modules, but these are not
+necessarily unique:
+<<isa.iterate.res>>=
+ncol(modules$rows)
+@
+
+It is also possibly that some seeds converge to an all-zero vector, we
+have \Sexpr{sum(apply(modules$rows==0, 2, all))} such module: % $
+<<all.zero.modules>>=
+sum(apply(modules$rows==0, 2, all))
+@
+
+The \Rfunction{isa.unique} function finds duplicate or very similar
+modules in a list returned by \Rfunction{isa.iterate} and keeps only a
+single one of them. It also eliminates all-zero modules and seeds
+that did not converge, the latter result all \verb+NA+
+columns in the \listentry{rows} and \listentry{columns} matrix:
+<<unique>>=
+modules2 <- isa.unique(normed.data, modules, cor.limit=0.9)
+@
+The \Rargument{cor.limit} argument specifies the correlation limit
+above which two modules are considered to be the same. Let us see how
+many modules are left:
+<<isa.unique.res>>=
+ncol(modules2$rows)
+@
+
+We still have \Sexpr{ncol(modules2$rows)} modules. Let us check %$
+whether each real module was find by the ISA. Remember that
+\Rfunction{isa.in.silico} stores the correct modules in the second and
+third entry of its return value.
+<<correctness>>=
+found.rows <- cor(data[[2]], modules2$rows!=0)
+found.cols <- cor(data[[3]], modules2$columns!=0)
+found <- pmin(found.rows,found.cols)
+apply(found, 1, max)
+@ 
+Out of the ten modules, \Sexpr{sum(apply(found, 1, max)-1<1e-14)} were
+correctly identified by the ISA.
 
@@
 
-TODO
+There are two features that we expect from a good biclustering
+algorithm. The first is sensitivity: it is able to find all
+biclusters in the input data. The second is specificity: it should not
+generate spurious biclusters, or in other words biclusters that are
+correlated just by chance.
+
+To address this problem, we have developed a robustness measure, a
+single scalar number that gives how correlated the rows and columns of
+a given module are. This measure is a simple generalization of the
+singular value decomposition (SVD) singular value, and can be
+calculated with the \Rfunction{robustness} function:
+<<robustness>>=
+rob <- robustness(normed.data, modules2$rows, modules2$columns)
+summary(rob)
+@ 
+As the user can see, the robustness of the \Sexpr{ncol(modules2$rows)} %$
+modules varies considerably, from \Sexpr{round(min(rob),2)} to
+\Sexpr{round(max(rob), 2)}. For more insight we create a boxplot for
+the robustness scores (Fig.~\ref{fig:robbox}).
+
+<<robbox,include=FALSE,results=hide,cache=TRUE>>=
+par(cex.lab=1.5, cex.axis=1.5)
+boxplot(rob, ylab="Robustness")
+@
+
+\begin{figure}
+\centering
+\setkeys{Gin}{width=0.7\textwidth}
+<<robbox2,fig=TRUE,echo=FALSE,results=hide,cache=TRUE>>=
+<<robbox>>
+@
+\caption{}
+\label{fig:robbox}
+\end{figure}
+
+One can use the robustness measure to filter out list of modules, in
+the following way. We permute the input matrix and run the same
+module finding procedure that we did for the original matrix. Then we
+calculate the robustness scores for both the real and the scrambled
+modules and eliminate the ``real'' modules having a robustness score
+that is less than the score of at least one scrambled module. Note,
+that this filtering can be done only if both sets of modules were
+found using the same threshold parameters. The procedure is
+implemented in the \Rfunction{isa.filter.robust} function:
+<<filter.robust>>=
+modules3 <- isa.filter.robust(data[[1]], normed.data, modules2, 
+                              perms=1, row.seeds=row.seeds)
+@
+We use the very same row seeds, as for the original matrix; this is
+not strictly requires. The \Rargument{perms} argument sets the number
+of permutations to perform.
+
+Note that the ISA run on the scrambled matrix usually takes longer,
+than on the real data, because it takes more steps for the ISA to
+converge.
+
+After the robustness filtering we have \Sexpr{ncol(modules3$rows)} %$
+modules left. This is still more than the ten correct modules, so let
+us take a closer look. First, let us create another plot with the 
+robustness scores of the remaining modules. 
+\Rfunction{isa.filter.robust} places the robustness scores in the seed
+data, so we don't need to calculate them again. See Fig.~\ref{fig:robbox2}.
+
+<<robbox3,include=FALSE,results=hide,cache=TRUE>>=
+plot(modules3$seeddata$rob, ylab="Robustness", cex.lab=1.5)
+@
+
+\begin{figure}
+\centering
+\setkeys{Gin}{width=0.7\textwidth}
+<<robbox4,fig=TRUE,echo=FALSE,results=hide,cache=TRUE>>=
+<<robbox3>>
+@
+\caption{}
+\label{fig:robbox2}
+\end{figure}
+
+The scatterplot shows that ten modules have in fact higher
+robustness scores than the rest. Let us check that these are indeed the
+ten ``real'' modules of the data.
+<<bestmods>>=
+bestmods <- which(modules3$seeddata$rob > 61)
+mod.cor <- pmin(cor(modules3$rows[,bestmods]!=0, data[[2]]),
+                cor(modules3$columns[,bestmods]!=0, data[[3]]))
+apply(mod.cor, 1, max)
+apply(mod.cor, 2, max)
+@
+
+Indeed, we have found the ten real modules, the ISA is both specific
+and sensitive. Of course for the analysis of real data it is much more
+difficult to filter the modules found by the biclustering algorithm,
+but the robustness based filtering still helps getting rid of spurious
+modules.
+
+Finally, let us take a look at some of the other modules, that were
+also kept after the robustness-based filtering. Here we plot nine of
+them.
+<<othermods,include=FALSE,results=hide,cache=TRUE>>=
+othermods <- which(modules3$seeddata$rob < 61)[1:3]
+layout(matrix(seq(length(othermods)+1), 2, byrow=TRUE))
+idx <- seq(ncol(data[[1]]), 1)
+par(mar=c(1,1,4,1))
+image(1-data[[1]][,idx], main="Input matrix", 
+      axes=FALSE, frame=TRUE)
+for (i in othermods) { 
+  par(mar=c(1,1,4,1))
+  image(1-outer(modules3$rows[,i], modules3$columns[,i])[,idx],
+        main=paste("Module", i), axes=FALSE, frame=TRUE)
+}
+@
+
+\begin{figure}
+\centering
+\setkeys{Gin}{width=0.7\textwidth}
+<<othermods2,fig=TRUE,echo=FALSE,results=hide,cache=TRUE>>=
+<<othermods>>
+@ 
+\end{figure}
 
@@
 
-<<results=tex,echo=FALSE>>=
+<<sessioninfo,results=tex,echo=FALSE>>=
 gsub(";", "; ", gsub("_", "\\\\_", sub("Locale:[ ]*\\\\verb\\|(.*)\\|", 

### 2009-04-28 isa/R/isa3.R
Corrected a bug in isa.iterate when different thresholds are supplied.

@@ -88,8 +88,8 @@ isa.iterate.default <- function(normed.data, row.seeds, col.seeds,
   if (length(thr.col) != 1 && length(thr.col) != no.seeds) {
     stop("`thr.col' does not have the right length")
   }
-  thr.row <- rep(thr.row, no.seeds)
-  thr.col <- rep(thr.col, no.seeds)
+  thr.row <- rep(thr.row, length=no.seeds)
+  thr.col <- rep(thr.col, length=no.seeds)
 
   ## Put the seeds together
   all.seeds <- matrix(ncol=0, nrow=nrow(normed.data$Ec))

### 2009-04-28 isa/inst/doc/ISA_tutorial.Rnw
isa vignette update, almost finished. Some captions missing and a proofreading is missing as well.

@@
 
-TODO
+In this section we show some examples that highlight some key features
+of the ISA: its resilience to noise and its ability to find
+overlapping modules.
 
@@
 
-TODO
+To test the behavior of ISA with noisy inputs we generate a number of
+in-silico data sets with different noise levels
+(Fig.~\ref{fig:noise}).
+<<noisegen,cache=TRUE>>=
+noise <- seq(0.1,1,by=0.1)
+data <- lapply(noise, function(x) isa.in.silico(500,200,10,noise=x))
+@
+
+<<noiseplot1,echo=FALSE,results=hide>>=
+png("noise.png", width=800, height=320)
+@ 
+<<noiseplot2>>=
+layout( matrix(1:10, nr=2, byrow=TRUE) )
+for (i in seq_along(data)) {
+  par(mar=c(1,1,1,1))
+  image(1-data[[i]][[1]], axes=FALSE, frame=TRUE)
+}
+@ 
+<<noiseplot3,echo=FALSE,results=hide>>=
+dev.off()
+@ 
+\begin{figure}
+  \centering
+  \includegraphics[width=\textwidth]{noise}
+  \caption{}
+  \label{fig:noise}
+\end{figure}
+
+Next, we run ISA with the default parameters on all the data
+sets. This might take a while.
+<<noise.isa,cache=TRUE>>=
+modules <- lapply(data, function(x) isa(x[[1]]))
+@
+
+Let us check the sensitivity of ISA, for every real module, we pick
+one from the ones ISA has found, the one that is the most correlated
+to it.
+<<sensitivity.noise>>=
+best <- lapply(seq_along(modules), function(i) {
+  cc <- pmin(cor(modules[[i]]$rows!=0, data[[i]][[2]]),
+             cor(modules[[i]]$columns!=0, data[[i]][[3]]))
+  apply(cc, 2, max)
+})
+best.mean <- sapply(best, mean)
+@
+
+Let's create boxplots for the different noise levels.
+
+<<sensbox,include=FALSE,results=hide>>=
+boxplot(best, names=noise, xlab="Noise level", ylab="Sensitivity",
+        cex.lab=1.5)
+@
+\begin{figure}
+\centering
+<<sensbox2,fig=TRUE,echo=FALSE,results=hide,width=10,height=5>>=
+<<sensbox>>
+@ 
+\caption{}
+\label{fig:sensbox}
+\end{figure}
+
+To test the specificity, we create boxplots for the correlation
+coefficients of all modules found by ISA and their closest real
+module.
+<<spec>>=
+spec <- lapply(seq_along(modules), function(i) {
+  cc <- pmin(cor(modules[[i]]$rows != 0, data[[i]][[2]]),
+             cor(modules[[i]]$columns != 0, data[[i]][[3]]))
+  apply(cc, 1, max)
+})
+@ 
+<<specbox,include=FALSE,results=hide>>=
+boxplot(spec, names=noise, xlab="Noise level", ylab="Specificity",
+        cex.lab=1.5)
+@
+\begin{figure}
+\centering
+<<specbox2,fig=TRUE,echo=FALSE,results=hide,width=10,height=5>>=
+<<specbox>>
+@
+\caption{}
+\label{fig:specbox}
+\end{figure}
 
@@
 
-TODO
+ISA biclusters might overlap in their rows, columns, or both. To
+illustrate this, we create an artificial data set with two overlapping
+blocks.
+
+<<set.seed.overlap>>=
+set.seed(1)
+@ 
+<<overlap>>=
+two.over <- isa.in.silico(100,100,2,40,40,overlap.row=10)
+@
+
+We run ISA on this input matrix, with two threshold parameters, the
+first thresholds are very mild, they are both zero.
+In other words, in each iteration step all rows/columns are kept in each
+that have a higher score than the average. The second set of
+thresholds are stricter, the scores have to be at least one standard
+deviation away from the mean to keep them.
+
+<<overlap.isa>>=
+ov.normed <- isa.normalize(two.over[[1]])
+ov.seeds <- generate.seeds(count=100, length=100)
+ov.modules.1 <- isa.iterate(ov.normed, ov.seeds, 
+                            thr.row=0, thr.col=0, direction="up")
+ov.modules.1 <- isa.unique(ov.normed, ov.modules.1)
+ov.modules.1 <- isa.filter.robust(two.over[[1]], ov.normed, ov.modules.1)
+ov.modules.2 <- isa.iterate(ov.normed, ov.seeds, 
+                            thr.row=1, thr.col=1, direction="up")
+ov.modules.2 <- isa.unique(ov.normed, ov.modules.2)
+ov.modules.2 <- isa.filter.robust(two.over[[1]], ov.normed, ov.modules.2)
+ncol(ov.modules.1$rows)
+ncol(ov.modules.2$rows)
+@
+
+ISA found \Sexpr{ncol(ov.modules.1$rows)} %$
+modules with the mild thresholds and \Sexpr{ncol(ov.modules.2$rows)} %$
+modules with the strict ones. Let us plot the original data and the
+modules found by ISA (Fig.~\ref{fig:overlapplot}).
+
+<<overlapplot,include=FALSE,results=hide,cache=TRUE>>=
+layout( rbind(1:4, c(8,5,6,7)) )
+par(mar=c(1,1,3,1))
+image(2-two.over[[1]], axes=FALSE, frame=TRUE, main="Input matrix")
+for (i in 1:3) {
+  par(mar=c(1,1,3,1))
+  image(1-outer(ov.modules.1$rows[,i]!=0, ov.modules.1$columns[,i]!=0),
+        axes=FALSE, frame=TRUE, main=paste("Module", i))
+}
+for (i in 1:3) { 
+  par(mar=c(1,1,3,1))
+  image( 1-outer(ov.modules.2$rows[,i]!=0, ov.modules.2$columns[,i]!=0),
+        axes=FALSE, frame=TRUE, main=paste("Module", i+3))
+}
+@
+
+\begin{figure}
+\centering
+<<overlapplot2,fig=TRUE,echo=FALSE,results=hide,cache=TRUE,width=8,height=5>>=
+<<overlapplot>>
+@ 
+\caption{}
+\label{fig:overlapplot}
+\end{figure}
+
+As expected the ISA modules are in general bigger with the mild
+thresholds. ISA correctly identifies the two overlapping modules and
+also the union of them. With the stricter thresholds it find the
+non-overlapping parts of the two modules, plus their overlap as a
+separate module.
+
+\section{Running ISA in parallel}
+
+Since a typical ISA analysis consists of using a range of row/column
+and these runs are independent of each other; it is trivial to
+parallelize the task by performing the iterations for different
+threshold parameters on different processors or computers. Here we
+show and example on how to do this easily with the \Rpackage{snow} and
+the \Rpackage{Rmpi} packages:
+<<parallelloadpackage>>=
+library(snow)
+library(Rmpi)
+@
+
+We generate some simple in-silico data.
+<<paralleldata>>=
+pdata <- isa.in.silico()
+@
+
+Create the MPI cluster, with eight working nodes. You need to have a
+working MPI installation for this. See more in the documentation of
+the \Rpackage{Rmpi} and \Rpackage{snow} packages.
+<<createcluster>>=
+clu <- makeMPIcluster(8)
+invisible(clusterEvalQ(clu, library(isa)))
+clusterExport(clu, "pdata")
+@
+
+Create a big matrix in which each row is a combination of the
+threshold parameters.
+<<parthr>>=
+thr <- seq(1,3,by=0.2)
+thr.list <- expand.grid(thr, thr)
+@
+
+First we run the ISA on a single processor only, and measure the
+running time.
+<<seqrun,cache=TRUE>>=
+system.time(modules <- isa(pdata[[1]], thr.row=thr, thr.col=thr))
+@
+
+Let us now do a parallel run, again, with measuring the running
+time. If you are really running this on multiple CPUs, then it is much
+faster.
+<<parrun,cache=TRUE>>=
+system.time(modules.par <- parApply(clu, thr.list, 1, function(x) {
+  isa(pdata[[1]], thr.row=x[1], thr.col=x[2])
+}))
+@
+
+Finally, stop the cluster.
+<<closecluster>>=
+stopCluster(clu)
+@
 

### 2009-04-30 isa/inst/doc/ISA_tutorial.Rnw
More isa vignette improvements, use bitmaps for big image plots.

@@
 input data, see Fig.~\ref{fig:block}.
-<<blockplot,include=FALSE,results=hide,cache=TRUE>>=
+
+<<blockplot.pre,echo=FALSE,results=hide>>=
+png("blockplot.png", 800, 600)
+@ 
+<<blockplot>>=
 print(image(Matrix(data)))
+@ 
+<<blockplot.post,echo=FALSE,results=hide>>=
+dev.off()
 @
@@
 \centering
-\setkeys{Gin}{width=0.7\textwidth}
-<<blockplot2,fig=TRUE,echo=FALSE,results=hide,cache=TRUE,width=8,height=6>>=
-<<blockplot>>
-@
+\includegraphics[width=\textwidth]{blockplot}
 \caption{Another artificial data matrix with two modules and a lot of
@@
 <<insilico10>>=
-data <- isa.in.silico(500, 200, 10)
+data <- isa.in.silico(200, 100, 10)
 @ 
-This will be a $500\times 200$ matrix, with 10 modules. By default
-``half'' of the matrix is filled with noise, so each module will be
-$25\times 10$ submatrix. See the input matrix on
-Fig.~\ref{fig:tenmodules}.
+This will be a $\Sexpr{nrow(data[[1]])}\times \Sexpr{ncol(data[[1]])}$
+matrix, with \Sexpr{ncol(data[[2]])} modules. By default ``half'' of
+the matrix is filled with noise, so each module will be a 
+$\Sexpr{round(nrow(data[[1]])/2/10)}\times 
+\Sexpr{round(ncol(data[[1]])/2/10)}$ submatrix. See the
+input matrix on Fig.~\ref{fig:tenmodules}.
 
-<<tenmods,include=FALSE,results=hide,cache=TRUE>>=
+<<tenmods.pre,echo=FALSE,results=hide>>=
+png("tenmods.png", 600, 800)
+@ 
+<<tenmods>>=
 print(image(Matrix(data[[1]])))
+@ 
+<<tenmods.post,echo=FALSE,results=hide>>=
+dev.off()
 @
@@
 \centering
-\setkeys{Gin}{width=0.7\textwidth}
-<<tenmods2,fig=TRUE,echo=FALSE,results=hide,cache=TRUE>>=
-<<tenmods>>
-@
+\includegraphics{tenmods}
 \caption{In-silico data with ten non-overlapping modules.}
@@
 @
-\caption{}
+\caption{Box-plot for the robustness scores of the modules found by
+  the ISA. Some modules have a much higher score than the rest.}
 \label{fig:robbox}
@@
 @
-\caption{}
+\caption{The robustness scores of ISA modules after filtering out some
+  modules with low robustness. It is still obvious that ten modules
+  have a significantly higher score than the rest.}
 \label{fig:robbox2}
@@
 <<bestmods>>=
-bestmods <- which(modules3$seeddata$rob > 61)
+bestmods <- which(modules3$seeddata$rob > 27.5)
 mod.cor <- pmin(cor(modules3$rows[,bestmods]!=0, data[[2]]),
@@
 them.
-<<othermods,include=FALSE,results=hide,cache=TRUE>>=
-othermods <- which(modules3$seeddata$rob < 61)[1:3]
-layout(matrix(seq(length(othermods)+1), 2, byrow=TRUE))
-idx <- seq(ncol(data[[1]]), 1)
-par(mar=c(1,1,4,1))
-image(1-data[[1]][,idx], main="Input matrix", 
-      axes=FALSE, frame=TRUE)
-for (i in othermods) { 
-  par(mar=c(1,1,4,1))
-  image(1-outer(modules3$rows[,i], modules3$columns[,i])[,idx],
-        main=paste("Module", i), axes=FALSE, frame=TRUE)
-}
+<<othermods.pre>>=
+png("othermods.png", 800, 800)
+@ 
+<<othermods>>=
+othermods <- which(modules3$seeddata$rob < 27.5)[1:5]
+print(plot.modules(modules3, othermods, data=data[[1]]))
+@ 
+<<othermods.post>>=
+dev.off()
 @
@@
 \centering
-\setkeys{Gin}{width=0.7\textwidth}
-<<othermods2,fig=TRUE,echo=FALSE,results=hide,cache=TRUE>>=
-<<othermods>>
-@ 
+\includegraphics{othermods}
+\caption{The input matrix and five modules with relatively low
+  robustness scores. As it turns out, each of these modules contains
+  in fact two blocks of the input matrix.}
+\label{fig:othermods}
 \end{figure}
@@
 
-<<noiseplot1,echo=FALSE,results=hide>>=
-png("noise.png", width=800, height=320)
+<<noiseplot.pre,echo=FALSE,results=hide>>=
+png("noise.png", width=800, height=600)
 @ 
-<<noiseplot2>>=
-layout( matrix(1:10, nr=2, byrow=TRUE) )
-for (i in seq_along(data)) {
-  par(mar=c(1,1,1,1))
-  image(1-data[[i]][[1]], axes=FALSE, frame=TRUE)
-}
+<<noiseplot>>=
+images(lapply(data, "[[", 1), names=as.character(noise), 
+       xlab="", ylab="")
 @ 
-<<noiseplot3,echo=FALSE,results=hide>>=
+<<noiseplot.post,echo=FALSE,results=hide>>=
 dev.off()
@@
   \includegraphics[width=\textwidth]{noise}
-  \caption{}
+  \caption{Modular data sets with different levels of noise. The data
+  sets are generated by adding normally distributed background noise
+  with given variance to the checkerboard matrix.}
   \label{fig:noise}
@@
 @ 
-\caption{}
+\caption{The sensitivity of the ISA in the function of noise. The
+  sensitivity score is calculated by finding the best (in terms of
+  Pearson correlation) ISA module for each block in the input matrix.
+
+  ISA is very sensitive to relatively large noise levels.
+}
 \label{fig:sensbox}
@@
 @
-\caption{}
+\caption{Specificity of the ISA modules, in the function of
+  noise. Specificity was calculated by finding the best -- again,
+  by highest Pearson correlation -- matching block in the input matrix
+  for each ISA module. These correlation coefficients are plotted in a
+  boxplot for each noise level.
+
+  Specificity for ISA decreases approximately linearly with the
+  increase of noise.
+}
 \label{fig:specbox}
@@
 
-<<overlapplot,include=FALSE,results=hide,cache=TRUE>>=
-layout( rbind(1:4, c(8,5,6,7)) )
-par(mar=c(1,1,3,1))
-image(2-two.over[[1]], axes=FALSE, frame=TRUE, main="Input matrix")
-for (i in 1:3) {
-  par(mar=c(1,1,3,1))
-  image(1-outer(ov.modules.1$rows[,i]!=0, ov.modules.1$columns[,i]!=0),
-        axes=FALSE, frame=TRUE, main=paste("Module", i))
-}
-for (i in 1:3) { 
-  par(mar=c(1,1,3,1))
-  image( 1-outer(ov.modules.2$rows[,i]!=0, ov.modules.2$columns[,i]!=0),
-        axes=FALSE, frame=TRUE, main=paste("Module", i+3))
-}
+<<overlapplot.pre,echo=FALSE,results=hide>>=
+png("overlapplot.png", 800, 600)
+@ 
+<<overlapplot>>=
+plot.modules(data=two.over[[1]],
+             list(rows=cbind(ov.modules.1$rows, ov.modules.2$rows),
+                  columns=cbind(ov.modules.1$columns,
+                    ov.modules.2$column)),
+             names=c("Input matrix", paste("Module", 1:6)))
+@
+<<overlapplot.post,echo=FALSE,results=hide>>=
+dev.off()
 @
@@
 \centering
-<<overlapplot2,fig=TRUE,echo=FALSE,results=hide,cache=TRUE,width=8,height=5>>=
-<<overlapplot>>
-@ 
-\caption{}
+\includegraphics{overlapplot}
+\caption{The input matrix with two overlapping blocks and the ISA
+  modules for this data set. The first three modules were found at
+  milder threshold parameters, that is why they are bigger. As you can
+  see for Modules #1 and #2, at milder thresholds there is a higher
+  probability that ISA will pick up some incorrect rows/columns.
+}
 \label{fig:overlapplot}

### 2009-04-30 isa/inst/doc/ISA_tutorial.Rnw
isa: added a method to convert ISA modules to Biclust objects from the biclust package.

@@
 them.
-<<othermods.pre>>=
+<<othermods.pre,echo=FALSE,results=hide>>=
 png("othermods.png", 800, 800)
@@
 @ 
-<<othermods.post>>=
+<<othermods.post,echo=FALSE,results=hide>>=
 dev.off()
@@
   Pearson correlation) ISA module for each block in the input matrix.
-
   ISA is very sensitive to relatively large noise levels.
@@
   boxplot for each noise level.
-
-  Specificity for ISA decreases approximately linearly with the
+  Specificity of ISA decreases approximately linearly with the
   increase of noise.
@@
   milder threshold parameters, that is why they are bigger. As you can
-  see for Modules #1 and #2, at milder thresholds there is a higher
+  see for Modules \#1 and \#2, at milder thresholds there is a higher
   probability that ISA will pick up some incorrect rows/columns.

### 2009-04-30 isa/R/biclust.R

@@ -1,0 +1,12 @@
+
+isa.biclust <- function(modules) {
+
+  if (!require(biclust)) {
+    stop("The `biclust' package is required for this")
+  }
+
+  new("Biclust", Parameters=list(seeddata=modules$seeddata,
+                   rundata=modules$rundata),
+      RowxNumber=modules$rows != 0,
+      NumberxCol=t(modules$columns != 0), Number=ncol(modules$rows))
+}

### 2009-05-01 isa/R/AllGenerics.R
Renamed plot.modules to plotModules.

@@ -17,5 +17,5 @@ setGeneric("isa.sweep",
 setGeneric("sweep.graph",
            function(sweep.result, ...) standardGeneric("sweep.graph"))
 
-setGeneric("plot.modules",
-           function(modules, ...) standardGeneric("plot.modules"))
+setGeneric("plotModules",
+           function(modules, ...) standardGeneric("plotModules"))

### 2009-05-01 isa/R/viz.R

@@ -23,14 +23,14 @@ images <- function(matrices, names=NULL, ...) {
             ...)
 } 
 
-setMethod("plot.modules", signature(modules="list"),
+setMethod("plotModules", signature(modules="list"),
           function(modules, ...)
-          plot.modules.default(modules, ...))
+          plotModules.default(modules, ...))
 
-plot.modules.default <- function(modules, to.plot=seq_len(ncol(modules$rows)),
-                                 data, binary=TRUE, names=NULL,
-                                 xlab="", ylab="",
-                                 ...) {
+plotModules.default <- function(modules, to.plot=seq_len(ncol(modules$rows)),
+                                data, binary=TRUE, names=NULL,
+                                xlab="", ylab="",
+                                ...) {
 
   require(lattice)
   

### 2009-05-01 isa/inst/doc/ISA_tutorial.Rnw
Updated isa vignette, a PDF file was also added.

@@
 <<blockplot>>=
-print(image(Matrix(data)))
+images(list(data), xlab="", ylab="", strip=FALSE)
 @ 
@@
 \centering
-\includegraphics[width=\textwidth]{blockplot}
+\includegraphics[width=0.6\textwidth]{blockplot}
 \caption{Another artificial data matrix with two modules and a lot of
@@
 <<tenmods.pre,echo=FALSE,results=hide>>=
-png("tenmods.png", 600, 800)
+png("tenmods.png", 600, 400)
 @ 
 <<tenmods>>=
-print(image(Matrix(data[[1]])))
+images(list(data[[1]]), strip=FALSE, xlab="", ylab="")
 @ 
@@
 \setkeys{Gin}{width=0.7\textwidth}
-<<robbox2,fig=TRUE,echo=FALSE,results=hide,cache=TRUE>>=
+<<robbox2,fig=TRUE,echo=FALSE,results=hide,cache=TRUE,width=7,height=4>>=
 <<robbox>>
@@
 \setkeys{Gin}{width=0.7\textwidth}
-<<robbox4,fig=TRUE,echo=FALSE,results=hide,cache=TRUE>>=
+<<robbox4,fig=TRUE,echo=FALSE,results=hide,cache=TRUE,width=7,height=4>>=
 <<robbox3>>
@@
 <<othermods.pre,echo=FALSE,results=hide>>=
-png("othermods.png", 800, 800)
+png("othermods.png", 400, 400)
 @ 
@@
 othermods <- which(modules3$seeddata$rob < 27.5)[1:5]
-print(plot.modules(modules3, othermods, data=data[[1]]))
+print(plotModules(modules3, othermods, data=data[[1]]))
 @ 
@@
 \centering
-\includegraphics{othermods}
+\includegraphics[width=0.6\textwidth]{othermods}
 \caption{The input matrix and five modules with relatively low
@@
 
-TODO
+Visualizing overlapping biclusters is quite a challenging task. We
+show simple methods that usually visualize a single bicluster 
+at a time. For some of these we will use the \Rpackage{biclust} R
+package.
+
+\subsubsection{The \Rpackage{biclust} package}
+
+The \Rpackage{biclust} R package implements several biclustering
+algorithms in a unified framework. It uses the class \Rclass{Biclust}
+to store a set of biclusters. The \Rfunction{isa.biclust} converts ISA
+modules to a \Rclass{Biclust} object. This requires the binarization
+of the modules, i.e. the ISA scores are lost, they are converted to
+zeros and ones:
+<<biclust,cache=TRUE>>=
+library(biclust)
+five.mods <- isa.in.silico(200,200,5,20,20)
+modules <- isa(five.mods[[1]],thr.row=2, thr.col=2)
+Bc <- isa.biclust(modules)
+Bc
+@
+
+\subsubsection{Image plots}
+
+There are some examples in this document that show how to create image
+plots for the modules and potentially also the input data. See
+e.g. Figs~\ref{fig:othermods} and
+\ref{fig:overlapplot}. These plots use the \Rfunction{plotModules}
+function, that directly takes the output of \Rfunction{isa} (and other
+functions with the same output format: \Rfunction{isa.iterate},
+Rfunction{isa.unique}, etc.
+
+In fact \Rfunction{plotModules} calls the
+\Rfunction{images} function to do its job. \Rfunction{images} takes a
+list of matrices, see Fig.~\ref{fig:noise} for an example.
+
+The \Rfunction{drawHeatmap} function of the \Rpackage{biclust} package
+can be also used to draw an image plot. This visualizes a single
+bicluster on top of the input matrix. It reorders the rows and columns
+of the input matrix to move the block of the bicluster to the top
+left corner of the input matrix (Fig.~\ref{fig:heatmap}).
+<<drawHeatmap.pre,echo=FALSE,results=hide>>=
+png("drawheatmap.png", 400,400)
+@ 
+<<drawHeatmap>>=
+drawHeatmap(five.mods[[1]], Bc, number=1,local=FALSE)
+@ 
+<<drawHeatmap.post,echo=FALSE,results=hide>>=
+dev.off()
+@ 
+\begin{figure}
+\centering
+\includegraphics[width=0.5\textwidth]{drawheatmap}
+\caption{Heatmap plot of the reordered input matrix. The rows and
+  columns corresponding to the first bicluster are moved to the
+  top-left corner of the matrix.}
+\label{fig:heatmap}
+\end{figure}
+
+\subsubsection{Profile plots}
+
+The \Rfunction{parallelCoordinates} function of the \Rpackage{biclust}
+package plots the profile of the rows (or columns) that are included
+in a bicluster with a different color. These plots visualize the
+difference between bicluster and the rest of the matrix, see
+Fig.\ref{fig:parallel} for an example.
+<<parallel.pre,include=TRUE>>=
+parallelCoordinates(five.mods[[1]], Bc, number=1, plotBoth=TRUE)
+@ 
+\begin{figure}
+\centering
+\setkeys{Gin}{width=0.6\textwidth}
+<<parallel,fig=TRUE,echo=FALSE,results=hide>>=
+<<parallel.pre>>
+@ 
+\caption{Profile plots for the first bicluster found by ISA. In this
+  artificial data set there is a clear distinction between the
+  background and the rows/columns of the module.}
+\label{fig:parallel}
+\end{figure}
+
+\subsubsection{Contrast bar plots}
+
+The \Rfunction{plotclust} function of the \Rpackage{biclust} package
+creates barplots for one or more biclusters. A single bar is the
+mean of the columns of the bicluster for a given row of the bicluster;
+or the mean of the columns of the background (i.e. the rest of the
+input matrix). The bigger the difference between the bars of the two
+colors, the better the bicluster. The results of the following code
+are in Fig.~\ref{fig:plotclust}.
+<<plotclust.pre,include=FALSE>>=
+plotclust(Bc, five.mods[[1]], Titel="")
+@
+\begin{figure}
+\centering
+\setkeys{Gin}{width=0.6\textwidth}
+<<plotclust,fig=TRUE,echo=FALSE,results=hide>>=
+<<plotclust.pre>>
+@ 
+\caption{Bar plots to show the difference between the biclusters and
+  the rest of the input matrix, for the first five modules found by
+  ISA.}
+\label{fig:plotclust}
+\end{figure}
 
@@
 
-In this section we show some examples that highlight some key features
+In this section we show examples that highlight key features
 of the ISA: its resilience to noise and its ability to find
@@
 <<overlapplot>>=
-plot.modules(data=two.over[[1]],
+plotModules(data=two.over[[1]],
             list(rows=cbind(ov.modules.1$rows, ov.modules.2$rows),
@@
 
+\bibitem[Kaiser, 2009]{biclust}
+Sebastian Kaiser, Rodrigo Santamaria, Roberto Theron, Luis Quintales
+and Friedrich Leisch. (2009). 
+\newblock {\em biclust: BiCluster Algorithms}. 
+\newblock R package version 0.7.2.
+
 \end{thebibliography}

### 2009-05-13 isa/src/isa.c
isa: removed C functions that are not used.

@@ -1,345 +1,7 @@
 
 #include <Rdefines.h>
 #include <math.h>
-
-/* TODO: handle empty matrix */
-
-SEXP beta_mean_sd(SEXP matrix) {
   
-  SEXP result, res_mean, res_std;
-  long int nrow=INTEGER(GET_DIM(matrix))[0];
-  long int ncol=INTEGER(GET_DIM(matrix))[1];
-  long int i, j;
-  
-  PROTECT(result=NEW_LIST(2));
-  SET_VECTOR_ELT(result, 0, NEW_NUMERIC(ncol));
-  SET_VECTOR_ELT(result, 1, NEW_NUMERIC(ncol));
-  res_mean=VECTOR_ELT(result, 0);
-  res_std=VECTOR_ELT(result, 1);
-
-  for (i=0; i<ncol; i++) {
-    double *data = REAL(matrix)+i*nrow;
-    double mean=data[0], std=0;
-    for (j=1; j<nrow; j++) {
-      std += j*pow((data[j]-mean), 2)/(j+1);
-      mean += (data[j]-mean)/(j+1);
-    }
-    std = sqrt(std / (nrow-1));
-    REAL(res_mean)[i]=mean;
-    REAL(res_std)[i]=std;
-  }
-  
-  UNPROTECT(1);
-  return result;
-}
-
-/*
-
-Test it in R:
-
-R
-library(beta)
-M <- matrix(runif(1000*1000), 1000, 1000)
-system.time(ms <- .Call("beta_mean_sd", M, PACKAGE="beta"))
-system.time(m <- apply(M, 2, mean))
-system.time(s <- apply(M, 2, sd))
-max(abs(ms[[2]] - s ))
-max(abs(ms[[1]] - m ))
-
-*/
-
-SEXP beta_threshold_cols_up(SEXP matrix, SEXP thr) {
-  
-  SEXP result;
-  long int nrow=INTEGER(GET_DIM(matrix))[0];
-  long int ncol=INTEGER(GET_DIM(matrix))[1];
-  long int i, j;
-  double *data, threshold;
-  
-  PROTECT(result=duplicate(matrix));
-  data=REAL(result);
-  for (i=0; i<ncol; i++) {
-    threshold=REAL(thr)[i];
-    for (j=0; j<nrow; j++) {
-      if (*data <= threshold) {
-	*data=0;
-      }
-      data++;
-    }
-  }
-  
-  UNPROTECT(1);
-  return result;
-}
-
-/* 
-
-Test in R:
-
-R
[…]

### 2009-05-13 isa/inst/doc/ISA_parallel.Rnw
isa: split parallel running example into another vignette.

--- 2009-04-30 isa/inst/doc/ISA_tutorial.Rnw
@@
-% \VignetteIndexEntry{The Iterative Signature Algorithm}
+% \VignetteIndexEntry{Running ISA in parallel}
 \documentclass{article}
@@
 
-\title{The Iterative Signature Algorithm}
+\title{Running ISA in parallel with the \Rpackage{snow} package}
 \author{G\'abor Cs\'ardi}
@@
 
-\tableofcontents
-
-<<Roptions,echo=FALSE,results=hide>>=
-## Some options for R
-options(width=60)
-@
-
-\section{For the impatient}
-
-To run the typical ISA work flow with default parameters on your data
-matrix, just load the \Rpackage{isa} package and call the
-\Rfunction{isa} function with your matrix as the single argument. The
-return value of \Rfunction{isa} is a named list, the members
-\texttt{rows} and \texttt{columns} contain the biclusters ISA have
-found. Every bicluster is given by a pair of columns from
-\texttt{rows} and \texttt{columns} (i.e. the first columns define the
-first bicluster, etc.) and the elements the biclusters are the
-non-zero elements in the columns of \texttt{rows} and
-\texttt{columns}.
-
-Please continue reading for a less dense tutorial.
-
-\section{Introduction}
-
-The Iterative Signature Algorithm (ISA) is biclustering method. Its 
-input is a matrix and its output is a set of biclusters: blocks of the
-potentially reordered input matrix, that fulfill some predefined
-criteria. A biclustering algorithm tries to find
-blocks that are different from the rest of the matrix, e.g. the values
-covered by the bicluster are all above or below the background.
-
-The ISA is developed to find biclusters (or modules as most of the ISA
-papers call them) that have correlated rows and columns. More
-precisely, the rows in the bicluster need to be only correlated across
-the columns of the bicluster and vice versa.
-
-Fig.~\ref{fig:twoclusters} shows possibly the simplest
-example of a (rather artificial) data matrix with very strong modular
-structure. It is a $20\times 20$ matrix and has two correlated blocks,
-each of them of size $10\times 10$.
-\begin{figure}
-<<toymatrix,echo=FALSE,fig=TRUE,width=6,height=3>>=
-library(isa)
-library(Matrix)
-data <- data2 <- isa.in.silico(20,20,2,10,10)[[1]]
-perm1 <- sample(seq_len(nrow(data)))
-data2 <- data2[perm1,]
-perm2 <- sample(seq_len(ncol(data)))
-data2 <- data2[,perm2]
-
-alldata <- structure(c(data2, data), dim=c(nrow(data),ncol(data),2), 
-                     dimnames=list(NULL,NULL,c("Original matrix",
-                     "Reordered matrix")))
-
-rx <- range(alldata, finite = TRUE)
-nn <- 100
-n0 <- min(nn, max(0, round((0 - rx[1])/(rx[2] - rx[1]) * nn)))
-col.regions <- c(colorRampPalette(c("blue3", "gray80"))(n0), 
-                 colorRampPalette(c("gray75", "red3"))(nn - n0))
-
-lp <- levelplot(alldata, xlab="", ylab="", 
-                scale=list(y=list(draw=FALSE),
-                  x=list(draw=FALSE)),
-                col.regions=col.regions, between=list(x=0.5))
-print(lp)
-@ 
-\caption{An artificial data matrix, on the left. On the right the
-  reordered data matrix with two blocks.}
-\label{fig:twoclusters}
-\end{figure}
-
-\section{How ISA works}
-
-Before showing an actual ISA tool chain, a few words about how the
-algorithm works are in order.
-
-\subsection{ISA iteration}
-
-ISA works in an iterative way. For an $E (m\times n)$
-input matrix it starts from a seed vector $r_0$, which is
-typically a sparse 0/1 vector of length $m$. The non-zero elements in
-the seed vector define a set of rows in $E$. Then the transposed of
-$E$, $E'$ is multiplied by $r_0$ and the result is thresholded.
-
-The thresholding is an important step of the ISA, without thresholding
-ISA would be equivalent to a (not too effective) numerical singular
-value decomposition (SVD) algorithm. Currently thresholding is done by
-calculating the mean and standard deviation of the vector and keeping
-only elements that are further than a given number of standard
-deviations from the mean. Based on the \Rargument{direction} parameter,
-this means keeping values that are significantly higher than the
-mean (``\emph{up}''), or keeping the ones that are significantly lower
-than the mean (``\emph{down}''); or keeping both
-(``\emph{updown}'').
-
-The thresholded vector $c_0$ is the (column)
-\emph{signature} of $r_0$. Then the (row) signature of
-$c_0$ is calculated, $E$ is multiplied by $c_0$ and then thresholded
-to get $r_1$.
-
-This iteration is performed until it converges, i.e. $r_{i-1}$
-and $r_i$ are \emph{close}, and $c_{i-1}$ and
-$c_i$ are also close. The convergence criteria,
-i.e. what \emph{close} means is by default defined by high Pearson
-correlation.
-
-It is very possible that the ISA finds the same modules more than once;
-two or more seeds might converge to the same module. The function
-\Rfunction{isa.unique} eliminates every module from the result of 
-\Rfunction{isa.iterate} that is very similar (in terms of
-Pearson correlation) to the one that was already found before.
-
-It might be also apparent from the description of ISA, that the
-biclusters are soft, i.e. they might have an overlap in their rows,
-columns, or both. It is also possible that some rows and/or
-columns of the input matrix are not found to be part of any ISA
-biclusters. Depending on the stringency parameters in the
-thresholding (i.e. how far the values should be from the mean), it
-might even happen that ISA does not find any biclusters.
-
-\subsection{Parameters}
-
-The two main parameters of ISA are the two thresholds (one for the
-rows and one for the columns). They basically define the stringency of
-the modules. If the row threshold is high, then the modules will have
-very similar rows. If it is mild, then modules will be bigger, with
-less similar rows than in the first case. The same applied to the
-column threshold and the columns of the modules.
-
-\subsection{Random seeding and smart seeding}
-
-By default (i.e. if the \Rfunction{isa} function is used) the ISA is
-performed from random sparse starting seeds, generated by the
-\Rfunction{generate.seeds} function. This way the algorithm is 
-completely unsupervised, but also stochastic: it might give different
-results for different runs.
-
-It is possible to use non-random seeds as well, if you have some
-knowledge about the data or are interested in a particular subset of
-rows/columns, then you can feed in your seeds into the
-\Rfunction{isa.iterate} function directly. In this case the
-algorithm is deterministic, for the same seed you will always get the
-same results. This can be considered as a semi-supervised approach.
-
-\subsection{Normalization}
-
-On in silico data we observed that ISA has the best performance if the
-input matrix is normalized (see \Rfunction{isa.normalize}). The
-normalization produces two matrices: $E_r$ and
-$E_c$. $E_r$ is calculated by transposing $E$ and
-centering and scaling its rows (see \Rfunction{scale}). $E_c$ is
-calculated by centering and scaling the rows of $E$. $E_r$ is
-used to calculate the column signature of rows and $E_c$ is used
-to calculate the signature of the columns.
-
-It is possible to use another normalization. In this case the user is
-requested to supply the normalized input data in a named list,
-including the two matrices of appropriate
-dimensions to the \Rfunction{isa.iterate} function.
-
-The \texttt{Er} entry of the list will be used for calculating the
-signature of the rows, \texttt{Ec} will be used for the signature of the
-columns. If you want to use the same matrix in both steps, then supply
-it twice, the first one transposed.
-
-\subsection{Row and column scores}
-
-In addition to finding biclusters in the input matrix, the ISA also
-assigns scores to the rows and columns, separately for each
-module. The scores are between minus one and one and they are by
-definition zero for the rows/columns that are not included in the
-module. For the non-zero entries, the further the score of a
-row/columns is from zero, the stronger the association between the
-row/column and the module. If the sign of two rows/columns are the
-same, then they are correlated, if they have opposite signs, then they
-are anti-correlated.
-
-\subsection{Robustness}
-As ISA is an unsupervised algorithm, it may very well find some
-modules, even if you feed in noise as an input matrix. To avoid these
-spurious modules we defined a robustness measure, a single number for
-a module that measures how well the rows and the columns are
-correlated. The robustness score is a generalization of the first
-singular value of the matrix. If there would be no thresholding, then
-the ISA would (almost always) converge to the leading SVD vector and
-the robustness score would be the corresponding singular value.
-
-It recommended that the user uses \Rfunction{isa.filter.robust} to
-run ISA on the scrambled input matrix with the same threshold
-parameters and then drop every module, which has a robustness score
-lower than the highest robustness score among modules found in the
-scrambled data.
-
-\section{A simple work flow}
-
-The simplest way to use ISA on your data is by calling the
-\Rfunction{isa} function with your input matrix as the single
-arguments. This function uses random seeding and has three optional
-arguments: 
-\begin{description}
-  \item[\Rargument{thr.row}] A numeric vector, the row thresholds to
-    use. By default this is $1,1.5,2,2.5,3$. 
-  \item[\Rargument{thr.col}] A numeric vector, the column thresholds
-    to use. By default this is $1,1.5,2,2.5,3$.
-  \item[\Rargument{no.seeds}] The number of random seed vectors to
-    generate. By default 100 seeds are generated.
-\end{description}
-
-The \Rfunction{isa} function runs the ISA algorithm for all threshold
-combinations of the supplied row and column thresholds. For every
-single run the same set of random seeds are used.
-
-Let us see an example. We load the \Rpackage{isa} generate some in
-silico data with the \Rfunction{isa.in.silico} function first:
-<<insilico,cache=TRUE>>=
-data <- isa.in.silico(200,200,2,50,50)[[1]]
-@
-The function call above generates the data matrix in
-Fig.~\ref{fig:twoclusters}. The first two arguments define the size of
-the data matrix ($200\times 200$), the third the number of modules (two)
-and the fourth and fifth give the size of the modules, both are
-$50\times 50$ submatrices in this case. By default
-\Rfunction{isa.in.silico} creates non-overlapping modules with some
-background noise.
-
-\Rfunction{isa.in.silico} returns a list with a list with three
-elements, we are most interested in the first now, that is the actual
-artificial data matrix. The second and the entries in the list give
-the correct module memberships of the elements, see the manual page of
-\Rfunction{isa.in.silico} for the details. Let's take a look at the
-input data, see Fig.~\ref{fig:block}.
-
-<<blockplot.pre,echo=FALSE,results=hide>>=
-png("blockplot.png", 800, 600)
-@ 
-<<blockplot>>=
-print(image(Matrix(data)))
-@ 
-<<blockplot.post,echo=FALSE,results=hide>>=
-dev.off()
-@
-
-\begin{figure}
-\centering
-\includegraphics[width=\textwidth]{blockplot}
-\caption{Another artificial data matrix with two modules and a lot of
-  noisy elements.}
-\label{fig:block}
-\end{figure}
-
-All we have to do now is calling the \Rfunction{isa} function on the
-data matrix. 
-<<seed,echo=FALSE>>=
-## Just to get the same results
-set.seed(10)
-@ 
-<<isa1,cache=TRUE>>=
-modules <- isa(data)
-names(modules)
-@
-
-The \Rfunction{isa} function returns a list with four
-elements. \listentry{rows} and \listentry{columns} define the
-modules. The non-zero elements in the first column of both define the
-first module, the second columns define the second module, etc.
-The number of columns in \listentry{rows} (and \listentry{columns})
-correspond to the number of modules.
-<<no.mods.>>=
-ncol(modules$rows)
-@
-ISA has found \Sexpr{ncol(modules$rows)} %$
-modules in the artificial data. This seem a bit surprising, as we
-expected only two modules, so let us take a look at the modules. We
-simply plot the row scores of the individual modules
-(Fig.~\ref{fig:simpleplot}). The first two modules corresponds to the
-two blocks in the input matrix. The third module is the union of them,
-with opposite score signs. The \Rfunction{isa} function finds modules
-containing anticorrelated rows/columns, as long as the
-anti-correlation is coordinated. I.e. in the third module, the first
-50 rows \emph{always} behave in an opposite way compared to the second
-50 rows. This is why they are collected into a single module.
-
-<<simpleplot,results=hide,include=FALSE>>=
-layout(cbind(seq(ncol(modules$rows))))
-sapply(seq(ncol(modules$rows)), function(x) {
-  par(mar=c(3,5,1,1))
-  plot(modules$rows[,x], ylim=c(-1,1), 
-       ylab="scores", xlab=NA, 
-       cex.lab=2, cex.axis=1.5)
-})
-@
-
-\begin{figure}
-\centering
-\setkeys{Gin}{width=0.7\textwidth}
-<<simpleplot2,fig=TRUE,echo=FALSE,results=hide,width=8,height=6>>=
-<<simpleplot>>
-@ 
-\caption{The row scores of the three modules found in the two-block
-  artificial data matrix. A score of zero means that the row is not
-  included in the module. Notice that in the third module the first
-  50 and the second 50 rows have opposite score signs, i.e. they are
-  anti-correlated.}
-\label{fig:simpleplot}
-\end{figure}
-
-The \listentry{rundata} entry of the return value of \Rfunction{isa}
-contains data about the ISA run, all the parameters (the default
-values in our case) are included here. It has the following members:
-<<isaresult.rundata>>=
-names(modules$rundata)
-@ 
-See the documentation of \Rfunction{isa} for more about these.
-
-Finally, the listentry{seeddata} entry is a data frame, it contains
-various data about the individual modules or the seeds that were used
-to find the module. There is one row for each module. It has the
-following columns:
-<<seedcols>>=
-colnames(modules$seeddata)
-@ 
-The columns \listentry{thr.row} and \listentry{thr.col} contain the
-row and columns thresholds that were used to find the module, while
-the \listentry{rob} column contains its robustness score:
-<<seeddata>>=
-modules$seeddata
-@
-
-Finally, let us show how to transform the result of the
-\Rfunction{isa} function to a list of biclusters. Each entry of the
-list will have two sublists, the first -- named \listentry{rows} --
-will contain the rows indices of the module, the second -- named
-\listentry{columns} -- the column indices:
-<<mymodules>>=
-mymodules <- lapply(seq(ncol(modules$rows)), function(x) {
-  list(rows=which(modules$rows[,x] != 0),
-       columns=which(modules$columns[,x] != 0))
-})
-length(mymodules)
-mymodules[[1]]$rows
-mymodules[[1]]$columns
-@
-
-\section{A detailed work flow}
-
-In this section we will perform each steps of the ISA analysis
-individually. This makes sense only if the user wants to adjust the
-parameters of some steps.
-
-Let us create some in-silico data to be analyzed:
-<<insilico10>>=
-data <- isa.in.silico(200, 100, 10)
-@ 
-This will be a $\Sexpr{nrow(data[[1]])}\times \Sexpr{ncol(data[[1]])}$
-matrix, with \Sexpr{ncol(data[[2]])} modules. By default ``half'' of
-the matrix is filled with noise, so each module will be a 
-$\Sexpr{round(nrow(data[[1]])/2/10)}\times 
-\Sexpr{round(ncol(data[[1]])/2/10)}$ submatrix. See the
-input matrix on Fig.~\ref{fig:tenmodules}.
-
-<<tenmods.pre,echo=FALSE,results=hide>>=
-png("tenmods.png", 600, 800)
-@ 
-<<tenmods>>=
-print(image(Matrix(data[[1]])))
-@ 
-<<tenmods.post,echo=FALSE,results=hide>>=
-dev.off()
-@
-
-\begin{figure}
-\centering
-\includegraphics{tenmods}
-\caption{In-silico data with ten non-overlapping modules.}
-\label{fig:tenmodules}
-\end{figure}
-
-\subsection{Preparing the data}
-
-One ISA iteration consists of two thresholded matrix
-multiplications. The data preparation step means producing these two
-matrices, $E_r$ and $E_c$ from the input matrix, $E$. The default
-behavior (i.e. what the \Rfunction{isa} function does) is to
-row-wise scale and center the transposed of $E$ to get $E_r$; and to
-row-wise scale and center $E$ to get $E_c$. The
-\Rfunction{isa.normalize} function performs the appropriate scaling
-and returns $E_r$ and $E_c$ in a list:
-<<normalize>>=
-normed.data <- isa.normalize(data[[1]])
-names(normed.data)
-dim(normed.data$Er)
-dim(normed.data$Ec)
-@
-
-Let us do a quick check that the rows of the matrices are indeed
-centered and scaled.
-<<checkscalecenter>>=
-max(abs(rowSums(normed.data$Er)))
-max(abs(rowSums(normed.data$Ec)))
-max(abs(apply(normed.data$Er, 1, sd)-1))
-max(abs(apply(normed.data$Ec, 1, sd)-1))
-@
-
-\subsection{Running the ISA}
-
-To run the ISA, we need to create the starting seeds first. The
-\Rfunction{isa} function uses sparse random seeds, generated by
-\Rfunction{generate.seeds}. We also have the choice of using
-non-random seeds, e.g. if you have a matrix with gene expression data
-measured on many samples you can use ``gene'' seeds that correspond to
-biological pathways; in this case ISA does a biased search to find
-modules related to the input pathways. Or, in a case/control
-experiment one can use seed vectors that correspond to cases and look
-for modules that are differentially regulated in the control and the
-case samples.
-
-For now, we will use thousand random seeds, with different sparseness:
-<<generate.seeds>>=
-row.seeds <- generate.seeds(length=nrow(data[[1]]), count=1000, 
-                            sparsity=c(2,5,10,100))
-@ 
-Note, that vector giving the sparsity is recycled, thus we will have
-four kind of seeds, some with 2, 5, 10 and 100 rows; 250 of each of
-them.
-
-We are ready to run the ISA now. Let us assume that we only want to
-find modules that are ``above'' the background. (In this artificial
-case we actually know that there are no modules below the background.)
-The \Rargument{direction} argument is thus set to \verb+"up"+. 
-<<isa.iterate>>=
-modules <- isa.iterate(normed.data, row.seeds=row.seeds, 
-                       thr.row=2, thr.col=2, direction="up")
-@
-
-Unlike \Rfunction{isa}, \Rfunction{isa.iterate} does not merge or
-filter the modules the input seeds converge to. Consequently, the
-\verb+modules+ object contains 1000 modules, but these are not
-necessarily unique:
-<<isa.iterate.res>>=
-ncol(modules$rows)
-@
-
-It is also possibly that some seeds converge to an all-zero vector, we
-have \Sexpr{sum(apply(modules$rows==0, 2, all))} such module: % $
-<<all.zero.modules>>=
-sum(apply(modules$rows==0, 2, all))
-@
-
-The \Rfunction{isa.unique} function finds duplicate or very similar
-modules in a list returned by \Rfunction{isa.iterate} and keeps only a
-single one of them. It also eliminates all-zero modules and seeds
-that did not converge, the latter result all \verb+NA+
-columns in the \listentry{rows} and \listentry{columns} matrix:
-<<unique>>=
-modules2 <- isa.unique(normed.data, modules, cor.limit=0.9)
-@
-The \Rargument{cor.limit} argument specifies the correlation limit
-above which two modules are considered to be the same. Let us see how
-many modules are left:
-<<isa.unique.res>>=
-ncol(modules2$rows)
-@
-
-We still have \Sexpr{ncol(modules2$rows)} modules. Let us check %$
-whether each real module was find by the ISA. Remember that
-\Rfunction{isa.in.silico} stores the correct modules in the second and
-third entry of its return value.
-<<correctness>>=
-found.rows <- cor(data[[2]], modules2$rows!=0)
-found.cols <- cor(data[[3]], modules2$columns!=0)
-found <- pmin(found.rows,found.cols)
-apply(found, 1, max)
-@ 
-Out of the ten modules, \Sexpr{sum(apply(found, 1, max)-1<1e-14)} were
-correctly identified by the ISA.
-
-\subsection{Robustness of biclusters, filtering the results}
-
-There are two features that we expect from a good biclustering
-algorithm. The first is sensitivity: it is able to find all
-biclusters in the input data. The second is specificity: it should not
-generate spurious biclusters, or in other words biclusters that are
-correlated just by chance.
-
-To address this problem, we have developed a robustness measure, a
-single scalar number that gives how correlated the rows and columns of
-a given module are. This measure is a simple generalization of the
-singular value decomposition (SVD) singular value, and can be
-calculated with the \Rfunction{robustness} function:
-<<robustness>>=
-rob <- robustness(normed.data, modules2$rows, modules2$columns)
-summary(rob)
-@ 
-As the user can see, the robustness of the \Sexpr{ncol(modules2$rows)} %$
-modules varies considerably, from \Sexpr{round(min(rob),2)} to
-\Sexpr{round(max(rob), 2)}. For more insight we create a boxplot for
-the robustness scores (Fig.~\ref{fig:robbox}).
-
-<<robbox,include=FALSE,results=hide,cache=TRUE>>=
-par(cex.lab=1.5, cex.axis=1.5)
-boxplot(rob, ylab="Robustness")
-@
-
-\begin{figure}
-\centering
-\setkeys{Gin}{width=0.7\textwidth}
-<<robbox2,fig=TRUE,echo=FALSE,results=hide,cache=TRUE>>=
-<<robbox>>
-@
-\caption{Box-plot for the robustness scores of the modules found by
-  the ISA. Some modules have a much higher score than the rest.}
-\label{fig:robbox}
-\end{figure}
-
-One can use the robustness measure to filter out list of modules, in
-the following way. We permute the input matrix and run the same
-module finding procedure that we did for the original matrix. Then we
-calculate the robustness scores for both the real and the scrambled
-modules and eliminate the ``real'' modules having a robustness score
-that is less than the score of at least one scrambled module. Note,
-that this filtering can be done only if both sets of modules were
-found using the same threshold parameters. The procedure is
-implemented in the \Rfunction{isa.filter.robust} function:
-<<filter.robust>>=
-modules3 <- isa.filter.robust(data[[1]], normed.data, modules2, 
-                              perms=1, row.seeds=row.seeds)
-@
-We use the very same row seeds, as for the original matrix; this is
-not strictly requires. The \Rargument{perms} argument sets the number
-of permutations to perform.
-
-Note that the ISA run on the scrambled matrix usually takes longer,
-than on the real data, because it takes more steps for the ISA to
-converge.
-
-After the robustness filtering we have \Sexpr{ncol(modules3$rows)} %$
-modules left. This is still more than the ten correct modules, so let
-us take a closer look. First, let us create another plot with the 
-robustness scores of the remaining modules. 
-\Rfunction{isa.filter.robust} places the robustness scores in the seed
-data, so we don't need to calculate them again. See Fig.~\ref{fig:robbox2}.
-
-<<robbox3,include=FALSE,results=hide,cache=TRUE>>=
-plot(modules3$seeddata$rob, ylab="Robustness", cex.lab=1.5)
-@
-
-\begin{figure}
-\centering
-\setkeys{Gin}{width=0.7\textwidth}
-<<robbox4,fig=TRUE,echo=FALSE,results=hide,cache=TRUE>>=
-<<robbox3>>
-@
-\caption{The robustness scores of ISA modules after filtering out some
-  modules with low robustness. It is still obvious that ten modules
-  have a significantly higher score than the rest.}
-\label{fig:robbox2}
-\end{figure}
-
-The scatterplot shows that ten modules have in fact higher
-robustness scores than the rest. Let us check that these are indeed the
-ten ``real'' modules of the data.
-<<bestmods>>=
-bestmods <- which(modules3$seeddata$rob > 27.5)
-mod.cor <- pmin(cor(modules3$rows[,bestmods]!=0, data[[2]]),
-                cor(modules3$columns[,bestmods]!=0, data[[3]]))
-apply(mod.cor, 1, max)
-apply(mod.cor, 2, max)
-@
-
-Indeed, we have found the ten real modules, the ISA is both specific
-and sensitive. Of course for the analysis of real data it is much more
-difficult to filter the modules found by the biclustering algorithm,
-but the robustness based filtering still helps getting rid of spurious
-modules.
-
-Finally, let us take a look at some of the other modules, that were
-also kept after the robustness-based filtering. Here we plot nine of
-them.
-<<othermods.pre>>=
-png("othermods.png", 800, 800)
-@ 
-<<othermods>>=
-othermods <- which(modules3$seeddata$rob < 27.5)[1:5]
-print(plot.modules(modules3, othermods, data=data[[1]]))
-@ 
-<<othermods.post>>=
-dev.off()
-@
-
-\begin{figure}
-\centering
-\includegraphics{othermods}
-\caption{The input matrix and five modules with relatively low
-  robustness scores. As it turns out, each of these modules contains
-  in fact two blocks of the input matrix.}
-\label{fig:othermods}
-\end{figure}
-
-\subsection{Visualize the results}
-
-TODO
-
-\section{Features of ISA}
-
-In this section we show some examples that highlight some key features
-of the ISA: its resilience to noise and its ability to find
-overlapping modules.
-
-\subsection{Resilience to noise}
-
-To test the behavior of ISA with noisy inputs we generate a number of
-in-silico data sets with different noise levels
-(Fig.~\ref{fig:noise}).
-<<noisegen,cache=TRUE>>=
-noise <- seq(0.1,1,by=0.1)
-data <- lapply(noise, function(x) isa.in.silico(500,200,10,noise=x))
-@
-
-<<noiseplot.pre,echo=FALSE,results=hide>>=
-png("noise.png", width=800, height=600)
-@ 
-<<noiseplot>>=
-images(lapply(data, "[[", 1), names=as.character(noise), 
-       xlab="", ylab="")
-@ 
-<<noiseplot.post,echo=FALSE,results=hide>>=
-dev.off()
-@ 
-\begin{figure}
-  \centering
-  \includegraphics[width=\textwidth]{noise}
-  \caption{Modular data sets with different levels of noise. The data
-  sets are generated by adding normally distributed background noise
-  with given variance to the checkerboard matrix.}
-  \label{fig:noise}
-\end{figure}
-
-Next, we run ISA with the default parameters on all the data
-sets. This might take a while.
-<<noise.isa,cache=TRUE>>=
-modules <- lapply(data, function(x) isa(x[[1]]))
-@
-
-Let us check the sensitivity of ISA, for every real module, we pick
-one from the ones ISA has found, the one that is the most correlated
-to it.
-<<sensitivity.noise>>=
-best <- lapply(seq_along(modules), function(i) {
-  cc <- pmin(cor(modules[[i]]$rows!=0, data[[i]][[2]]),
-             cor(modules[[i]]$columns!=0, data[[i]][[3]]))
-  apply(cc, 2, max)
-})
-best.mean <- sapply(best, mean)
-@
-
-Let's create boxplots for the different noise levels.
-
-<<sensbox,include=FALSE,results=hide>>=
-boxplot(best, names=noise, xlab="Noise level", ylab="Sensitivity",
-        cex.lab=1.5)
-@
-\begin{figure}
-\centering
-<<sensbox2,fig=TRUE,echo=FALSE,results=hide,width=10,height=5>>=
-<<sensbox>>
-@ 
-\caption{The sensitivity of the ISA in the function of noise. The
-  sensitivity score is calculated by finding the best (in terms of
-  Pearson correlation) ISA module for each block in the input matrix.
-
-  ISA is very sensitive to relatively large noise levels.
-}
-\label{fig:sensbox}
-\end{figure}
-
-To test the specificity, we create boxplots for the correlation
-coefficients of all modules found by ISA and their closest real
-module.
-<<spec>>=
-spec <- lapply(seq_along(modules), function(i) {
-  cc <- pmin(cor(modules[[i]]$rows != 0, data[[i]][[2]]),
-             cor(modules[[i]]$columns != 0, data[[i]][[3]]))
-  apply(cc, 1, max)
-})
-@ 
-<<specbox,include=FALSE,results=hide>>=
-boxplot(spec, names=noise, xlab="Noise level", ylab="Specificity",
-        cex.lab=1.5)
-@
-\begin{figure}
-\centering
-<<specbox2,fig=TRUE,echo=FALSE,results=hide,width=10,height=5>>=
-<<specbox>>
-@
-\caption{Specificity of the ISA modules, in the function of
-  noise. Specificity was calculated by finding the best -- again,
-  by highest Pearson correlation -- matching block in the input matrix
-  for each ISA module. These correlation coefficients are plotted in a
-  boxplot for each noise level.
-
-  Specificity for ISA decreases approximately linearly with the
-  increase of noise.
-}
-\label{fig:specbox}
-\end{figure}
-
-\subsection{Finding overlapping biclusters}
-
-ISA biclusters might overlap in their rows, columns, or both. To
-illustrate this, we create an artificial data set with two overlapping
-blocks.
-
-<<set.seed.overlap>>=
-set.seed(1)
-@ 
-<<overlap>>=
-two.over <- isa.in.silico(100,100,2,40,40,overlap.row=10)
-@
-
-We run ISA on this input matrix, with two threshold parameters, the
-first thresholds are very mild, they are both zero.
-In other words, in each iteration step all rows/columns are kept in each
-that have a higher score than the average. The second set of
-thresholds are stricter, the scores have to be at least one standard
-deviation away from the mean to keep them.
-
-<<overlap.isa>>=
-ov.normed <- isa.normalize(two.over[[1]])
-ov.seeds <- generate.seeds(count=100, length=100)
-ov.modules.1 <- isa.iterate(ov.normed, ov.seeds, 
-                            thr.row=0, thr.col=0, direction="up")
-ov.modules.1 <- isa.unique(ov.normed, ov.modules.1)
-ov.modules.1 <- isa.filter.robust(two.over[[1]], ov.normed, ov.modules.1)
-ov.modules.2 <- isa.iterate(ov.normed, ov.seeds, 
-                            thr.row=1, thr.col=1, direction="up")
-ov.modules.2 <- isa.unique(ov.normed, ov.modules.2)
-ov.modules.2 <- isa.filter.robust(two.over[[1]], ov.normed, ov.modules.2)
-ncol(ov.modules.1$rows)
-ncol(ov.modules.2$rows)
-@
-
-ISA found \Sexpr{ncol(ov.modules.1$rows)} %$
-modules with the mild thresholds and \Sexpr{ncol(ov.modules.2$rows)} %$
-modules with the strict ones. Let us plot the original data and the
-modules found by ISA (Fig.~\ref{fig:overlapplot}).
-
-<<overlapplot.pre,echo=FALSE,results=hide>>=
-png("overlapplot.png", 800, 600)
-@ 
-<<overlapplot>>=
-plot.modules(data=two.over[[1]],
-             list(rows=cbind(ov.modules.1$rows, ov.modules.2$rows),
-                  columns=cbind(ov.modules.1$columns,
-                    ov.modules.2$column)),
-             names=c("Input matrix", paste("Module", 1:6)))
-@
-<<overlapplot.post,echo=FALSE,results=hide>>=
-dev.off()
-@
-
-\begin{figure}
-\centering
-\includegraphics{overlapplot}
-\caption{The input matrix with two overlapping blocks and the ISA
-  modules for this data set. The first three modules were found at
-  milder threshold parameters, that is why they are bigger. As you can
-  see for Modules #1 and #2, at milder thresholds there is a higher
-  probability that ISA will pick up some incorrect rows/columns.
-}
-\label{fig:overlapplot}
-\end{figure}
-
-As expected the ISA modules are in general bigger with the mild
-thresholds. ISA correctly identifies the two overlapping modules and
-also the union of them. With the stricter thresholds it find the
-non-overlapping parts of the two modules, plus their overlap as a
-separate module.
+% \tableofcontents
 
 \section{Running ISA in parallel}
+
+In this document we show a little example on how to speed up ISA
+analysis by running the ISA iterations in parallel, on a computer
+cluster, or a multiprocessor machine.
 
@@
 <<parallelloadpackage>>=
+library(isa)
 library(snow)

### 2009-05-13 isa/inst/doc/ISA_tutorial.Rnw
isa: split parallel running example into another vignette.

@@
 
-\section{Running ISA in parallel}
-
-Since a typical ISA analysis consists of using a range of row/column
-and these runs are independent of each other; it is trivial to
-parallelize the task by performing the iterations for different
-threshold parameters on different processors or computers. Here we
-show and example on how to do this easily with the \Rpackage{snow} and
-the \Rpackage{Rmpi} packages:
-<<parallelloadpackage>>=
-library(snow)
-library(Rmpi)
-@
-
-We generate some simple in-silico data.
-<<paralleldata>>=
-pdata <- isa.in.silico()
-@
-
-Create the MPI cluster, with eight working nodes. You need to have a
-working MPI installation for this. See more in the documentation of
-the \Rpackage{Rmpi} and \Rpackage{snow} packages.
-<<createcluster>>=
-clu <- makeMPIcluster(8)
-invisible(clusterEvalQ(clu, library(isa)))
-clusterExport(clu, "pdata")
-@
-
-Create a big matrix in which each row is a combination of the
-threshold parameters.
-<<parthr>>=
-thr <- seq(1,3,by=0.2)
-thr.list <- expand.grid(thr, thr)
-@
-
-First we run the ISA on a single processor only, and measure the
-running time.
-<<seqrun,cache=TRUE>>=
-system.time(modules <- isa(pdata[[1]], thr.row=thr, thr.col=thr))
-@
-
-Let us now do a parallel run, again, with measuring the running
-time. If you are really running this on multiple CPUs, then it is much
-faster.
-<<parrun,cache=TRUE>>=
-system.time(modules.par <- parApply(clu, thr.list, 1, function(x) {
-  isa(pdata[[1]], thr.row=x[1], thr.col=x[2])
-}))
-@
-
-Finally, stop the cluster.
-<<closecluster>>=
-stopCluster(clu)
-@
-
 \section{Session information}

### 2009-05-14 isa2/inst/doc/ISA_tutorial.Rnw
isa package renamed to isa2.

--- 2009-05-13 isa/inst/doc/ISA_tutorial.Rnw
@@
 To run the typical ISA work flow with default parameters on your data
-matrix, just load the \Rpackage{isa} package and call the
+matrix, just load the \Rpackage{isa2} package and call the
 \Rfunction{isa} function with your matrix as the single argument. The
@@
 <<toymatrix,echo=FALSE,fig=TRUE,width=6,height=3>>=
-library(isa)
+library(isa2)
 library(Matrix)
@@
 
-Let us see an example. We load the \Rpackage{isa} generate some in
+Let us see an example. We load the \Rpackage{isa2} generate some in
 silico data with the \Rfunction{isa.in.silico} function first:
@@
 Cs\'ardi, G. (Apr 1, 2009).
-\newblock {\em isa: The Iterative Signature Algorithm}.
+\newblock {\em isa2: The Iterative Signature Algorithm}.
 \newblock R package version 0.1.

### 2009-05-14 isa2/R/isa3.R

@@ -237,11 +237,11 @@ isa.step <- function(normed.data, rows, thr.row, thr.col, direction) {
 
   filter <- function(x, t, dir) {
     if (dir=="updown") {
-      x <- .Call("beta_filter_updown_vart", x, as.double(t), PACKAGE="isa")
+      x <- .Call("beta_filter_updown_vart", x, as.double(t), PACKAGE="isa2")
     } else if (dir=="up") {
-      x <- .Call("beta_filter_up_vart", x, as.double(t), PACKAGE="isa")
+      x <- .Call("beta_filter_up_vart", x, as.double(t), PACKAGE="isa2")
     } else { ## dir=="down"
-      x <- .Call("beta_filter_down_vart", x, as.double(t), PACKAGE="isa")
+      x <- .Call("beta_filter_down_vart", x, as.double(t), PACKAGE="isa2")
     }
   }
 
@@ -337,11 +337,11 @@ isa.row.from.col <- function(normed.data, col.seeds, thr.row, direction) {
 
   filter <- function(x, t, dir) {
     if (dir=="updown") {
-      x <- .Call("beta_filter_updown_vart", x, as.double(t), PACKAGE="isa")
+      x <- .Call("beta_filter_updown_vart", x, as.double(t), PACKAGE="isa2")
     } else if (dir=="up") {
-      x <- .Call("beta_filter_up_vart", x, as.double(t), PACKAGE="isa")
+      x <- .Call("beta_filter_up_vart", x, as.double(t), PACKAGE="isa2")
     } else { ## dir=="down"
-      x <- .Call("beta_filter_down_vart", x, as.double(t), PACKAGE="isa")
+      x <- .Call("beta_filter_down_vart", x, as.double(t), PACKAGE="isa2")
     }
   }
 

### 2009-05-14 isa2/R/zzz.R

@@ -1,5 +1,5 @@
 .onLoad <- function(dir, package) {
-  library.dynam("isa", package, dir, local=FALSE);
+  library.dynam("isa2", package, dir, local=FALSE);
 
   ## Default ISA options
   isa.options[["verbose"]] <-  FALSE
@@ -8,9 +8,9 @@
 }
 
 .onUnload <- function(libpath) {
-  library.dynam.unload("isa", libpath)
+  library.dynam.unload("isa2", libpath)
 }
 
 .Last.lib <- function(libpath) {
-  isa::.onUnload(libpath)
+  isa2::.onUnload(libpath)
 }

### 2009-05-29 isa2/inst/doc/ISA_tutorial.Rnw
isa2: updated vignettes.

@@
 example of a (rather artificial) data matrix with very strong modular
-structure. It is a $20\times 20$ matrix and has two correlated blocks,
+structure. It is a $20\times 20$ matrix and (after reordering its rows
+and columns) it has two correlated blocks,
 each of them of size $10\times 10$.
@@
 $c_i$ are also close. The convergence criteria,
-i.e. what \emph{close} means is by default defined by high Pearson
+i.e. what \emph{close} means, is by default defined by high Pearson
 correlation.
 
-It is very possible that the ISA finds the same modules more than once;
+It is very possible that the ISA finds the same module more than once;
 two or more seeds might converge to the same module. The function
@@
 very similar rows. If it is mild, then modules will be bigger, with
-less similar rows than in the first case. The same applied to the
+less similar rows than in the first case. The same applies to the
 column threshold and the columns of the modules.
@@
 
-It is possible to use non-random seeds as well, if you have some
+It is possible to use non-random seeds as well. If you have some
 knowledge about the data or are interested in a particular subset of
@@
 algorithm is deterministic, for the same seed you will always get the
-same results. This can be considered as a semi-supervised approach.
+same results. Using smart (i.e. non-random) seeds can be considered as
+a semi-supervised approach.
 
@@
 
-On in silico data we observed that ISA has the best performance if the
+Using in silico data we observed that ISA has the best performance if the
 input matrix is normalized (see \Rfunction{isa.normalize}). The
@@
 $E_c$. $E_r$ is calculated by transposing $E$ and
-centering and scaling its rows (see \Rfunction{scale}). $E_c$ is
+centering and scaling its rows (see the \Rfunction{scale} R function). $E_c$ is
 calculated by centering and scaling the rows of $E$. $E_r$ is
@@
 correlated. The robustness score is a generalization of the first
-singular value of the matrix. If there would be no thresholding, then
+singular value of the matrix. If there would be no thresholding during
+the ISA iteration, then
 the ISA would (almost always) converge to the leading SVD vector and
@@
 
-It recommended that the user uses \Rfunction{isa.filter.robust} to
+It is recommended that the user uses \Rfunction{isa.filter.robust} to
 run ISA on the scrambled input matrix with the same threshold
@@
 \Rfunction{isa} function with your input matrix as the single
-arguments. This function uses random seeding and has three optional
+argument. This function uses random seeding and has three optional
 arguments: 
@@
   \item[\Rargument{thr.row}] A numeric vector, the row thresholds to
-    use. By default this is $1,1.5,2,2.5,3$. 
+    use. By default this is
+    \Sexpr{paste(eval(formals(isa2:::isa.default)$thr.row), collapse=", ")}. % $
   \item[\Rargument{thr.col}] A numeric vector, the column thresholds
-    to use. By default this is $1,1.5,2,2.5,3$.
+    to use. By default this is
+    \Sexpr{paste(eval(formals(isa2:::isa.default)$thr.col), collapse=", ")}. % $
   \item[\Rargument{no.seeds}] The number of random seed vectors to
-    generate. By default 100 seeds are generated.
+    generate. By default 
+    \Sexpr{formals(isa2:::isa.default)$no.seeds} seeds are generated. % $
 \end{description}
@@
 
-Let us see an example. We load the \Rpackage{isa2} generate some in
+Let us see an example. We load the \Rpackage{isa2} package, and
+generate some in 
 silico data with the \Rfunction{isa.in.silico} function first:
 <<insilico,cache=TRUE>>=
+library(isa2)
 data <- isa.in.silico(200,200,2,50,50)[[1]]
 @
-The function call above generates the data matrix in
-Fig.~\ref{fig:twoclusters}. The first two arguments define the size of
+The function call above generates a data matrix similar to the one in
+Fig.~\ref{fig:twoclusters}, but a bigger one. The first two arguments define the size of
 the data matrix ($200\times 200$), the third the number of modules (two)
@@
 
-\Rfunction{isa.in.silico} returns a list with a list with three
+\Rfunction{isa.in.silico} returns a list with three
 elements, we are most interested in the first now, that is the actual
-artificial data matrix. The second and the entries in the list give
+artificial data matrix. The second and third entries in the list give
 the correct module memberships of the elements, see the manual page of
@@
 two blocks in the input matrix. The third module is the union of them,
-with opposite score signs. The \Rfunction{isa} function finds modules
+with opposite score signs. The \Rfunction{isa} function also finds modules
 containing anticorrelated rows/columns, as long as the
@@
 50 rows \emph{always} behave in an opposite way compared to the second
-50 rows. This is why they are collected into a single module.
+50 rows. This is why they are collected into a single module. The
+following code creates Fig.~\ref{fig:simpleplot}.
 
@@
 The \listentry{rundata} entry of the return value of \Rfunction{isa}
-contains data about the ISA run, all the parameters (the default
+contains information about the ISA run, all the parameters (the default
 values in our case) are included here. It has the following members:
@@
 
-Finally, the listentry{seeddata} entry is a data frame, it contains
-various data about the individual modules or the seeds that were used
-to find the module. There is one row for each module. It has the
+Finally, the \listentry{seeddata} entry is a data frame, it contains
+various information about the individual modules and the seeds that
+were used to find them. There is one row for each module. It has the
 following columns:
@@
 @
+See the documentation of the \Rfunction{isa} function for more
+information about these and other fields.
 
@@
 individually. This makes sense only if the user wants to adjust the
-parameters of some steps.
+parameters of some steps. Otherwise a simple call to the
+\Rfunction{isa} function would do.
 
@@
 experiment one can use seed vectors that correspond to cases and look
-for modules that are differentially regulated in the control and the
+for modules that are different in the control and the
 case samples.
@@
 single one of them. It also eliminates all-zero modules and seeds
-that did not converge, the latter result all \verb+NA+
-columns in the \listentry{rows} and \listentry{columns} matrix:
+that did not converge, \verb+NA+
+columns in the \listentry{rows} and \listentry{columns} matrix
+correspond to non-convegent seeds:
 <<unique>>=
@@
 @
-We use the very same row seeds, as for the original matrix; this is
-not strictly requires. The \Rargument{perms} argument sets the number
+We used the very same row seeds, as for the original matrix; this is
+not strictly required. The \Rargument{perms} argument sets the number
 of permutations to perform.
@@
 \caption{The robustness scores of ISA modules after filtering out some
-  modules with low robustness. It is still obvious that ten modules
-  have a significantly higher score than the rest.}
+  modules with low robustness. Ten modules
+  have a higher score than the rest.}
 \label{fig:robbox2}
@@
 <<bestmods>>=
-bestmods <- which(modules3$seeddata$rob > 27.5)
+bestmods <- order(modules3$seeddata$rob, decreasing=TRUE)[1:10]
 mod.cor <- pmin(cor(modules3$rows[,bestmods]!=0, data[[2]]),
@@
 Indeed, we have found the ten real modules, the ISA is both specific
-and sensitive. Of course for the analysis of real data it is much more
+and sensitive. If one analyzes real data, then it is much more
 difficult to filter the modules found by the biclustering algorithm,
@@
 Finally, let us take a look at some of the other modules, that were
-also kept after the robustness-based filtering. Here we plot nine of
-them.
+also kept after the robustness-based filtering. Here we plot five of
+them, together with the input matrix.
 <<othermods.pre,echo=FALSE,results=hide>>=
@@
 <<othermods>>=
-othermods <- which(modules3$seeddata$rob < 27.5)[1:5]
+othermods <- order(modules3$seeddata$rob)[1:5]
 print(plotModules(modules3, othermods, data=data[[1]]))
@@
   robustness scores. As it turns out, each of these modules contains
-  in fact two blocks of the input matrix.}
+  two blocks of the input matrix.}
 \label{fig:othermods}
@@
 
-Visualizing overlapping biclusters is quite a challenging task. We
+Visualizing overlapping biclusters is a challenging task. We
 show simple methods that usually visualize a single bicluster 
@@
 functions with the same output format: \Rfunction{isa.iterate},
-Rfunction{isa.unique}, etc.
-
-In fact \Rfunction{plotModules} calls the
+\Rfunction{isa.unique}, etc. (In fact \Rfunction{plotModules} calls the
 \Rfunction{images} function to do its job. \Rfunction{images} takes a
-list of matrices, see Fig.~\ref{fig:noise} for an example.
+list of matrices, see Fig.~\ref{fig:noise} for an example.)
 
@@
 in a bicluster with a different color. These plots visualize the
-difference between bicluster and the rest of the matrix, see
-Fig.\ref{fig:parallel} for an example.
+difference between the bicluster and the rest of the matrix, see
+Fig.~\ref{fig:parallel} for an example.
 <<parallel.pre,include=TRUE>>=
@@
 \centering
-\setkeys{Gin}{width=0.6\textwidth}
-<<plotclust,fig=TRUE,echo=FALSE,results=hide>>=
+\setkeys{Gin}{width=0.8\textwidth}
+<<plotclust,fig=TRUE,echo=FALSE,results=hide,width=10,height=7>>=
 <<plotclust.pre>>
 @ 
-\caption{Bar plots to show the difference between the biclusters and
+\caption{Bar plots show the difference between the biclusters and
   the rest of the input matrix, for the first five modules found by
@@
 
-Let's create boxplots for the different noise levels.
+Let's create boxplots for the different noise
+levels. (Fig.~\ref{fig:sensbox}.)
 
@@
   Pearson correlation) ISA module for each block in the input matrix.
-  ISA is very sensitive to relatively large noise levels.
+  ISA performs very well, even in the presence of relatively high
+  noise levels.
 }
@@
 coefficients of all modules found by ISA and their closest real
-module.
+module. (Fig.~\ref{fig:specbox}.)
 <<spec>>=
@@
 thresholds. ISA correctly identifies the two overlapping modules and
-also the union of them. With the stricter thresholds it find the
+also the union of them. With the stricter thresholds it finds the
 non-overlapping parts of the two modules, plus their overlap as a

### 2009-06-10 eisa/inst/doc/ISA_tutorial.Rnw
eisa vignette extended a bit.

@@
 
-\title{The Iterative Signature Algorithm}
+\title{The Iterative Signature Algorithm on Gene Expression data}
 \author{G\'abor Cs\'ardi}
@@
 
-In this tutorial we will use the data in the \Rpackage{ALL}
-package. This is a data set from a clinical trial in acute
-lymphoblastic leukemia and it contains 128 samples altogether.
-
 \section{Preparing the data}
@@
 
-First, we load the required packages and the data to analyze.
-<<>>=
-library(isa)
+First, we load the required packages and the data to analyze. ISA is
+implemented in the \Rpackage{eisa} and \Rpackage{isa2} packages, see
+Section{sec:isapackages} for a more elaborated summary about the two
+packages. It is enough to load the \Rpackage{eisa} package,
+\Rpackage{isa2} and other required packages are loaded automatically.
+\Rpackage{eisa} 
+<<load the packages>>=
+library(eisa)
+@
+
+In this tutorial we will use the data in the \Rpackage{ALL}
+package. This is a data set from a clinical trial in acute
+lymphoblastic leukemia and it contains 128 samples altogether.
+<<load the data>>=
 library(ALL)
@@
 
-\subsection{Filtering the genes}
+\section{Simple ISA runs}
 
-<<>>=
-library(genefilter)
-f1 <- pOverA(0.25, log2(100))
-f2 <- function(x) (IQR(x) > 0.5)
-ff <- filterfun(f1, f2)
-selected <- genefilter(ALL, ff)
-ALL2 <- ALL[selected,]
+The simplest way to run ISA is to choose the two threshold parameters
+and then call the \Rfunction{isa} function on the
+\Rclass{ExpressionSet} object:
+<<simple ISA>>=
+thr.gene <- 2.7
+thr.cond <- 1.4
+set.seed(1) # to get the same results, always
+biclusts <- isa(ALL, thr.gene=thr.gene, thr.cond=thr.cond)
 @
 
-\subsection{Generating seeds for the ISA}
+This first applies a non-specific filter to the data set filters the
+data set and then runs ISA from
+\Sexpr{formals(eisa:::isa.ExpressionSet)$no.seeds} % $
+random seeds (the default). See Section~\ref{sec:detailed-isa} if the
+default parameters are not appropriate for you and need more control.
 
-<<>>=
-g <- generate.seeds(count=100, length=nrow(ALL2), gs=2/nrow(ALL2))
+\section{Inspect the result}
+
+The \Rfunction{isa} function returns an \Rclass{ISAModules} object. By
+typing in its name we can get a brief summary of the results:
+<<type in name of ISAModules object>>=
+biclusts
 @
 
-\subsection{Normalize the data}
+There are various other \Rclass{ISAModules} methods that help to
+access the modules themselves and the ISA parameters that were used
+for the run.
 
-<<>>=
-ALL2.normed <- isa.normalize(ALL2)
+% TODO: length, dim, featureNames, sampleNames, pData, 
+% annotation, organism, indexing
+
+The \Rfunction{getNoFeatures} function returns a numeric vector, the
+number of features (probesets in our case) in each module. Similarly,
+\Rfunction{getNoSamples} returns a numeric vector, the number of
+samples in each module. 
+<<number of features and samples>>=
+getNoFeatures(biclusts)
+getNoSamples(biclusts)
 @
 
-\section{Running the ISA}
-
-<<>>=
-thr.gene <- 2.7
-thr.cond <- 1.4
+\Rfunction{getFeatureNames} lists the probes (more precisely,
+the feature names coming from the \Rclass{ExpressionSet} object) in
+the modules. It returns a list, here we just print the first entry in
+the list.
+<<list genes in modules>>=
+getFeatureNames(biclusts)[[1]]
 @
 
-<<>>=
-biclusts <- isa(ALL2.normed, thr.gene=thr.gene, thr.cond=thr.cond)
+The \Rfunction{getSampleNames} function does the same for the samples.
+Again, the sample names are taken from the \Rclass{ExpressionSet}
+object that was passed to \Rfunction{isa}:
+<<list conditions in modules>>=
+getSampleNames(biclusts)[[1]]
 @
 
+Remember, that the ISA biclustering is not binary, every feature (and
+similarly, every sample) has a score between -1 and 1; the further the
+score is from zero the stronger the association between the feature
+(or sample) and the module. If two features both have scores with the
+same sign, then they are correlated, if the sign of their scores are
+opposite, then they are anti-correlated. You can query the scores of
+the features with the \Rfunction{getFeatureScores} function, and
+similarly, the \Rfunction{getSampleScores} function queries the sample
+scores. You can supply the modules you want to query as an optional
+argument:
+<<query scores>>=
+getFeatureScores(biclusts, 3)
+getSampleScores(biclusts, 3)
+@
+
+You can also query the scores in a matrix form, that is probably
+better if you need many or all of them at the same time. The
+\Rfunction{getFeatureMatrix} and \RFunction{getSampleMatrix} functions
+are defined for this:
+<<query all scores>>=
+dim(getFeatureMatrix(biclusts))
+dim(getSampleMatrix(biclusts))
+@
+
+Objects from the \Rclass{ISAModules} class store various information
+about the ISA run and the convergence of the seeds. Information
+associated with the individual seeds can be queried with the
+\Rfunction{seedData} function, it returns a data frame, with as many
+rows as the number of seeds and various seed-level information,
+e.g. the number of iterations required for the seed to converge. See
+the manual page of \Rfunction{isa} for details.
+<<seed data>>=
+seedData(biclusts)
+@
+
+The \Rfunction{runData} function returns additional information about
+the ISA run, seed the \Rfunction{isa} manual page for details.
+<<run data>>=
+runData(biclusts)
+@
+
+\section{Enrichment calculations}
+
+The \Rpackage{eisa} package provides some functions to perform
+enrichment tests for the genes in the various modules against various
+databases. These tests are usually simplified and less sophisticated
+versions than the ones in the \Rpackage{Category}, \Rpackage{GOstats}
+or \Rpackages{topGO} packages, but they are much faster and this is
+important if we need to perform them for many modules.
+
+\subsection{Gene Ontology}
+
+To perform enrichment analysis against the Gene Ontology database, all
+you have to do is to supply your \Rclass{ISAModules} object to the
+\Rfunction{isa.GO} function.
+
+\subsection{KEGG Pathway Database}
+
+\subsection{Chromosomes}
+
+\subsection{Predicted $\mu$RNA targets 
+  from the TargetScan database}
+
+\subsection{Transcription regulation from the DBD database}
+
 \section{Visualizing the results}
+
+\subsection{Generating a HTML summary for the modules}
+
+\section{The \Rpackage{isa2} and \Rpackage{eisa} packages}%
+\label{sec:isapackages}
+
+ISA and its companion functions for visualization, functional
+enrichment calculation, etc. are distributed in two separate packages,
+\Rpackage{isa2} and \Rpackage{eisa}.
+
+\Rpackage{isa2} contains the implementation of ISA itself, and
+\Rpackage{eisa} specifically deals with supplying expression data to
+\Rpackage{isa2} and visualizing the results.
+
+The two packages provide some functions that share the same name, and
+the right one is selected based on the arguments supplied. E.g. both
+\Rpackage{isa2} and \Rpackage{eisa} has a function called
+\Rfunction{isa}. It the first argument of \Rfunction{isa} is an
+\Rclass{ExpressionSet}, then the \Rpackage{eisa} version is called. If
+the first argument is a matrix, then the \Rpackage{isa} version is
+called.
+
+This also means that there are two manual pages for the two
+\Rfunction{isa} functions. I.e. if one types `\verb+?isa+', (and both
+\Rpackage{isa2} and \Rpackage{eisa} are loaded, then one
+has to choose between the two versions. To avoid this, the
+\Rpackage{eisa} manual pages can also be referenced by supplying and
+`\verb+e+' prefix: `\verb+?eisa+' or `\verb+?eisa.unique+'.
+
+\section{Finer control over ISA parameters}%
+\label{sec:detailed-isa}
 

### 2009-06-30 isa2/R/isa3.R
isa2: prenormalize by default.

@@ -2,7 +2,7 @@
 setMethod("isa.normalize", signature(data="matrix"),
           function(data, ...) isa.normalize.default(data, ...))
 
-isa.normalize.default <- function(data, prenormalize=FALSE) {
+isa.normalize.default <- function(data, prenormalize=TRUE) {
 
   isa.status("ISA normalization", "in")
 

### 2009-06-30 eisa/R/autogen.R
eisa: visualization functions have somewhat nicer interfaces now.

@@ -230,7 +230,7 @@ autogen.modules <- function(nm, isares, modules=seq_len(length(isares)),
                             template=system.file("autogen", package="eisa"),
                             GO=NULL, KEGG=NULL, miRNA=NULL, CHR=NULL, DBD=NULL,
                             cond.to.include=NULL,
-                            markup=numeric(), markdown=numeric(),
+                            cond.col="white",
                             sep=NULL, seed=NULL, drive.BP=NULL,
                             drive.CC=NULL, drive.MF=NULL, drive.KEGG=NULL,
                             drive.miRNA=NULL, drive.DBD=NULL, drive.CHR=NULL) {
@@ -275,7 +275,7 @@ autogen.modules <- function(nm, isares, modules=seq_len(length(isares)),
     isa.autogen.module(nm, isares, x, target.dir=target.dir, template=template,
                        GO=GO, KEGG=KEGG, miRNA=miRNA, DBD=DBD, CHR=CHR,
                        cond.to.include=cond.to.include,
-                       markup=markup, markdown=markdown, sep=sep,
+                       cond.col=cond.col, sep=sep,
                        seed=seed, drive.BP=drive.BP, drive.CC=drive.CC,
                        drive.MF=drive.MF, drive.KEGG=drive.KEGG,
                        drive.miRNA=drive.miRNA, drive.DBD=drive.DBD,
@@ -290,7 +290,7 @@ autogen.modules <- function(nm, isares, modules=seq_len(length(isares)),
 
 isa.autogen.module <- function(nm, isares, module, target.dir, template,
                                GO, KEGG, miRNA, CHR, DBD, cond.to.include,
-                               markup, markdown, sep=NULL,
+                               cond.col="white", sep=NULL,
                                seed=NULL, drive.BP=NULL, drive.CC=NULL,
                                drive.MF=NULL, drive.KEGG=NULL,
                                drive.miRNA=NULL, drive.DBD=NULL,
@@ -710,8 +710,8 @@ isa.autogen.module <- function(nm, isares, module, target.dir, template,
 
   print(paste("Module", m, "expression graph"))
 
-  ep <- exp.plot.create(nexp, getFeatureMatrix(isares, m)[[1]],
-                        getSampleMatrix(isares, m)[[1]], normalize=FALSE)
[…]

### 2009-06-30 eisa/R/viz.R

@@ -7,8 +7,8 @@ gograph <- function(table, colbar.length=30, label.cex=1, GOGRAPHS=NULL,
   require(GO.db)
   require(igraph)
 
-  terms <- as.character(table[,1])
-  pval <- table[,2]
+  terms <- rownames(table)
+  pval <- table[,1]
 
   #######################
   # Check which GO tree
@@ -174,9 +174,9 @@ gograph <- function(table, colbar.length=30, label.cex=1, GOGRAPHS=NULL,
   g2
 }
 
-gograph.plot <- function(graph) {
+gograph.plot <- function(graph, coords=FALSE, ...) {
 
-  require(TeachingDemos)
+  if (coords) require(TeachingDemos)
   
   if (dev.cur() == 1) {
     device <- options("device")[[1]]
@@ -187,15 +187,19 @@ gograph.plot <- function(graph) {
        ylim=c(0,graph$height), axes=FALSE)
   par(xpd=TRUE)
   plot(graph, add=TRUE, asp=FALSE, rescale=FALSE, vertex.label=V(graph)$abbrv,
-       vertex.size=V(graph)$size*200, vertex.size2=V(graph)$size2*200)
+       vertex.size=V(graph)$size*200, vertex.size2=V(graph)$size2*200, ...)
 
   text(graph$layout[,1]-V(graph)$size, graph$layout[,2], V(graph)$plabel,
        pos=2, cex=1, col="blue", font=2)
   text(graph$layout[,1]+V(graph)$size, graph$layout[,2], V(graph)$label,
        pos=4, cex=1, col="darkgreen", family="mono")
 
-  co <- cnvrt.coords(graph$layout[,1]-V(graph)$size,
-                     graph$layout[,2]+V(graph)$size2)$tdev
+  if (coords) {
+    cnvrt.coords(graph$layout[,1]-V(graph)$size,
+                 graph$layout[,2]+V(graph)$size2)$tdev
+  } else {
+    invisible(NULL)
+  }
 }
 
 
@@ -351,30 +355,54 @@ exp.plot <- function(epo, scores=TRUE) {
 # coords <- exp.plot(ep)
 # dev.off()
 
-cond.plot <- function(nm, genes, thr, markup, markdown, ylim=c(-1.2,1.5), all=TRUE,
+cond.plot <- function(modules, number, exprs, nm,
+                      col="white", all=TRUE,
                       sep=NULL, sepcol=NULL, val=TRUE, srt=90,
[…]

### 2009-06-30 eisa/inst/doc/EISA_tutorial.Rnw
eisa: vignette significantly updated.

\documentclass{article}
\usepackage{ragged2e}
\usepackage{url}

\newcommand{\Rfunction}[1]{\texttt{#1}}
\newcommand{\Rpackage}[1]{\texttt{#1}}
\newcommand{\Rclass}[1]{\texttt{#1}}
\newcommand{\Rargument}[1]{\textsl{#1}}
\newcommand{\filename}[1]{\texttt{#1}}
\newcommand{\variable}[1]{\texttt{#1}}

\SweaveOpts{cache=TRUE}

\begin{document}

\title{The Iterative Signature Algorithm for Gene Expression data}
\author{G\'abor Cs\'ardi}
\maketitle

\tableofcontents

\RaggedRight

\section{Introduction}

The Iterative Signature Algorithm (ISA) is biclustering method. The
input of a biclustering method a matrix and its output is a set of
biclusters that fulfill some criteria; a bicluster is a block of the
potentially reordered input matrix.

Most commonly, this algorithm is used on microarray expression data,
to find gene sets that are coexpressed across a subset of the original
samples. In the original ISA paper the biclusters are called
transription modules (TM), we will often refer them under this name in
the following.

\section{Preparing the data}

\subsection{Loading the data}

First, we load the required packages and the data to analyze. ISA is
implemented in the \Rpackage{eisa} and \Rpackage{isa2} packages, see
Section{sec:isapackages} for a more elaborated summary about the two
packages. It is enough to load the \Rpackage{eisa} package,
\Rpackage{isa2} and other required packages are loaded automatically.
\Rpackage{eisa} 
<<load the packages>>=
library(eisa)
@

In this tutorial we will use the data in the \Rpackage{ALL}
package. 
<<load the data>>=
library(ALL)
library(hgu95av2.db)
library(affy)
data(ALL)
@
This is a data set from a clinical trial in acute
lymphoblastic leukemia and it contains \Sexpr{ncol(ALL)} samples
altogether.

\section{Simple ISA runs}

The simplest way to run ISA is to choose the two threshold parameters
and then call the \Rfunction{isa} function on the
\Rclass{ExpressionSet} object:
<<simple ISA>>=
thr.gene <- 2.7
thr.cond <- 1.4
set.seed(1) # to get the same results, always
modules <- isa(ALL, thr.gene=thr.gene, thr.cond=thr.cond)
@

This first applies a non-specific filter to the data set filters the
data set and then runs ISA from
\Sexpr{formals(eisa:::isa.ExpressionSet)$no.seeds} % $
random seeds (the default). See Section~\ref{sec:detailed-isa} if the
default parameters are not appropriate for you and need more control.

\section{Inspect the result}

The \Rfunction{isa} function returns an \Rclass{ISAModules} object. By
typing in its name we can get a brief summary of the results:
<<type in name of ISAModules object>>=
modules
@

There are various other \Rclass{ISAModules} methods that help to
access the modules themselves and the ISA parameters that were used
for the run.

% TODO: length, dim, featureNames, sampleNames, pData, 
% annotation, organism, indexing

The \Rfunction{getNoFeatures} function returns a numeric vector, the
number of features (probesets in our case) in each module. Similarly,
\Rfunction{getNoSamples} returns a numeric vector, the number of
samples in each module. 
<<number of features and samples>>=
getNoFeatures(modules)
getNoSamples(modules)
@

\Rfunction{getFeatureNames} lists the probes (more precisely,
the feature names coming from the \Rclass{ExpressionSet} object) in
the modules. It returns a list, here we just print the first entry in
the list.
<<list genes in modules>>=
getFeatureNames(modules)[[1]]
@

The \Rfunction{getSampleNames} function does the same for the samples.
Again, the sample names are taken from the \Rclass{ExpressionSet}
object that was passed to \Rfunction{isa}:
<<list conditions in modules>>=
getSampleNames(modules)[[1]]
@

Remember, that the ISA biclustering is not binary, every feature (and
similarly, every sample) has a score between -1 and 1; the further the
score is from zero the stronger the association between the feature
(or sample) and the module. If two features both have scores with the
same sign, then they are correlated, if the sign of their scores are
opposite, then they are anti-correlated. You can query the scores of
the features with the \Rfunction{getFeatureScores} function, and
similarly, the \Rfunction{getSampleScores} function queries the sample
scores. You can supply the modules you want to query as an optional
argument:
<<query scores>>=
getFeatureScores(modules, 3)
getSampleScores(modules, 3)
@

You can also query the scores in a matrix form, that is probably
better if you need many or all of them at the same time. The
\Rfunction{getFeatureMatrix} and \Rfunction{getSampleMatrix} functions
are defined for this:
<<query all scores>>=
dim(getFeatureMatrix(modules))
dim(getSampleMatrix(modules))
@

Objects from the \Rclass{ISAModules} class store various information
about the ISA run and the convergence of the seeds. Information
associated with the individual seeds can be queried with the
\Rfunction{seedData} function, it returns a data frame, with as many
rows as the number of seeds and various seed-level information,
e.g. the number of iterations required for the seed to converge. See
the manual page of \Rfunction{isa} for details.
<<seed data>>=
seedData(modules)
@

The \Rfunction{runData} function returns additional information about
the ISA run, seed the \Rfunction{isa} manual page for details.
<<run data>>=
runData(modules)
@

\section{Enrichment calculations}

The \Rpackage{eisa} package provides some functions to perform
enrichment tests for the genes in the various modules against various
databases. These tests are usually simplified and less sophisticated
versions than the ones in the \Rpackage{Category}, \Rpackage{GOstats}
or \Rpackage{topGO} packages, but they are much faster and this is
important if we need to perform them for many modules.

\subsection{Gene Ontology}

To perform enrichment analysis against the Gene Ontology database, all
you have to do is to supply your \Rclass{ISAModules} object to the
\Rfunction{isa.GO} function.
<<GO enrichment>>=
GO <- isa.GO(modules)
@

The \Rfunction{isa.GO} requires an annotation package that maps the
Entrez identifiers of the studied organism to GO terms. For human, the
\Rpackage{org.Hs.eg.db} package is required, for the mouse the
\Rpackage{org.Mm.eg.db} package, etc.

The \variable{GO} object is a list with three elements, these
correspond to the ontologies: biological function, cellular component
and molecular function, in this order.
<<list GO result>>=
GO
@

We can see the number of categories tested, this is different for each
ontology, as they have different number of terms. The gene universe
size is also different, because it contains only genes that have at
least one annotation in the given category.

For extracting the results themselves, the \Rfunction{summary}
function can be used, this converts them to a simple data frame. The
$p$-value limit can be supplied to \Rfunction{summary}. Note, that
since \Rfunction{isa.GO} calculates enrichment for many gene sets
(i.e. for all biclusters), \Rfunction{summary} returns a list of data
frames, one for each bicluster. A table for the first module:
<<GO summary>>=
summary(GO[[1]], p=0.001)[[1]]
@

The \texttt{drive} column includes the Entrez ids of the genes that
are included in the intersection of the bicluster and the GO
category. They can be also obtained with the
\Rfunction{geneIdsByCategory} function. The following returns the
genes in the first module and the second GO BP category.
<<GO gene ids by category>>=
geneIdsByCategory(GO[[1]])[[1]][[2]]
@

You can use the \Rpackage{GO.db} package to obtain more information
about the enriched GO categories.
<<GO info>>=
sigCategories(GO[[1]])[[1]]
library(GO.db)
mget(sigCategories(GO[[1]])[[1]], GOTERM)
@

In addition, the following functions are implemented to work on the
objects returned by \Rfunction{isa.GO}: \Rfunction{htmlReport},
\Rfunction{pvalues}, \Rfunction{geneCounts}, \Rfunction{oddsRatios},
\Rfunction{exprectedCounts}, \Rfunction{universeCounts},
\Rfunction{universeMappedCount}, \Rfunction{geneMappedCount},
\Rfunction{geneIdUniverse}. These functions do essentially the same as
they counterparts for \Rclass{GOHyperGResult} objects, see the
documentation of the \Rpackage{GOstats} package. The only difference
is, that since here we are testing a list of gene sets (=biclusters),
they calculate the results for all gene sets and return a list.

\subsubsection{Multiple testing correction}

By default, the \Rfunction{isa.GO} function performs multiple testing
correction using the Holm method, this can be changed via the
\texttt{correction} and \texttt{correction.method} arguments. See the
manual page of the \Rfunction{p.adjust} function for the possible
multiple testing correction schemes.

\subsection{KEGG Pathway Database}

Enrichment calculation against the KEGG pathway goes essentially the
same way as for the Gene Ontology, this time we use the
\Rfunction{isa.KEGG} function:
<<KEGG enrichment>>=
KEGG <- isa.KEGG(modules)
KEGG
summary(KEGG)[[1]]
library(KEGG.db)
mget(sigCategories(KEGG)[[1]], KEGGPATHID2NAME)
@

The functions mentioned in the Gene ontology enrichment section can be
used for KEGG, as well.

\subsection{Chromosomes}

The \Rpackage{eisa} includes a simple way to check whether the genes
in a bicluster are associated with a chromosome. See the
\Rfunction{isa.CHR} function:
<<CHR enrichment>>=
CHR <- isa.CHR(modules)
summary(CHR,p=0.05)[[11]]
@

The third bicluster has \Sexpr{summary(CHR,p=0.05)[[2]]$Count[1]} % $
genes on chromosome \Sexpr{rownames(summary(CHR,p=0.05)[[2]])[1]}.

The functions mentioned in the Gene ontology enrichment section can be
used for chromosome enrichment, as well.

\subsection{Predicted $\mu$RNA targets 
  from the TargetScan database}

$\mu$RNAs are short RNA molecules that regulate gene expression.
TargetScan is a data based of predicted target genes of $\mu$RNAs, for
severral organisms. There are two R packages that incorporate this
database, one for human and another one for mouse, right now they can
be downloaded from
\url{http://www2.unil.ch/cbg/index.php?title=Building_BioConductor_Annotation_Packages}.
The \Rpackage{targetscan.Hs.eg.db} package is for human, the
\Rpackage{targetscan.Mm.eg.db} package is for mouse.

The enrichment calculation itself is basically the same as for GO and
KEGG, but the \Rfunction{isa.miRNA} function should be used:
<<miRNA enrichment>>=
if (require(targetscan.Hs.eg.db)) {
  miRNA <- isa.miRNA(modules)
  summary(miRNA, p=1)[[7]]
}
@

%\subsection{Transcription regulation from the DBD database}

%TODO

\section{Visualizing the results}

Visualizing overlapping biclusters is a challenging task. We
show simple methods that usually visualize a single bicluster 
at a time. For some of these we will use the \Rpackage{biclust} R
package.

\subsection{The \Rpackage{biclust} package}

The \Rpackage{biclust} R package implements several biclustering
algorithms in a unified framework. It uses the class \Rclass{Biclust}
to store a set of biclusters. The \Rfunction{eisa.biclust} converts ISA
modules to a \Rclass{Biclust} object. This requires the binarization
of the modules, i.e. the ISA scores are lost, they are converted to
zeros and ones:
<<biclust,cache=TRUE>>=
library(biclust)
Bc <- eisa.biclust(modules)
Bc
@

\subsection{Image plots}

The easiest way to create a heatmap of a single module is to call the
\Rfunction{ISA2heatmap} function. You need to specify which module you
want to plot and also the \Rclass{ExpressionSet} object that is being
analyzed. Note that by default ISA normalizes the expression data
before running the module detection; the raw, non-normalized values
are plotted in the next example. If you want to plot the normalized
values, then you need to do the normalization ``by hand'', before
calling ISA, see Section~\ref{sec:detailed-isa}.
<<heatmap,fig=TRUE>>=
ISA2heatmap(modules, 1, ALL)
@

\Rfunction{ISA2heatmap} simply calls the \Rfunction{heatmap} function,
and passes additional arguments to it. See the manual of
\Rfunction{heatmap} for details.

You can also use the \Rpackage{biclust} package to create image
plots, by calling the \Rfunction{drawHeatmap} function. The result is
in Fig.~\ref{fig:drawHeatmap}.
<<heatmap biclust pre,echo=FALSE,results=hide>>=
CairoPNG("heatmap.png", 400, 400 )
@ 
<<heatmap>>=
drawHeatmap(exprs(ALL), Bc, number=1) 
@
<<heatmap post,echo=FALSE,results=hide>>=
dev.off() 
@ 
\begin{figure}
  \centering
  \includegraphics[width=0.7\textwidth]{heatmap}
  \caption{TODO}
  \label{fig:drawHeatmap}
\end{figure}

\subsection{Profile plots}

Profile plots visualize the difference between the genes (or samples)
in the modules and the rest of the expression data. A profile plot
contains a line plot for every single gene (or sample) and the genes
that belong to the module have a different color, see
Fig.~\ref{fig:parallel}.
<<parallel.pre,echo=FALSE,results=hide>>=
CairoPNG("parallel.png", 400, 400 )
@ 
<<parallel>>=
parallelCoordinates(exprs(ALL), Bc, number=1, plotBoth=TRUE, pch=NA)
@
<<parallel.post,echo=FALSE,results=hide>>=
dev.off()
@ 
\begin{figure}
\centering
\includegraphics[width=0.7\textwidth]{parallel}
\caption{Profile plots for the first bicluster found by ISA.}
\label{fig:parallel}
\end{figure}

\subsection{Gene ontology tree plots}

The GO database is organized in a hierarchical fashion, in a tree-like
structure, where the broadest category sits in the root of the tree
and broader categories are subdivided into more specific
subcategories. But the GO is not exactly a tree, as the same category
can be the subcategory of more than one broader categories: e.g. the
``Golgi vesicle transport'' category is part of both
``vesicle-mediated transport'' and ``intracellular
transport''.

The \Rpackage{eisa} package provides some functions to plot parts of
the GO graph that is related to a transcription module. The
\Rfunction{gograph} function creates an object that is a
representation of such a plot. Its input is a table with the GO
categories to plot and their enrichment $p$-values. (Additional
columns are silently ignored.) Here is how to use
it on the previously calculated enrichment scores:
<<goplot1>>=
goplot.2 <- gograph(summary(GO[[1]], p=0.05)[[2]])
@

\Rfunction{goplot} uses the \Rpackage{igraph} package to create a
graph with associated meta data:
<<gograph object>>=
summary(goplot.2)
@ 
The \texttt{width} and \texttt{height} graph attributes contain the
suggested width and height of the graph, if plotted to a bitmap
device. (The graph attributes of an \Rpackage{igraph} graph can be
queried with the `\texttt{\$}' selector:
<<gograph object 2>>=
goplot.2$width
goplot.2$height
@

Let's plot the graph, we can do this with the \Rfunction{gograph.plot}
function.
<<goplot2,eval=FALSE>>=
x11(width=10, height=10*goplot.2$height/goplot.2$width)
gograph.plot(goplot.2)
@
\setkeys{Gin}{width=\textwidth}
<<goplot2-real,echo=FALSE,width=10,height=2.5,fig=TRUE>>=
ps.options(fonts=c("serif", "mono"))
gograph.plot(goplot.2)
@

Because the GO is not really a tree, \Rfunction{gograph} ``unfolds''
it into a tree by including categories more than once, if needed. It
also abbreviates the names of the GO categories to make them fit on
the tree. The graph object contains the full name of the category as
well. The full and abbreviated names can be listed by querying the
appropriate vertex attributes of the graph. Here they are for the
first five categories:
<<goplot names>>=
V(goplot.2)$abbrv[1:5]
lapply(V(goplot.2)$desc[1:5], strwrap)
@ 
\Rfunction{gograph.plot} colors the categories according to the
supplied enrichment $p$-values, the minus $\log_{10}$ $p$-value is
also added to the plot, see the bold blue numbers.

\subsection{Sample score plots}

In many studies, especially the case-control ones, it is useful to
plot the sample scores of a module. For example if the sample scores
significantly differ for two groups of samples (e.g. cases versus
controls), then the genes in the module can be used as discriminators
between the two groups.

The \Rfunction{cond.plot} function can plot the sample scores,
potentially including the scores before the ISA filtering. Let us plot
the scores for the second module, the color code denotes B-cell
vs. T-cell leukemia.

<<condplot,fig=TRUE,width=10,height=4>>=
col <- ifelse( grepl("^B", pData(modules)$BT), "orange", "darkolivegreen")
cond.plot(modules, 2, ALL, col=col)
@

It is clear that the \Sexpr{getNoFeatures(modules)[2]} genes included
in this transcription module can separate B-cell and T-cell leukemia
samples.

\subsection{Generating a HTML summary for the modules}

\section{How ISA works}

Before showing an actual ISA tool chain, a few words about how the
algorithm works are in order.

\subsection{ISA iteration}

ISA works in an iterative way. For an $E (m\times n)$
input matrix it starts from a seed vector $r_0$, which is
typically a sparse 0/1 vector of length $m$. The non-zero elements in
the seed vector define a set of genes in $E$. Then the transposed of
$E$, $E'$ is multiplied by $r_0$ and the result is thresholded.

The thresholding is an important step of the ISA, without thresholding
ISA would be equivalent to a (not too effective) numerical singular
value decomposition (SVD) algorithm. Currently thresholding is done by
calculating the mean and standard deviation of the vector and keeping
only elements that are further than a given number of standard
deviations from the mean. Based on the \Rargument{direction} parameter,
this means keeping values that are significantly higher than the
mean (``\emph{up}''), or keeping the ones that are significantly lower
than the mean (``\emph{down}''); or keeping both
(``\emph{updown}'').

The thresholded vector $c_0$ is the (sample)
\emph{signature} of $r_0$. Then the (gene) signature of
$c_0$ is calculated, $E$ is multiplied by $c_0$ and then thresholded
to get $r_1$.

This iteration is performed until it converges, i.e. $r_{i-1}$
and $r_i$ are \emph{close}, and $c_{i-1}$ and
$c_i$ are also close. The convergence criteria,
i.e. what \emph{close} means, is by default defined by high Pearson
correlation.

It is very possible that the ISA finds the same module more than once;
two or more seeds might converge to the same module. The function
\Rfunction{isa.unique} eliminates every module from the result of 
\Rfunction{isa.iterate} that is very similar (in terms of
Pearson correlation) to the one that was already found before.

It might be also apparent from the description of ISA, that the
biclusters are soft, i.e. they might have an overlap in their genes,
samples, or both. It is also possible that some genes and/or
samples of the input matrix are not found to be part of any ISA
biclusters. Depending on the stringency parameters in the
thresholding (i.e. how far the values should be from the mean), it
might even happen that ISA does not find any biclusters.

\subsection{Parameters}

The two main parameters of ISA are the two thresholds (one for the
genes and one for the samples). They basically define the stringency of
the modules. If the gene threshold is high, then the modules will have
very similar genes. If it is mild, then modules will be bigger, with
less similar genes than in the first case. The same applies to the
sample threshold and the samples of the modules.

\subsection{Random seeding and smart seeding}

By default (i.e. if the \Rfunction{isa} function is used) the ISA is
performed from random sparse starting seeds, generated by the
\Rfunction{generate.seeds} function. This way the algorithm is 
completely unsupervised, but also stochastic: it might give different
results for different runs.

It is possible to use non-random seeds as well. If you have some
knowledge about the data or are interested in a particular subset of
genes/samples, then you can feed in your seeds into the
\Rfunction{isa.iterate} function directly. In this case the
algorithm is deterministic, for the same seed you will always get the
same results. Using smart (i.e. non-random) seeds can be considered as
a semi-supervised approach.

\subsection{Normalization}

Using in silico data we observed that ISA has the best performance if the
input matrix is normalized (see \Rfunction{isa.normalize}). The
normalization produces two matrices: $E_r$ and
$E_c$. $E_r$ is calculated by transposing $E$ and
centering and scaling its genes (see the \Rfunction{scale} R function). $E_c$ is
calculated by centering and scaling the genes of $E$. $E_r$ is
used to calculate the sample signature of genes and $E_c$ is used
to calculate the signature of the samples.

It is possible to use another normalization. In this case the user is
requested to supply the normalized input data in a named list,
including the two matrices of appropriate
dimensions to the \Rfunction{isa.iterate} function.

The \texttt{Er} entry of the list will be used for calculating the
signature of the genes, \texttt{Ec} will be used for the signature of the
samples. If you want to use the same matrix in both steps, then supply
it twice, the first one transposed.

\subsection{Gene and sample scores}

In addition to finding biclusters in the input matrix, the ISA also
assigns scores to the genes and samples, separately for each
module. The scores are between minus one and one and they are by
definition zero for the genes/samples that are not included in the
module. For the non-zero entries, the further the score of a
gene/samples is from zero, the stronger the association between the
gene/sample and the module. If the sign of two genes/samples are the
same, then they are correlated, if they have opposite signs, then they
are anti-correlated.

\section{Bicluster coherence and robustness measures}

\subsection{Coherence}

Madeira and Oliviera\cite{madeira04} define various coherence scores
for biclusters, these measure how well the rows and or columns are
correlated. It is possible to use these measures for ISA as well,
after converting the output of ISA to a \Rclass{biclust} object. 
Here are the measures for the first bicluster:
<<coherence>>=
constantVariance(exprs(ALL), Bc, number=1)
additiveVariance(exprs(ALL), Bc, number=1)
multiplicativeVariance(exprs(ALL), Bc, number=1)
signVariance(exprs(ALL), Bc, number=1)
@

You can use \Rfunction{sapply} to perform the calculation for many or
all modules, e.g. for this data set `constant variance' and `additive
variance' are not the same:
<<coherence all>>=
cv <- sapply(seq_len(Bc@Number), 
             function(x) constantVariance(exprs(ALL), Bc, number=x))
av <- sapply(seq_len(Bc@Number), 
             function(x) additiveVariance(exprs(ALL), Bc, number=x))
cor(av, cv)
@
Please see the manual pages of these functions and the paper cited
above for more details.

\subsection{Robustness}

The \Rpackage{eisa} package uses a measure that is related to
coherence; it is called robustness. Robustness is a generalization of
the singular value of a matrix. If there were no thresholding during
the ISA iteration, then ISA would be equivalent to a numerical method
for singular value decomposition and robustness would be indeed the
same the principal singular value of the input matrix.

If the \Rfunction{isa} function was used to find the transcription
modules, then the robustness measure is used to filter the results.
This is done by first scrambling the input matrix and then running ISA
on it. As ISA is an unsupervised algorithm it usually finds some
(although less and smaller) modules even in such a scrambled data
set. Then the robustness scores are calculated for the proper and the
scrambled modules and only (proper) modules that have a higher score
than the highest scrambled module are kept. The robustness scores are
stored in the seed data during this process, so you can check them
later:
<<robustness>>=
seedData(modules)$rob
@

\section{The \Rpackage{isa2} and \Rpackage{eisa} packages}%
\label{sec:isapackages}

ISA and its companion functions for visualization, functional
enrichment calculation, etc. are distributed in two separate packages,
\Rpackage{isa2} and \Rpackage{eisa}.

\Rpackage{isa2} contains the implementation of ISA itself, and
\Rpackage{eisa} specifically deals with supplying expression data to
\Rpackage{isa2} and visualizing the results.

The two packages provide some functions that share the same name, and
the right one is selected based on the arguments supplied. E.g. both
\Rpackage{isa2} and \Rpackage{eisa} has a function called
\Rfunction{isa}. It the first argument of \Rfunction{isa} is an
\Rclass{ExpressionSet}, then the \Rpackage{eisa} version is called. If
the first argument is a matrix, then the \Rpackage{isa} version is
called.

This also means that there are two manual pages for the two
\Rfunction{isa} functions. I.e. if one types `\verb+?isa+', (and both
\Rpackage{isa2} and \Rpackage{eisa} are loaded, then one
has to choose between the two versions. To avoid this, the
\Rpackage{eisa} manual pages can also be referenced by supplying and
`\verb+e+' prefix: `\verb+?eisa+' or `\verb+?eisa.unique+'.

\section{Finer control over ISA parameters}%
\label{sec:detailed-isa}

\section{Session information}

The version number of R and packages loaded for generating this
vignette were:

<<sessioninfo,results=tex,echo=FALSE>>=
toLatex(sessionInfo())
@

\bibliographystyle{apalike}
\bibliography{EISA}
\nocite{*}

\end{document}

### 2009-07-01 isanote/modular.tex
Added skeleton of the application note.

\documentclass{bioinfo}
\usepackage{natbib}
\copyrightyear{2009}
\pubyear{2009}

\begin{document}
\firstpage{1}

\title[Modular analysis]{Modular analysis of gene expression data with
  GNU R or Matlab}
\author[G\'abor Cs\'ardi \textit{et~al}]{G\'abor Cs\'ardi\,,$^{1,2}$,
  Zolt\'an Kutalik\,$^{1,2}$ and Sven Bergmann\,$^{1,2}$}
\address{$^{1}$Department of Medical Genetics, and 
  $^{2}$Swiss Institute of Bioinformatics,
  University of Lausanne, Rue de Bugnon 27 - DGM 328, CH-1005 Lausanne,
  Switzerland.}

\history{Received on XXXXX; revised on XXXXX; accepted on XXXXX}

\editor{Associate Editor: XXXXXXX}

\maketitle

\begin{abstract}

\section{Summary:}
\section{Availability:}
\section{Contact:} \href{Sven.Bergmann@unil.ch}{Sven.Bergmann@unil.ch}
\end{abstract}

\section{Introduction}

What is biclustering.
What is the ISA.

\section{Methods}

Normalization.

Random and smart seeding.

The ISA iteration.

Merging the modules.

The robustness measure and filtering the modules.

Module trees.

\section{R Implementation}

Two packages: isa2 does the computation, eisa does everything related
to biology.

\subsection{Workflow}

Two workflows, the simple workflow does everything with default
parameters. In the detailed workflow the user has access to every step
of the analysis.

\subsection{Visualization}

Expression plots, GO tree plots, module tree plots, HTML summary
generation.

\subsection{Connection to other software}

The biclust package.

\section{Matlab implementation}

\section*{Acknowledgement}

\paragraph{Funding\textcolon}

\paragraph{Conflict of interest\textcolon} None declared.

\bibliographystyle{natbib}
\bibliography{isa}
\nocite{*}

\end{document}

### 2009-07-14 eisa/inst/doc/EISA_tutorial.Rnw
Some eisa vignette improvements.

@@
 \RaggedRight
+
+<<echo=FALSE,print=FALSE>>=
+options(width=60)
+@
 
@@
 <<GO summary>>=
-summary(GO[[1]], p=0.001)[[1]]
+summary(GO[[1]], p=0.001)[[2]][,-6]
 @
 
-The \texttt{drive} column includes the Entrez ids of the genes that
-are included in the intersection of the bicluster and the GO
-category. They can be also obtained with the
+We omitted the sixth column of the result, because it is very wide and
+would look bad in this vignette. This column is called
+\variable{drive} and lists the Entrez IDs of the genes that are in the
+intersection of the bicluster and the GO category; or in other
+words, the genes that drive the enrichment.
+These genes can also be obtained with the
 \Rfunction{geneIdsByCategory} function. The following returns the
-genes in the first module and the second GO BP category.
+genes in the second module and the third GO BP category. (The GO
+categories are ordered according to the enrichment $p$-values, just
+like in the output of \Rfunction{summary}.)
 <<GO gene ids by category>>=
-geneIdsByCategory(GO[[1]])[[1]][[2]]
+geneIdsByCategory(GO[[1]])[[2]][[3]]
 @
@@
 <<GO info>>=
-sigCategories(GO[[1]])[[1]]
+sigCategories(GO[[1]])[[2]]
 library(GO.db)
-mget(sigCategories(GO[[1]])[[1]], GOTERM)
+mget(sigCategories(GO[[1]])[[2]], GOTERM)
 @
@@
 KEGG
-summary(KEGG)[[1]]
+summary(KEGG)[[4]]
 library(KEGG.db)
-mget(sigCategories(KEGG)[[1]], KEGGPATHID2NAME)
+mget(sigCategories(KEGG)[[4]], KEGGPATHID2NAME)
 @
@@
 CHR <- isa.CHR(modules)
-summary(CHR,p=0.05)[[11]]
+summary(CHR,p=0.05)[[11]][,-6]
 @
 
-The third bicluster has \Sexpr{summary(CHR,p=0.05)[[2]]$Count[1]} % $
-genes on chromosome \Sexpr{rownames(summary(CHR,p=0.05)[[2]])[1]}.
+The eleventh bicluster has \Sexpr{summary(CHR,p=0.05)[[11]]$Count[1]} % $
+genes on chromosome
+\Sexpr{rownames(summary(CHR,p=0.05)[[11]])[1]}. Here is a list of the
+genes:
+<<CHR enrichment list>>=
+unlist(mget(geneIdsByCategory(CHR)[[11]][[1]], org.Hs.egSYMBOL))
+@
 
@@
   miRNA <- isa.miRNA(modules)
-  summary(miRNA, p=1)[[7]]
+  summary(miRNA, p=0.05)[[11]]
 }
@@
 algorithms in a unified framework. It uses the class \Rclass{Biclust}
-to store a set of biclusters. The \Rfunction{eisa.biclust} converts ISA
+to store a set of biclusters. The \Rfunction{eisa.biclust} function converts ISA
 modules to a \Rclass{Biclust} object. This requires the binarization
@@
 <<heatmap>>=
-drawHeatmap(exprs(ALL), Bc, number=1) 
+drawHeatmap(exprs(ALL[featureNames(modules),]), Bc, number=1) 
 @
@@
 <<parallel>>=
-parallelCoordinates(exprs(ALL), Bc, number=1, plotBoth=TRUE, pch=NA)
+ex <- samp.exprs(isa.normalize(ALL[featureNames(modules),]))
+parallelCoordinates(ex, Bc, number=2, pch=NA)
 @
@@
 summary(goplot.2)
+list.vertex.attributes(goplot.2)
 @ 
@@
 \subsection{Generating a HTML summary for the modules}
+
+The \Rfunction{autogen.table} function creates a summary
 

### 2009-07-15 eisa/inst/doc/EISA_tutorial.Rnw
eisa vignette update, use profile.plot instead of parallelCoordinates.

@@
 <<parallel>>=
-ex <- samp.exprs(isa.normalize(ALL[featureNames(modules),]))
-parallelCoordinates(ex, Bc, number=2, pch=NA)
+profile.plot(modules, 2, ALL, plot="both")
 @
@@
 \includegraphics[width=0.7\textwidth]{parallel}
-\caption{Profile plots for the first bicluster found by ISA.}
+\caption{Profile plots for the second bicluster found by ISA.}
 \label{fig:parallel}
 \end{figure}
+
+The \Rfunction{profile.plot} function has several options to set the
+plot colors and styles, please see the manual for the details. This
+function was inspired by the \Rfunction{parallelCoordinates} function
+in the \Rpackage{biclust} package.
 

### 2009-07-15 eisa/inst/doc/EISA_tutorial.Rnw
eisa: some functions were renamed, and we don't use generics any more.

@@
 set.seed(1) # to get the same results, always
-modules <- isa(ALL, thr.gene=thr.gene, thr.cond=thr.cond)
+modules <- ISA(ALL, thr.gene=thr.gene, thr.cond=thr.cond)
 @
@@
 data set and then runs ISA from
-\Sexpr{formals(eisa:::isa.ExpressionSet)$no.seeds} % $
+\Sexpr{formals(eisa:::ISA)$no.seeds} % $
 random seeds (the default). See Section~\ref{sec:detailed-isa} if the
@@
 you have to do is to supply your \Rclass{ISAModules} object to the
-\Rfunction{isa.GO} function.
+\Rfunction{ISA.GO} function.
 <<GO enrichment>>=
-GO <- isa.GO(modules)
+GO <- ISA.GO(modules)
 @
 
-The \Rfunction{isa.GO} requires an annotation package that maps the
+The \Rfunction{ISA.GO} requires an annotation package that maps the
 Entrez identifiers of the studied organism to GO terms. For human, the
@@
 $p$-value limit can be supplied to \Rfunction{summary}. Note, that
-since \Rfunction{isa.GO} calculates enrichment for many gene sets
+since \Rfunction{ISA.GO} calculates enrichment for many gene sets
 (i.e. for all biclusters), \Rfunction{summary} returns a list of data
@@
 In addition, the following functions are implemented to work on the
-objects returned by \Rfunction{isa.GO}: \Rfunction{htmlReport},
+objects returned by \Rfunction{ISA.GO}: \Rfunction{htmlReport},
 \Rfunction{pvalues}, \Rfunction{geneCounts}, \Rfunction{oddsRatios},
@@
 
-By default, the \Rfunction{isa.GO} function performs multiple testing
+By default, the \Rfunction{ISA.GO} function performs multiple testing
 correction using the Holm method, this can be changed via the
@@
 same way as for the Gene Ontology, this time we use the
-\Rfunction{isa.KEGG} function:
+\Rfunction{ISA.KEGG} function:
 <<KEGG enrichment>>=
-KEGG <- isa.KEGG(modules)
+KEGG <- ISA.KEGG(modules)
 KEGG
@@
 in a bicluster are associated with a chromosome. See the
-\Rfunction{isa.CHR} function:
+\Rfunction{ISA.CHR} function:
 <<CHR enrichment>>=
-CHR <- isa.CHR(modules)
+CHR <- ISA.CHR(modules)
 summary(CHR,p=0.05)[[11]][,-6]
@@
 The enrichment calculation itself is basically the same as for GO and
-KEGG, but the \Rfunction{isa.miRNA} function should be used:
+KEGG, but the \Rfunction{ISA.miRNA} function should be used:
 <<miRNA enrichment>>=
 if (require(targetscan.Hs.eg.db)) {
-  miRNA <- isa.miRNA(modules)
+  miRNA <- ISA.miRNA(modules)
   summary(miRNA, p=0.05)[[11]]
@@
 algorithms in a unified framework. It uses the class \Rclass{Biclust}
-to store a set of biclusters. The \Rfunction{eisa.biclust} function converts ISA
+to store a set of biclusters. The \Rfunction{ISA.biclust} function converts ISA
 modules to a \Rclass{Biclust} object. This requires the binarization
@@
 library(biclust)
-Bc <- eisa.biclust(modules)
+Bc <- ISA.biclust(modules)
 Bc
@@
 two or more seeds might converge to the same module. The function
-\Rfunction{isa.unique} eliminates every module from the result of 
-\Rfunction{isa.iterate} that is very similar (in terms of
+\Rfunction{ISA.unique} eliminates every module from the result of 
+\Rfunction{ISA.iterate} that is very similar (in terms of
 Pearson correlation) to the one that was already found before.
@@
 genes/samples, then you can feed in your seeds into the
-\Rfunction{isa.iterate} function directly. In this case the
+\Rfunction{ISA.iterate} function directly. In this case the
 algorithm is deterministic, for the same seed you will always get the
@@
 Using in silico data we observed that ISA has the best performance if the
-input matrix is normalized (see \Rfunction{isa.normalize}). The
+input matrix is normalized (see \Rfunction{ISA.normalize}). The
 normalization produces two matrices: $E_r$ and
@@
 including the two matrices of appropriate
-dimensions to the \Rfunction{isa.iterate} function.
+dimensions to the \Rfunction{ISA.iterate} function.
 
@@
 
-The two packages provide some functions that share the same name, and
-the right one is selected based on the arguments supplied. E.g. both
-\Rpackage{isa2} and \Rpackage{eisa} has a function called
-\Rfunction{isa}. It the first argument of \Rfunction{isa} is an
-\Rclass{ExpressionSet}, then the \Rpackage{eisa} version is called. If
-the first argument is a matrix, then the \Rpackage{isa} version is
-called.
-
-This also means that there are two manual pages for the two
-\Rfunction{isa} functions. I.e. if one types `\verb+?isa+', (and both
-\Rpackage{isa2} and \Rpackage{eisa} are loaded, then one
-has to choose between the two versions. To avoid this, the
-\Rpackage{eisa} manual pages can also be referenced by supplying and
-`\verb+e+' prefix: `\verb+?eisa+' or `\verb+?eisa.unique+'.
-
 \section{Finer control over ISA parameters}%

### 2009-07-15 eisa/R/CHR.R

@@ -267,7 +267,7 @@ setMethod("sigCategories", signature(r="CHRListHyperGResult"),
             })
           })
 
-isa.CHR <- function(modules,
+ISA.CHR <- function(modules,
                     org=organism(modules),
                     shortorg=abbreviate(org,2),
                     ann=annotation(modules),

### 2009-07-15 eisa/R/GO.R

@@ -359,7 +359,7 @@ setMethod("sigCategories", signature(r="GOListHyperGResult"),
             })
           })
 
-isa.GO <- function(modules,
+ISA.GO <- function(modules,
                    org=organism(modules),
                    shortorg=abbreviate(org, 2),
                    ann=annotation(modules),

### 2009-07-15 eisa/R/ISAExpressionSet-methods.R

@@ -1,6 +1,6 @@
 
 setMethod("feat.exprs", signature(object="ISAExpressionSet"),
-          function(object) assayDataElement(object, "exprs"))
+          function(object) assayDataElement(object, "er.exprs"))
 
 setMethod("samp.exprs", signature(object="ISAExpressionSet"),
           function(object) assayDataElement(object, "ec.exprs"))

### 2009-07-15 eisa/R/KEGG.R

@@ -263,7 +263,7 @@ setMethod("sigCategories", signature(r="KEGGListHyperGResult"),
             })
           })
 
-isa.KEGG <- function(modules,
+ISA.KEGG <- function(modules,
                      org=organism(modules),
                      shortorg=abbreviate(org, 2),
                      ann=annotation(modules),

### 2009-07-15 eisa/R/autogen.R

@@ -1,9 +1,13 @@
 
-autogen.table <- function(nm, isares, target.dir,
-                          modules=seq_len(length(isares)),
-                          template=system.file("autogen", package="eisa"),
-                          GO=NULL, KEGG=NULL, miRNA=NULL, CHR=NULL, DBD=NULL,
-                          htmltitle=NULL, notes=NULL, seed=NULL) {
+ISA.html <- function(...) {
+  ## TODO: call ISA.html.table and ISA.html.modules
+}
+
+ISA.html.table <- function(nm, isares, target.dir,
+                           modules=seq_len(length(isares)),
+                           template=system.file("autogen", package="eisa"),
+                           GO=NULL, KEGG=NULL, miRNA=NULL, CHR=NULL, DBD=NULL,
+                           htmltitle=NULL, notes=NULL, seed=NULL) {
 
   isa2:::isa.status("Creating HTML module table", "in")
   
@@ -225,15 +229,15 @@ autogen.table <- function(nm, isares, target.dir,
 }  
                           
 
-autogen.modules <- function(nm, isares, modules=seq_len(length(isares)),
-                            target.dir,
-                            template=system.file("autogen", package="eisa"),
-                            GO=NULL, KEGG=NULL, miRNA=NULL, CHR=NULL, DBD=NULL,
-                            cond.to.include=NULL,
-                            cond.col="white",
-                            sep=NULL, seed=NULL, drive.BP=NULL,
-                            drive.CC=NULL, drive.MF=NULL, drive.KEGG=NULL,
-                            drive.miRNA=NULL, drive.DBD=NULL, drive.CHR=NULL) {
+ISA.html.modules <- function(nm, isares, modules=seq_len(length(isares)),
+                             target.dir,
+                             template=system.file("autogen", package="eisa"),
+                             GO=NULL, KEGG=NULL, miRNA=NULL, CHR=NULL, DBD=NULL,
+                             cond.to.include=NULL,
+                             cond.col="white",
+                             sep=NULL, seed=NULL, drive.BP=NULL,
+                             drive.CC=NULL, drive.MF=NULL, drive.KEGG=NULL,
+                             drive.miRNA=NULL, drive.DBD=NULL, drive.CHR=NULL) {
 
   isa2:::isa.status("Generating module pages", "in")
   

### 2009-07-15 eisa/R/eisa.R

@@ -30,14 +30,19 @@ ISAModules.to.isa.result <- function(isaresult) {
 
 } 
 
-setMethod("isa", signature(data="ExpressionSet"),
-          function(data, ...) isa.ExpressionSet(data, ...))
+eisa.get.nm <- function(data) {
+  if (is(data, "ISAExpressionSet")) {
+    data
+  } else {
+    ISA.normalize(data)
+  }
+}
 
-isa.ExpressionSet <- function(data,
-                              flist=filterfun(function(x) IQR(x) > 0.5),
-                              uniqueEntrez=TRUE,
-                              thr.gene=seq(2,4,by=0.5), thr.cond=seq(1,3,by=0.5),
-                              no.seeds=100) {
+ISA <- function(data,
+                 flist=filterfun(function(x) IQR(x) > 0.5),
+                 uniqueEntrez=TRUE,
+                 thr.gene=seq(2,4,by=0.5), thr.cond=seq(1,3,by=0.5),
+                 no.seeds=100) {
 
   isa2:::isa.status("ISA on an ExpressionSet", "in")
   
@@ -75,115 +80,103 @@ isa.ExpressionSet <- function(data,
   modules
 }
 
-setMethod("isa.normalize", signature(data="ExpressionSet"),
-          function(data, ...) isa.normalize.ExpressionSet(data, ...))
-
-isa.normalize.ExpressionSet <- function(data, prenormalize=FALSE) {
+ISA.normalize <- function(data, prenormalize=FALSE) {
 
   isa2:::isa.status("Normalizing ExpressionSet", "in")
-
+  
   data <- as(data, "ISAExpressionSet")
   normed.data.exprs <- isa2::isa.normalize(exprs(data),
                                           prenormalize=prenormalize)
 
-  data@assayData <- assayDataNew(exprs=t(normed.data.exprs$Er),
+  data@assayData <- assayDataNew(exprs=exprs(data),
+                                 er.exprs=t(normed.data.exprs$Er),
                                  ec.exprs=normed.data.exprs$Ec)
 
-  attr(data, "prenormalize") <- attr(normed.data.exprs, "prenormalize")
-  attr(data, "hasNA") <- attr(normed.data.exprs, "hasNA")
+  data@prenormalized <- attr(normed.data.exprs, "prenormalize")
+  data@hasNA <- attr(normed.data.exprs, "hasNA")
 
[…]

### 2009-07-15 eisa/R/miRNA.R

@@ -281,7 +281,7 @@ setMethod("sigCategories", signature(r="miRNAListHyperGResult"),
             })
           })
 
-isa.miRNA <- function(modules,
+ISA.miRNA <- function(modules,
                       org=organism(modules),
                       shortorg=abbreviate(org, 2),
                       ann=annotation(modules),

### 2009-07-15 eisa/R/viz.R

@@ -355,35 +355,25 @@ exp.plot <- function(epo, scores=TRUE) {
 # coords <- exp.plot(ep)
 # dev.off()
 
-cond.plot <- function(modules, number, exprs, nm,
+cond.plot <- function(modules, number, eset,
                       col="white", all=TRUE,
                       sep=NULL, sepcol=NULL, val=TRUE, srt=90,
                       adj.above=c(0,0.5), adj.below=c(1,0.5),
-                      plot.only=seq_len(dim(nm)[2]), ...) {
+                      plot.only=seq_len(ncol(eset)), ...) {
 
   isa2:::isa.status("Creating a condition plot", "in")
 
-  if (missing(exprs) && missing(nm)) {
-    stop("Either `exprs' or `nm' is required")
-  }
+  eset <- eisa.get.nm(eset)
+  eset <- eset[featureNames(modules),]
+  nm <- list(t(feat.exprs(eset)), samp.exprs(eset))
 
-  if (!missing(exprs) && !missing(nm)) {
-    warning("`exprs' ignored, because `nm' was also given")
-  }
-  if (missing(nm)) {
-    exprs <- exprs[ featureNames(modules), ]
-    nm <- isa.normalize(exprs)
-  } else {
-    nm <- nm[ featureNames(modules), ]
-  }
-  
   genes <- getFeatureMatrix(modules, mods=number)
   samp <- getSampleMatrix(modules, mods=number)
   thr <- seedData(modules)$thr.col[number]
   
   ## Calculate all condition scores, might not be correct for
   ## oscillating modules
-  scores <- as.vector(t(exprs(nm)) %*% genes)
+  scores <- as.vector(nm[[1]] %*% genes)
   msc <- mean(scores)
   ssc <- sd(scores)
   thr1 <- msc + thr * ssc
@@ -555,12 +545,34 @@ mnplot <- function(x, eset, data, group, ...) {
   invisible(tts)
 }
 
-ISAmnplot <- function(modules, module, eset, data=annotation(x), group, ...) {
-  x <- modules$rundata$features[ modules$genes[,module] != 0 ]
+select.eset <- function(eset, norm=c("raw", "feature", "sample")) {
+  norm <- match.arg(norm)
+  if (norm=="raw") {
+    eset <- exprs(eset)
+  } else if (norm=="feature") {
+    eset <- eisa.get.nm(eset)
+    eset <- feat.exprs(eset)
[…]

### 2009-08-05 eisa/R/AllGenerics.R
eisa: renamed 'organism' function to 'getOrganism', package 'annotate' defines a generic called 'organism', too.

@@ -13,7 +13,7 @@ setGeneric("drive<-", function(r, value) standardGeneric("drive<-"))
 
 setGeneric("seedData", function(object) standardGeneric("seedData"))
 setGeneric("runData", function(object) standardGeneric("runData"))
-setGeneric("organism", function(object) standardGeneric("organism"))
+setGeneric("getOrganism", function(object) standardGeneric("getOrganism"))
 
 setGeneric("getFeatures", function(object, ...) standardGeneric("getFeatures"))
 setGeneric("getSamples", function(object, ...) standardGeneric("getSamples"))

### 2009-08-05 eisa/R/CHR.R

@@ -268,7 +268,7 @@ setMethod("sigCategories", signature(r="CHRListHyperGResult"),
           })
 
 ISA.CHR <- function(modules,
-                    org=organism(modules),
+                    org=getOrganism(modules),
                     shortorg=abbreviate(org,2),
                     ann=annotation(modules),
                     features=featureNames(modules),

### 2009-08-05 eisa/R/GO.R

@@ -360,7 +360,7 @@ setMethod("sigCategories", signature(r="GOListHyperGResult"),
           })
 
 ISA.GO <- function(modules,
-                   org=organism(modules),
+                   org=getOrganism(modules),
                    shortorg=abbreviate(org, 2),
                    ann=annotation(modules),
                    features=featureNames(modules),

### 2009-08-05 eisa/R/ISAModules-methods.R

@@ -54,7 +54,7 @@ setMethod("annotation", signature(object="ISAModules"),
             object@rundata$annotation
           })
 
-setMethod("organism", signature(object="ISAModules"),
+setMethod("getOrganism", signature(object="ISAModules"),
           function(object) {
             object@rundata$organism
           })

### 2009-08-05 eisa/R/KEGG.R

@@ -264,7 +264,7 @@ setMethod("sigCategories", signature(r="KEGGListHyperGResult"),
           })
 
 ISA.KEGG <- function(modules,
-                     org=organism(modules),
+                     org=getOrganism(modules),
                      shortorg=abbreviate(org, 2),
                      ann=annotation(modules),
                      features=featureNames(modules),

### 2009-08-05 eisa/R/autogen.R

@@ -13,7 +13,7 @@ ISA.html.table <- function(nm, isares, target.dir,
   
   chip <- annotation(isares)
   library(paste(sep="", chip, ".db"), character.only=TRUE)
-  organism <- organism(isares)
+  organism <- getOrganism(isares)
 
   ################################################
   ## GO
@@ -312,7 +312,7 @@ isa.autogen.module <- function(nm, isares, module, target.dir, template,
 
   chip <- annotation(isares)
   library(paste(sep="", chip, ".db"),  character.only=TRUE)
-  organism <- organism(isares)
+  organism <- getOrganism(isares)
   short.organism <- organism
   require(paste(sep="", "org.", abbreviate(organism, 2), ".eg.db"),
           character.only=TRUE)
@@ -939,7 +939,7 @@ isa.autogen.module <- function(nm, isares, module, target.dir, template,
   
   valid <- !is.na(entrezIds)
 
-  if (organism(isares) == "Homo sapiens") {
+  if (getOrganism(isares) == "Homo sapiens") {
     html <- paste(sep="", "<a href=\"http://www.genecards.org/cgi-bin/carddisp.pl?gene=", entrezIds[valid],
                   "\" class=\"tag", val[valid],
                   "\">", entrezIds[valid], "<span>", longname[valid], ", score: ", orig.val[valid], "</span></a>")

### 2009-08-05 eisa/R/dbd.R

@@ -282,7 +282,7 @@ setMethod("sigCategories", signature(r="DBDListHyperGResult"),
           })
 
 ISA.DBD <- function(modules,
-                    org=organism(modules),
+                    org=getOrganism(modules),
                     shortorg=abbreviate(org, 2),
                     ann=annotation(modules),
                     features=featureNames(modules),

### 2009-08-05 eisa/R/miRNA.R

@@ -282,7 +282,7 @@ setMethod("sigCategories", signature(r="miRNAListHyperGResult"),
           })
 
 ISA.miRNA <- function(modules,
-                      org=organism(modules),
+                      org=getOrganism(modules),
                       shortorg=abbreviate(org, 2),
                       ann=annotation(modules),
                       features=featureNames(modules),

### 2009-08-05 eisa/inst/doc/EISA_tutorial.Rnw
EISA vignette improvements.

@@
 
-% TODO: length, dim, featureNames, sampleNames, pData, 
-% annotation, organism, indexing
+Calling \Rfunction{length} on \variable{modules} returns the number of
+ISA modules in the set, \Rfunction{dim} gives the dimension of the
+input expression matrix: the number of features and the number of
+samples.
+<<accessors 1>>=
+length(modules)
+dim(modules)
+@
+
+Functions \Rfunction{featureNames} and \Rfunction{sampleNames} return
+the names of the features and samples, just like the functions with
+the same name for an \Rclass{ExpressionSet}
+<<accessors 2>>=
+featureNames(modules)[1:5]
+sampleNames(modules)[1:5]
+@
 
@@
 \Rfunction{getNoSamples} returns a numeric vector, the number of
-samples in each module. 
+samples in each module. Similarly, \Rfunction{pData} returns the
+phenotype data of the expression set as a data frame. The
+\Rfunction{getOrganism} function returns the scientific name of the
+organism, \Rfunction{annotation} the name of the chip. For the former
+the appropriate annotation must be installed.
 <<number of features and samples>>=
@@
 getNoSamples(modules)
+colnames(pData(modules))
+getOrganism(modules)
+annotation(modules)
+@
+
+The double bracket indexing operator (`\verb+[[+') can be used to
+select some modules from the complete set, the result is another,
+smaller \Rclass{ISAModules} object. The following selects the first
+five modules.
+<<indexing>>=
+modules[[1:5]]
+@
+
+The single bracket indexing operator can be used to restrict an
+\Rclass{ISAModules} object to a subset of features and/or
+samples. E.g. selecting all features that map to a gene on
+chromosome~1 can be done with 
+<<indexing 2>>=
+chr <- get(paste(annotation(modules), sep="", "CHR"))
+entrez <- sapply(mget(featureNames(modules), chr), 
+                 function(x) "1" %in% x)
+modules[entrez,]
+@
+
+Similarly, selecting all B-cell samples can be performed with
+<<indexing 3>>=
+modules[ ,grep("^B", pData(modules)$BT)]
 @
@@
 TargetScan is a data based of predicted target genes of $\mu$RNAs, for
-severral organisms. There are two R packages that incorporate this
+several organisms. There are two R packages that incorporate this
 database, one for human and another one for mouse, right now they can
@@
   \includegraphics[width=0.7\textwidth]{heatmap}
-  \caption{TODO}
+  \caption{Heatmap of the first transcription module.}
   \label{fig:drawHeatmap}
@@
 
+The \Rfunction{ISA} function takes care of all steps performed during
+a modular study, and for each step it uses parameters work reasonably
+well. In some cases, however, one wants to access these steps
+individually, to use custom parameters instead of the defaults.
+
+In this section, we will still use the acute lymphoblastic leukemia
+gene expression data from the \Rpackage{ALL} package.
+
+\subsection{Non-specific filtering}
+
+The first step of the analysis typically involves non-specific
+filtering of the probesets. The aim is to eliminate the probesets that
+do not show variation across the samples, as they only contribute
+noise to the data.
+
+By default (i.e. if the \Rfunction{ISA} function is called) this is
+performed using the \Rpackage{genefilter} package, and the default
+filter is based on the inter-quantile ratio of the probesets'
+expression values, a robust measure of variance. Other possible
+filtering techniques include using the AffyMetrix present/absent calls
+produced by the \Rfunction{mas5calls} function of the \Rpackage{affy}
+package, but this requires the raw data, so in this vignette we use a
+simple method based on variance and minimum expression value: only
+probesets that have a variance of at least \variable{varLimit} and
+that have at least \variable{kLimit} samples with expression values
+over \variable{ALimit} are kept.
+<<filtering>>=
+varLimit <- 0.5
+kLimit <- 4
+ALimit <- 5
+flist <- filterfun(function(x) var(x)>varLimit, kOverA(kLimit,ALimit))
+ALL.filt <- ALL[genefilter(ALL, flist), ]
+@
+
+The original expression set had \Sexpr{nrow(ALL)} features, the
+filtered one has only \Sexpr{nrow(ALL.filt)}.
+
+\subsection{Entrez Id matching}
+
+In this step we match the probesets to Entrez Id and remove the ones
+that don't map to any Entrez genes.
+<<Entrez>>=
+ann <- annotation(ALL.filt)
+library(paste(ann, sep=".", "db"), character.only=TRUE)
+ENTREZ <- get( paste(ann, sep="", "ENTREZID") )
+EntrezIds <- mget(featureNames(ALL.filt), ENTREZ)
+keep <- sapply(EntrezIds, function(x) length(x) >= 1 && !is.na(x))
+ALL.filt.2 <- ALL.filt[keep,]
+@
+
+To reduce ambiguity in the interpretation of the results, we might
+also want to keep only single probeset for each Entrez genes.
+<<Entrez unique>>=
+vari <- apply(exprs(ALL.filt.2), 1, var)
+larg <- findLargest(featureNames(ALL.filt.2), vari, data=annotation(ALL.filt.2))
+ALL.filt.3 <- ALL.filt.2[larg,]
+@
+
+\subsection{Normalizing the data}
+
+The ISA works best, if the expression matrix is scaled and
+centered. In fact, the two sub-steps of an ISA step require expression
+matrices that are normalized differently. The
+\Rfunction{ISA.normalize} function can be used to calculate the
+normalized expression matrices; it returns an \Rclass{ExpressionSet}
+object including three expression matrices: the original raw
+expression, the row-wise normalized and the column-wise normalized
+expression matrix.
+<<normed>>=
+ALL.normed <- ISA.normalize(ALL.filt.3)
+ls(assayData(ALL.normed))
+@
+
+\subsection{Generating starting seeds for the ISA}
+
+The ISA is an iterative algorithm that starts with a set of input
+seeds. An input seed is basically a set of probesets and the ISA stepwise
+refines this set by 1) including other probesets in the set that are
+coexpressed with the input probesets and 2) removing probesets from it
+that are not coexpressed with the rest of the input set.
+
+The \Rfunction{generate.seeds} generates a set of random seeds (i.e. a
+set of random gene sets). See its documentation if you need to change
+the sparsity of the seeds.
+
+<<seeds>>=
+random.seeds <- generate.seeds(length=nrow(ALL.normed), count=100)
+@
+
+In addition to random seeds, it is possible to start the ISA iteration
+from ``educated'' seeds, i.e. gene sets the user is interested in, or
+a set of samples that are supposed to have coexpressed genes. We
+create another set of starting seeds here, based on the type of acute
+lymphoblastic leukemia: ``B'', ``B1'', ``B2'', ``B3'', ``B4'' or
+``T'', ``T1'', ``T2'', ``T3'' and ``T4''.
+<<smart seeds>>=
+type <- as.character(pData(ALL.normed)$BT)
+ss1 <- ifelse(grepl("^B", type), -1, 1)
+ss2 <- ifelse(grepl("^B1", type), 1, 0)
+ss3 <- ifelse(grepl("^B2", type), 1, 0)
+ss4 <- ifelse(grepl("^B3", type), 1, 0)
+ss5 <- ifelse(grepl("^B4", type), 1, 0)
+ss6 <- ifelse(grepl("^T1", type), 1, 0)
+ss7 <- ifelse(grepl("^T2", type), 1, 0)
+ss8 <- ifelse(grepl("^T3", type), 1, 0)
+ss9 <- ifelse(grepl("^T4", type), 1, 0)
+smart.seeds <- cbind(ss1, ss2, ss3, ss4, ss5, ss6, ss7, ss8, ss9)
+@
+
+The \variable{ss1} seed includes all samples, but their sign is
+opposite for B-cell leukemia samples and T-cell samples. This way ISA
+is looking for sets of genes that are differently regulated in these
+two groups of samples. \variable{ss2} contains only B1 type samples,
+so here we look for genes that are specific to this variant of the
+disease. The other seeds are similar, for the other subtypes.
+
+\subsection{Performing the ISA iteration}
+
+We perform the ISA iterations for our two sets of seeds separately.
+<<iteration>>=
+modules1 <- ISA.iterate(ALL.normed, row.seeds=random.seeds, 
+                        thr.row=2, thr.col=2)
+modules2 <- ISA.iterate(ALL.normed, col.seeds=smart.seeds,
+                        thr.row=2, thr.col=2)
+@
+
+\subsection{Dropping non-unique modules}
+
+\Rfunction{ISA.iterate} returns the same number of ``modules'' as the
+number of input seeds; these, however, do not always correspond to
+meaningful modules, the input seeds can converge to an all-zero
+vector, or occasionally they may not converge at all. It is also
+possible that two or more input seeds converge to the same module.
+
+The \Rfunction{ISA.unique} function eliminates the all-zero or
+non-convergent input seeds and keeps only one instance of the
+duplicated ones.
+<<unique>>=
+modules1.unique <- ISA.unique(ALL.normed, modules1)
+modules2.unique <- ISA.unique(ALL.normed, modules2)
+@
+
+\Sexpr{length(modules1.unique)} modules were kept for the first set of
+seeds and \Sexpr{length(modules2.unique)} for the second set.
+
+\subsection{Dropping non-robust modules}
+
+The \Rfunction{ISA.filter.robust} function filters a set of modules by
+running ISA with the same parameters on a scrambled data set and then
+calculating a robustness score, both for the real modules and the ones
+from the scrambled data. The highest robustness score obtained from
+the scrambled data is used as a threshold to filter the real modules.
+<<robust>>=
+modules1.robust <- ISA.filter.robust(ALL.normed, modules1.unique)
+modules2.robust <- ISA.filter.robust(ALL.normed, modules2.unique)
+@
+
+We still have \Sexpr{length(modules1.robust)} modules for
+the first set of seeds and \Sexpr{length(modules2.unique)} for the
+second set.
+
+\subsection{Differentially regulated modules}
+
+Now we check whether the ISA modules that we found can be used as
+classifiers for the different types of ALL. For this we use the sample
+scores of the modules, the score of a sample in a given module is the
+(weighted) average of the expression of the genes of the module in
+the sample in question.
+
+Let's first check, whether any of the modules can distinguish between
+T-cell and B-cell ALL samples.
+<<B vs T>>=
+scores1 <- getSampleMatrix(modules1.robust)
+tt1 <- colttests(scores1, as.factor(substr(type, 1, 1)))
+scores2 <- getSampleMatrix(modules2.robust)
+tt2 <- colttests(scores2, as.factor(substr(type, 1, 1)))
+sign1 <- which(p.adjust(tt1$p.value, "holm") < 0.05)
+sign2 <- which(p.adjust(tt2$p.value, "holm") < 0.05)
+@
+
+For the first set of samples \Sexpr{length(sign1)} modules show
+significant difference for T-cell and B-cell samples, for the second
+(smart) set \Sexpr{length(sign2)} of them.
+
+Let's make some condition plots for the best separating modules from
+each set.
+<<conditions-plots-dysreg,fig=TRUE,width=10,height=8>>=
+color <- ifelse(grepl("T", type), "orange", "darkolivegreen")
+layout(cbind(1:2))
+cond.plot(modules1.robust, which.min(tt1$p.value), ALL.normed, col=color, 
+          main="Best separator, random seeds")
+cond.plot(modules2.robust, which.min(tt2$p.value), ALL.normed, col=color,
+          main="Best separator, smart seeds")
+@
+
+Let's extract the modules that are good separators.
+<<extract separators>>=
+modules1.TB <- modules1.robust[[sign1]]
+modules2.TB <- modules2.robust[[sign2]]
+@
+
+\subsection{Enrichment calculations}
+
+We can check our separator modules against the Gene Ontology
+categories and the pathways in the KEGG database to find dysregulated
+GO categories and/or KEGG pathways.
+<<dysreg enrichment>>=
+GO.dysreg1 <- ISA.GO(modules1.TB)
+GO.dysreg2 <- ISA.GO(modules2.TB)
+KEGG.dysreg1 <- ISA.KEGG(modules1.TB)
+KEGG.dysreg2 <- ISA.KEGG(modules2.TB)
+@
+
+Let's collect the significantly enriched GO categories and KEGG
+pathways.
+<<dysreg enriched categories>>=
+gocats <- unique(unlist(lapply(c(GO.dysreg1, GO.dysreg2), function(x) 
+                               unique(unlist(sigCategories(x))))))
+keggp  <- unique(unlist(lapply(list(KEGG.dysreg1, KEGG.dysreg2), function(x)
+                        unique(unlist(sigCategories(x))))))
+library(GO.db)
+library(KEGG.db)
+sapply(mget(gocats, GOTERM), Term)
+mget(keggp, KEGGPATHID2NAME)
+@
+
+\subsection{More separator modules}
+
+Of course we can search for modules that can separate the samples of
+the ALL subtypes as well, e.g. let's try to find some that
+differentiate between type B1 and other B types. 
+<<B1 vs others>>=
+keep <- grepl("^B[1234]", type)
+type.B <- ifelse(type[keep]=="B1", "B1", "Bx")
+scores.B1.1 <- scores1[keep,]
+scores.B1.2 <- scores2[keep,]
+tt.B1.1 <- colttests(scores.B1.1, as.factor(type.B))
+tt.B1.2 <- colttests(scores.B1.2, as.factor(type.B))
+min(p.adjust(na.omit(tt.B1.1$p.value)))
+min(p.adjust(na.omit(tt.B1.2$p.value)))
+@
+
+Both module sets seem to have some good separators, let us make some
+condition plots.
+<<B1-cond-plots,fig=TRUE,width=10,height=8>>=
+color <- ifelse(type == "B1", "green", 
+                ifelse(grepl("^T", type), "orange", "darkolivegreen"))
+layout(cbind(1:2))
+cond.plot(modules1.robust, which.min(tt.B1.1$p.value), ALL.normed, col=color, 
+          main="Best B1 separator, random seeds")
+cond.plot(modules2.robust, which.min(tt.B1.2$p.value), ALL.normed, col=color,
+          main="Best B1 separator, smart seeds")
+@
+
+From the plot it seems that these two modules are essentially the
+same, one is the opposite of the others.
+<<check B1 same module>>=
+B1.cor <- c(cor(scores1[,which.min(tt.B1.1$p.value)],
+                scores2[,which.min(tt.B1.2$p.value)]),
+            cor(getFeatureMatrix(modules1.robust, 
+                                 mods=which.min(tt.B1.1$p.value)),
+                getFeatureMatrix(modules2.robust, 
+                                 mods=which.min(tt.B1.2$p.value))))
+B1.cor
+@
+
+Indeed, they are almost the same, their Pearson correlation is around
+\Sexpr{round(B1.cor[1],3)} for the sample scores and
+\Sexpr{round(B1.cor[2],3)} for the feature (=gene) scores.
+
 \section{Session information}

### 2009-08-06 eisa/inst/doc/EISA_tutorial.Rnw
eisa vignette: do not use Cairo package and add a VignetteEntry.

@@
+% \VignetteIndexEntry{The Iterative Signature Algorithm for Gene Expression Data}
 \documentclass{article}
@@
 <<heatmap biclust pre,echo=FALSE,results=hide>>=
-CairoPNG("heatmap.png", 400, 400 )
+png("heatmap.png", 400, 400 )
 @ 
@@
 <<parallel.pre,echo=FALSE,results=hide>>=
-CairoPNG("parallel.png", 400, 400 )
+png("parallel.png", 400, 400 )
 @ 

### 2009-08-06 eisa/inst/doc/EISA_tutorial.Rnw
eisa vignette: put graphics into subdir, use Cairo if available.

@@
 \SweaveOpts{cache=TRUE}
+\SweaveOpts{prefix.string=graphics/plot}
 
@@
 
-<<echo=FALSE,print=FALSE>>=
+<<set width,echo=FALSE,print=FALSE>>=
 options(width=60)
@@
 <<heatmap biclust pre,echo=FALSE,results=hide>>=
-png("heatmap.png", 400, 400 )
+if (require(Cairo)) {
+  CairoPNG("graphics/heatmap.png", 400, 400 )  
+} else {
+  png("graphics/heatmap.png", 400, 400 )
+}
 @ 
@@
   \centering
-  \includegraphics[width=0.7\textwidth]{heatmap}
+  \includegraphics[width=0.7\textwidth]{graphics/heatmap}
   \caption{Heatmap of the first transcription module.}
@@
 <<parallel.pre,echo=FALSE,results=hide>>=
-png("parallel.png", 400, 400 )
+if (require(Cairo)) {
+  CairoPNG("graphics/parallel.png", 400, 400 )
+} else {
+  png("graphics/parallel.png", 400, 400 )
+}
 @ 
@@
 \centering
-\includegraphics[width=0.7\textwidth]{parallel}
+\includegraphics[width=0.7\textwidth]{graphics/parallel}
 \caption{Profile plots for the second bicluster found by ISA.}

### 2009-08-06 isa2/inst/doc/ISA_tutorial.Rnw
Update vignettes, do not mess with sessionInfo() because the vignette fails if sessionInfo() to toLatex() output is changed.
isa2: put vignette images into subdir.

@@
 \newcommand{\listentry}[1]{\texttt{#1}}
+
+\SweaveOpts{prefix.string=graphics/plot}
 
@@
 <<blockplot.pre,echo=FALSE,results=hide>>=
-png("blockplot.png", 800, 600)
+png("graphics/blockplot.png", 800, 600)
 @ 
@@
 <<tenmods.pre,echo=FALSE,results=hide>>=
-png("tenmods.png", 600, 400)
+png("graphics/tenmods.png", 600, 400)
 @ 
@@
 <<othermods.pre,echo=FALSE,results=hide>>=
-png("othermods.png", 400, 400)
+png("graphics/othermods.png", 400, 400)
 @ 
@@
 <<drawHeatmap.pre,echo=FALSE,results=hide>>=
-png("drawheatmap.png", 400,400)
+png("graphics/drawheatmap.png", 400,400)
 @ 
@@
 <<noiseplot.pre,echo=FALSE,results=hide>>=
-png("noise.png", width=800, height=600)
+png("graphics/noise.png", width=800, height=600)
 @ 
@@
 <<overlapplot.pre,echo=FALSE,results=hide>>=
-png("overlapplot.png", 800, 600)
+png("graphics/overlapplot.png", 800, 600)
 @ 
@@
 <<sessioninfo,results=tex,echo=FALSE>>=
-gsub(";", "; ", gsub("_", "\\\\_", sub("Locale:[ ]*\\\\verb\\|(.*)\\|", 
-                                       "\\\\texttt\\{\\1\\}", 
-                                       toLatex(sessionInfo()))))
+toLatex(sessionInfo())
 @

### 2009-08-06 isa2/inst/doc/ISA_tutorial.Rnw
isa2 vignette: use Cairo for graphics, if available.
Corrected images in isa2 vignette.

@@
 options(width=60)
+if (require(Cairo)) { png <- CairoPNG }
 @
@@
 silico data with the \Rfunction{isa.in.silico} function first:
+<<seed,echo=FALSE>>=
+## Just to get the same results
+set.seed(10)
+@ 
 <<insilico,cache=TRUE>>=
@@
 \centering
-\includegraphics[width=0.6\textwidth]{blockplot}
+\includegraphics[width=0.6\textwidth]{graphics/blockplot}
 \caption{Another artificial data matrix with two modules and a lot of
@@
 data matrix. 
-<<seed,echo=FALSE>>=
-## Just to get the same results
-set.seed(10)
-@ 
 <<isa1,cache=TRUE>>=
@@
 
-<<simpleplot,results=hide,include=FALSE>>=
+<<simpleplot,results=hide,include=FALSE,eval=FALSE>>=
 layout(cbind(seq(ncol(modules$rows))))
@@
 \centering
-\includegraphics{tenmods}
+\includegraphics{graphics/tenmods}
 \caption{In-silico data with ten non-overlapping modules.}
@@
 \centering
-\includegraphics[width=0.6\textwidth]{othermods}
+\includegraphics[width=0.6\textwidth]{graphics/othermods}
 \caption{The input matrix and five modules with relatively low
@@
 \centering
-\includegraphics[width=0.5\textwidth]{drawheatmap}
+\includegraphics[width=0.5\textwidth]{graphics/drawheatmap}
 \caption{Heatmap plot of the reordered input matrix. The rows and
@@
   \centering
-  \includegraphics[width=\textwidth]{noise}
+  \includegraphics[width=\textwidth]{graphics/noise}
   \caption{Modular data sets with different levels of noise. The data
@@
 \centering
-\includegraphics{overlapplot}
+\includegraphics{graphics/overlapplot}
 \caption{The input matrix with two overlapping blocks and the ISA

### 2009-08-06 RExpressionView/R/AllGenerics.R
Make ExpressionView a separate R package.

@@ -1,0 +1,2 @@
+setGeneric("orderModules",
+           function(modules, ...) standardGeneric("orderModules"))

### 2009-08-06 RExpressionView/R/isaorder.R

@@ -1,5 +1,7 @@
-setMethod("orderModules", signature(modules="list"),
-	function(modules, ...) orderModules.default(modules, ...))
+if (require(isa2)) {
+  setMethod("orderModules", signature(modules="list"),
+            function(modules, ...) orderModules.default(modules, ...))
+}
 
 orderModules.default <- function(modules, debuglevel=0, timelimit=60) {
 	#isa.status("ISA ordering", "in")

### 2009-08-06 RExpressionView/R/order.R

@@ -1,5 +1,7 @@
-setMethod("orderModules", signature(modules="ISAModules"),
-          function(modules, ...) orderModules.ISAModules(modules, ...))
+if (require(eisa)) {
+  setMethod("orderModules", signature(modules="ISAModules"),
+            function(modules, ...) orderModules.ISAModules(modules, ...))
+}
 
 orderModules.ISAModules <- function(modules, debuglevel=0, timelimit=60) {
 	isamodules <- ISAModules.to.isa.result(modules)

### 2009-08-06 RExpressionView/R/zzz.R

@@ -1,0 +1,11 @@
+.onLoad <- function(dir, package) {
+  library.dynam("ExpressionView", package, dir, local=FALSE);
+}
+
+.onUnload <- function(libpath) {
+  library.dynam.unload("ExpressionView", libpath)
+}
+
+.Last.lib <- function(libpath) {
+  isa2::.onUnload(libpath)
+}

### 2009-08-06 isa2/R/AllGenerics.R

@@ -19,6 +19,4 @@ setGeneric("sweep.graph",
 
 setGeneric("plotModules",
            function(modules, ...) standardGeneric("plotModules"))
-setGeneric("orderModules",
-           function(modules, ...) standardGeneric("orderModules"))
 

### 2009-08-10 eisa/inst/doc/EISA_tutorial.Rnw
eisa vignette updates.

@@
 CHR <- ISA.CHR(modules)
-summary(CHR,p=0.05)[[11]][,-6]
+summary(CHR,p=0.05)[[2]][,-6]
 @
 
-The eleventh bicluster has \Sexpr{summary(CHR,p=0.05)[[11]]$Count[1]} % $
+The second bicluster has \Sexpr{summary(CHR,p=0.05)[[2]]$Count[1]} % $
 genes on chromosome
-\Sexpr{rownames(summary(CHR,p=0.05)[[11]])[1]}. Here is a list of the
+\Sexpr{rownames(summary(CHR,p=0.05)[[2]])[1]}. Here is a list of the
 genes:
 <<CHR enrichment list>>=
-unlist(mget(geneIdsByCategory(CHR)[[11]][[1]], org.Hs.egSYMBOL))
+unlist(mget(geneIdsByCategory(CHR)[[2]][[1]], org.Hs.egSYMBOL))
 @
@@
   miRNA <- ISA.miRNA(modules)
-  summary(miRNA, p=0.05)[[11]]
+  summary(miRNA, p=0.1)[[7]]
 }
@@
 <<goplot1>>=
-goplot.2 <- gograph(summary(GO[[1]], p=0.05)[[2]])
+goplot.2 <- gograph(summary(GO[[1]], p=0.05)[[1]])
 @

### 2009-09-09 eisa/R/viz.R
eisa: corrected a bug in ISA2heatmap.

@@ -699,6 +699,8 @@ ISA2heatmap <- function(modules, module, eset,
                         norm=c("raw", "feature", "sample"),
                         scale=c("none", "row", "column"), ...) {
 
+  norm <- match.arg(norm)
+  scale <- match.arg(scale)
   eset <- select.eset(eset, modules, norm)
   x <- getFeatureNames(modules, module)[[1]]
   y <- getSampleNames (modules, module)[[1]]

### 2009-09-16 RExpressionView/R/open.R
RExpressionView: Added open.R file.

@@ -1,0 +1,8 @@
+open.ev <- function(modules, eset, order) {
+  filename <- paste(tempfile(), sep="", ".ged")
+  export.ev(modules, eset, order, filename=filename)
+  swf <- system.file("ExpressionView.html", package="ExpressionView")
+  url <- URLencode(paste("file://", swf, sep="", "?filename=", filename))
+  browseURL(url)
+}
+

### 2009-09-22 expressionviewnote/expressionview.tex
ExpressionView: Added Application Note to repository.

\documentclass[round]{bioinfo}

\usepackage{color}
\newcommand{\cred}{\color{red}}
%\usepackage[pdftex,colorlinks,bookmarks=false,citecolor=blue,linkcolor=red,urlcolor=blue]{hyperref}

\copyrightyear{2009}
\pubyear{2009}

\begin{document}

\firstpage{1}
\application{}
\title[ExpressionView]{ExpressionView - an interactive viewer for biclusters identified in gene expression data}
\author[Andreas L\"uscher, G\'abor Cs\'ardi, Aitana Morton de Lachapelle, Zolt\'an Kutalik, and Sven Bergmann]{Andreas L\"uscher$^1$, G\'abor Cs\'ardi$^{1,2}$, Aitana Morton de Lachapelle$^{1,2}$, Zolt\'an Kutalik$^{1,2}$, and Sven Bergmann$^{1,2}$\footnote{to whom correspondence should be addressed}}
\address{
	$^{1}$Swiss Institute of Bioinformatics, Lausanne, Switzerland\\
	$^{2}$Department of Medical Genetics, University of Lausanne, Lausanne, Switzerland
}

\history{Received on XXXXX; revised on XXXXX; accepted on XXXXX}

\editor{Associate Editor: XXXXXXX}

\maketitle

\begin{abstract}

\section{Summary:}
ExpressionView is a Bioconductor package that provides an interactive environment to explore biclusters identified in gene expression data. A sophisticated ordering algorithm is used to present the biclusters in a visually appealing layout that provides an intuitive summary of the results. From this overview, the user can select individual biclusters and access all the biologically relevant data associated with it. The package is aimed to facilitate the exchange between bioinformaticians and life scientists who are not familiar with Bioconductor and the R language.

\section{Availability:}
\href{http://www.bioconductor.org/packages/2.0/bioc/html/ExpressionView.htmI}{Bioconductor package}

\section{Contact:} 
\href{andreas.luescher@a3.epfl.ch}{Andreas L\"uscher}

\section{Supplementary information:}
Screenshots, tutorials and sample data sets can be found on the ExpressionView \href{http://maya:7575/ExpressionView}{website}.

\end{abstract}

\section{Introduction}
Clustering genes according to their expression profiles is an important task in analysing microarray data. While there is a multitude of biclustering software available~\citep{?}, none of them provides an intuitive interface that would allow the user to interactively explore the obtained results. ExpressionView is designed to close this gap, hereby facilitating the exchange between bioinformaticians and life scientists without consolidated programming experience. Implemented as an R package, the user can apply all the powerful microarray analysis tools provided by the Bioconductor~\citep{gentleman04} project before exporting the data to a platform independent Flash applet that allows him and his co-workers to  visualise the biclusters together with the underlying gene expression data.

\section{Package design and workflow}
The ExpressionView package contains two independent parts - an ordering algorithm implemented in C++ and a Flash applet written in ActionScript and Adobe Flex. Fig.~\ref{fig:workflow} schematically summarises the typical workflow that uses the genomic data available as a Bioconductor ExpressionSet and combines it with biclustering results to produce an ExpressionView data file that can be explored with the Flash applet. In the following, we describe the different steps in more detail.
\begin{figure*}[!tpb]
\centerline{\includegraphics[width=\linewidth]{fig1-crop}}
\caption{How to use ExpressionView: Starting from gene expression data in the form of a Bioconductor ExpressionSet, the user first runs a biclustering algorithm to determine co-regulated groups of genes and samples. In a second step, the rows and columns of the gene expression matrix are rearranged to produce an easily readable overview of the results. The last step consist in combining the gene expression data and its associated metadata with the results from the biclustering and produce an ExpressionView data file that can be explored with the interactive Flash applet.}\label{fig:workflow}
\end{figure*}
For users who prefer to install the viewer as a standalone program, we also provide a platform independent Adobe AIR version that can be downloaded from the \href{http://maya:7575/ExpressionView}{website}.

\subsection{Gene expression data and biclusters}
ExpressionView is designed to work with gene expression data in the form of a Bioconductor ExpressionSet. This class provides a user-friendly way to access the actual gene expression matrix and its associated metadata. There is a variety of biclustering algorithms described in the literature~\citep{madeira04}, several of which have been implemented as R packages. ExpressionView can treat results obtained by the iterative signature algorithm (ISA)~\citep{bergmann03,csardi09} and the methods available in the Biclust package~\citep{kaiser08}. Since the structure of biclustering results is independent of the algorithm, an extension to other methods is straightforward.

\subsection{Order}
To present the tens of possibly overlapping biclusters in a visually appealing form, it is necessary to regroup the rows and columns of the gene expression matrix in such a way that biclusters form contiguous rectangles. Since for more than two mutually overlapping biclusters, it is in general impossible to find such an arrangement, one has to make concessions. In contrast to \cite{grothaus06}, who proposed to repeat rows and columns as necessary to achieve this goal, we prefer to optimise the arrangement within the original data by maximising the area of the largest contiguous biclusters.

Since this reordering is an interesting problem on its own, which to the best of our knowledge has not been studied in the literature, we briefly outline our strategy here. Noting that the rows and the columns can be ordered separately, the problem reduces to finding the optimal arrangement of a set of $n$ elements that are part of at least one of $m$ clusters. The quality of the order, which we call fragmentation, is defined as the sum over the maximal number of neighbouring elements in every cluster and the tasks thus consists in maximizing this quantity by changing the order of the elements. Starting from an initial configuration determined by similarity, this is achieved by either shifting well aligned subsequences of a given cluster, herby enlarging the longest contiguous part, or permuting the individual elements of a given subcluster.

To determine the efficiency of this method, we have studied a large number of perfectly orderable, but initially scrambled, test cases. We find that in  almost every situation, the proposed algorithm finds a solution that recovers more than 99\% of the optimal fragmentation and in most cases, it actually recovers the correct alignment. In these test samples, execution time increases as ${\mathcal O}(m^\alpha)$, where $\alpha \in [4, 4.5]$, depending on the number of elements $n$.

\subsection{Export}
Once the optimal order is determined, the program rearranges the gene expression matrix accordingly and exports all the relevant information into a single uncompressed XML file. This format has the advantage of being self-explanatory and extensible. In addition, the files can be read and edited with any text editor. The actual gene expression data is rounded to two digit precision and encoded in Base64. The structure of the XML file is described in more detail on the ExpressionView \href{http://maya:7575/ExpressionView}{website}. In most cases, these files are of the order of a few Megabytes and can thus easily be sent to co-workers.

\subsection{Visualisation}
The core application of this package is the Flash applet that allows the user to interactively explore the gene expression data, the biclusters, and the results of the enrichment calculations. Relying on the flash architecture has the advantage of producing a platform and browser independent applet that can be launched from any computer with even the most restrictive user rights. A screenshot is shown in Fig.~\ref{fig:workflow}. The interface is divided in two parts: On the left-hand side, the user finds the gene expression data in the common heat map form, on top of which the biclusters are overlaid. On the right-hand side, the metadata associated with the gene expression data and the results of the enrichment calculations are shown. Wherever possible, these elements are linked to the corresponding databases. The interface essentially behaves as an image viewer, allowing the user to zoom and pan around the expression data, getting instant feedback on the selected item. The currently selected region can be exported to a PDF file, helpful to prepare figures for presentations or articles, or a simple text file.\\

%\section*{Acknowledgement}

\paragraph{Funding\textcolon} This work was supported by ...

\paragraph{Conflict of interest\textcolon} none declared.

%\bibliography{../../expressionview}
%\bibliographystyle{plainnat}

\begin{thebibliography}{6}

\bibitem[Bergmann et~al.(2003)]{bergmann03}
Bergmann,S., Ihmels,J. and Barkai,N. (2003)
Iterative signature algorithm for the analysis of large-scale gene expression data.
\href{http://dx.doi.org/10.1103/PhysRevE.67.031902 }{\emph{Physical Review E}, {\bf 67}, 031902.}

\bibitem[Cs{\'a}rdi et~al.(2009)]{csardi09}
Cs{\'a}rdi,G., Kutalik,Z. and Bergmann,S. (2009)
Eisa.
\href{}{\emph{Bioinformatics}, ...}

\bibitem[Gentleman et~al.(2004)]{gentleman04}
Gentleman,R.C., \emph{et al.} (2004)
Bioconductor: open software development for computational biology and bioinformatics.
\href{http://dx.doi.org/10.1186/gb-2004-5-10-r80}{\emph{Genome Biology}, {\bf 5}, R80.}

\bibitem[Grothaus et~al.(2006)]{grothaus06}
Grothaus,G., Mufti,A., and Murali,TM. (2006)
Automatic layout and visualization of biclusters.
\href{http://dx.doi.org/10.1186/1748-7188-1-15}{\emph{Algorithms for Molecular Biology}, {\bf 1}, 15.}

\bibitem[Kaiser and Leisch(2008)]{kaiser08}
Kaiser,S. and Leisch,F. (2008)
A toolbox for bicluster analysis in r.
\href{http://epub.ub.uni-muenchen.de/3293/}{Technical Report 028, Department of Statistics, University of Munich}.

\bibitem[Madeira and Oliveira(2004)]{madeira04}
Madeira,S.C. and Oliveira,A.L. (2004)
Biclustering algorithms for biological data analysis: a survey.
\emph{IEEE/ACM Transactions on Computational Biology and Bioinformatics}, {\bf 1}, 24 -- 45.

\end{thebibliography}

\end{document}

### 2009-09-22 expressionviewnote/expressionview.tex

@@
 \section{Summary:}
-ExpressionView is a Bioconductor package that provides an interactive environment to explore biclusters identified in gene expression data.
+ExpressionView is an R package that provides an interactive environment to explore biclusters identified in gene expression data.
 A sophisticated ordering algorithm is used to present the biclusters in a visually appealing layout that provides an intuitive summary of the results.
 From this overview, the user can select individual biclusters and access all the biologically relevant data associated with it.
-The package is aimed to facilitate the exchange between bioinformaticians and life scientists who are not familiar with Bioconductor and the R language.
+The package is aimed to facilitate the collaboration between bioinformaticians and life scientists who are not familiar with the R language.
 
 \section{Availability:} 
-\href{http://www.bioconductor.org/packages/2.0/bioc/html/ExpressionView.htmI}{Bioconductor package}
+\href{http://www.google.com}{R package with open source license}
 
@@
 \section{Introduction}
-Clustering genes according to their expression profiles is an important task in analysing microarray data.
-While there is a multitude of biclustering software available~\citep{?}, none of them provides an intuitive interface that would allow the user to interactively explore the obtained results.
-ExpressionView is designed to close this gap, hereby facilitating the exchange between bioinformaticians and life scientists without consolidated programming experience.
-Implemented as an R package, the user can apply all the powerful microarray analysis tools provided by the Bioconductor~\citep{gentleman04} project before exporting the data to a platform independent Flash applet that allows him and his co-workers to  visualise the biclusters together with the underlying gene expression data.
+Biclustering is an unsupervised learning technique used to analyse microarray data.
+It consists in identifying co-regulated groups of genes and conditions according to their expression profile.
+While there is a multitude of biclustering software available~\citep{madeira04}, packages with intuitive interfaces that allow the user to interactively explore the results are sparse~\citep{santamaria08}.
+ExpressionView is designed to close this gap, hereby facilitating the collaboration between bioinformaticians and life scientists without consolidated programming experience.
+Implemented as an R package, the user can apply all the powerful microarray analysis tools provided by the Bioconductor~\citep{gentleman04} project before exporting the data to a platform independent Adobe Flash applet that allows him and his co-workers to visualise the biclusters together with the underlying gene expression data.
 
 \section{Package design and workflow}
-The ExpressionView package contains two independent parts - an ordering algorithm implemented in C++ and a Flash applet written in ActionScript and Adobe Flex.
-Fig.~\ref{fig:workflow} schematically summarises the typical workflow that uses the genomic data available as a Bioconductor ExpressionSet and combines it with biclustering results to produce an ExpressionView data file that can be explored with the Flash applet.
+The ExpressionView package contains two independent parts: An ordering algorithm implemented in C++ and a Adobe Flash applet written in ActionScript and Adobe Flex.
+Fig.~\ref{fig:workflow} schematically summarises the typical workflow that uses the genomic data available as a Bioconductor ExpressionSet and combines it with the biclustering results to produce an ExpressionView data file that can be explored with the Flash applet.
 In the following, we describe the different steps in more detail.
@@
 This class provides a user-friendly way to access the actual gene expression matrix and its associated metadata.
-There is a variety of biclustering algorithms described in the literature~\citep{madeira04}, several of which have been implemented as R packages.
+There is a variety of biclustering algorithms described in the literature~\citep{madeira04,prelic06}, several of which have been implemented as R packages.
 ExpressionView can treat results obtained by the iterative signature algorithm (ISA)~\citep{bergmann03,csardi09} and the methods available in the Biclust package~\citep{kaiser08}.
@@
 \subsection{Order}
-To present the tens of possibly overlapping biclusters in a visually appealing form, it is necessary to regroup the rows and columns of the gene expression matrix in such a way that biclusters form contiguous rectangles.
+To present the tens of possibly overlapping biclusters in a visually appealing form, it is necessary to reorder the rows and columns of the gene expression matrix in such a way that biclusters form contiguous rectangles.
 Since for more than two mutually overlapping biclusters, it is in general impossible to find such an arrangement, one has to make concessions.
@@
 Noting that the rows and the columns can be ordered separately, the problem reduces to finding the optimal arrangement of a set of $n$ elements that are part of at least one of $m$ clusters.
-The quality of the order, which we call fragmentation, is defined as the sum over the maximal number of neighbouring elements in every cluster and the tasks thus consists in maximizing this quantity by changing the order of the elements.
+The quality of the order is defined as the sum over the maximal number of neighbouring elements in every cluster and the tasks thus consists in maximizing this quantity by changing the order of the elements.
 Starting from an initial configuration determined by similarity, this is achieved by either shifting well aligned subsequences of a given cluster, herby enlarging the longest contiguous part, or permuting the individual elements of a given subcluster.
@@
 To determine the efficiency of this method, we have studied a large number of perfectly orderable, but initially scrambled, test cases.
-We find that in  almost every situation, the proposed algorithm finds a solution that recovers more than 99\% of the optimal fragmentation and in most cases, it actually recovers the correct alignment.
-In these test samples, execution time increases as ${\mathcal O}(m^\alpha)$, where $\alpha \in [4, 4.5]$, depending on the number of elements $n$.
+We find that in almost every situation, the proposed algorithm finds an order that recovers more than 99\% of the score of the optimal solution and in most cases, it actually recovers the correct alignment.
+In these test samples, execution time increases polynomially with the number of clusters $m$ as ${\mathcal O}(m^\alpha)$, where $\alpha \in [4, 4.5]$, almost independent of the number of elements $n$.
 
 \subsection{Export}
-Once the optimal order is determined, the program rearranges the gene expression matrix accordingly and exports all the relevant information into a single uncompressed XML file.
+Once the optimal order is determined, the program rearranges the gene expression matrix accordingly and exports all the relevant information into an XML file.
 This format has the advantage of being self-explanatory and extensible.
@@
 \subsection{Visualisation}
-The core application of this package is the Flash applet that allows the user to interactively explore the gene expression data, the biclusters, and the results of the enrichment calculations.
-Relying on the flash architecture has the advantage of producing a platform and browser independent applet that can be launched from any computer with even the most restrictive user rights.
+The core application of this package is the Flash applet that allows the user to interactively explore the gene expression data and the biclusters.
+Relying on the Flash architecture has the advantage of producing a platform and browser independent applet that can be launched from any computer with even the most restrictive user rights.
 A screenshot is shown in Fig.~\ref{fig:workflow}.
 The interface is divided in two parts: On the left-hand side, the user finds the gene expression data in the common heat map form, on top of which the biclusters are overlaid.
-On the right-hand side, the metadata associated with the gene expression data and the results of the enrichment calculations are shown.
+On the right-hand side, the metadata associated with the gene expression data and the results of the enrichment calculations (GO and KEGG) are shown.
 Wherever possible, these elements are linked to the corresponding databases.
@@
 Madeira,S.C. and Oliveira,A.L. (2004)
-Biclustering algorithms for biological data analysis: a survey.
-\emph{IEEE/ACM Transactions on Computational Biology and Bioinformatics}, {\bf 1}, 24 -- 45.
+Biclustering algorithms for biological data analysis: A survey.
+\href{http://dx.doi.org/10.1109/TCBB.2004.2}{\emph{IEEE/ACM Transactions on Computational Biology and Bioinformatics}, {\bf 1}, 24 -- 45.}
+
+\bibitem[Prelic et~al.(2006)]{prelic06}
+Prelic,A., \emph{et~al.} 
+A systematic comparison and evaluation of biclustering methods for gene expression data 
+\href{http://dx.doi.org/10.1093/bioinformatics/bt1060}{\emph{Bioinformatics}, {\bf 22}(9), 1122 -- 1129.}
+
+\bibitem[Santamar\'ia et~al.(2008)]{santamaria08}
+Santamar\'ia,R., Ther\'on,R. and Quintales,L. (2008)
+BicOverlapper: A tool for bicluster visualization. 
+\href{http://dx.doi.org/10.1093/bioinformatics/btn076}{\emph{Bioinformatics}, {\bf 24}, 1212 -- 1213.}
 

### 2009-09-22 eisa/R/AllGenerics.R
eisa: eliminate dot from function names

@@ -1,6 +1,6 @@
 
-setGeneric("feat.exprs", function(object) standardGeneric("feat.exprs"))
-setGeneric("samp.exprs", function(object) standardGeneric("samp.exprs"))
+setGeneric("featExprs", function(object) standardGeneric("featExprs"))
+setGeneric("sampExprs", function(object) standardGeneric("sampExprs"))
 setGeneric("prenormalized", function(object) standardGeneric("prenormalized"))
 setGeneric("prenormalized<-", function(object, value) standardGeneric("prenormalized<-"))
 setGeneric("hasNA", function(object) standardGeneric("hasNA"))

### 2009-09-22 eisa/R/CHR.R

@@ -54,7 +54,7 @@ setMethod("hyperGTest",
             do.call(new, c("CHRListHyperGResult", res))
           })
 
-ISA.CHR <- function(modules,
+ISACHR <- function(modules,
                     ann=annotation(modules),
                     features=featureNames(modules),
                     hgCutoff=0.05,

### 2009-09-22 eisa/R/GO.R

@@ -155,7 +155,7 @@ setMethod("htmlReport", signature(r="GOListHyperGResult"),
             }
           })  
 
-ISA.GO <- function(modules,
+ISAGO <- function(modules,
                    ann=annotation(modules),
                    features=featureNames(modules),
                    hgCutoff=0.05,

### 2009-09-22 eisa/R/ISAExpressionSet-methods.R

@@ -1,8 +1,8 @@
 
-setMethod("feat.exprs", signature(object="ISAExpressionSet"),
+setMethod("featExprs", signature(object="ISAExpressionSet"),
           function(object) assayDataElement(object, "er.exprs"))
 
-setMethod("samp.exprs", signature(object="ISAExpressionSet"),
+setMethod("sampExprs", signature(object="ISAExpressionSet"),
           function(object) assayDataElement(object, "ec.exprs"))
 
 setMethod("prenormalized", signature(object="ISAExpressionSet"),

### 2009-09-22 eisa/R/ISAModules-methods.R

@@ -128,7 +128,7 @@ setMethod("getSampleMatrix", signature(object="ISAModules"),
 setMethod("getFullFeatureMatrix", signature(object="ISAModules"),
           function(object, eset, mods) {
             eset <- eisa.get.nm(eset, object)
-            nm2 <- samp.exprs(eset)
+            nm2 <- sampExprs(eset)
             if (missing(mods)) {
               genes <- getFeatureMatrix(object)
               samp  <- getSampleMatrix(object)
@@ -152,7 +152,7 @@ setMethod("getFullFeatureMatrix", signature(object="ISAModules"),
 setMethod("getFullSampleMatrix", signature(object="ISAModules"),
           function(object, eset, mods) {
             eset <- eisa.get.nm(eset, object)
-            nm1 <- t(feat.exprs(eset))
+            nm1 <- t(featExprs(eset))
             if (missing(mods)) {
               genes <- getFeatureMatrix(object)
               samp  <- getSampleMatrix(object)

### 2009-09-22 eisa/R/KEGG.R

@@ -68,7 +68,7 @@ setMethod("htmlReport", signature(r="KEGGListHyperGResult"),
             }
           })  
 
-ISA.KEGG <- function(modules,
+ISAKEGG <- function(modules,
                      ann=annotation(modules),
                      features=featureNames(modules),
                      hgCutoff=0.05,

### 2009-09-22 eisa/R/autogen.R

@@ -1,15 +1,15 @@
 
-ISA.html <- function(eset, modules, target.dir,
+ISAHTML <- function(eset, modules, target.dir,
                      template=system.file("autogen", package="eisa"),
                      GO, KEGG, miRNA=NULL, CHR=NULL, htmltitle=NULL,
                      notes=NULL, seed=NULL,
                      cond.to.include=NULL, cond.col="white", sep=NULL) {
 
-  ISA.html.table(modules=modules, target.dir=target.dir, template=template,
+  ISAHTMLTable(modules=modules, target.dir=target.dir, template=template,
                  GO=GO, KEGG=KEGG, miRNA=miRNA, CHR=CHR, htmltitle=htmltitle,
                  notes=notes, seed=seed)
 
-  ISA.html.modules(eset=eset, modules=modules, target.dir=target.dir,
+  ISAHTMLModules(eset=eset, modules=modules, target.dir=target.dir,
                    template=template, GO=GO, KEGG=KEGG, miRNA=miRNA, CHR=CHR,
                    cond.to.include=cond.to.include,
                    cond.col=cond.col, sep=sep, seed=seed)
@@ -50,7 +50,7 @@ isa.autogen.create.dirs <- function(template, target.dir) {
   }
 }
 
-ISA.html.table <- function(modules, target.dir,
+ISAHTMLTable <- function(modules, target.dir,
                            which=seq_len(length(modules)),
                            template=system.file("autogen", package="eisa"),
                            GO=NULL, KEGG=NULL, miRNA=NULL, CHR=NULL,
@@ -187,7 +187,7 @@ ISA.html.table <- function(modules, target.dir,
 }  
                           
 
-ISA.html.modules <- function(eset, modules, which=seq_len(length(modules)),
+ISAHTMLModules <- function(eset, modules, which=seq_len(length(modules)),
                              target.dir,
                              template=system.file("autogen", package="eisa"),
                              GO=NULL, KEGG=NULL, miRNA=NULL, CHR=NULL,
@@ -524,7 +524,7 @@ isa.autogen.module <- function(eset, modules, which, target.dir, template,
                    colbar.length=20, label.cex=1,
                    go.terms=go.terms, GOGRAPHS=GOGRAPHS)
     png(file=filename, width=gop$width*4, height=gop$height*4)
-    co <- gograph.plot(gop, coords=TRUE)
+    co <- gographPlot(gop, coords=TRUE)
[…]

### 2009-09-22 eisa/R/eisa.R

@@ -35,9 +35,9 @@ eisa.get.nm <- function(data, modules) {
     data
   } else {
     if (!is.null(modules)) {
-      ISA.normalize(data, prenormalize=runData(modules)$prenormalize)
+      ISANormalize(data, prenormalize=runData(modules)$prenormalize)
     } else {
-      ISA.normalize(data)
+      ISANormalize(data)
     }
   }
 }
@@ -84,7 +84,7 @@ ISA <- function(data,
   modules
 }
 
-ISA.normalize <- function(data, prenormalize=FALSE) {
+ISANormalize <- function(data, prenormalize=FALSE) {
 
   isa2:::isa.status("Normalizing ExpressionSet", "in")
   
@@ -104,15 +104,15 @@ ISA.normalize <- function(data, prenormalize=FALSE) {
   data
 }
 
-ISA.iterate <- function(data, feature.seeds, sample.seeds,
-                        thr.feat, thr.samp=thr.feat, ...) {
+ISAIterate <- function(data, feature.seeds, sample.seeds,
+                       thr.feat, thr.samp=thr.feat, ...) {
   
   isa2:::isa.status("Starting ISA iteration on an ExpressionSet", "in")
 
   data <- eisa.get.nm(data, modules=NULL)
   
-  nm <- list(Er=t(feat.exprs(data)),
-             Ec=samp.exprs(data))
+  nm <- list(Er=t(featExprs(data)),
+             Ec=sampExprs(data))
 
   attr(nm, "prenormalize") <- prenormalized(data)
   attr(nm, "hasNA") <- hasNA(data)
@@ -138,14 +138,14 @@ ISA.iterate <- function(data, feature.seeds, sample.seeds,
   modules
 }
 
-ISA.unique <- function(data, isaresult, ...) {
+ISAUnique <- function(data, isaresult, ...) {
 
   isa2:::isa.status("Creating unique ISA module set for expression data", "in")
   
   data <- eisa.get.nm(data, isaresult)
   
-  nm <- list(Er=t(feat.exprs(data)),
-             Ec=samp.exprs(data))
+  nm <- list(Er=t(featExprs(data)),
+             Ec=sampExprs(data))
 
   res <- ISAModules.to.isa.result(isaresult)
 
@@ -158,15 +158,15 @@ ISA.unique <- function(data, isaresult, ...) {
   modules
 }
 
-ISA.sweep <- function(data, isaresult, ...) {
[…]

### 2009-09-22 eisa/R/miRNA.R

@@ -63,7 +63,7 @@ setMethod("hyperGTest",
             do.call(new, c("miRNAListHyperGResult", res))
           })
 
-ISA.miRNA <- function(modules,
+ISAmiRNA <- function(modules,
                       ann=annotation(modules),
                       features=featureNames(modules),
                       hgCutoff=0.05,

### 2009-09-22 eisa/R/viz.R

@@ -308,7 +308,7 @@ function(x,y=NULL,input=c('usr','plt','fig','dev','tdev')) {
 
 }
 
-gograph.plot <- function(graph, coords=FALSE, ...) {
+gographPlot <- function(graph, coords=FALSE, ...) {
 
   if (dev.cur() == 1) {
     device <- options("device")[[1]]
@@ -484,7 +484,7 @@ expPlot <- function(epo, scores=TRUE) {
        cond.height=epo$cond.height.px)
 }  
 
-cond.plot <- function(modules, number, eset,
+condPlot <- function(modules, number, eset,
                       col="white", all=TRUE,
                       sep=NULL, sepcol="grey", val=TRUE, srt=90,
                       adj.above=c(0,0.5), adj.below=c(1,0.5),
@@ -493,7 +493,7 @@ cond.plot <- function(modules, number, eset,
   isa2:::isa.status("Creating a condition plot", "in")
 
   eset <- eisa.get.nm(eset, modules)
-  nm1 <- t(feat.exprs(eset))
+  nm1 <- t(featExprs(eset))
 
   genes <- getFeatureMatrix(modules, mods=number)
   samp <- getSampleMatrix(modules, mods=number)
@@ -623,7 +623,7 @@ overlap <- function(modules, algorithm=c("mds", "fr", "drl"), edge.limit=0.5) {
   g
 }
 
-overlap.plot <- function(graph, xsize=400, ysize=400,
+overlapPlot <- function(graph, xsize=400, ysize=400,
                          vertex.size=20, vertex.size2=10, ...) {
 
   l <- cbind(V(graph)$x, V(graph)$y)
@@ -672,10 +672,10 @@ select.eset <- function(eset, modules, norm=c("raw", "feature", "sample")) {
     eset <- exprs(eset)
   } else if (norm=="feature") {
     eset <- eisa.get.nm(eset, modules)
-    eset <- feat.exprs(eset)
+    eset <- featExprs(eset)
   } else if (norm=="sample") {
     eset <- eisa.get.nm(eset, modules)
-    eset <- samp.exprs(eset)
+    eset <- sampExprs(eset)
   }
   eset
 }

### 2009-09-29 RExpressionView/R/order.R
(R)ExpressionView: Added vignette for ordering. Added R help files.

@@ -11,7 +11,7 @@ OrderEV.ISAModules <- function(modules, initialorder=NULL, debuglevel=0, maxtime
 	if ( !is.null(initialorder) ) {
 		initialorder <- list(rows=initialorder$genes, cols=initialorder$samples, status=initialorder$status)
 	}
-	resp <- order.clusters(isamodules, initialorder, debuglevel, maxtime)
+	resp <- OrderEV(isamodules, initialorder, debuglevel, maxtime)
 	res <- list(genes=resp$rows, samples=resp$cols, status=list(genes=resp$status[[1]], samples=resp$status[[2]]))
 	res
 }
@@ -56,40 +56,28 @@ OrderEV.list <- function(modules, initialorder=NULL, debuglevel=0, maxtime=60) {
 			row.map[[mod + 1]] <- c(1:sum(modules[[1]][,mod] != 0))
 			col.map[[mod + 1]] <- c(1:sum(modules[[2]][,mod] != 0))
 		}
-		initialorder <- list(rows=row.map, cols=col.map, status=list(vector("numeric",no.mods+1),vector("numeric",no.mods+1)))
+		initialorder <- list(rows=row.map, cols=col.map, status=list(vector("numeric",no.mods+1), vector("numeric",no.mods+1)))
 	}
-	
-	clusters = list(matrix(as.integer(0), no.rows, no.mods), matrix(as.integer(0), no.cols, no.mods))
-	intersections = list()
-	
-	# determine intersecting modules
+
+	clusters <- list(matrix(as.integer(modules[[1]] != 0), nrow=dim(modules[[1]])[1]), 
+					matrix(as.integer(modules[[2]] != 0), nrow=dim(modules[[2]])[1]))
+
+	intersections <- list()
 	for ( mod in 1:no.mods ) {
-	
-		intersection <- list()
+
+		temp <- list()
 		for ( i in 1:2 ) {
 
-			contains <- mat.or.vec(0, 1)
-			for ( slot in 1:no.slots[i] ) {
-				if ( modules[[i]][slot, mod] != 0 ) {
-					clusters[[i]][slot, mod] <- as.integer(1)
-					
-					for ( modp in 1:no.mods ) {
-						if ( modp != mod ) {
-							if ( modules[[i]][slot, modp] != 0 ) {
-								contains <- append(contains, modp)
-							}
-						}
-					}
-					
+			temp[[i]] <- vector("integer", 0)
[…]

### 2009-09-29 RExpressionView/src/orderclusters.cpp

@@ -227,7 +227,7 @@ void Clusters::prearrange() {
 
 		double bestfitness = getfitness();
 
-		for ( int slot2 = 0; slot2 <= slot1; slot2++ ) {		
+		for ( int slot2 = 0; slot2 <= slot1; slot2++ ) {
 			swap(slot1, slot2);
 			double fitness = getfitness();
 			if ( fitness <= bestfitness ) {
@@ -236,9 +236,13 @@ void Clusters::prearrange() {
 				bestfitness = fitness;
 			}
 		}
+		
+		if ( elapsedtime() > maxtime && maxtime != 0 ) {
+			break;
+		}
 
 	}
-	
+		
 	if ( getfitness() < initialfitness ) {
 		order = oldorder;
 		if ( debug > 1 ) {
@@ -259,7 +263,7 @@ double Clusters::getfitness() {
 			if ( data[order[slot]][cluster] ) {
 				length += multiplicity[order[slot]];
 				if ( slot == nSlots - 1 && length > maxlength ) {
-					maxlength = length;					
+					maxlength = length;
 				}
 			} else {
 				if ( length > maxlength ) {
@@ -397,6 +401,11 @@ int Clusters::reposition(int cluster) {
 			
 			start = end + 1;
 			
+			if ( elapsedtime() > maxtime && maxtime != 0 ) {
+				go = false; 
+				break;
+			}
+			
 		}
 	
 	} while ( go );
@@ -470,6 +479,10 @@ int Clusters::exchange(int cluster, int what) {
 		
 		if ( bestpermutation != initialpermutation ) {
 			result = 1;
+		}
+		
+		if ( elapsedtime() > maxtime && maxtime != 0 ) {
+			break;
 		}
 		
 	}

### 2009-09-29 eisa/R/autogen.R
HTML generation: made some logos smaller.

@@ -717,8 +717,6 @@ isa.autogen.module <- function(eset, modules, which, target.dir, template,
   jfname <- paste(sep="", target.dir, "/module-", m, ".js")
   cat(jlines, file=jfname, sep="\n")
 
-  print("Done.")
-  
   isa2:::isa.status("DONE", "out")
 }
 

### 2009-09-29 expressionviewnote/expressionview.tex

@@
 We find that in almost every situation, the proposed algorithm finds an order that recovers more than 99\% of the score of the optimal solution and in most cases, it actually recovers the correct alignment.
-In these test samples, execution time increases polynomially with the number of clusters $m$ as ${\mathcal O}(m^\alpha)$, where $\alpha \in [4, 4.5]$, almost independent of the number of elements $n$.
+For random samples, which are more representative for actual gene expression data than orderable situations, execution time increases polynomially with the number of clusters $m$ as ${\mathcal O}(m^\alpha)$, where $\alpha \in [1.6, 2]$, almost independent of the number of elements $n$.
+For a given number of clusters, we find ${\mathcal O}(n^\alpha)$, with $\alpha \in [2.5, 2.7]$.
 
@@
 The interface is divided in two parts: On the left-hand side, the user finds the gene expression data in the common heat map form, on top of which the biclusters are overlaid.
-On the right-hand side, the metadata associated with the gene expression data and the results of the enrichment calculations (GO and KEGG) are shown.
+On the right-hand side, the metadata associated with the gene expression data and the results of the enrichment calculations [GO (Gene Ontology)~\citep{ashburner00} and KEGG (Kyoto Encyclopedia of Genes and Genomes)~\citep{kanehisa04}] are shown.
 Wherever possible, these elements are linked to the corresponding databases.
@@
 
-%\bibliography{../../expressionview}
-%\bibliographystyle{plainnat}
+\begin{thebibliography}{6}
 
-\begin{thebibliography}{6}
+\bibitem[Ashburner et~al.(2000)]{ashburner00}
+Ashburner,M. \emph{et al.} (2000)
+Gene Ontology: tool for the unification of biology.
+\href{http://dx.doi.org/10.1038/75556}{\emph{Nature Genetics}, {\bf 25}, 25 -- 29.}
 
@@
 Eisa.
-\href{}{\emph{Bioinformatics}, ...}
+\href{http://dx.doi.org/}{\emph{Bioinformatics}, ...}
 
@@
 Kaiser,S. and Leisch,F. (2008)
-A toolbox for bicluster analysis in r.
+A toolbox for bicluster analysis in R.
 \href{http://epub.ub.uni-muenchen.de/3293/}{Technical Report 028, Department of Statistics, University of Munich}.
+
+\bibitem[Kanehisa et~al.(2004)]{kanehisa04}
+Kanehisa,M. \emph{et al.} (2004)
+The KEGG resource for deciphering the genome.
+\href{http://dx.doi.org/10.1093/nar/gkh063}{\emph{Nucleic Acids Res.}, {\bf 32} (Database special issue), D277.}
 

### 2009-09-30 eisa/R/viz.R
eisa profilePlot updates.

@@ -712,17 +712,44 @@ ISA2heatmap <- function(modules, module, eset,
 }
 
 profilePlot <- function(modules, module, eset,
-                        norm=c("raw", "feature", "sample"),
                         plot=c("samples", "features", "both"),
+                        norm="default",
                         background=TRUE, col=gray(0.7), col.mod=1,
-                        type="l", type.mod=type, order=FALSE,
+                        type="l", type.mod=type,
+                        mean=TRUE, meancol="green", meancol.mod="red",
                         xlabs=c("Features","Samples"),
                         ylab="Expression", ...) {
 
-  data <- select.eset(eset, modules, norm)
+  plot <- match.arg(plot)
+
+  # How to normalize the expression matrix for the plots
+  if (any(!norm %in% c("default", "feature", "sample", "raw"))) {
+    stop("`norm' must be one of `default', `feature', `sample' or `raw'")
+  }
+  norm <- rep(norm, length=2)
+  if (norm[1]=="default") {
+    if (plot=="samples" || plot=="both") {
+      norm[1] <- "feature"
+    } else {
+      norm[1] <- "sample"
+    }
+  }
+  if (norm[2]=="default" && plot=="both") {
+    norm[2] <- "sample"
+  }
+   
+  data <- select.eset(eset, modules, norm[1])
+  if (norm[2] != norm[1] && plot=="both") {
+    data2 <- select.eset(eset, modules, norm[2])
+  } else {
+    data2 <- data
+  }
   
   if (is(data, "ExpressionSet") || !is.null(rownames(data))) {
     data <- data[featureNames(modules),]
+  }
+  if (is(data2, "ExpressionSet") || !is.null(rownames(data2))) {
+    data2 <- data2[featureNames(modules),]
   }
 
   if (!all(dim(data) == dim(modules))) {
@@ -744,10 +771,6 @@ profilePlot <- function(modules, module, eset,
     xlim <- c(1, length(xx))
     ylim <- range(data)
     
-    if (order) {
-      data <- data[ order(rowMeans(data[,yy])), ]
-    }
-    
     par(mar=c(2,4,1,1)+0.1)
     plot(NA, type="n", xlim=xlim, ylim=ylim, xlab=NA,
          ylab=ylab, axes=FALSE, ...)
[…]

### 2009-09-30 eisa/R/eisa.R
Sweep trees: added vertical lines and modified default vertex color.

@@ -197,12 +197,27 @@ ISASweepGraphPlot <- function(graph, vertex.label=V(graph)$id,
                               edge.label=NA, asp=FALSE, rescale=FALSE,
                               xlim=range(graph$layout[,1]),
                               ylim=range(graph$layout[,2]),
+                              thresholds=TRUE,
+                              xlab=NA, ylab=NA,
                               ...) {
 
+  plot(NA, xlim=xlim, ylim=ylim, xlab=xlab, ylab=ylab, axes=FALSE)
+  
+  if (thresholds) {
+    la <- sort(unique(V(G)$thr))
+    at <- sort(unique(G$layout[,1]))
+    axis(1, at=at, labels=la)
+    axis(3, at=at, labels=la)
+    par(xpd=FALSE)
+    abline(v=at, col="lightgrey", lty=2)
+    par(xpd=TRUE)
+  }
+
   tmp <- plot(graph, asp=asp, rescale=rescale,
-              xlim=xlim, ylim=ylim,
+              xlim=xlim, ylim=ylim, add=TRUE,
               vertex.label=vertex.label, edge.label=edge.label,
               vertex.label.cex=vertex.label.cex, ...)
+
   if (any(!is.na(vertex.label.topleft))) {
     text(G$layout[,1]-V(G)$size/200, G$layout[,2], pos=3,
          vertex.label.topleft, cex=vertex.label.cex)

### 2009-09-30 isa2/R/isa3.R

@@ -581,6 +581,7 @@ sweep.graph.default <- function(sweep.result) {
   G$width <- length(unique(G$layout[,1])) * 2
   G$height <- (max(G$layout[,2])-min(G$layout[,2])) * 0.4
 
+  V(G)$color <- "lightgray"
   V(G)$shape <- "vrectangle"
   V(G)$size  <- 16
   V(G)$size2 <- offs * 1.5

### 2009-09-30 RExpressionView/R/export.R
REV updates, ISA.GO and ISA.KEGG were renamed.

@@ -52,14 +52,14 @@ ExportEV.ISAModules <- function(modules, eset, order, go=NULL, kegg=NULL, filena
 
 	go.table <- toTable(GOTERM)
 	if ( is.null(go) ) {
-		gos <- ISA.GO(modules)
+		gos <- ISAGO(modules)
 	} else {
 		gos <- go
 	}
 
 	kegg.table <- toTable(KEGGPATHID2NAME)
 	if ( is.null(kegg) ) {
-		keggs <- ISA.KEGG(modules)
+		keggs <- ISAKEGG(modules)
 	} else {
 		keggs <- kegg
 	}

### 2009-10-01 homepage/generate.py
Homepage improvements.

@@ -11,8 +11,11 @@ import glob
 import os.path
 
 template = [string.Template(line) for line in open("template.html")]
-tokens = { "ISA_VERSION": "0.1", "EISA_VERSION": "0.1", 
-           "EXPRESSIONVIEW_VERSION": "0.1" }
+tokens = { "ISA_VERSION": "0.1", 
+           "EISA_VERSION": "0.1", 
+           "EXPRESSIONVIEW_VERSION": "0.1",
+           "TARGETSCAN_VERSION": "5.0-1",
+           }
 
 files = glob.glob('*.html.in')
 

### 2009-10-02 eisa/R/autogen.R
Various eisa HTML generation updates to make it work with the biclust package.

@@ -3,7 +3,8 @@ ISAHTML <- function(eset, modules, target.dir,
                      template=system.file("autogen", package="eisa"),
                      GO, KEGG, miRNA=NULL, CHR=NULL, htmltitle=NULL,
                      notes=NULL, seed=NULL,
-                     cond.to.include=NULL, cond.col="white", sep=NULL) {
+                     cond.to.include=NULL, cond.col="white", sep=NULL,
+                     condPlot=TRUE) {
 
   ISAHTMLTable(modules=modules, target.dir=target.dir, template=template,
                  GO=GO, KEGG=KEGG, miRNA=miRNA, CHR=CHR, htmltitle=htmltitle,
@@ -12,7 +13,7 @@ ISAHTML <- function(eset, modules, target.dir,
   ISAHTMLModules(eset=eset, modules=modules, target.dir=target.dir,
                    template=template, GO=GO, KEGG=KEGG, miRNA=miRNA, CHR=CHR,
                    cond.to.include=cond.to.include,
-                   cond.col=cond.col, sep=sep, seed=seed)
+                   cond.col=cond.col, sep=sep, seed=seed, condPlot=condPlot)
 
   invisible(NULL)
 }
@@ -193,7 +194,7 @@ ISAHTMLModules <- function(eset, modules, which=seq_len(length(modules)),
                              GO=NULL, KEGG=NULL, miRNA=NULL, CHR=NULL,
                              cond.to.include=NULL,
                              cond.col="white",
-                             sep=NULL, seed=NULL) {
+                             sep=NULL, seed=NULL, condPlot=TRUE) {
 
   isa2:::isa.status("Generating module pages", "in")
 
@@ -266,7 +267,7 @@ ISAHTMLModules <- function(eset, modules, which=seq_len(length(modules)),
                        target.dir=target.dir, template=template,
                        GO=GO, KEGG=KEGG, miRNA=miRNA, CHR=CHR,
                        cond.to.include=cond.to.include,
-                       cond.col=cond.col, sep=sep,
+                       cond.col=cond.col, sep=sep, condPlot=condPlot,
                        seed=seed, drive.BP=drive.BP, drive.CC=drive.CC,
                        drive.MF=drive.MF, drive.KEGG=drive.KEGG,
                        drive.miRNA=drive.miRNA, drive.CHR=drive.CHR,
[…]

### 2009-10-02 eisa/R/eisa.R
Update to make eisa visualization tools work with biclust results.

@@ -35,7 +35,9 @@ eisa.get.nm <- function(data, modules) {
     data
   } else {
     if (!is.null(modules)) {
-      ISANormalize(data, prenormalize=runData(modules)$prenormalize)
+      pre <- runData(modules)$prenormalize
+      if (is.null(pre)) { pre <- FALSE }
+      ISANormalize(data, prenormalize=pre)
     } else {
       ISANormalize(data)
     }

### 2009-10-06 isanote/modular.tex
ISA note updated, a figure was added.

@@
 \pubyear{2009}
+
+\newcommand{\Rpackage}{`\texttt{#1}'}
 
@@
   $^{2}$Swiss Institute of Bioinformatics,
-  University of Lausanne, Rue de Bugnon 27 - DGM 328, CH-1005 Lausanne,
+  University of Lausanne, Rue de Bugnon 27, CH-1005 Lausanne,
   Switzerland.}
@@
 
-What is biclustering.
-What is the ISA.
+% What is biclustering.
+
+Biclustering is the idea of clustering the rows and the columns of a
+data matrix, together, usually not independently. In other words, the
+goal of a biclustering algorithm is to find blocks in the reordered
+data matrix that have, such that the blocks have correlated rows and
+columns. It is important to note that these blocks (or biclusters)
+have context, i.e. the rows only need to be correlated across the
+columns of the block, but not across other columns of the data matrix;
+and vice-versa.
+
+When it comes to gene expression data, a bicluster means a subset of
+genes ($G$) (or probes, probe sets) and a subset of samples ($S$),
+such that genes $G$ are exactly coexpressed across $S$ and samples $S$
+have correlated expression profile exactly across $G$.
+
+% What is the ISA.
+
+The Iterative Signature Algorithm (ISA)~\cite{sa,isa,ismod} is a 
+powerful biclustering algorithm. It is able to identify biclusters
+that overlap and it is resilient to noise.
+
+In this short note, we introduce the \Rpackage{isa2} and
+\Rpackage{eisa} GNU R~\cite{R} packages; these implement the ISA
+biclustering method, provide a convenient interface to run it using
+the BioConductor~\cite{BioC} software package, and also contain
+visualization tools.
 
@@
 
-Normalization.
+\begin{figure*}
+\centering
+\includegraphics[width=\textwidth]{isa2workflow3}
+\caption{}
+\label{fig:workflow}
+\end{figure*}
 
-Random and smart seeding.
+% Batch correction
 
-The ISA iteration.
+% Normalization.
 
-Merging the modules.
+% Random and smart seeding.
 
-The robustness measure and filtering the modules.
+% The ISA iteration.
 
-Module trees.
+% Merging the modules.
+
+% The robustness measure and filtering the modules.
+
+% Module trees.
 
@@
 
-\subsection{Workflow}
+\subsection{Work flow}
 
@@
 
+\subsubsection{The simple work flow}
+
+\subsubsection{The detailed work flow}
+
 \subsection{Visualization}
@@
 Expression plots, GO tree plots, module tree plots, HTML summary
-generation.
+generation. ExpressionView.
 

### 2009-10-06 isanote/plots.R

@@ -1,0 +1,71 @@
+
+library(eisa)
+library(ALL)
+library(hgu95av2.db)
+library(affy)
+library(GO.db)
+data(ALL)
+
+thr.gene <- 2.7
+thr.cond <- 1.4
+set.seed(1) # to get the same results, always
+modules <- ISA(ALL, thr.gene=thr.gene, thr.cond=thr.cond)
+
+# Condition plot
+col <- ifelse( grepl("^B", pData(modules)$BT), "darkolivegreen", "orange")
+pdf("condplot.pdf", width=10, height=5)
+condPlot(modules, 1, ALL, col=col)
+dev.off()
+
+# Profile plot
+png("profileplot.png")
+profilePlot(modules, 2, ALL, plot="both")
+dev.off()
+pdf("profileplot.pdf")
+profilePlot(modules, 2, ALL, plot="both")
+dev.off()
+
+# Expression plot
+ep <- expPlotCreate(ALL, modules, 5)
+pdf("expplot.pdf")
+expPlot(ep)
+dev.off()
+
+# Another module set for the module tree
+
+varLimit <- 0.5
+kLimit <- 4
+ALimit <- 5
+flist <- filterfun(function(x) var(x)>varLimit, kOverA(kLimit,ALimit))
+ALL.filt <- ALL[genefilter(ALL, flist), ]
+ALL.filt2 <- ALL.filt[, grepl("^B", ALL.filt$BT)]
+
+set.seed(1)
+modules <- ISA(ALL.filt2, flist=NA, thr.gene=seq(2,4,by=0.5), thr.cond=1)
+
+modules2 <- ISASweep(ALL.filt2, modules)
+
+G <- ISASweepGraph(modules2)
+
+GO <- ISAGO(modules2)
+p.bp <- sapply(summary(GO$BP), function(x) as.integer(-log10(x$Pvalue[1])))
+c.bp <- d.bp <- sapply(sigCategories(GO$BP), function(x) x[1])
+d.bp[!is.na(c.bp)] <- sapply(mget(na.omit(c.bp), GOTERM), Term)
+d.bp <- abbreviate(d.bp, 6)
+colbar <- hcl(h=260, c=35, l=seq(30, 100, length=20))
+colbar <- c("#FFFFFF", rev(colbar))
+col <- colbar[p.bp]
+
+G2 <- delete.vertices(G, match(c(19,24,27,28,29,30,32,33,35:41), V(G)$id)-1)
+G2$layout <- G2$layout[V(G2)$id,]
+
[…]

### 2009-10-09 RExpressionView/src/orderclusters.h
RExpressionView: Optimized fitness function.

@@ -16,13 +16,20 @@ class Clusters {
 		vector<set<int> > redundantslots;
 		vector<int> optimallength;
 		double optimalfitness;
-		
+
+		vector<vector<int> > maximalcluster;
+		set<int> modifiedslots;
+		set<int> recalculateclusters;
+
 		int random(int, int);
 		void swap(int, int);
 		void shift(int, int, int);
 		int findbestposition(int, int);
 
+		void getclusters();
+
 		double getfitness();
+		double getfullfitness();
 		double getsimilarity(int, int);
 		double getoptimalfitness();
 
@@ -61,6 +68,5 @@ class Clusters {
 		double quality();
 		
 };
-
 
 #endif

### 2009-10-09 RExpressionView/R/export.R
RExpressionView: Updated R functions. Fixed the maxtime bug in OrderEV.

@@ -22,7 +22,7 @@ if (require(eisa)) {
 	setMethod("ExportEV", signature(modules="ISAModules"), function(modules, ...) ExportEV.ISAModules(modules, ...))
 }
 
-ExportEV.ISAModules <- function(modules, eset, order, go=NULL, kegg=NULL, filename="", norm=c("sample", "feature", "raw")) {
+ExportEV.ISAModules <- function(modules, eset, order, filename="", norm=c("sample", "feature", "raw"), go=NULL, kegg=NULL) {
 	
 	if ( filename == "" ) {
 		con <- file(file.choose(TRUE), open="w")

### 2009-10-15 isanote/modular.tex
ISA application note updates, almost finished.

@@
 
-\newcommand{\Rpackage}{`\texttt{#1}'}
+\newcommand{\Rpackage}[1]{`\texttt{#1}'}
+\newcommand{\Rclass}[1]{\textsl{#1}}
+\newcommand{\subfig}[1]{\textbf{#1}}
 
@@
 \title[Modular analysis]{Modular analysis of gene expression data with
-  GNU R or Matlab}
+  GNU R}
 \author[G\'abor Cs\'ardi \textit{et~al}]{G\'abor Cs\'ardi\,,$^{1,2}$,
@@
 \begin{abstract}
-
-\section{Summary:}
+TODO
+\section{Summary:} TODO
 \section{Availability:}
+\href{http://www.unil.ch/cbg/homepage/software.html}%
+{http://www.unil.ch/cbg/homepage/software.html}
 \section{Contact:} \href{Sven.Bergmann@unil.ch}{Sven.Bergmann@unil.ch}
@@
 Biclustering is the idea of clustering the rows and the columns of a
-data matrix, together, usually not independently. In other words, the
+data matrix together, usually not independently. In other words, the
 goal of a biclustering algorithm is to find blocks in the reordered
 data matrix that have, such that the blocks have correlated rows and
-columns. It is important to note that these blocks (or biclusters)
-have context, i.e. the rows only need to be correlated across the
-columns of the block, but not across other columns of the data matrix;
-and vice-versa.
+columns. It is important to note that these blocks (or biclusters,
+modules) have context, i.e. the rows only need to be correlated across
+the columns of the block, but not across other columns of the data
+matrix; and vice-versa.
 
@@
 
-The Iterative Signature Algorithm (ISA)~\cite{sa,isa,ismod} is a 
+The Iterative Signature Algorithm (ISA), \cite{sa,isa,isamod}, is a 
 powerful biclustering algorithm. It is able to identify biclusters
@@
 In this short note, we introduce the \Rpackage{isa2} and
-\Rpackage{eisa} GNU R~\cite{R} packages; these implement the ISA
+\Rpackage{eisa} GNU R, \cite{R}, packages; these implement the ISA
 biclustering method, provide a convenient interface to run it using
-the BioConductor~\cite{BioC} software package, and also contain
+the BioConductor, \cite{BioC}, software package, and also contain
 visualization tools.
@@
 \centering
-\includegraphics[width=\textwidth]{isa2workflow3}
-\caption{}
+\includegraphics[width=0.9\textwidth]{isa2workflow3}
+\caption{Work flow of a typical modular analysis study, with the
+  \Rpackage{eisa} package (left); and four bicluster visualization
+  techniques (right). The input of the \Rpackage{eisa} pipeline is a
+  BioConductor \Rclass{ExpressionSet} object. This is normalized to
+  $Z$-scores (1); the ISA iteration is performed on it using smart or
+  random seeds (2); and the biclusters are then merged and filtered
+  based on the robustness measure (3). The \Rpackage{eisa} package
+  contains tools to perform many GO, KEGG, etc. enrichment calculation
+  quickly (4), and several bicluster visualization tools (5).
+  Subfigures \subfig{A}, \subfig{B}, \subfig{C} and \subfig{D} were generated
+  using the acute lymphoblastic leukemia data set, see
+  \cite{chiaretti04} and the \Rpackage{ALL} R package.
+  \subfig{(A)} A condition plot for a simple bicluster. Each sample is
+  represented as a bar, its height is its score in the bicluster. 
+  Samples above the red (none in this case) and samples
+  below the green line are part of the biclusters. The coloring
+  indicates cases and controls. This module can separate control and
+  case samples.
+  \subfig{(B)} A module tree. Each rectangle is a bicluster. See the
+  definition of the module tree in the text. The modules are colored
+  according to their Gene Ontology enrichment $p$-values, the codes of
+  the enriched GO categories are shown in the top-left corner of the
+  rectangles. The top-right corner shows the number of genes and
+  conditions in the bicluster. The numbers on the horizontal axes are
+  the gene thresholds used for finding the modules.
+  \subfig{(C)} Heatmap for a single module, showing correlated
+  expression and the genes and samples. The red lines are the gene and
+  sample scores.
+  \subfig{(D)} Profile plot, for a single module. On the top plot the
+  expression of all samples are plotted, the samples of the module
+  with black. The red and green lines show the mean of the two
+  groups. The bottom plot is the same for the genes of the module.
+}
 \label{fig:workflow}
@@
 
+Let us first explain the briefly the steps of a typical modular
+analysis study for gene expression data.
+
 % Batch correction
+\paragraph{Batch correction}
+Often, the ISA is applied to samples from different experiments. In
+this case it is crucial to address the question of experimental
+variation. Several methods exist to solve this problem,
+see e.g. \cite{johnson07} for an algorithm that has a GNU R
+implementation.
+
+% Filtering
+\paragraph{Gene filtering}
+Depending on the application, it can be advantageous to remove genes
+that are not expressed in any of the samples, especially because the
+ISA normalization (next step) tends to amplify their effect.
 
 % Normalization. 
+\paragraph{ISA normalization}
+ISA is an iterative algorithm. In each step it computes weighted sums
+of expression levels for a number of genes or a number of
+samples. Since different genes typically show different levels of base
+expression and variance, it is important to standardize expression
+levels to $Z$-scores. The ISA uses two sets of $Z$-scores, one
+calculated for each gene across all samples, the other for each sample
+across all genes.
 
 % Random and smart seeding.
+\paragraph{Random and smart seeding}
+The ISA used with random seeds is an unsupervised algorithm, it uses no
+external knowledge to find the biclusters. ISA with smart seeds is
+biased towards genes or pathways of interest, or sample annotation.
+This latter form can be considered as a semi-supervised approach.
 
 % The ISA iteration. 
+\paragraph{The ISA iteration}
+This step is essentially performing the ISA and finding the biclusters
+from the starting seeds.
 
 % Merging the modules.
+\paragraph{Merging the modules}
+It is possible that more seeds converge to the same, or very similar
+biclusters. This step eliminates such duplicates.
 
 % The robustness measure and filtering the modules. 
+\paragraph{Robustness of biclusters}
+To access the significance of a bicluster, we designed a robustness
+measure, that can be used to filter out spurious modules. This is done
+based on performing ISA on the scrambled input data.
 
 % Module trees.
+\paragraph{Module trees}
+The ISA works with two stringency threshold parameters, the gene
+threshold and the sample threshold. ISA modules can be organized into
+a directed graph, a module tree, in which there is an edge from module
+$A$ to module $B$, if the ISA converges to module $B$ from module $A$,
+with the same threshold parameters  that were used to find module
+$B$. A module tree provides a hierarchical modular description of a
+data set.
 
-\section{R Implementation}
+\section{Implementation}
 
-Two packages: isa2 does the computation, eisa does everything related
-to biology.
+The ISA and accompanying visualization tools are implemented in two R
+packages. The \Rpackage{isa2} package contains the implementation of
+the basic ISA itself; this package can be used to analyze any tabular
+data. The \Rpackage{eisa} package builds on \Rpackage{isa2}. It adds
+support to standard BioConductor gene expression data structures; and
+contains gene expression specific visualization tools, see some of
+these in Fig.~\ref{fig:workflow}.
 
-\subsection{Work flow}
+Both the \Rpackage{isa2} and \Rpackage{eisa} packages support two
+workflows. The \emph{simple workflow} involves a single R function
+call and runs all ISA steps (1-3 on Fig.~\ref{fig:workflow}.) with
+their default parameters.
 
-Two workflows, the simple workflow does everything with default
-parameters. In the detailed workflow the user has access to every step
-of the analysis.
+The \emph{detailed workflow} means running every step of the modular
+analysis separately, possibly with non-default parameters. This allows
+the users to tailor the ISA completely according their needs.
 
-\subsubsection{The simple work flow}
+For users that prefer Matlab over R, we created a Matlab package for
+ISA, please see the ISA homepage for more information.
 
-\subsubsection{The detailed work flow}
+% \subsubsection{Visualization}
 
-\subsection{Visualization}
+The \Rpackage{eisa} package implements a set of visualization
+techniques for biclusters, some of these are specific to ISA, others
+are not. Please see Fig.~\ref{fig:workflow} for some of these.
 
-Expression plots, GO tree plots, module tree plots, HTML summary
-generation. ExpressionView.
+The \Rpackage{ExpressionView} R package implements an interactive tool
+for visualizing ISA (or other) overlapping biclusters.
 
-\subsection{Connection to other software}
+% \subsubsection{Connection to other software}
 
-The biclust package.
-
-\section{Matlab implementation}
+The \Rpackage{biclust} package, \cite{biclust}, implements a number of
+biclustering algorithms in a unified framework. The \Rpackage{eisa}
+package include tools to convert between \Rpackage{biclust} and ISA
+biclusters. This allows the cross-talk of the functions in the two
+packages.
 
@@
 
-\paragraph{Funding\textcolon}
+TODO
+
+\paragraph{Funding\textcolon} TODO
 
@@
 \bibliography{isa}
-\nocite{*}
 

### 2009-10-15 isanote/modular.tex
isanote updated, seems to be ready.

@@
 
-\title[Modular analysis]{Modular analysis of gene expression data with
-  GNU R}
+\title[Modular analysis]{Modular analysis of gene expression data with R}
 \author[G\'abor Cs\'ardi \textit{et~al}]{G\'abor Cs\'ardi\,,$^{1,2}$,
@@
 \begin{abstract}
-TODO
-\section{Summary:} TODO
-\section{Availability:}
-\href{http://www.unil.ch/cbg/homepage/software.html}%
-{http://www.unil.ch/cbg/homepage/software.html}
-\section{Contact:} \href{Sven.Bergmann@unil.ch}{Sven.Bergmann@unil.ch}
-\end{abstract}
+\section{Summary:} 
+Biclustering means clustering the rows and the columns of a
+data matrix together, usually not independently. A bicluster is, then,
+a block of the reordered input matrix. When it comes to gene
+expression data, a bicluster means a subset of genes ($G$) (or probes,
+probe sets) and a subset of samples ($S$), such that genes $G$ are
+exactly coexpressed across $S$ and samples $S$ have correlated
+expression profile exactly across $G$.
 
-\section{Introduction}
-
-% What is biclustering.
-
-Biclustering is the idea of clustering the rows and the columns of a
-data matrix together, usually not independently. In other words, the
-goal of a biclustering algorithm is to find blocks in the reordered
-data matrix that have, such that the blocks have correlated rows and
-columns. It is important to note that these blocks (or biclusters,
-modules) have context, i.e. the rows only need to be correlated across
-the columns of the block, but not across other columns of the data
-matrix; and vice-versa.
-
-When it comes to gene expression data, a bicluster means a subset of
-genes ($G$) (or probes, probe sets) and a subset of samples ($S$),
-such that genes $G$ are exactly coexpressed across $S$ and samples $S$
-have correlated expression profile exactly across $G$.
-
-% What is the ISA.
-
-The Iterative Signature Algorithm (ISA), \cite{sa,isa,isamod}, is a 
+The Iterative Signature Algorithm (ISA), \cite{sa,isa}, is a 
 powerful biclustering algorithm. It is able to identify biclusters
@@
 
-\section{Methods}
+\section{Availability:}
+\href{http://www.unil.ch/cbg/homepage/software.html}%
+{http://www.unil.ch/cbg/homepage/software.html}
+\section{Contact:} \href{Sven.Bergmann@unil.ch}{Sven.Bergmann@unil.ch}
+\end{abstract}
+
+\section{Introduction}
+
+% How the ISA works
+
+The ISA iteration is started from an input seed, this is 
+refined it by adding and/or removing more genes and/or
+samples until the process converges to a stable bicluster. The
+procedure is typically performed to a large number of seeds; these can
+be either chosen randomly (random seeding, unsupervised method), or
+based on the data (smart seeding, semi-supervised method).
+
+The output of ISA is a set of potentially overlapping
+biclusters. These biclusters are not binary, both genes and samples of
+a bicluster have scores, normalized between minus one and one. Genes
+and samples with scores further away from zero have a stronger
+association to the module.
+
+In this short note has two parts, in Section~\ref{sec:methods} we look
+over the steps of a typical modular analysis, and in
+Section~\ref{sec:implementation} introduce the R packages that
+implement the ISA.
+
+\section{Methods}%
+\label{sec:methods}
 
@@
 \centering
-\includegraphics[width=0.9\textwidth]{isa2workflow3}
+\includegraphics[width=\textwidth]{isa2workflow3}
 \caption{Work flow of a typical modular analysis study, with the
@@
   below the green line are part of the biclusters. The coloring
-  indicates cases and controls. This module can separate control and
-  case samples.
-  \subfig{(B)} A module tree. Each rectangle is a bicluster. See the
-  definition of the module tree in the text. The modules are colored
+  indicates cases and controls.
+  \subfig{(B)} A module tree. Each rectangle is a bicluster; see the
+  definition of the edges in the text. The modules are colored
   according to their Gene Ontology enrichment $p$-values, the codes of
@@
   rectangles. The top-right corner shows the number of genes and
-  conditions in the bicluster. The numbers on the horizontal axes are
-  the gene thresholds used for finding the modules.
+  conditions in the bicluster. The gene thresholds used for finding
+  the modules is shown on the horizontal axes.
   \subfig{(C)} Heatmap for a single module, showing correlated
-  expression and the genes and samples. The red lines are the gene and
+  expression for the genes and samples. The red lines are the gene and
   sample scores.
   \subfig{(D)} Profile plot, for a single module. On the top plot the
-  expression of all samples are plotted, the samples of the module
+  expression of all samples is plotted, the samples of the module
   with black. The red and green lines show the mean of the two
@@
 
-Let us first explain the briefly the steps of a typical modular
+Let us first explain briefly the steps of a typical modular
 analysis study for gene expression data.
@@
 % Batch correction
-\paragraph{Batch correction}
+\emph{Batch correction.}
 Often, the ISA is applied to samples from different experiments. In
@@
 % Filtering
-\paragraph{Gene filtering}
+\emph{Gene filtering.}
 Depending on the application, it can be advantageous to remove genes
@@
 % Normalization. 
-\paragraph{ISA normalization}
+\emph{ISA normalization.}
 ISA is an iterative algorithm. In each step it computes weighted sums
@@
 % Random and smart seeding.
-\paragraph{Random and smart seeding}
-The ISA used with random seeds is an unsupervised algorithm, it uses no
-external knowledge to find the biclusters. ISA with smart seeds is
-biased towards genes or pathways of interest, or sample annotation.
-This latter form can be considered as a semi-supervised approach.
+\emph{Random and smart seeding.}
+As mentioned previously, ISA can be performed with random or smart
+seeds, based on the application.
 
 % The ISA iteration. 
-\paragraph{The ISA iteration}
+\emph{The ISA iteration.}
 This step is essentially performing the ISA and finding the biclusters
@@
 % Merging the modules.
-\paragraph{Merging the modules}
+\emph{Merging the modules.}
 It is possible that more seeds converge to the same, or very similar
@@
 % The robustness measure and filtering the modules. 
-\paragraph{Robustness of biclusters}
+\emph{Robustness of biclusters.}
 To access the significance of a bicluster, we designed a robustness
@@
 % Module trees.
-\paragraph{Module trees}
+\emph{Module trees.}
 The ISA works with two stringency threshold parameters, the gene
@@
 
-\section{Implementation}
+\section{Implementation}%
+\label{sec:implementation}
 
@@
 TODO
-
 \paragraph{Funding\textcolon} TODO
-
 \paragraph{Conflict of interest\textcolon} None declared.

### 2009-10-29 RExpressionView/R/AllGenerics.R
RExpressionView: Updated documentation, tutorials and description of ordering algorithm.

@@ -1,5 +1,5 @@
 setGeneric("OrderEV",
-           function(modules, ...) standardGeneric("OrderEV"))
+           function(biclusters, ...) standardGeneric("OrderEV"))
 
 setGeneric("ExportEV",
-           function(modules, ...) standardGeneric("ExportEV"))
+           function(biclusters, ...) standardGeneric("ExportEV"))

### 2009-10-30 eisa/inst/doc/EISA_biclust.Rnw
Simplify Makefile, reorganize HTML generation from the vignettes.

% \VignetteIndexEntry{The eisa and the biclust packages}
\documentclass{article}
\usepackage{ragged2e}
\usepackage{url}

\newcommand{\Rfunction}[1]{\texttt{#1()}}
\newcommand{\Rpackage}[1]{\texttt{#1}}
\newcommand{\Rclass}[1]{\texttt{#1}}
\newcommand{\Rargument}[1]{\textsl{#1}}
\newcommand{\filename}[1]{\texttt{#1}}
\newcommand{\variable}[1]{\texttt{#1}}

\SweaveOpts{cache=TRUE}
\SweaveOpts{prefix.string=plot}

\setlength{\parindent}{2em}

\begin{document}

\title{The \Rpackage{eisa} and \Rpackage{biclust} packages}
\author{G\'abor Cs\'ardi}
\maketitle

\tableofcontents

\RaggedRight

<<set width,echo=FALSE,print=FALSE>>=
options(width=60)
options(continue=" ")
@

\section{Introduction}

Biclustering is technique that simultaneously clusters the rows and
columns of a matrix~\cite{madeira04}. In other words, the problem is
finding blocks in the reordered input matrix that exhibit correlated
behavior, both across the rows and columns of the block.

Biclustering is used increasingly in the analysis of gene expression
data sets, because it reduces the complexity of the data: instead of
tens of thousands of individual genes, one can focus on a handful of
biclusters, in which the genes behave similarly.

The Iterative Signature Algorithm (ISA)~\cite{isa} is a biclustering
method, that can efficiently find potentially overlapping biclusters
(modules, according to the ISA terminology) in a matrix. The ISA is
implemented in the \Rpackage{eisa} package~\cite{eisapackage} it uses
standard BioConductor classes and includes a number of visualization
tools as well.

The \Rpackage{biclust} R package~\cite{biclust} is a general
biclustering package, it contains several biclustering methods, and
these can be invoked with a common interface. It provides a set of
visualization tools for the results.

In this short document, we show examples on how to use the
visualization tools of \Rpackage{eisa} for the biclusters found with
\Rpackage{biclust}, and vice-versa.

\section{From \Rclass{Biclust} to \Rclass{ISAModules}}

For all examples in this document, we will use the acute lymphoblastic
leukemia data set, that is included in the standard BioConductor
\Rpackage{ALL} package. Let's load this data set and the required
packages first.
<<packages>>=
library(biclust)
library(eisa)
library(ALL)
data(ALL)
@

Next, we select a subset of the genes in the data set. We do this to speed 
up the computation for our simple examples. We select the genes that are 
annotated to involved in immune system processes, according to the Gene
Ontology database.

<<filter>>=
library(GO.db)
library(hgu95av2.db)
gotable <- toTable(GOTERM)
myterms <- unique(gotable$go_id[gotable$Term %in% c("immune system process")])
myprobes <- unique(unlist(mget(myterms, hgu95av2GO2ALLPROBES)))
ALL.filtered <- ALL[myprobes,]
@

We have kept only \Sexpr{nrow(ALL.filtered)} probes:
<<filtered>>=
nrow(ALL.filtered)
@

For consistent results, we set the random seed.
<<seed>>=
set.seed(0xf00)
@

Next, we apply the Plaid Model Biclustering method~\cite{turner03} to the 
reduced data set.
<<bcplaid>>=
Bc <- biclust(exprs(ALL.filtered), BCPlaid(),
              fit.model = ~m + a + b, verbose = FALSE)
@ 
The method finds \Sexpr{Bc@Number} biclusters, and returns a 
\Rclass{Biclust} object:
<<bcplaid result>>=
class(Bc)
Bc
@

Now we will convert the \Rclass{Biclust} object to an \Rclass{ISAModules} 
object, that is used in the \Rpackage{eisa} package. To help
some \Rpackage{eisa} functions, we add the name of the annotation package to 
the parameters stored in the \Rclass{Biclust} object, this is always adwised.
The procedure makes use of the probe and sample names that are kept and stored 
in the \Rclass{Biclust} object, this information will be used later, e.g. for 
the enrichment analysis. The conversion itself can be performed with
the usual \Rfunction{as} function. 
<<convert bc to isa>>=
Bc@Parameters$annotation <- annotation(ALL.filtered)
modules <- as(Bc, "ISAModules")
modules
@

Now we are able apply the usual \Rclass{ISAModules} methods to 
the biclusters. See more about these functions in the documentation of the 
\Rpackage{eisa} package.

Doing some enrichment analysis is easy:
<<biclust enrichment>>=
library(KEGG.db)
KEGG <- ISAKEGG(modules)
sigCategories(KEGG)[[2]]
unlist(mget(sigCategories(KEGG)[[2]], KEGGPATHID2NAME))
@

The \Rfunction{ISA2heatmap} function creates a heatmap for a module. Let us
annotate the heatmap with the leukemia sample type, white means B-cell, black 
means T-cell leukemia. See Fig.~\ref{fig:heatmap}.
<<heatmap,eval=FALSE>>=
col <- ifelse(grepl("^B", ALL.filtered$BT), "white", "black")
modcol <- col[ getSamples(modules, 2)[[1]] ]
ISA2heatmap(modules, 2, ALL.filtered, 
            ColSideColors=modcol)
@

\begin{figure}
\centering
<<heatmap-real,fig=TRUE,echo=FALSE>>=
<<heatmap>>
@
\caption{Heatmap of the second module, found with the Plaid Model
  biclustering algorithm. The black squares denote the T-cell samples;
  all samples in the module belong to T-cell samples.}
\label{fig:heatmap}
\end{figure}

It turns out, that all samples in the second bicluster belong to
patients with T-cell leukemia.

Profile plots visualize the mean expression levels, both for the
genes/samples in the module and in the background (i.e. the background
means all genes and samples \emph{not} in the module). 
See Fig.~\ref{fig:profile}.
<<profilePlot,eval=FALSE>>=
profilePlot(modules, 2, ALL, plot="both")
@
\begin{figure}
\centering
<<profilePlot-real,fig=TRUE,echo=FALSE>>=
<<profilePlot>>
@ 
\caption{Profile plot for the second module. The red lines show the
  average exression of the samples/genes in the module. The green
  lines show the same for the samples/genes not in the module.}
\label{fig:profile}
\end{figure}

The \Rfunction{gograph} and \Rfunction{gographPlot} functions 
create a plot of the part of the Gene Ontology tree that contains 
the enriched categories. See Fig.~\ref{fig:gotree}.
<<psoptions,results=hide,echo=FALSE>>=
ps.options(fonts=c("serif", "mono"))
@
<<GOtreeplot,eval=FALSE>>=
library(GO.db)
GO <- ISAGO(modules)
gog <- gograph(summary(GO$CC)[[2]])
summary(gog)
gographPlot(gog)
@ 
\begin{figure}
\centering
<<GOtreeplot-real,fig=TRUE,width=10,height=4,echo=FALSE>>=
<<GOtreeplot>>
@
\caption{Part of the Gene Ontology tree, Cellular Components ontology.
  The plot includes all terms with significant enrichment for the
  second module, and their parent terms, up to the most general term.
}
\label{fig:gotree}
\end{figure}

The \Rfunction{ISAHTML} function creates a HTML overview of all modules.
<<html>>=
CHR <- ISACHR(modules)
htmldir <- tempdir()
ISAHTML(eset=ALL.filtered, modules=modules, target.dir=htmldir, 
        GO=GO, KEGG=KEGG, CHR=CHR, condPlot=FALSE)
if (interactive()) {
  browseURL(URLencode(paste("file://", htmldir, "/index.html", sep="")))
}
@

The \Rfunction{ISAmnplot} funtion plots group means of expression
levels againts each other, for all genes in the module. Here we plot
the mean expression of the B-cell samples against the T-cell samples, 
for the second module. See Fig.~\ref{fig:mnplot}.
<<mnplot,eval=FALSE>>=
group <- ifelse(grepl("^B", ALL.filtered$BT), "B-cell", "T-cell")
ISAmnplot(modules, 2, ALL.filtered, norm="raw", group=group)
@ 
\begin{figure}
\centering
<<mnplot-real,fig=TRUE,echo=FALSE>>=
<<mnplot>>
@ 
\caption{Group means against each other, for B-cell and T-cell
  samples, for all genes in the second bicluster.}
\label{fig:mnplot}
\end{figure}

\section{From \Rclass{ISAModules} to \Rclass{Biclust}}

It is also possible to convert an \Rclass{ISAModules} object to a 
\Rclass{Biclust} object, but this involves some information loss. The reason
for this is, that ISA biclusters are not binary, but the genes and 
the samples both have scores between minus one and one; whereas 
\Rclass{Biclust} biclusters are required to be binary.

We make use of the small sample set of modules that is included in the
\Rpackage{eisa} package. These were generated for the ALL data set.
<<load isa data>>=
data(ALLModules)
ALLModules
@

The conversion from \Rclass{ISAModules} to \Rclass{Biclust} can be done
the usual way:
<<isa to biclust>>=
BcMods <- as(ALLModules, "Biclust")
BcMods
@

The usual methods of the \Rclass{Biclust} class can be applied to 
\texttt{BcMods} now. E.g. we can calculate the coherence of the 
biclusters:
<<coherence>>=
data <- exprs(ALL[featureNames(ALLModules),])
constantVariance(data, BcMods, 1)
additiveVariance(data, BcMods, 1)
multiplicativeVariance(data, BcMods, 1)
signVariance(data, BcMods, 1)
@

As another example, we calculate these coherence measures for all 
modules and compare them to the ISA robustness measure.
<<coherence vs robustness>>=
cV <- sapply(1:BcMods@Number, function(x) constantVariance(data, BcMods, x))
aV <- sapply(1:BcMods@Number, function(x) additiveVariance(data, BcMods, x))
mV <- sapply(1:BcMods@Number, function(x) multiplicativeVariance(data, BcMods, x))
sV <- sapply(1:BcMods@Number, function(x) signVariance(data, BcMods, x))
rob <- ISARobustness(ALL, ALLModules)
@

Let's create a pairs-plot to visualize the relationship of these measures
for our data set, the result is in Fig.~\ref{fig:coherence}.
<<pairs,eval=FALSE>>=
pairs( cbind(cV, aV, mV, sV, rob) )
@
\begin{figure}
\centering
<<pairs-real,fig=TRUE,echo=FALSE>>=
<<pairs>>
@ 
\caption{Relationship of the various bicluster coherence measueres and
  the ISA robustness measure. They show high correlation.}
\label{fig:coherence}
\end{figure}

\section{More information}

For more information about the ISA, please see the references
below. The ISA homepage at
\url{http://www.unil.ch/cbg/homepage/software.html} has example data
sets, and all ISA related tutorials and papers.

\section{Session information}

The version number of R and packages loaded for generating this
vignette were:

<<sessioninfo,results=tex,echo=FALSE>>=
toLatex(sessionInfo())
@

\bibliographystyle{apalike}
\bibliography{EISA}
\nocite{*}

\end{document}

### 2009-10-30 eisa/inst/doc/EISA_module_trees.Rnw
Simplify Makefile, reorganize HTML generation from the vignettes.

% \VignetteIndexEntry{ISA module trees}
\documentclass{article}
\usepackage{ragged2e}
\usepackage{url}

\newenvironment{narrow}[2]{%
  \begin{list}{}{%
  \setlength{\topsep}{0pt}%
  \setlength{\leftmargin}{#1}%
  \setlength{\rightmargin}{#2}%
  \setlength{\listparindent}{\parindent}%
  \setlength{\itemindent}{\parindent}%
  \setlength{\parsep}{\parskip}}%
\item[]}{\end{list}}

\newcommand{\Rfunction}[1]{\texttt{#1()}}
\newcommand{\Rpackage}[1]{\texttt{#1}}
\newcommand{\Rclass}[1]{\texttt{#1}}
\newcommand{\Rargument}[1]{\textsl{#1}}
\newcommand{\filename}[1]{\texttt{#1}}
\newcommand{\variable}[1]{\texttt{#1}}

\SweaveOpts{cache=TRUE}
\SweaveOpts{prefix.string=plot}

\setlength{\parindent}{2em}

\begin{document}

\title{Transcription Module Trees from Gene Expression Data}
\author{G\'abor Cs\'ardi}
\maketitle

\tableofcontents

\RaggedRight

<<set width,echo=FALSE,print=FALSE>>=
options(width=60)
options(continue=" ")
@

\section{Introduction}

The Iterative Signature Algorithm (ISA) finds transcription modules in
gene expression data. A module is a block of the reordered gene
expression matrix.

The ISA has two threshold parameters: the gene (or feature) threshold
and the conditions (or sample) threshold. These essentially specify
the stringency of the biclustering. If the gene threshold is high,
then modules will have less genes, only the one that are consistently
far above or below the mean expression level will be included. The
same is true for the sample threshold and the samples in the modules.

It is possible to run the ISA using different threshold parameters and
define a hierarchical organization of the found modules. We call such
a description a sweep graph or sweep tree. The vertices (or nodes) of
the sweep graph are the modules and there is a directed edge from
module A to module B if the ISA converges to module B, when started
from module A, and the thresholds that were used to find module B are
used.

A sweep tree can be generated by keeping one threshold parameter
fixed, and vary the other one. We first find modules at a stringent
threshold value and then check where these converge at less stringent
thresholds.

This document shows how it is possible to create a sweep tree with the
\Rpackage{eisa} package and the tree can be visualized.

\section{Running the ISA}

We will use the acute lymphoblastic leukemia data set from the
\Rpackage{ALL} package.

<<load data,cache=FALSE>>=
library(ALL)
library(hgu95av2.db)
library(genefilter)
library(eisa)
library(GO.db)
data(ALL)
@

<<filter limits>>=
varLimit <- 0.5
kLimit <- 4
ALimit <- 5
@

First we filter the genes and keep only the ones that have an
expression level of at least \Sexpr{ALimit} in at least \Sexpr{kLimit}
samples; and that have a variance of at least \Sexpr{varLimit} across
the samples.

<<filtering the data>>=
flist <- filterfun(function(x) var(x)>varLimit, kOverA(kLimit,ALimit))
ALL.filt <- ALL[genefilter(ALL, flist), ]
@

We remove the T-cell samples as well, and use only the B-cell leukemia
samples.

<<keep B cell data only>>=
ALL.filt2 <- ALL.filt[, grepl("^B", ALL.filt$BT)]
@

Next, we run the ISA, with keeping the condition threshold fixed, but
varying the gene threshold.

<<run ISA>>=
set.seed(2)
modules <- ISA(ALL.filt2, flist=NA, thr.gene=seq(2,4,by=0.5), thr.cond=1)
@

\section{Generating the module tree}

The \Rfunction{ISASweep} function can be used to create a sweep tree
from a set of ISA modules. This essentially means finding the edges
between the modules and occasionally involves finding new modules as
well.

<<sweep>>=
modules2 <- ISASweep(ALL.filt2, modules)
@

\Rfunction{ISASweep} returns an \Rclass{ISAModules} object, but adds
some new seed data that defines the edges of the sweep graph, plus
possibly some new modules:
<<sweep result>>=
modules
modules2
colnames(seedData(modules))
colnames(seedData(modules2))
@

The \texttt{father} column contains pointers to the parent vertices in
the graph and the \texttt{level} column gives levels defined based on
the gene thresholds, higher levels indicate less stringent thresholds.

This data can be converted into a graph and then plotted using the
\Rpackage{igraph} package. The \Rfunction{ISASweepGraph} function
creates the graph.

<<getting the graph>>=
G <- ISASweepGraph(modules2)
@

$G$ has a lot of attributes:
<<G>>=
list.graph.attributes(G)
list.vertex.attributes(G)
@

The \texttt{layout} graph attribute contains a two-column matrix with
carefully calculated coordinates for plotting the graph. The
\texttt{width} and \texttt{height} graph attributes contain the
suggested size of the plot in inches.
<<G size>>=
G$width
G$height
@

The \texttt{id} vertex attribute contains the ids of the modules. 
The other vertex attributes essentially define the look of the graph,
when it is plotted.

See the documentation of the \Rpackage{igraph} package for more about
igraph graph objects.

\section{Plotting the sweep tree}

\subsection{Simple plots}

The \Rfunction{ISASweepGraphPlot} function can be used to plot the
graph returned by \Rfunction{ISASweepGraph}. We will use the suggested
size for the plot, and leave all plotting options at their default
values. The result is in Fig.~\ref{fig:moduletree}.

<<plotting,eval=FALSE>>=
X11(width=G$width, height=G$height)
@ 
<<plotting2,eval=FALSE>>=
ISASweepGraphPlot(G)
@

\begin{figure}
\begin{narrow}{-1cm}{-1cm}
\setkeys{Gin}{width=\linewidth}
<<plot,fig=TRUE,width=10,height=10,echo=FALSE>>=
ps.options(fonts=c("serif", "mono"))
<<plotting2>>
@ 
\caption{An ISA module tree. Each rectangle symbolizes a transcription
  module, modules in the same column were found at the same gene
  threshold. An arrow between two modules indicates that one is the
  ``generalization'' of the other, in the sense that ISA converges to
  the more general (=bigger) module at the milder threshold.}
\label{fig:moduletree}
\end{narrow}
\end{figure}

\subsection{Customized plots}

Let us show a little example on how to customize the module tree
plot. We will add some annotation to the module, based on Gene
Ontology enrichment calculations. Let us perform the hypergeometric
test for the enrichment calculation. The $p$-values are automatically
corrected using the Benjamini-Hochberg method.

<<enrichment>>=
GO <- ISAGO(modules2)
@

We will color the vertices according to their enrichment in the
Biological Process GO ontology. We also create some labels for them,
the abbreviated GO terms will be used for this.

<<customize>>=
p.bp <- sapply(summary(GO$BP), function(x) as.integer(-log10(x$Pvalue[1])))
c.bp <- d.bp <- sapply(sigCategories(GO$BP), function(x) x[1])
d.bp[!is.na(c.bp)] <- sapply(mget(na.omit(c.bp), GOTERM), Term)
d.bp <- abbreviate(d.bp, 6)
colbar <- hcl(h=260, c=35, l=seq(30, 100, length=20))
colbar <- c("#FFFFFF", rev(colbar))
col <- colbar[p.bp]
@

<<plotting3,eval=FALSE>>=
X11(width=G$width, height=G$height)
@

We are ready to plot the sweep tree now. We put the abbreviated GO
terms to the top left corner, and the size of the module (i.e. number
of genes and number of samples) in the top right corner. We also
increase the height of the vertices a bit.

<<plotting4,eval=FALSE>>=
ISASweepGraphPlot(G, vertex.color=col, vertex.size2=50, 
                  vertex.label.topleft=d.bp,
                  vertex.label.topright=paste(V(G)$noFeatures, sep=",", 
                    V(G)$noSamples))
@

Finally, we add a key for the abbreviated names at the top right
corner. The result is in Fig.~\ref{fig:moduletree2}.

<<key,eval=FALSE>>=
key <- na.omit(d.bp)[unique(names(na.omit(d.bp)))]
key2 <- paste(key, sep=": ", names(key))
legend("topright", key2)
@

\begin{figure}
\begin{narrow}{-1cm}{-1cm}
\setkeys{Gin}{width=\linewidth}
<<plot2,fig=TRUE,width=10,height=10,echo=FALSE>>=
<<plotting4>>
<<key>>
@ 
\caption{The same module tree, but with some annotation added. The
  modules are colored according to their Gene Ontology Biological
  Process enrichment $p$-values, darker colors correspond to more
  significant enrichment. The enriched GO terms and the sizes of the
  modules were also added.}
\label{fig:moduletree2}
\end{narrow}
\end{figure}

\section{More information}

For more information about the ISA, please see the references
below. The ISA homepage at
\url{http://www.unil.ch/cbg/homepage/software.html} has example data
sets, and all ISA related tutorials and papers.

\section{Session information}

The version number of R and packages loaded for generating this
vignette were:

<<sessioninfo,results=tex,echo=FALSE>>=
toLatex(sessionInfo())
@

\bibliographystyle{apalike}
\bibliography{EISA}
\nocite{*}

\end{document}

### 2009-10-30 eisa/inst/doc/EISA_tutorial.Rnw
Simplify Makefile, reorganize HTML generation from the vignettes.

@@
 
-\newcommand{\Rfunction}[1]{\texttt{#1}}
+\newcommand{\Rfunction}[1]{\texttt{#1()}}
 \newcommand{\Rpackage}[1]{\texttt{#1}}
@@
 \SweaveOpts{cache=TRUE}
-\SweaveOpts{prefix.string=graphics/plot}
+\SweaveOpts{prefix.string=plot}
+
+\setlength{\parindent}{2em}
 
@@
 options(width=60)
+options(continue=" ")
 @
@@
 The Iterative Signature Algorithm (ISA) is biclustering method. The
-input of a biclustering method a matrix and its output is a set of
-biclusters that fulfill some criteria; a bicluster is a block of the
+input of a biclustering method a matrix and the output is a set of
+biclusters that fulfill some criteria. A bicluster is a block of the
 potentially reordered input matrix.
@@
 to find gene sets that are coexpressed across a subset of the original
-samples. In the original ISA paper the biclusters are called
+samples. In the ISA papers the biclusters are called
 transription modules (TM), we will often refer them under this name in
@@
 
+This tutorial specifically deals with the modular analysis of gene
+expression data. Section~\ref{sec:isa} gives a short summary of how
+ISA works. If you need more information of the underlying math
+or want to apply it to other data, then please see the referenced
+papers, the vignette titled ``The Iterative Signature Algorithm'' in
+the \Rpackage{isa2} R package, or the ISA homepage at
+\url{http://www.unil.ch/cbg/homepage/software.html}.
+
 \section{Preparing the data}
@@
 implemented in the \Rpackage{eisa} and \Rpackage{isa2} packages, see
-Section{sec:isapackages} for a more elaborated summary about the two
-packages. It is enough to load the \Rpackage{eisa} package,
-\Rpackage{isa2} and other required packages are loaded automatically.
-\Rpackage{eisa} 
-<<load the packages>>=
+Section~\ref{sec:isapackages} for a more elaborated summary about the
+two packages. It is enough to load the \Rpackage{eisa} package,
+\Rpackage{isa2} and other required packages are loaded automatically:
+<<load the packages,cache=FALSE>>=
 library(eisa)
@@
 package. 
-<<load the data>>=
+<<load the data,cache=FALSE>>=
 library(ALL)
@@
 and then call the \Rfunction{isa} function on the
-\Rclass{ExpressionSet} object:
+\Rclass{ExpressionSet} object. The threshold parameters tune the size
+of the modules, less stringent (i.e. smaller) values result bigger,
+less correlated modules. The optimal values depend on your data and
+some experimentation is needed to determine them.
+
+Since running ISA might take a couple of minutes and the results 
+depend on the random number generator used, the ISA run is commented out 
+and we just load a precomputed set of modules in this vignette.
 <<simple ISA>>=
@@
 set.seed(1) # to get the same results, always
-modules <- ISA(ALL, thr.gene=thr.gene, thr.cond=thr.cond)
-@
-
-This first applies a non-specific filter to the data set filters the
-data set and then runs ISA from
-\Sexpr{formals(eisa:::ISA)$no.seeds} % $
+# modules <- ISA(ALL, thr.gene=thr.gene, thr.cond=thr.cond)
+data(ALLModulesSmall)
+modules <- ALLModulesSmall
+@
+
+This first applies a non-specific filter to the data set and then runs
+ISA from \Sexpr{formals(eisa:::ISA)$no.seeds} % $
 random seeds (the default). See Section~\ref{sec:detailed-isa} if the
@@
 ISA modules in the set, \Rfunction{dim} gives the dimension of the
-input expression matrix: the number of features and the number of
-samples.
+input expression matrix: the number of features (after the filtering)
+and the number of samples:
 <<accessors 1>>=
@@
 the names of the features and samples, just like the functions with
-the same name for an \Rclass{ExpressionSet}
+the same name for an \Rclass{ExpressionSet}:
 <<accessors 2>>=
@@
 \Rfunction{getNoSamples} returns a numeric vector, the number of
-samples in each module. Similarly, \Rfunction{pData} returns the
+samples in each module. \Rfunction{pData} returns the
 phenotype data of the expression set as a data frame. The
@@
 organism, \Rfunction{annotation} the name of the chip. For the former
-the appropriate annotation must be installed.
+the appropriate annotation package must be installed.
 <<number of features and samples>>=
@@
 
-\Rfunction{getFeatureNames} lists the probes (more precisely,
+\Rfunction{getFeatureNames} lists the probesets (more precisely,
 the feature names coming from the \Rclass{ExpressionSet} object) in
@@
 
-Remember, that the ISA biclustering is not binary, every feature (and
+ISA biclustering is not binary, every feature (and
 similarly, every sample) has a score between -1 and 1; the further the
@@
 e.g. the number of iterations required for the seed to converge. See
-the manual page of \Rfunction{isa} for details.
+the manual page of \Rfunction{ISA} for details.
 <<seed data>>=
@@
 The \Rpackage{eisa} package provides some functions to perform
-enrichment tests for the genes in the various modules against various
-databases. These tests are usually simplified and less sophisticated
-versions than the ones in the \Rpackage{Category}, \Rpackage{GOstats}
-or \Rpackage{topGO} packages, but they are much faster and this is
-important if we need to perform them for many modules.
+enrichment tests for the gene sets corresponding to the ISA modules
+against various databases. These tests are usually simplified and less
+sophisticated versions than the ones in the \Rpackage{Category},
+\Rpackage{GOstats} or \Rpackage{topGO} packages, but they are much
+faster and this is important if we need to perform them for many
+modules.
 
@@
 you have to do is to supply your \Rclass{ISAModules} object to the
-\Rfunction{ISA.GO} function.
+\Rfunction{ISAGO} function.
 <<GO enrichment>>=
-GO <- ISA.GO(modules)
-@
-
-The \Rfunction{ISA.GO} requires an annotation package that maps the
-Entrez identifiers of the studied organism to GO terms. For human, the
-\Rpackage{org.Hs.eg.db} package is required, for the mouse the
-\Rpackage{org.Mm.eg.db} package, etc.
+GO <- ISAGO(modules)
+@
+
+The \Rfunction{ISAGO} function requires the annotation package on the
+chip, e.g. for the ALL data, the \Rpackage{hgu95av2.db} package is
+required.
 
 The \variable{GO} object is a list with three elements, these
-correspond to the ontologies: biological function, cellular component
-and molecular function, in this order.
+correspond to the GO ontologies, they are: biological function,
+cellular component and molecular function, in this order.
 <<list GO result>>=
@@
 For extracting the results themselves, the \Rfunction{summary}
-function can be used, this converts them to a simple data frame. The
+function can be used, this converts them to a simple data frame. A
 $p$-value limit can be supplied to \Rfunction{summary}. Note, that
-since \Rfunction{ISA.GO} calculates enrichment for many gene sets
+since \Rfunction{ISAGO} calculates enrichment for many gene sets
 (i.e. for all biclusters), \Rfunction{summary} returns a list of data
@@
 <<GO summary>>=
-summary(GO[[1]], p=0.001)[[2]][,-6]
+summary(GO$BP, p=0.001)[[1]][,-6]
 @
@@
 \Rfunction{geneIdsByCategory} function. The following returns the
-genes in the second module and the third GO BP category. (The GO
+genes in the first module and the third GO BP category. (The GO
 categories are ordered according to the enrichment $p$-values, just
@@
 <<GO gene ids by category>>=
-geneIdsByCategory(GO[[1]])[[2]][[3]]
+geneIdsByCategory(GO$BP)[[1]][[3]]
 @
@@
 about the enriched GO categories.
-<<GO info>>=
-sigCategories(GO[[1]])[[2]]
+<<GO info,cache=FALSE>>=
+sigCategories(GO$BP)[[1]]
 library(GO.db)
-mget(sigCategories(GO[[1]])[[2]], GOTERM)
+mget(sigCategories(GO$BP)[[1]], GOTERM)
 @
@@
 In addition, the following functions are implemented to work on the
-objects returned by \Rfunction{ISA.GO}: \Rfunction{htmlReport},
+objects returned by \Rfunction{ISAGO}: \Rfunction{htmlReport},
 \Rfunction{pvalues}, \Rfunction{geneCounts}, \Rfunction{oddsRatios},
@@
 is, that since here we are testing a list of gene sets (=biclusters),
-they calculate the results for all gene sets and return a list.
+they calculate the results for all gene sets and usually return lists.
 
@@
 
-By default, the \Rfunction{ISA.GO} function performs multiple testing
+By default, the \Rfunction{ISAGO} function performs multiple testing
 correction using the Holm method, this can be changed via the
 \texttt{correction} and \texttt{correction.method} arguments. See the
-manual page of the \Rfunction{p.adjust} function for the possible
-multiple testing correction schemes.
+manual page of the \Rfunction{ISAGO} function for details, and also
+the \Rfunction{p.adjust} function for the possible multiple testing
+correction schemes.
 
@@
 
-Enrichment calculation against the KEGG pathway goes essentially the
-same way as for the Gene Ontology, this time we use the
-\Rfunction{ISA.KEGG} function:
-<<KEGG enrichment>>=
-KEGG <- ISA.KEGG(modules)
+Enrichment calculation against the KEGG pathway database goes
+essentially the same way as for the Gene Ontology, this time we use
+the \Rfunction{ISAKEGG} function:
+<<KEGG enrichment,cache=FALSE>>=
+KEGG <- ISAKEGG(modules)
 KEGG
-summary(KEGG)[[4]]
+summary(KEGG, p=1e-3)[[1]]
 library(KEGG.db)
-mget(sigCategories(KEGG)[[4]], KEGGPATHID2NAME)
+mget(sigCategories(KEGG)[[1]], KEGGPATHID2NAME)
 @
@@
 in a bicluster are associated with a chromosome. See the
-\Rfunction{ISA.CHR} function:
+\Rfunction{ISACHR} function:
 <<CHR enrichment>>=
-CHR <- ISA.CHR(modules)
+CHR <- ISACHR(modules)
 summary(CHR,p=0.05)[[2]][,-6]
@@
 $\mu$RNAs are short RNA molecules that regulate gene expression.
-TargetScan is a data based of predicted target genes of $\mu$RNAs, for
+TargetScan is a database of predicted target genes of $\mu$RNAs, for
 several organisms. There are two R packages that incorporate this
-database, one for human and another one for mouse, right now they can
+data, one for human and another one for mouse, right now they can
 be downloaded from
-\url{http://www2.unil.ch/cbg/index.php?title=Building_BioConductor_Annotation_Packages}.
+\url{http://www.unil.ch/cbg/homepage/software.html}.
 The \Rpackage{targetscan.Hs.eg.db} package is for human, the
@@
 The enrichment calculation itself is basically the same as for GO and
-KEGG, but the \Rfunction{ISA.miRNA} function should be used:
+KEGG, but the \Rfunction{ISAmiRNA} function should be used:
 <<miRNA enrichment>>=
 if (require(targetscan.Hs.eg.db)) {
-  miRNA <- ISA.miRNA(modules)
+  miRNA <- ISAmiRNA(modules)
   summary(miRNA, p=0.1)[[7]]
@@
 
-\subsection{The \Rpackage{biclust} package}
+\subsection{The \Rpackage{biclust} package}%
+\label{sec:biclust}
 
@@
 algorithms in a unified framework. It uses the class \Rclass{Biclust}
-to store a set of biclusters. The \Rfunction{ISA.biclust} function converts ISA
-modules to a \Rclass{Biclust} object. This requires the binarization
-of the modules, i.e. the ISA scores are lost, they are converted to
-zeros and ones:
-<<biclust,cache=TRUE>>=
+to store a set of biclusters. The standard R \Rfunction{as} function
+can be used to convert ISA modules to a \Rclass{Biclust} object. This
+requires the binarization of the modules, i.e. the ISA scores are
+lost, they are converted to zeros and ones:
+<<biclust,cache=FALSE>>=
 library(biclust)
-Bc <- ISA.biclust(modules)
+Bc <- as(modules, "Biclust")
 Bc
@@
 analyzed. Note that by default ISA normalizes the expression data
-before running the module detection; the raw, non-normalized values
-are plotted in the next example. If you want to plot the normalized
-values, then you need to do the normalization ``by hand'', before
-calling ISA, see Section~\ref{sec:detailed-isa}.
-<<heatmap,fig=TRUE>>=
-ISA2heatmap(modules, 1, ALL)
-@
+before running the module detection, but \Rfunction{ISA2heatmap} plots
+the raw values by default, optionally scaled. 
+If you want to plot the normalized values, then you need supply the
+\Rargument{norm} argument to \Rfunction{ISA2heatmap}, see the manual
+for details. The result is in Fig.~\ref{fig:heatmap}.
+<<heatmap,results=hide,include=FALSE,eval=FALSE>>=
+ISA2heatmap(modules, 1, ALL, norm="sample", scale="none")
+@
+
+\begin{figure}
+\centering
+<<heatmap-real,fig=TRUE,echo=FALSE,results=hide>>=
+<<heatmap>>
+@
+\caption{Heatmap of the first ISA module, plotted via the
+\Rfunction{ISA2heatmap} function. The expression values are
+sample-wised normalized.}
+\label{fig:heatmap}
+\end{figure}
 
@@
 if (require(Cairo)) {
-  CairoPNG("graphics/heatmap.png", 400, 400 )  
+  CairoPNG("plot-heatmap.png", 400, 400 )  
 } else {
-  png("graphics/heatmap.png", 400, 400 )
+  png("plot-heatmap.png", 400, 400 )
 }
@@
   \centering
-  \includegraphics[width=0.7\textwidth]{graphics/heatmap}
-  \caption{Heatmap of the first transcription module.}
+  \includegraphics[width=0.7\textwidth]{plot-heatmap.png}
+  \caption{Heatmap of the first transcription module, plotted with the
+  \Rfunction{drawHeatmap} function from the \Rpackage{biclust}
+  package.}
   \label{fig:drawHeatmap}
@@
 if (require(Cairo)) {
-  CairoPNG("graphics/parallel.png", 400, 400 )
+  CairoPNG("plot-parallel.png", 400, 400 )
 } else {
-  png("graphics/parallel.png", 400, 400 )
+  png("plot-parallel.png", 400, 400 )
 }
@@
 <<parallel>>=
-profile.plot(modules, 2, ALL, plot="both")
+profilePlot(modules, 2, ALL, plot="both")
 @
@@
 \centering
-\includegraphics[width=0.7\textwidth]{graphics/parallel}
+\includegraphics[width=0.7\textwidth]{plot-parallel.png}
 \caption{Profile plots for the second bicluster found by ISA.}
@@
 
-The \Rpackage{eisa} package provides some functions to plot parts of
+The \Rpackage{eisa} package provides functions to plot parts of
 the GO graph that is related to a transcription module. The
@@
 <<goplot1>>=
-goplot.2 <- gograph(summary(GO[[1]], p=0.05)[[1]])
-@
+goplot.2 <- gograph(summary(GO$BP, p=0.05)[[1]])
+@ 
+% $
 
 \Rfunction{goplot} uses the \Rpackage{igraph} package to create a
-graph with associated meta data:
+graph with the associated meta data:
 <<gograph object>>=
 summary(goplot.2)
-list.vertex.attributes(goplot.2)
 @ 
 The \texttt{width} and \texttt{height} graph attributes contain the
-suggested width and height of the graph, if plotted to a bitmap
+suggested width and height of the graph in pixels, if plotted to a bitmap
 device. (The graph attributes of an \Rpackage{igraph} graph can be
@@
 
-Let's plot the graph, we can do this with the \Rfunction{gograph.plot}
-function.
+Let's plot the graph, we can do this with the \Rfunction{gographPlot}
+function, see Fig.~\ref{fig:gotree}.
 <<goplot2,eval=FALSE>>=
 x11(width=10, height=10*goplot.2$height/goplot.2$width)
-gograph.plot(goplot.2)
+gographPlot(goplot.2)
 @
 \setkeys{Gin}{width=\textwidth}
-<<goplot2-real,echo=FALSE,width=10,height=2.5,fig=TRUE>>=
+<<goplot2-real,results=hide,include=FALSE,eval=FALSE,echo=FALSE>>=
 ps.options(fonts=c("serif", "mono"))
-gograph.plot(goplot.2)
-@
-
-Because the GO is not really a tree, \Rfunction{gograph} ``unfolds''
+gographPlot(goplot.2)
+@
+
+\begin{figure}
+\centering
+<<goplot2-real-real,width=10,height=5,fig=TRUE,results=hide,echo=FALSE,cache=FALSE>>=
+<<goplot2-real>>
+@
+\caption{Part of the Gene ontology database, ``Biological Function''
+ontology, that contains all GO terms enriched by a transcription module.}
+\label{fig:gotree}
+\end{figure}
+
+Because the GO is not a tree, \Rfunction{gograph} ``unfolds''
 it into a tree by including categories more than once, if needed. It
 also abbreviates the names of the GO categories to make them fit on
-the tree. The graph object contains the full name of the category as
+the plot. The graph object contains the full name of the category as
 well. The full and abbreviated names can be listed by querying the
@@
 @ 
-\Rfunction{gograph.plot} colors the categories according to the
+\Rfunction{gographPlot} colors the categories according to the
 supplied enrichment $p$-values, the minus $\log_{10}$ $p$-value is
@@
 
-The \Rfunction{cond.plot} function can plot the sample scores,
+The \Rfunction{condPlot} function can plot the sample scores,
 potentially including the scores before the ISA filtering. Let us plot
 the scores for the second module, the color code denotes B-cell
-vs. T-cell leukemia.
-
-<<condplot,fig=TRUE,width=10,height=4>>=
-col <- ifelse( grepl("^B", pData(modules)$BT), "orange", "darkolivegreen")
-cond.plot(modules, 2, ALL, col=col)
-@
+vs. T-cell leukemia (Fig.~\ref{fig:condplot}).
+
+<<condplot,results=hide,include=FALSE,eval=FALSE>>=
+col <- ifelse( grepl("^B", pData(modules)$BT), "darkolivegreen", "orange")
+condPlot(modules, 2, ALL, col=col)
+@
+% $
+
+\begin{figure}
+\centering
+<<condplot-real,fig=TRUE,width=10,height=4,echo=FALSE,results=hide>>=
+<<condplot>>
+@
+\caption{Condition plot for the ALL data and the second module. Every
+  sample is represented as a bar, and its ISA score is plotted. The
+  scores are also printed at the top or bottom of the bars. The scores
+  corresponding to the ISA thresholds are represented as horizontal
+  lines. The dashed line is the mean of the condition scores. B-cell
+  samples are dark green, T-cell samples are orange.
+}
+\label{fig:condplot}
+\end{figure}
 
@@
 
-The \Rfunction{autogen.table} function creates a summary
-
-\section{How ISA works}
-
-Before showing an actual ISA tool chain, a few words about how the
-algorithm works are in order.
+The \Rfunction{ISAHTMLTable} function creates a summary of a set of
+ISA modules, as an HTML table. The summary optionally includes the
+results of the enrichment calculations as well. Fig.~\ref{fig:HTML}
+shows a screenshot of the table created for the
+\Sexpr{length(modules)} modules we have found earlier.
+<<HTML1>>=
+htmldir <- tempdir()
+ISAHTMLTable(modules=modules, target.dir=htmldir, 
+             GO=GO, KEGG=KEGG, CHR=CHR)
+@ 
+<<HTML2,eval=FALSE>>=
+if (interactive()) {
+  browseURL(URLencode(paste("file://", htmldir, "/maintable.html", sep="")))
+}
+@
+
+\begin{figure}
+\centering
+\includegraphics[width=\textwidth]{plot-HTMLTable.png}
+\caption{A HTML summary table showing information for the first couple
+of modules found in the ALL data set. The ``Thr.'' column shows the
+ISA thresholds that were used to find the module; the next two columns
+are the number of features (genes) and the number of samples
+(conditions) in the modules. The other columns show the most
+significantly enriched GO categories, KEGG pathways and
+chromosomes. The small grey numbers are the enrichment $p$-value, the
+number of genes expected to be hit by chance, the number of hits and
+the size of the category (within the respective gene universe), in
+this order.}
+\label{fig:HTML}
+\end{figure}
+
+The Rfunction{ISAHTMLModules} generates a HTML page for every single
+module with the following information on it: 
+\begin{itemize}
+\item An expression plot of the genes and samples in the module,
+  including the ISA scores. This is done by calling \Rfunction{expPlot}.
+\item Overlap plot of all modules, see \Rfunction{overlapPlot}.
+\item  Gene Ontology tree plots for the enriched GO terms,
+  separately for the three ontologies., These are produced by
+  calling \Rfunction{gograph}.
+\item Tables for the enriched Gene Ontology terms, separately for
+  the three ontologies.
+\item A table for the enriched KEGG pathways.
+\item A table for the enriched miRNA families.
+\item The list of genes in the module.
+\item The list of samples in the module.
+\item A condition plot, see the \Rfunction{condPlot} function.
+\end{itemize}
+<<HTML modules>>=
+ISAHTMLModules(eset=ALL, modules=modules, target.dir=htmldir,
+               GO=GO, KEGG=KEGG, CHR=CHR)
+@ 
+The rows of the summary table generated by \Rfunction{ISAHTMLTable}
+link to the module pages generated by \Rfunction{ISAHTMLModules}.
+
+Both the summary and the module pages can be generated with the
+\Rfunction{ISAHTML} function.
+
+\subsection{The ExpressionView package}
+
+The previously mentioned visualization techniques focus on a single
+module only. The \Rpackage{ExpressionView} package visualizes a set of
+possibly overlapping biclusters together.
+
+The \Rpackage{ExpressionView} package~\cite{ev} tries to reorder the
+rows and columns of the expression matrix in such a way that the
+genes/samples that appear together in a module, are placed right
+beside/above each other. The package directly supports the
+\Rclass{ISAModules} and \Rclass{ExpressionSet} classes, see its
+documentation for details.
+
+\section{How ISA works}%
+\label{sec:isa}
 
@@
 two or more seeds might converge to the same module. The function
-\Rfunction{ISA.unique} eliminates every module from the result of 
-\Rfunction{ISA.iterate} that is very similar (in terms of
+\Rfunction{ISAUnique} eliminates every module from the result of 
+\Rfunction{ISAIterate} that is very similar (in terms of
 Pearson correlation) to the one that was already found before.
@@
 
-By default (i.e. if the \Rfunction{isa} function is used) the ISA is
+By default (i.e. if the \Rfunction{ISA} function is used) the ISA is
 performed from random sparse starting seeds, generated by the
@@
 genes/samples, then you can feed in your seeds into the
-\Rfunction{ISA.iterate} function directly. In this case the
+\Rfunction{ISAIterate} function directly. In this case the
 algorithm is deterministic, for the same seed you will always get the
 same results. Using smart (i.e. non-random) seeds can be considered as
-a semi-supervised approach.
+a semi-supervised approach. We show an example of using smart seeds in
+Section~\ref{sec:detailed-isa}.
 
@@
 Using in silico data we observed that ISA has the best performance if the
-input matrix is normalized (see \Rfunction{ISA.normalize}). The
+input matrix is normalized (see \Rfunction{ISANormalize}). The
 normalization produces two matrices: $E_r$ and
@@
 
-It is possible to use another normalization. In this case the user is
-requested to supply the normalized input data in a named list,
-including the two matrices of appropriate
-dimensions to the \Rfunction{ISA.iterate} function.
-
-The \texttt{Er} entry of the list will be used for calculating the
-signature of the genes, \texttt{Ec} will be used for the signature of the
-samples. If you want to use the same matrix in both steps, then supply
-it twice, the first one transposed.
+It is possible to use another normalization, or not to use
+normalization at all; the user has to construct an
+\Rclass{ISAExpressionSet} object containing the three matrices
+corresponding to the raw data, the gene-wise normalized data and the
+sample-wise normalized data. This object can be passed to the
+\Rfunction{ISAIterate} function. The matrices are not required to be
+different, the user can supply the raw data matrix three times, if
+desired.
 
@@
 gene/samples is from zero, the stronger the association between the
-gene/sample and the module. If the sign of two genes/samples are the
+gene/sample and the module. If the signs of two genes/samples are the
 same, then they are correlated, if they have opposite signs, then they
@@
 after converting the output of ISA to a \Rclass{biclust} object. 
+We use the \texttt{Bc} object that was created in Section~\ref{sec:biclust}. 
 Here are the measures for the first bicluster:
@@
 the ISA iteration, then ISA would be equivalent to a numerical method
-for singular value decomposition and robustness would be indeed the
+for singular value decomposition and robustness would be the
 same the principal singular value of the input matrix.
@@
 If the \Rfunction{isa} function was used to find the transcription
-modules, then the robustness measure is used to filter the results.
+modules, then the robustness measure is used automatically to filter the results.
 This is done by first scrambling the input matrix and then running ISA
@@
 ISA and its companion functions for visualization, functional
-enrichment calculation, etc. are distributed in two separate packages,
+enrichment calculation, etc. are distributed in two separate R packages,
 \Rpackage{isa2} and \Rpackage{eisa}.
@@
 The \Rfunction{ISA} function takes care of all steps performed during
-a modular study, and for each step it uses parameters work reasonably
+a modular study, and for each step it uses parameters, that work reasonably
 well. In some cases, however, one wants to access these steps
@@
 that don't map to any Entrez genes.
-<<Entrez>>=
+<<Entrez,cache=FALSE>>=
 ann <- annotation(ALL.filt)
@@
 To reduce ambiguity in the interpretation of the results, we might
-also want to keep only single probeset for each Entrez genes.
+also want to keep only single probeset for each Entrez gene.
 <<Entrez unique>>=
@@
 matrices that are normalized differently. The
-\Rfunction{ISA.normalize} function can be used to calculate the
-normalized expression matrices; it returns an \Rclass{ExpressionSet}
-object including three expression matrices: the original raw
-expression, the row-wise normalized and the column-wise normalized
-expression matrix.
+\Rfunction{ISANormalize} function can be used to calculate the
+normalized expression matrices; it returns an \Rclass{ISAExpressionSet}
+object. This is a subclass of \Rclass{ExpressionSet}, and contains
+three expression matrices: the original raw expression, the row-wise
+(=gene-wise) normalized and the column-wise (=sample-wise) normalized
+expression matrix. The normalized expression matrices can be queried
+with the \Rfunction{featExprs} and \Rfunction{sampExprs} functions.
 <<normed>>=
-ALL.normed <- ISA.normalize(ALL.filt.3)
+ALL.normed <- ISANormalize(ALL.filt.3)
 ls(assayData(ALL.normed))
+dim(featExprs(ALL.normed))
+dim(sampExprs(ALL.normed))
 @
@@
 
-The \Rfunction{generate.seeds} generates a set of random seeds (i.e. a
-set of random gene sets). See its documentation if you need to change
-the sparsity of the seeds.
+The \Rfunction{generate.seeds} function generates a set of random
+seeds (i.e. a set of random gene sets). See its documentation if you
+need to change the sparsity of the seeds.
 
 <<seeds>>=
+set.seed(3)
 random.seeds <- generate.seeds(length=nrow(ALL.normed), count=100)
@@
 <<iteration>>=
-modules1 <- ISA.iterate(ALL.normed, row.seeds=random.seeds, 
-                        thr.row=2, thr.col=2)
-modules2 <- ISA.iterate(ALL.normed, col.seeds=smart.seeds,
-                        thr.row=2, thr.col=2)
+modules1 <- ISAIterate(ALL.normed, feature.seeds=random.seeds, 
+                        thr.feat=1.5, thr.samp=1.8, convergence="cor")
+modules2 <- ISAIterate(ALL.normed, sample.seeds=smart.seeds,
+                        thr.feat=1.5, thr.samp=1.8, convergence="cor")
 @
@@
 
-\Rfunction{ISA.iterate} returns the same number of ``modules'' as the
-number of input seeds; these, however, do not always correspond to
-meaningful modules, the input seeds can converge to an all-zero
+\Rfunction{ISAIterate} returns the same number of modules as the
+number of input seeds; these are, however, not always meaningful,
+the input seeds can converge to an all-zero
 vector, or occasionally they may not converge at all. It is also
@@
 
-The \Rfunction{ISA.unique} function eliminates the all-zero or
+The \Rfunction{ISAUnique} function eliminates the all-zero or
 non-convergent input seeds and keeps only one instance of the
@@
 <<unique>>=
-modules1.unique <- ISA.unique(ALL.normed, modules1)
-modules2.unique <- ISA.unique(ALL.normed, modules2)
+modules1.unique <- ISAUnique(ALL.normed, modules1)
+modules2.unique <- ISAUnique(ALL.normed, modules2)
 @
@@
 
-The \Rfunction{ISA.filter.robust} function filters a set of modules by
-running ISA with the same parameters on a scrambled data set and then
+The \Rfunction{ISAFilterRobust} function filters a set of modules by
+running ISA with the same parameters on the scrambled data set and then
 calculating a robustness score, both for the real modules and the ones
@@
 <<robust>>=
-modules1.robust <- ISA.filter.robust(ALL.normed, modules1.unique)
-modules2.robust <- ISA.filter.robust(ALL.normed, modules2.unique)
+modules1.robust <- ISAFilterRobust(ALL.normed, modules1.unique)
+modules2.robust <- ISAFilterRobust(ALL.normed, modules2.unique)
 @
@@
 classifiers for the different types of ALL. For this we use the sample
-scores of the modules, the score of a sample in a given module is the
-(weighted) average of the expression of the genes of the module in
-the sample in question.
+scores of the modules. The score of a sample is the weighted average
+of the normalized expression level of the modules genes. Thus, a good
+classifier module should have high sample scores for one type of ALL
+and low for the other. Here we perform a $t$-test to quantify this
+difference.
 
 Let's first check, whether any of the modules can distinguish between
-T-cell and B-cell ALL samples.
+T-cell and B-cell ALL samples. We perform $t$-tests for both sets of
+modules.
 <<B vs T>>=
@@
 
-For the first set of samples \Sexpr{length(sign1)} modules show
+For the first (random) set of seeds \Sexpr{length(sign1)} modules show
 significant difference for T-cell and B-cell samples, for the second
@@
 Let's make some condition plots for the best separating modules from
-each set.
-<<conditions-plots-dysreg,fig=TRUE,width=10,height=8>>=
+each set (Fig.~\ref{fig:dysregmodules}).
+<<conditions-plots-dysreg,results=hide,include=FALSE,eval=FALSE>>=
 color <- ifelse(grepl("T", type), "orange", "darkolivegreen")
 layout(cbind(1:2))
-cond.plot(modules1.robust, which.min(tt1$p.value), ALL.normed, col=color, 
+condPlot(modules1.robust, which.min(tt1$p.value), ALL.normed, col=color, 
           main="Best separator, random seeds")
-cond.plot(modules2.robust, which.min(tt2$p.value), ALL.normed, col=color,
+condPlot(modules2.robust, which.min(tt2$p.value), ALL.normed, col=color,
           main="Best separator, smart seeds")
 @
+
+\begin{figure}
+<<conditions-plots-dysreg-real,fig=TRUE,width=10,height=8,echo=FALSE,results=hide>>=
+<<conditions-plots-dysreg>>
+@
+\caption{Condition plots for the best separator module found with
+  random seeds and the one found with the smart seeds. They are
+  essentially the same, with some minor differences. 
+}
+\label{fig:dysregmodules}
+\end{figure}
 
@@
 
+As it turns out, one modules found using the smart seeds was also
+found with random seeds.
+
 \subsection{Enrichment calculations}
@@
 <<dysreg enrichment>>=
-GO.dysreg1 <- ISA.GO(modules1.TB)
-GO.dysreg2 <- ISA.GO(modules2.TB)
-KEGG.dysreg1 <- ISA.KEGG(modules1.TB)
-KEGG.dysreg2 <- ISA.KEGG(modules2.TB)
+GO.dysreg1 <- ISAGO(modules1.TB)
+GO.dysreg2 <- ISAGO(modules2.TB)
+KEGG.dysreg1 <- ISAKEGG(modules1.TB)
+KEGG.dysreg2 <- ISAKEGG(modules2.TB)
 @
@@
                         unique(unlist(sigCategories(x))))))
-library(GO.db)
-library(KEGG.db)
 sapply(mget(gocats, GOTERM), Term)
@@
 Both module sets seem to have some good separators, let us make some
-condition plots.
-<<B1-cond-plots,fig=TRUE,width=10,height=8>>=
+condition plots (Fig.~\ref{fig:B1condplots}).
+<<B1-cond-plots,results=hide,include=FALSE,eval=FALSE>>=
 color <- ifelse(type == "B1", "green", 
@@
 layout(cbind(1:2))
-cond.plot(modules1.robust, which.min(tt.B1.1$p.value), ALL.normed, col=color, 
+condPlot(modules1.robust, which.min(tt.B1.1$p.value), ALL.normed, col=color, 
           main="Best B1 separator, random seeds")
-cond.plot(modules2.robust, which.min(tt.B1.2$p.value), ALL.normed, col=color,
+condPlot(modules2.robust, which.min(tt.B1.2$p.value), ALL.normed, col=color,
           main="Best B1 separator, smart seeds")
 @
+
+\begin{figure}
+\centering
+<<B1-cond-plots-real,fig=TRUE,echo=FALSE,results=hide,width=10,height=8>>=
+<<B1-cond-plots>>
+@ 
+\caption{Condition plots for the modules that best separate the B1-type 
+  ALL samples from the other B-cell ALL samples. The two modules are essentially
+  the same.}
+\label{fig:B1condplots}
+\end{figure}
 
@@
 
+\section{More information}
+
+For more information about the ISA, please see the references
+below. The ISA homepage at
+\url{http://www.unil.ch/cbg/homepage/software.html} has example data
+sets, and all ISA related tutorials and papers.
+
 \section{Session information}

### 2009-10-30 isa2/inst/doc/ISA_tutorial.Rnw
Simplify Makefile, reorganize HTML generation from the vignettes.

--- 2009-05-29 isa2/inst/doc/ISA_tutorial.Rnw
@@
 \usepackage{ragged2e}
+\usepackage{url}
 
-\newcommand{\Rfunction}[1]{\texttt{#1}}
+\newcommand{\Rfunction}[1]{\texttt{#1()}}
 \newcommand{\Rpackage}[1]{\texttt{#1}}
@@
 \newcommand{\filename}[1]{\texttt{#1}}
-\newcommand{\listentry}[1]{\texttt{#1}}
+\newcommand{\listentry}[1]{`\texttt{#1}'}
+
+\SweaveOpts{prefix.string=plot}
 
@@
 options(width=60)
+options(continue=" ")
+if (require(Cairo)) { png <- CairoPNG }
 @
@@
 \texttt{rows} and \texttt{columns} (i.e. the first columns define the
-first bicluster, etc.) and the elements the biclusters are the
+first bicluster, etc.) and the elements of the biclusters are the
 non-zero elements in the columns of \texttt{rows} and
@@
 
-The Iterative Signature Algorithm (ISA) is biclustering method. Its 
-input is a matrix and its output is a set of biclusters: blocks of the
-potentially reordered input matrix, that fulfill some predefined
-criteria. A biclustering algorithm tries to find
-blocks that are different from the rest of the matrix, e.g. the values
-covered by the bicluster are all above or below the background.
+The Iterative Signature Algorithm (ISA)~\cite{sa,isa,isamod} is
+biclustering method. Its input is a matrix and its output is a set of
+biclusters: blocks of the potentially reordered input matrix, that
+fulfill some predefined criteria. A biclustering algorithm typically
+tries to find blocks that are different from the rest of the matrix,
+e.g. the values covered by the bicluster are all above or below the
+background.
 
@@
 example of a (rather artificial) data matrix with very strong modular
-structure. It is a $20\times 20$ matrix and (after reordering its rows
-and columns) it has two correlated blocks,
-each of them of size $10\times 10$.
+structure. It is a $20\times 20$ matrix and---after reordering its rows
+and columns---it has two correlated blocks,
+each of size $10\times 10$.
 \begin{figure}
@@
 
-ISA works in an iterative way. For an $E (m\times n)$
+ISA works in an iterative way. For an $E$ $(m\times n)$
 input matrix it starts from a seed vector $r_0$, which is
 typically a sparse 0/1 vector of length $m$. The non-zero elements in
-the seed vector define a set of rows in $E$. Then the transposed of
+the seed vector define a set of rows in $E$. First, the transposed of
 $E$, $E'$ is multiplied by $r_0$ and the result is thresholded.
@@
 this means keeping values that are significantly higher than the
-mean (``\emph{up}''), or keeping the ones that are significantly lower
-than the mean (``\emph{down}''); or keeping both
-(``\emph{updown}'').
+mean (direction=``\emph{up}''), or keeping the ones that are
+significantly lower than the mean (direction=``\emph{down}''); or
+keeping both (direction=``\emph{updown}'').
 
@@
 
+The \Rfunction{isa.iterate} function performs the ISA iteration from a
+given set of seed vectors.
+
 It is very possible that the ISA finds the same module more than once;
-two or more seeds might converge to the same module. The function
+two or more seed vectors might converge to the same module. The function
 \Rfunction{isa.unique} eliminates every module from the result of 
 \Rfunction{isa.iterate} that is very similar (in terms of
-Pearson correlation) to the one that was already found before.
+Pearson correlation) to one that was already found from a different
+seed.
 
-It might be also apparent from the description of ISA, that the
+The \Rfunction{isa} function performs the whole ISA workflow, this
+includes running \Rfunction{isa.iterate} and \Rfunction{isa.unique}.
+
+It might be also apparent, that the ISA
 biclusters are soft, i.e. they might have an overlap in their rows,
 columns, or both. It is also possible that some rows and/or
-columns of the input matrix are not found to be part of any ISA
+columns of the input matrix are not found to be part of any
 biclusters. Depending on the stringency parameters in the
@@
 normalization produces two matrices: $E_r$ and
-$E_c$. $E_r$ is calculated by transposing $E$ and
+$E_c$. $E_r$ is calculated by transposing $E$, then
 centering and scaling its rows (see the \Rfunction{scale} R function). $E_c$ is
 calculated by centering and scaling the rows of $E$. $E_r$ is
-used to calculate the column signature of rows and $E_c$ is used
-to calculate the signature of the columns.
+used to calculate the (column) signature of rows and $E_c$ is used
+to calculate the (row) signature of the columns.
 
@@
 row/columns is from zero, the stronger the association between the
-row/column and the module. If the sign of two rows/columns are the
+row/column and the module (i.e. the other rows/columns of the
+module). If the sign of two rows/columns are the 
 same, then they are correlated, if they have opposite signs, then they
@@
 As ISA is an unsupervised algorithm, it may very well find some
-modules, even if you feed in noise as an input matrix. To avoid these
+modules, even if you feed in noise as the input matrix. To avoid these
 spurious modules we defined a robustness measure, a single number for
@@
 
-Let us see an example. We load the \Rpackage{isa2} package, and
-generate some in 
+Let us see an example. We load the \Rpackage{isa2}
+package~\cite{isapackage}, and generate some in 
 silico data with the \Rfunction{isa.in.silico} function first:
 <<insilico,cache=TRUE>>=
+set.seed(10)
 library(isa2)
@@
 The function call above generates a data matrix similar to the one in
-Fig.~\ref{fig:twoclusters}, but a bigger one. The first two arguments define the size of
+Fig.~\ref{fig:twoclusters}, but bigger. The first two arguments define the size of
 the data matrix ($200\times 200$), the third the number of modules (two)
@@
 <<blockplot.pre,echo=FALSE,results=hide>>=
-png("blockplot.png", 800, 600)
+png("blockplot.png", 600, 400)
 @ 
@@
 \centering
-\includegraphics[width=0.6\textwidth]{blockplot}
+\includegraphics[width=0.6\textwidth]{blockplot.png}
 \caption{Another artificial data matrix with two modules and a lot of
@@
 ISA has found \Sexpr{ncol(modules$rows)} %$
-modules in the artificial data. This seem a bit surprising, as we
-expected only two modules, so let us take a look at the modules. We
+modules in the artificial data. This seems a bit surprising, as we
+expected only two, so let us take a closer look. We
 simply plot the row scores of the individual modules
@@
 The \listentry{rundata} entry of the return value of \Rfunction{isa}
-contains information about the ISA run, all the parameters (the default
+contains information about the ISA run, all the parameters (with the default
 values in our case) are included here. It has the following members:
@@
 
-In this section we will perform each steps of the ISA analysis
+In this section we will perform each step of the ISA analysis
 individually. This makes sense only if the user wants to adjust the
@@
 \centering
-\includegraphics{tenmods}
+\includegraphics{tenmods.png}
 \caption{In-silico data with ten non-overlapping modules.}
@@
 
-To run the ISA, we need to create the starting seeds first. The
-\Rfunction{isa} function uses sparse random seeds, generated by
+To run the ISA, we need to create some starting seeds. The
+\Rfunction{isa} function uses sparse random seeds, produced by
 \Rfunction{generate.seeds}. We also have the choice of using
@@
 Note, that vector giving the sparsity is recycled, thus we will have
-four kind of seeds, some with 2, 5, 10 and 100 rows; 250 of each of
-them.
+four kind of seeds, with 2, 5, 10 and 100 rows; 250 of each kind.
 
@@
 
-It is also possibly that some seeds converge to an all-zero vector, we
-have \Sexpr{sum(apply(modules$rows==0, 2, all))} such module: % $
+It is also possible that some seeds converge to an all-zero vector, we
+have \Sexpr{sum(apply(modules$rows==0, 2, all))} such modules: % $
 <<all.zero.modules>>=
@@
 single one of them. It also eliminates all-zero modules and seeds
-that did not converge, \verb+NA+
+that did not converge; \verb+NA+
 columns in the \listentry{rows} and \listentry{columns} matrix
@@
 @ 
-As the user can see, the robustness of the \Sexpr{ncol(modules2$rows)} %$
+The robustness of the \Sexpr{ncol(modules2$rows)} %$
 modules varies considerably, from \Sexpr{round(min(rob),2)} to
@@
 
-One can use the robustness measure to filter out list of modules, in
+One can use the robustness measure to filter a list of modules, in
 the following way. We permute the input matrix and run the same
@@
 \centering
-\includegraphics[width=0.6\textwidth]{othermods}
+\includegraphics[width=0.6\textwidth]{othermods.png}
 \caption{The input matrix and five modules with relatively low
@@
 
+It turns out that these are double-modules, each is a union of two
+real modules. ISA finds these, since they also have correlated rows
+and columns, although the correlation is lower than for the
+single-modules. The double-modules are typically found at lower
+thresholds.
+
 \subsection{Visualize the results}
@@
 at a time. For some of these we will use the \Rpackage{biclust} R
-package.
+package~\cite{biclust}.
 
@@
 algorithms in a unified framework. It uses the class \Rclass{Biclust}
-to store a set of biclusters. The \Rfunction{isa.biclust} converts ISA
-modules to a \Rclass{Biclust} object. This requires the binarization
-of the modules, i.e. the ISA scores are lost, they are converted to
-zeros and ones:
+to store a set of biclusters. The \Rfunction{isa.biclust} function
+converts ISA modules to a \Rclass{Biclust} object. This requires the
+binarization of the modules, i.e. the ISA scores are lost, they are
+converted to zeros and ones:
 <<biclust,cache=TRUE>>=
@@
 functions with the same output format: \Rfunction{isa.iterate},
-\Rfunction{isa.unique}, etc. (In fact \Rfunction{plotModules} calls the
+\Rfunction{isa.unique}, etc.) In fact \Rfunction{plotModules} calls the
 \Rfunction{images} function to do its job. \Rfunction{images} takes a
-list of matrices, see Fig.~\ref{fig:noise} for an example.)
+list of matrices, see Fig.~\ref{fig:noise} for an example.
 
@@
 \centering
-\includegraphics[width=0.5\textwidth]{drawheatmap}
+\includegraphics[width=0.5\textwidth]{drawheatmap.png}
 \caption{Heatmap plot of the reordered input matrix. The rows and
@@
 <<noiseplot.pre,echo=FALSE,results=hide>>=
-png("noise.png", width=800, height=600)
+png("noise.png", width=600, height=400)
 @ 
@@
   \centering
-  \includegraphics[width=\textwidth]{noise}
+  \includegraphics[width=\textwidth]{noise.png}
   \caption{Modular data sets with different levels of noise. The data
@@
   Pearson correlation) ISA module for each block in the input matrix.
+  Each boxplot is contains ten points, one for each module.
   ISA performs very well, even in the presence of relatively high
@@
 blocks.
-
-<<set.seed.overlap>>=
+<<overlap>>=
 set.seed(1)
-@ 
-<<overlap>>=
 two.over <- isa.in.silico(100,100,2,40,40,overlap.row=10)
@@
 ov.seeds <- generate.seeds(count=100, length=100)
-ov.modules.1 <- isa.iterate(ov.normed, ov.seeds, 
+ov.modules.1 <- isa.iterate(ov.normed, ov.seeds, convergence="cor",
                             thr.row=0, thr.col=0, direction="up")
@@
 ov.modules.1 <- isa.filter.robust(two.over[[1]], ov.normed, ov.modules.1)
-ov.modules.2 <- isa.iterate(ov.normed, ov.seeds, 
+ov.modules.2 <- isa.iterate(ov.normed, ov.seeds, convergence="cor",
                             thr.row=1, thr.col=1, direction="up")
@@
 <<overlapplot.pre,echo=FALSE,results=hide>>=
-png("overlapplot.png", 800, 600)
+png("overlapplot.png", 600, 400)
 @ 
 <<overlapplot>>=
+no.modules <- ncol(ov.modules.1$rows) + ncol(ov.modules.2$rows)
 plotModules(data=two.over[[1]],
@@
                    ov.modules.2$column)),
-            names=c("Input matrix", paste("Module", 1:6)))
+            names=c("Input matrix", paste("Module", seq_len(no.modules))))
 @
@@
 \centering
-\includegraphics{overlapplot}
+\includegraphics{overlapplot.png}
 \caption{The input matrix with two overlapping blocks and the ISA
@@
 
+\section{More information}
+
+For more information about the ISA please see the references at the
+end of this paper. The ISA homepage is at
+\url{http://www.unil.ch/cbg/homepage/software.html}.
+
+For analyzing gene expression data with ISA, we suggest using
+BioConductor~\cite{bioc} and the \Rpackage{eisa}~\cite{eisapackage} R
+package.
+
+If you want to run ISA on a computer cluster or a multi-processor
+machine, then see the vignette titled ``Running ISA in parallel with
+the \Rpackage{snow} pacakge'', in the \Rpackage{isa2} package.
+
 \section{Session information}
@@
 
+\bibitem[Gentleman, 2004]{bioc}
+R. Gentleman, V. J. Carey, D. M. Bates, B.Bolstad, M.
+Dettling, S. Dudoit, B. Ellis, L. Gautier, Y. Ge, and others (2004).
+\newblock Bioconductor: Open software development for computational biology and
+  bioinformatics.
+\newblock {\em Genome Biology}, Vol. 5, R80
+
 \bibitem[Cs\'ardi, 2009]{isapackage}
@@
 \newblock {\em isa2: The Iterative Signature Algorithm}.
+\newblock R package version 0.1.
+
+\bibitem[Cs\'ardi, 2009]{eisapackage}
+Cs\'ardi, G. (Sep 15, 2009).
+\newblock {\em eisa: The Iterative Signature Algorithm for Gene
+  Expression Data}.
 \newblock R package version 0.1.

### 2009-11-16 expressionviewnote/expressionview.tex
expressionviewnote: Updated figure to represent new layout.

@@
 \newcommand{\cred}{\color{red}}
-%\usepackage[pdftex,colorlinks,bookmarks=false,citecolor=blue,linkcolor=red,urlcolor=blue]{hyperref}
+
+\usepackage[pdftex,colorlinks,bookmarks=false,citecolor=blue,linkcolor=red,urlcolor=blue]{hyperref}
+\newcommand{\thefirstpage}{1} \newcommand{\thelastpage}{1}
 
@@
 \section{Availability:} 
-\href{http://www.google.com}{R package with open source license}
+\href{http://www2.unil.ch/cbg/index.php?title=ExpressionView}{R package with open source license}
 
@@
 \section{Supplementary information:}
-Screenshots, tutorials and sample data sets can be found on the ExpressionView \href{http://maya:7575/ExpressionView}{website}.
+Screenshots, tutorials and sample data sets can be found on the ExpressionView \href{http://www2.unil.ch/cbg/index.php?title=ExpressionView}{website}.
 
@@
 \begin{figure*}[!tpb]
-\centerline{\includegraphics[width=\linewidth]{fig1-crop}}
+\centerline{\includegraphics[width=0.9\linewidth]{fig1-crop}}
 \caption{How to use ExpressionView: Starting from gene expression data in the form of a Bioconductor ExpressionSet, the user first runs a biclustering algorithm to determine co-regulated groups of genes and samples.
@@
 \end{figure*}
-For users who prefer to install the viewer as a standalone program, we also provide a platform independent Adobe AIR version that can be downloaded from the \href{http://maya:7575/ExpressionView}{website}.
+For users who prefer to install the viewer as a standalone program, we also provide a platform independent Adobe AIR version that can be downloaded from the \href{http://www2.unil.ch/cbg/index.php?title=ExpressionView}{website}.
 
@@
 The actual gene expression data is rounded to two digit precision and encoded in Base64.
-The structure of the XML file is described in more detail on the ExpressionView \href{http://maya:7575/ExpressionView}{website}.
+The structure of the XML file is described in more detail on the ExpressionView \href{http://www2.unil.ch/cbg/index.php?title=ExpressionView}{website}.
 In most cases, these files are of the order of a few Megabytes and can thus easily be sent to co-workers.

### 2009-11-16 RExpressionView/R/export.R
RExpressionView: Updated documentation.

@@ -10,7 +10,7 @@ ExportEV.Biclust <- function(biclusters, eset, order, filename=NULL, norm=c("sam
 	eisamodules <- as(biclusters, "ISAModules")
 	eisamodules@rundata$annotation <- annotation(eset)
 	eisamodules@rundata$prenormalize <- FALSE
-	export.ev(eisamodules, eset, order, filename, norm)
+	ExportEV(eisamodules, eset, order, filename, norm)
 }
 
 
@@ -55,10 +55,10 @@ ExportEV.ISAModules <- function(biclusters, eset, order=NULL, filename=NULL, nor
 	writeLines("\t</summary>", con)
 
 	go.table <- toTable(GOTERM)
-	gos <- ISA.GO(biclusters)
+	gos <- ISAGO(biclusters)
 
 	kegg.table <- toTable(KEGGPATHID2NAME)
-	keggs <- ISA.KEGG(biclusters)
+	keggs <- ISAKEGG(biclusters)
 		
 	# get gene info
 	ann <- annotation(eset)
@@ -134,7 +134,6 @@ ExportEV.ISAModules <- function(biclusters, eset, order=NULL, filename=NULL, nor
 		for ( sample in 1:nSamples ) {
 			
 			writeLines("\t\t<sample>", con)
-					
 				writeLines(paste("\t\t\t<id>", sample, "</id>", sep=""), con)
 				writeLines("\t\t\t<score/>", con)
 				writeLines(paste("\t\t\t<name>", Samples[sample], "</name>", sep=""), con)
@@ -363,14 +362,19 @@ ExportEV.list <- function(biclusters, data, order=NULL, filename=NULL, descripti
 	
 	cnamesdata <- colnames(data)
 	rnamesdata <- rownames(data)
-	Genes <- cnamesdata[geneMaps[[1]]];
-	Samples <- rnamesdata[sampleMaps[[1]]];
+	Genes <- cnamesdata[geneMaps[[1]]]
+	Samples <- rnamesdata[sampleMaps[[1]]]
 	
 	geneLabels <- description$collabels
-	geneLabelsData <- as.matrix(description$coldata[Genes, rownames(geneLabels)])
-
+	geneLabelsData <- NULL
+	if ( !is.null(description$coldata) ) {
+		geneLabelsData <- as.matrix(description$coldata[Genes, rownames(geneLabels)])
+	}
 	sampleLabels <- description$rowlabels
[…]

### 2009-11-16 RExpressionView/R/order.R

@@ -40,10 +40,16 @@ if (require(isa2)) {
 }
 
 OrderEV.list <- function(biclusters, initialorder=NULL, debuglevel=0, maxtime=60) {	
+	if( !is.null(initialorder) && sum(optimalorder$status[[1]], optimalorder$status[[2]]) == 
+		length(optimalorder$status[[1]])+length(optimalorder$status[[2]]) ) {
+		cat("initial order is already fully ordered.\n")
+		return(initialorder)
+	}
+	
 	no.bics <- ncol(biclusters$rows)
 	no.rows <- nrow(biclusters$rows)
 	no.cols <- nrow(biclusters$columns)
-		
+	
 	no.slots <- c(no.rows, no.cols)
 	
 	if ( is.null(initialorder) ) {
@@ -95,21 +101,17 @@ OrderEV.list <- function(biclusters, initialorder=NULL, debuglevel=0, maxtime=60
 	}
 			
 	timelimits <- list()
-	#cat("timelimits:\n")
 	for ( i in 1:2 ) {
 		timelimits[[i]] <- list(maxtime * estimates[[i]][[1]] / estimatedtotal)
 		if ( maxtime > 0 && timelimits[[i]][[1]] < 1 ) {
 			timelimits[[i]][[1]] <- 1
 		}
-		#cat(timelimits[[i]][[1]], " ")
 		for ( mod in 1:no.bics ) {
 			timelimits[[i]][[mod+1]] <- maxtime * estimates[[i]][[mod+1]] / estimatedtotal
 			if ( maxtime > 0 && timelimits[[i]][[mod+1]] < 1 ) {
 				timelimits[[i]][[mod+1]] <- 1
 			}
-		#	cat(timelimits[[i]][[mod+1]], " ")
 		}
-		#cat("\n")
 	}
 	
 	for ( i in 1:2 ) {
@@ -186,8 +188,8 @@ OrderEV.list <- function(biclusters, initialorder=NULL, debuglevel=0, maxtime=60
 	cat("ordering done.                                  \r\n")
 	flush.console()
 
-	res = list(rows=row.map, cols=col.map, status=list(rows=initialorder$status[[1]], cols=initialorder$status[[2]]))
+	res <- list(rows=row.map, cols=col.map, status=list(rows=initialorder$status[[1]], cols=initialorder$status[[2]]))
 
-	res
+	return(res)
 
 }

### 2009-11-16 RExpressionView/src/orderclusters.cpp

@@ -1,3 +1,19 @@
+//     ExpressionView - A package to visualize biclusters
+//     Copyright (C) 2009 Computational Biology Group, University of Lausanne
+// 
+//     This program is free software: you can redistribute it and/or modify
+//     it under the terms of the GNU General Public License as published by
+//     the Free Software Foundation, either version 3 of the License, or
+//     (at your option) any later version.
+// 
+//     This program is distributed in the hope that it will be useful,
+//     but WITHOUT ANY WARRANTY; without even the implied warranty of
+//     MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE. See the
+//     GNU General Public License for more details.
+// 
+//     You should have received a copy of the GNU General Public License
+//     along with this program.  If not, see <http://www.gnu.org/licenses/>.
+
 #include <vector>
 #include <set>
 #include <map>

### 2009-11-16 RExpressionView/src/orderclusters.h

@@ -1,3 +1,19 @@
+//     ExpressionView - A package to visualize biclusters
+//     Copyright (C) 2009 Computational Biology Group, University of Lausanne
+// 
+//     This program is free software: you can redistribute it and/or modify
+//     it under the terms of the GNU General Public License as published by
+//     the Free Software Foundation, either version 3 of the License, or
+//     (at your option) any later version.
+// 
+//     This program is distributed in the hope that it will be useful,
+//     but WITHOUT ANY WARRANTY; without even the implied warranty of
+//     MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE. See the
+//     GNU General Public License for more details.
+// 
+//     You should have received a copy of the GNU General Public License
+//     along with this program.  If not, see <http://www.gnu.org/licenses/>.
+
 #ifndef __order_clusters_h__
 #define __order_clusters_h__
 

### 2009-11-25 RExpressionView/inst/doc/ExpressionView.ordering.Rnw
RExpressionView: Upgrade to BioC standards.

% \VignetteIndexEntry{How the ordering algorithm works}
\documentclass{article}
\usepackage{ragged2e}
\usepackage{hyperref}
\usepackage{url}
\usepackage[margin=3cm]{geometry}

\newcommand{\Rfunction}[1]{\texttt{#1}}
\newcommand{\Rpackage}[1]{\texttt{#1}}
\newcommand{\Rclass}[1]{\texttt{#1}}
\newcommand{\Rargument}[1]{\textsl{#1}}
\newcommand{\filename}[1]{\texttt{#1}}
\newcommand{\variable}[1]{\texttt{#1}}

\newcommand{\code}[1]{\texttt{#1}}

\SweaveOpts{cache=TRUE}

\begin{document}

\title{Ordering algorithm used in ExpressionView}
\author{Andreas L\"uscher}
\maketitle

\tableofcontents

\RaggedRight

<<set width,echo=FALSE,print=FALSE>>=
options(width=60)
options(continue=" ")
@

\section{Introduction}
Clustering genes according to their expression profiles is an important task 
in analyzing microarray data. To present the mutually overlapping biclusters
in a visually appealing layout, the gene expression matrix has to be arranged
in such a way that biclusters form contiguous rectangles. For more than two biclusters,
this is in general impossible. Here, we present an algorithm that finds the arrangement
that maximizes the areas of the largest contiguous parts of the biclusters.

\section{Problem}
The problem can be formulated very generally:
Given a matrix in which each element is part of at least one bicluster, arrange the rows and columns
in such a way that the biclusters are easily recognizable on a two-dimensional plot.
Since biclusters are rectangular, rows and columns can be ordered separately.
The problem thus reduces to finding the optimal arrangement of a set of $n$ elements 
that are part of at least one of $m$ clusters. In what follows, we refer to the $n$ elements as
slots.

\section{Quantifying the order}
There are several ways of quantifying a ``visually appealing layout''. We have chosen to
define the quality of the order $Q$ as the sum over the maximal number of neighboring elements 
in every cluster, as illustrated in Fig.~\ref{fig:ordering}.
\begin{figure} 
	\begin{center} 
		\includegraphics{figures/ordering.pdf}
	\end{center} 
	\caption{The initial configuration (a) with a score $Q$=58 can be rearranged by modifying the order of the slots. The optimal 
	solution (b) maximizes $Q$.} 
	\label{fig:ordering} 
\end{figure}

\section{Optimizing the order}
Optimizing the order thus consists in maximizing $Q$ by changing the order of the slots.
To achieve this goal we use a brute-force approach that 
\begin{itemize}
	\item{shifts parts of a given cluster to better positions and}
	\item{permutes elements contained in a given cluster.}
\end{itemize}

The \code{C++} code contains the functions \code{reposition} and \code{exchange} that 
execute these two basic moves.

Before optimizing the arrangement, the matrix is simplified (\code{simplify}) to get rid of duplicate slots.
The multiplicity of each slot is taken into account when calculating the score $Q$. 
At the end of the ordering the initial dimensions of the data are restored with the \code{complexify} 
function.

The \code{C++} program can also be compiled to a standalone executable by defining \code{-DSTANDALONE}
to exclude R-specific functions. You can also define \code{-DTIMING} to measure the time spent in the function
that calculates $Q$ (\code{getfitness}).

\section{Calculation time}
To get an idea of the complexity of the algorithm, we determined the scaling of the calculation time 
with the dimensions of the data, i.e., the number of clusters $m$ and the slots $n$. The results, obtained as
the average over at least 40 runs, are summarized in Figs.~\ref{fig:clusters} and \ref{fig:slots}. 
\begin{figure} 
	\begin{center} 
		\includegraphics{figures/nclusters.pdf}
	\end{center} 
	\caption{Scaling of the execution time as a function of the number of clusters. Data points connected
	by dashed lines represent results obtained from orderable samples. The second set is obtained
	from random samples and is thus more representative of gene expression data. Both sets scale polynomially
	with the number of clusters $m$ as $\mathcal{O}(m^\alpha)$, with $\alpha \in [1.6, 2]$ for random samples and
	$\alpha \in [2.3, 4.2]$ in the case of orderable samples.} 
	\label{fig:clusters} 
\end{figure}
We consider orderable and completely random samples. In both cases, the time increases polynomially with the number
of clusters as $\mathcal{O}(m^\alpha)$, but with different exponents, see Fig.~\ref{fig:clusters}: For random samples, we find 
$\alpha \in [1.6, 2]$, almost independent of the number of elements $n$, while for orderable samples, 
$\alpha \in [2.3, 4.3]$. We note that despite of having a slower increase of computation time with $m$, 
it takes considerably more time to order random samples.

The scaling of the computation time with the number of slots $n$ is shown in Fig.~\ref{fig:slots}. 
\begin{figure} 
	\begin{center} 
		\includegraphics{figures/nslots.pdf}
	\end{center} 
	\caption{Scaling of the execution time as a function of the number of slots. The second set is obtained
	from random samples and is thus more representative of gene expression data. The random set scales polynomially
	with the number of slots $n$ as $\mathcal{O}(n^\beta)$, with $\beta \in [2.5, 2.7]$. The orderable data set shows an initially
	polynomial increase and then saturates for larger $n$.} 
	\label{fig:slots} 
\end{figure}
Here, only the random samples exhibit a polynomial dependence $\mathcal{O}(n^\beta)$, with $\beta \in [2.5, 2.7]$. 
The orderable batch saturates at larger $n$. This is due to the fact that by increasing the number of slots $n$ while keeping
the number of clusters constant, the complexity of the problem stays roughly the same because most of the additional slots are 
equal to those already contained in the data.

\section{Possible improvements}
More than 90\% of the calculation time is spent to recalculate the length of the longest contiguous subclusters. Optimizing the \code{getfitness} routine
thus has a direct impact on the overall performance of the algorithm. Since the different clusters can be treated independently, it is
straightforward to parallelize the calculation, maybe even on the graphic chip.

\bibliographystyle{unsrt}
\bibliography{ExpressionView}

\end{document}

### 2009-11-25 RExpressionView/inst/doc/ExpressionView.tutorial.Rnw
RExpressionView: Upgrade to BioC standards.

% \VignetteIndexEntry{ExpressionView}
\documentclass{article}
\usepackage{ragged2e}
\usepackage{hyperref}
\usepackage{url}
\usepackage[margin=3cm]{geometry}

\newcommand{\Rfunction}[1]{\texttt{#1}}
\newcommand{\Rpackage}[1]{\texttt{#1}}
\newcommand{\Rclass}[1]{\texttt{#1}}
\newcommand{\Rargument}[1]{\textsl{#1}}
\newcommand{\filename}[1]{\texttt{#1}}
\newcommand{\variable}[1]{\texttt{#1}}

%\SweaveOpts{prefix.string=graphics/plot}

\begin{document}

\title{ExpressionView}
\author{Andreas L\"uscher}
\maketitle

\tableofcontents

\RaggedRight

<<set width,echo=FALSE,print=FALSE>>=
options(width=60)
options(continue=" ")
@

\section{Introduction}
Clustering genes according to their expression profiles is an important task 
in analyzing microarray data. In this tutorial, we explain how to use ExpressionView, 
an R package designed to interactively explore biclusters identified in gene 
expression data in conjunction with the Iterative Signature Algorithm (ISA)~\cite{bergmann03} 
and the biclustering methods available in the \Rpackage{Biclust} package~\cite{kaiser08}.

\section{Loading the gene expression data}
The \Rpackage{ExpressionView} package requires the gene expression data to be available in the
form of a \Rpackage{Bioconductor} \Rclass{ExpressionSet}. In this tutorial we will use the 
\Rpackage{Bioconductor} sample data from a 
clinical trial in acute lymphoblastic leukemia provided by the \Rpackage{ALL} package.
<<load the data>>=
library(ALL)
library(hgu95av2.db)
data(ALL)
@
The data set contains \Sexpr{ncol(ALL)} samples and \Sexpr{nrow(ALL)} features.

\section{Find biclusters}
There are many biclustering algorithms described in the literature~\cite{madeira04}.
All of them aim to reduce the complexity of the gene expression data by identifying 
suitable groups of genes and conditions that are co-regulated. In what follows we
show how to use two of the freely available biclustering methods.

\subsection{Iterative Signature Algorithm (ISA)}
The ISA~\cite{bergmann03} for gene expression data is implemented in the \Rpackage{eisa} package:
<<load the packages>>=
library(eisa)
@

To run the ISA for the given data set, simply call the \Rfunction{ISA} function on the
\Rclass{ExpressionSet} object:
<<ISA, eval=FALSE>>=
set.seed(5) # initialize random number generator to get always the same results
modules <- ISA(ALL)
@ 
Depending on your computing resources, this should take roughly two minutes. If you do not
want to wait that long, you can shorten the calculation by selecting the thresholds for genes and
conditions:
<<fast ISA, cache=FALSE>>=
threshold.genes <- 2.7 
threshold.conditions <- 1.4 
set.seed(5)
modules <- ISA(ALL, thr.gene=threshold.genes, thr.cond=threshold.conditions)
@
If you leave the thresholds undefined, as in the first example, the ISA runs with the 
default values, i.e., \variable{thr.gene=c(2,2.5,3,3.5,4)} and \variable{thr.cond=c(1,1.5,2,2.5,3)}.
In both cases, the random number generator is initialized manually using \Rfunction{set.seed(5)}, 
to give reproducible results. The \Rfunction{isa} function returns an \Rclass{ISAModules} object. 
Typing its name returns a brief summary of the results:
<<ISAModules summary>>=
modules
@

The ISA biclustering is not binary, every gene (and
similarly, every sample or condition) has a score between -1 and 1; the further the
score is from zero, the stronger the association between the gene
(or sample) and the module. If two features both have scores with the
same sign, they are correlated, otherwise, they are anti-correlated.

\subsection{Algorithms of the \Rpackage{Biclust} package}
The \Rpackage{biclust} package implements several biclustering
algorithms in a unified framework. It uses the \Rclass{Biclust} class 
to store a set of biclusters. Let us use the Plaid Model Bicluster Algorithm~\cite{turner04}
on the \Rpackage{ALL} data set
<<biclust bcplaid,cache=FALSE>>=
library(biclust)
biclusters <- biclust(exprs(ALL), BCPlaid(), fit.model = ~m + a + b, verbose = FALSE)
@

The standard R function \Rfunction{as} 
can be used to convert \Rclass{ISAModules} to \Rclass{Biclust} objects. This
requires the binarization of the modules, i.e., the ISA scores are
lost since they are converted to zeros and ones:
<<biclust>>=
as(modules, "Biclust")
@

The conversion also works in the other direction. Starting from a \Rclass{Biclust} object, just use
<<isa>>=
as(biclusters, "ISAModules")
@
to get an \Rclass{ISAModules} object.

\subsection{External clustering programs}
Since the structure of biclustering results is independent of the applied method, it is straightforward to
import results obtained from external clustering programs and convert them to \Rclass{ISAModules}.
To illustrate the conversion, let us consider the sample data and {\bf randomly} assign the  \Sexpr{nrow(ALL)} genes 
and  \Sexpr{ncol(ALL)} samples to  \Sexpr{length(modules)} modules. The resulting modules can be described by two
binary matrices
<<random-modules>>=
modules.genes <- matrix(as.integer(runif(dim(ALL)[1]*length(modules),0,1)>0.8),nrow=dim(ALL)[1])
modules.conditions <- matrix(as.integer(runif(dim(ALL)[2]*length(modules),0,1)>0.8),nrow=dim(ALL)[2])
@
indicating if a given gene \variable{i} is contained in module \variable{j} if \variable{modules.genes[i,j]$\ne$0}.
Using these matrices, it is straightforward to create an \Rclass{ISAModules} object:
<<toisa>>=
new("ISAModules",
genes=modules.genes, conditions=modules.conditions,rundata=data.frame(),seeddata=data.frame())
@

\section{Order}
To present the tens of possibly overlapping biclusters in a visually appealing form, 
it is necessary to regroup the rows and columns of the gene expression matrix in 
such a way that biclusters form contiguous rectangles. Since for more than two 
mutually overlapping biclusters, it is in general impossible to find such an arrangement,
 one has to make concessions. In contrast to~\cite{grothaus06}, who proposed to repeat 
 rows and columns as necessary to achieve this goal, we prefer to optimize the 
 arrangement within the original data by maximizing the area of the largest contiguous biclusters.

The \Rfunction{OrderEV} function implemented in the \Rpackage{ExpressionView} package
determines the optimal order of the gene expression matrix for a given set of biclusters:
<<expressionview, results=HIDE>>=
library(ExpressionView)
optimalorder <- OrderEV(modules)
@

The result is a list containing various mappings between the original data and the optimal arrangement.
Note that the genes and the samples can be ordered separately. Apart form reordering the 
full gene expression matrix, the algorithm also determines the best arrangement of individual biclusters.
The mapping of the genes and the samples contained in bicluster \variable{i} can be accessed by 
<<gene maps, eval=FALSE>>=
optimalorder$genes[i+1]
optimalorder$samples[i+1]
@
Here, \variable{i=0}, returns the map of the complete data set. By default, the \Rfunction{OrderEV}
runs for roughly one minute, this might not be sufficient to find an appropriate order for data 
containing many overlapping biclusters. The status of the ordering is stored in 
<<status ordering>>=
optimalorder$status
@
If the status is set to \variable{1}, the algorithm has found the optimal solution. A \variable{0} indicates that the
the calculation could not be terminated within the given timeframe. The \Rfunction{OrderEV} 
accepts two additional parameters to circumvent the problem of partial alignment:
One can start the ordering from a given initial configuration, i.e., the result of a previous arrangement
by defining the \Rargument{initialorder} argument
<<status ordering, eval=FALSE>>=
optimalorderp <- OrderEV(modules, initialorder=optimalorder, maxtime=120)
@
and one can increase the time limit by specifying \Rargument{maxtime}. Note that the time is indicated in seconds
and cannot be  smaller than 1.

\section{Export}
The \Rfunction{ExportEV} function allows the user to combine the available data and export it to an XML file that can be read by the Flash applet:
<<export, eval=FALSE>>=
ExportEV(modules, ALL, optimalorder, filename="file.evf")
@
The function gathers the data contained in the \Rclass{ExpressionSet} \variable{ALL}, orders it according to the optimal arrangement 
\variable{optimalorder} and adds the biclusters defined in \variable{modules}. The output is an uncompressed XML file that can be opened with any text viewer. We have chosen to use the extension \variable{.evf} (for ExpressionView file) for the data files. This extension is associated with the stand-alone version of the viewer, so that one can simply double-click on such a file to launch the program and load the data. The file association is the reason why we do not use the \variable{.xml} extension. A description of the XML layout can be found on the  href{http://www2.unil.ch/cbg/index.php?title=ExpressionView}{website}. Before exporting the data, the 
\Rfunction{ExportEV} function automatically calculates GO~\cite{ashburner00} and KEGG~\cite{kanehisa04} enrichments for the given biclusters.

\section{Visualize}
The ExpressionView Flash applet can be launched from the R environment:
<<export, eval=FALSE>>=
OpenEV()
@
Video tutorials describing how to use the applet can be found on the ExpressionView \\
\href{http://www2.unil.ch/cbg/index.php?title=ExpressionView}{website}.

\section{Session information}

The version number of R and packages loaded for generating this
vignette were:

<<sessioninfo,results=tex,echo=FALSE>>=
toLatex(sessionInfo())
@

\bibliographystyle{unsrt}
\bibliography{ExpressionView}

\end{document}

### 2009-11-25 RExpressionView/R/AllGenerics.R

@@ -1,5 +1,5 @@
 setGeneric("OrderEV",
-           function(biclusters, ...) standardGeneric("OrderEV"))
+           function(biclusters, initialorder, maxtime=60, debuglevel=0) standardGeneric("OrderEV"))
 
 setGeneric("ExportEV",
-           function(biclusters, ...) standardGeneric("ExportEV"))
+           function(biclusters, eset, order, filename, norm, description) standardGeneric("ExportEV"))

### 2009-11-25 RExpressionView/R/export.R

@@ -3,14 +3,17 @@
 ########################################################################
 
 if (require(biclust)) {
- 	setMethod("ExportEV", signature(biclusters="Biclust"), function(biclusters, ...) ExportEV.Biclust(biclusters, ...))
+ 	setMethod("ExportEV", signature(biclusters="Biclust"), 
+ 		function(biclusters, eset, order, filename, norm, description) 
+ 		ExportEV.Biclust(biclusters, eset, order, filename, norm, description)
+ 	)
 }
 
-ExportEV.Biclust <- function(biclusters, eset, order, filename=NULL, norm=c("sample", "feature", "raw")) {
+ExportEV.Biclust <- function(biclusters, eset, order, filename, norm, description) {
 	eisamodules <- as(biclusters, "ISAModules")
 	eisamodules@rundata$annotation <- annotation(eset)
 	eisamodules@rundata$prenormalize <- FALSE
-	ExportEV(eisamodules, eset, order, filename, norm)
+	ExportEV(eisamodules, eset, order, filename, norm, description)
 }
 
 
@@ -19,21 +22,29 @@ ExportEV.Biclust <- function(biclusters, eset, order, filename=NULL, norm=c("sam
 ########################################################################
 
 if (require(eisa)) {
-	setMethod("ExportEV", signature(biclusters="ISAModules"), function(biclusters, ...) ExportEV.ISAModules(biclusters, ...))
+	setMethod("ExportEV", signature(biclusters="ISAModules"), 
+		function(biclusters, eset, order, filename, norm, description) 
+		ExportEV.ISAModules(biclusters, eset, order, filename, norm, description)
+	)
 }
 
-ExportEV.ISAModules <- function(biclusters, eset, order=NULL, filename=NULL, norm=c("sample", "feature", "raw")) {
+ExportEV.ISAModules <- function(biclusters, eset, order, filename, norm, description) {
 
-	if ( is.null(order) ) {
-		order = OrderEV(biclusters)
+	if ( missing(order) ) {
+		order <- OrderEV(biclusters)
 	}
 
-	if ( is.null(filename) ) {
+	if ( missing(filename) ) {
 		con <- file(file.choose(TRUE), open="w")
 	} else {
 		con <- file(filename, open="w", blocking = TRUE)
 	}
-		
+	
+	if ( missing(norm) ) {
+		#norm <- c("sample", "feature", "raw")
+		norm <- "sample"
+	}
+	
[…]

### 2009-11-25 RExpressionView/R/open.R

@@ -1,18 +1,0 @@
-OpenEV <- function() {
-	swf <- system.file("ExpressionView.html", package="ExpressionView")
-	url <- URLencode(paste("file://", swf, sep=""))
-	browseURL(url)
-}
-
-#OpenEV <- function(modules=NULL, eset=NULL, order=NULL) {
-#	swf <- system.file("ExpressionView.html", package="ExpressionView")
-#	if ( is.null(modules) ) {
-#		url <- URLencode(paste("file://", swf, sep=""))
-#	} else {
-#		filename <- paste(tempfile(), sep="", ".xml")
-#		ExportEV(modules, eset, order, filename=filename)
-#		url <- URLencode(paste("file://", swf, sep="", "?filename=", filename))
-#	}
-#	browseURL(url)
-#}
-

### 2009-11-25 RExpressionView/R/order.R

@@ -3,15 +3,28 @@
 ########################################################################
 
 if (require(eisa)) {
-	setMethod("OrderEV", signature(biclusters="ISAModules"), function(biclusters, ...) OrderEV.ISAModules(biclusters, ...))
+	setMethod("OrderEV", signature(biclusters="ISAModules"), 
+		function(biclusters, initialorder, maxtime, debuglevel) 
+		OrderEV.ISAModules(biclusters, initialorder, maxtime, debuglevel)
+	)
 }
 
-OrderEV.ISAModules <- function(biclusters, initialorder=NULL, debuglevel=0, maxtime=60) {
-	isamodules <- eisa:::ISAModules.to.isa.result(biclusters)
-	if ( !is.null(initialorder) ) {
+OrderEV.ISAModules <- function(biclusters, initialorder, maxtime, debuglevel) {
+
+	if ( missing(initialorder) ) {
+		initialorder <- NULL	
+	} else {
 		initialorder <- list(rows=initialorder$genes, cols=initialorder$samples, status=initialorder$status)
 	}
-	resp <- OrderEV(isamodules, initialorder, debuglevel, maxtime)
+	if ( missing(maxtime) ) {
+		maxtime <- 60
+	}
+	if ( missing(debuglevel) ) {
+		debuglevel <- 0
+	}
+	
+	isamodules <- eisa:::ISAModules.to.isa.result(biclusters)
+	resp <- OrderEV(isamodules, initialorder, maxtime, debuglevel)
 	res <- list(genes=resp$rows, samples=resp$cols, status=list(genes=resp$status[[1]], samples=resp$status[[2]]))
 	res
 }
@@ -22,12 +35,25 @@ OrderEV.ISAModules <- function(biclusters, initialorder=NULL, debuglevel=0, maxt
 ########################################################################
 
 if (require(biclust)) {
-	setMethod("OrderEV", signature(biclusters="Biclust"), function(biclusters, ...) OrderEV.Biclust(biclusters, ...))
+	setMethod("OrderEV", signature(biclusters="Biclust"), 
+		function(biclusters, initialorder, maxtime, debuglevel) 
+		OrderEV.Biclust(biclusters, initialorder, maxtime, debuglevel)
+	)
 }
 
-OrderEV.Biclust <- function(biclusters, initialorder=NULL, debuglevel=0, maxtime=60) {
[…]

### 2009-11-29 isanote/modular.tex
Added SIB to isanote acknowledgement.
isanote: added Sven's changes.

--- 2009-10-06 isanote/modular.tex
@@
 
-\title[Modular analysis]{Modular analysis of gene expression data with
-  GNU R or Matlab}
+\title[Modular analysis]{Modular analysis of gene expression data with R}
 \author[G\'abor Cs\'ardi \textit{et~al}]{G\'abor Cs\'ardi\,,$^{1,2}$,
@@
 \begin{abstract}
+\section{Summary:}
+Large sets of data, like expression profile from many samples, require
+analytic tools to reduce their complexity. Classical (bi-)clustering
+algorithms typically attribute elements (genes, arrays) to disjoint groups
+(``clusters''). Yet, in some cases overlapping cluster assignments would suit
+the biological reality much better.
 
-\section{Summary:}
+The Iterative Signature Algorithm (ISA) was designed to overcome this and
+other limitations of standard clustering algorithms. It aims to reduce the
+complexity of very large sets of data by decomposing it into so-called
+``modules''. In the context of gene expression data these modules consist of
+subsets of genes that exhibit a coherent expression profile only over a
+subset of microarray experiments. Genes and arrays may be attributed to
+multiple modules and the level of required coherence can be varied resulting
+in different ``resolutions'' of the modular mapping. Since the ISA does not
+rely on the computation of correlation matrices (like many other tools), it
+is extremely fast even for very large datsets.
+
+In this short note, we introduce two BioConductor~\cite{BioC} software
+package written in GNU~R: The~\Rpackage{isa2} includes a highly optimized
+implementation of the ISA and~\Rpackage{eisa} provide a convenient interface
+to run the ISA and visualize its output.
+
 \section{Availability:}
+\href{http://www.unil.ch/cbg/homepage/software.html}%
+{http://www.unil.ch/cbg/homepage/software.html}
 \section{Contact:} \href{Sven.Bergmann@unil.ch}{Sven.Bergmann@unil.ch}
@@
 
-% What is biclustering.
+% How the ISA works
+The ISA can be applied to identify coherent substructures (i.e. modules)
+from any rectangular matrix of data. To be specific we consider here the
+case of transcriptomics data corresponding to a selection of gene-expression
+profiles from a collection of samples. The method has been described in
+detail in Refs~\cite{BioISA,PREISA}. Here only give a brief summary:
 
-Biclustering is the idea of clustering the rows and the columns of a
-data matrix, together, usually not independently. In other words, the
-goal of a biclustering algorithm is to find blocks in the reordered
-data matrix that have, such that the blocks have correlated rows and
-columns. It is important to note that these blocks (or biclusters)
-have context, i.e. the rows only need to be correlated across the
-columns of the block, but not across other columns of the data matrix;
-and vice-versa.
+The ISA identifies modules by an iterative procedure. The algorithm starts
+from an input seed (corresponding to some set of genes or samples) which are
+refined at each iteration by adding and/or removing genes and/or samples
+until the process converges to a stable set, which is referred to as a
+transcription module. This iterative procedure is typically performed to a
+large number of seeds. In the unsupervised approach these seeds are chosen
+randomly to sample uniformly the immense search space. We also implemented a
+semi-supervised method, to which we refer as ``smart seeding'', where the
+seeds a biased to start with certain sets of genes or samples for which
+there is prior knowledge.
 
-When it comes to gene expression data, a bicluster means a subset of
-genes ($G$) (or probes, probe sets) and a subset of samples ($S$),
-such that genes $G$ are exactly coexpressed across $S$ and samples $S$
-have correlated expression profile exactly across $G$.
+The output of ISA is a collection of potentially overlapping modules. Each
+gene and each sample is attributed a score that reflects the strength of the
+association with the module.
 
-% What is the ISA.
+In the following we first outline how our software tool is commonly for a
+modular analysis (Section~\ref{sec:methods}) and then we describe in more
+detail the R packages that implement the ISA and visualize its output
+(Section~\ref{sec:implementation}).
 
-The Iterative Signature Algorithm (ISA)~\cite{sa,isa,ismod} is a 
-powerful biclustering algorithm. It is able to identify biclusters
-that overlap and it is resilient to noise.
-
-In this short note, we introduce the \Rpackage{isa2} and
-\Rpackage{eisa} GNU R~\cite{R} packages; these implement the ISA
-biclustering method, provide a convenient interface to run it using
-the BioConductor~\cite{BioC} software package, and also contain
-visualization tools.
-
-\section{Methods}
+\section{Methods}%
+\label{sec:methods}
 
@@
 \includegraphics[width=\textwidth]{isa2workflow3}
-\caption{}
-\label{fig:workflow}
-\end{figure*}
+\caption{Work flow of a typical modular analysis study, with the
+  \Rpackage{eisa} package (left); and four bicluster visualization
+  techniques (right). The input of the \Rpackagescriptional regulation one often combines several
+microarray experiments into a single dataset. In such a case it is important
+to properly normalize such data and reduce the bias due to constituent
+datasets. Several methods address this challenge,
+see e.g.~\cite{johnson07} for an algorithm that has a GNU R implementation.
+(S: Do we have a standard flag to call their software?)
 
-% Batch correction
+% Filtering
+\emph{Gene filtering.}
+Genes that have very low expression levels in all samples, carry little if
+any information and may reflect ineffective array probes, etc. Since these
+genes are likely to contribute only noise to the analysis, we integrated the
+possibility to remove such genes. (S: Describe how this is done, and if it
+requires setting a threshold!)
 
 % Normalization.
+\emph{ISA normalization.}
+In each iteration of the ISA weighted sums of expression levels over either
+genes or samples are computed.  for a number of . Since different genes
+typically show different levels of base
+expression and variance, it is important to standardize expression
+levels to $Z$-scores. The ISA uses two se, the gene
+threshold and the sample threshold. ISA modules can be organized into
+a directed graph, a module tree, in which there is an edge from module
+$A$ to module $B$, if the ISA converges to module $B$ from module $A$,
+with the same threshold parameters  that were used to find module
+$B$. A module tree provides a hierarchical modular description of a
+data set.
 
-% Random and smart seeding.
+\section{Implementation}%
+\label{sec:implementation}
 
-% The ISA iteration.
+The ISA and accompanying visualization tools are implemented in two R
+packages. The \Rpackage{isa2} package contains the implementation of
+the basic ISA itself; this package can be used to analyze any tabular
+data. The \Rpackage{eisa} package builds on \Rpackage{isa2}. It adds
+support to standard BioConductor gene expression data structures; and
+contains gene expression specific visualization tools, see some of
+these in Fig.~\ref{fig:workflow}.
 
-% Merging the modules.
+Both the \Rpackage{isa2} and \Rpackage{eisa} packages support two
+workflows. The \emph{simple workflow} involves a single R function
+call and runs all ISA steps (1-3 on Fig.~\ref{fig:workflow}.) with
+their default parameters.
 
-% The robustness measure and filtering the modules.
+The \emph{detailed workflow} means running every step of the modular
+analysis separately, possibly with non-default parameters. This allows
+the users to tailor the ISA completely according their needs.
 
-% Module trees.
+For users that prefer Matlab over R, we created a Matlab package for
+ISA, please see the ISA homepage for more information.
 
-\section{R Implementation}
+% \subsubsection{Visualization}
 
-Two packages: isa2 does the computation, eisa does everything related
-to biology.
+The \Rpackage{eisa} package implements a set of visualization
+techniques for biclusters, some of these are specific to ISA, others
+are not. Please see Fig.~\ref{fig:workflow} for some of these.
 
-\subsection{Work flow}
+The \Rpackage{ExpressionView} R package implements an interactive tool
+for visualizing ISA (or other) overlapping biclusters.
 
-Two workflows, the simple workflow does everything with default
-parameters. In the detailed workflow the user has access to every step
-of the analysis.
+% \subsubsection{Connection to other software}
 
-\subsubsection{The simple work flow}
-
-\subsubsection{The detailed work flow}
-
-\subsection{Visualization}
-
-Expression plots, GO tree plots, module tree plots, HTML summary
-generation. ExpressionView.
-
-\subsection{Connection to other software}
-
-The biclust package.
-
-\section{Matlab implementation}
+The \Rpackage{biclust} package, \cite{biclust}, implements a number of
+biclustering algorithms in a unified framework. The \Rpackage{eisa}
+package include tools to convert between \Rpackage{biclust} and ISA
+biclusters. This allows the cross-talk of the functions in the two
+packages.
 
@@
 
-\paragraph{Funding\textcolon}
-
+\paragraph{Funding\textcolon} The authors are greatful to the Swiss
+Institute of Bioinformatics for support.
 \paragraph{Conflict of interest\textcolon} None declared.

### 2009-11-29 isanote/modular.tex
Added Sven's changes, this time properly.

@@
 algorithms typically attribute elements (genes, arrays) to disjoint groups
-(``clusters''). Yet, in some cases overlapping cluster assignments would suit
+("clusters"). Yet, in some cases overlapping cluster assignments would suit
 the biological reality much better.
@@
 complexity of very large sets of data by decomposing it into so-called
-``modules''. In the context of gene expression data these modules consist of
+"modules". In the context of gene expression data these modules consist of
 subsets of genes that exhibit a coherent expression profile only over a
@@
 multiple modules and the level of required coherence can be varied resulting
-in different ``resolutions'' of the modular mapping. Since the ISA does not
+in different "resolutions" of the modular mapping. Since the ISA does not
 rely on the computation of correlation matrices (like many other tools), it
@@
 randomly to sample uniformly the immense search space. We also implemented a
-semi-supervised method, to which we refer as ``smart seeding'', where the
+semi-supervised method, to which we refer as "smart seeding", where the
 seeds a biased to start with certain sets of genes or samples for which
@@
   \Rpackage{eisa} package (left); and four bicluster visualization
-  techniques (right). The input of the \Rpackagescriptional regulation one often combines several
+  techniques (right). The input of the \Rpackage{eisa} pipeline is a
+  BioConductor \Rclass{ExpressionSet} object. This is normalized to
+  $Z$-scores (1); the ISA iterations are initiated using "smart" or
+  random seeds (2); and the biclusters are then merged and filtered
+  based on the robustness measure (3). The \Rpackage{eisa} package
+  contains tools to perform many GO, KEGG, etc. enrichment calculation
+  quickly (4), and several bicluster visualization tools (5).
+  Subfigures \subfig{A}, \subfig{B}, \subfig{C} and \subfig{D} were
+generated
+  using the acute lymphoblastic leukemia data set, see
+  \cite{chiaretti04} and the \Rpackage{ALL} R package.
+  \subfig{(A)} A condition plot for a simple bicluster. Each sample is
+  represented as a bar, its height is its score in the bicluster.
+  Samples above the red (none in this case) and samples
+  below the green line are part of the biclusters. The coloring
+  indicates cases and controls.
+  \subfig{(B)} A module tree. Each rectangle is a bicluster; see the
+  definition of the edges in the text. The modules are colored
+  according to their Gene Ontology enrichment $p$-values, the codes of
+  the enriched GO categories are shown in the top-left corner of the
+  rectangles. The top-right corner shows the number of genes and
+  conditions in the bicluster. The gene thresholds used for finding
+  the modules is shown on the horizontal axes.
+  \subfig{(C)} Heatmap for a single module, showing correlated
+  expression for the genes and samples. The red lines are the gene and
+  sample scores.
+  \subfig{(D)} Profile plot, for a single module. On the top plot the
+  expression of all samples is plotted, the samples of the module
+  with black. The red and green lines show the mean of the two
+  groups. The bottom plot is the same for the genes of the module.
+}
+\label{fig:workflow}
+\end{figure*}
+
+A typical modular analysis for gene expression data includes the following
+steps:
+
+% Batch correction
+\emph{Batch correction.}
+In order study the global organization of a transcription program including
+many aspects of transcriptional regulation one often combines several
 microarray experiments into a single dataset. In such a case it is important
@@
 expression and variance, it is important to standardize expression
-levels to $Z$-scores. The ISA uses two se, the gene
+levels to $Z$-scores. The ISA uses two sets of $Z$-scores, one
+calculated for each gene across all samples, the other for each sample
+across all genes.
+
+% Random and smart seeding.
+\emph{Random and smart seeding.}
+As mentioned previously, ISA can be performed with random or smart
+seeds, based on the application.
+
+% The ISA iteration.
+\emph{The ISA iteration.}
+This step is essentially performing the ISA and finding the biclusters
+from the starting seeds.
+
+% Merging the modules.
+\emph{Merging the modules.}
+It is possible that more seeds converge to the same, or very similar
+biclusters. This step eliminates such duplicates.
+
+% The robustness measure and filtering the modules.
+\emph{Robustness of biclusters.}
+To access the significance of a bicluster, we designed a robustness
+measure, that can be used to filter out spurious modules. This is done
+based on performing ISA on the scrambled input data.
+
+% Module trees.
+\emph{Module trees.}
+The ISA works with two stringency threshold parameters, the gene
 threshold and the sample threshold. ISA modules can be organized into

### 2009-11-29 isanote/modular.tex
isanote updated.

@@
 \usepackage{natbib}
+\usepackage{url}
 \copyrightyear{2009}
@@
 algorithms typically attribute elements (genes, arrays) to disjoint groups
-("clusters"). Yet, in some cases overlapping cluster assignments would suit
+(``clusters''). Yet, in some cases overlapping cluster assignments would suit
 the biological reality much better.
@@
 complexity of very large sets of data by decomposing it into so-called
-"modules". In the context of gene expression data these modules consist of
+``modules''. In the context of gene expression data these modules consist of
 subsets of genes that exhibit a coherent expression profile only over a
@@
 multiple modules and the level of required coherence can be varied resulting
-in different "resolutions" of the modular mapping. Since the ISA does not
+in different ``resolutions'' of the modular mapping. Since the ISA does not
 rely on the computation of correlation matrices (like many other tools), it
@@
 
-In this short note, we introduce two BioConductor~\cite{BioC} software
-package written in GNU~R: The~\Rpackage{isa2} includes a highly optimized
-implementation of the ISA and~\Rpackage{eisa} provide a convenient interface
+In this short note, we introduce two BioConductor, \cite{BioC}, software
+packages written in GNU~R: The~\Rpackage{isa2} includes a highly optimized
+implementation of the ISA and~\Rpackage{eisa} provides a convenient interface
 to run the ISA and visualize its output.
@@
 \section{Availability:}
-\href{http://www.unil.ch/cbg/homepage/software.html}%
-{http://www.unil.ch/cbg/homepage/software.html}
+The ISA homepage is at \href{http://www.unil.ch/cbg/homepage/software.html}%
+{\url{http://www.unil.ch/cbg/homepage/software.html}}
 \section{Contact:} \href{Sven.Bergmann@unil.ch}{Sven.Bergmann@unil.ch}
@@
 profiles from a collection of samples. The method has been described in
-detail in Refs~\cite{BioISA,PREISA}. Here only give a brief summary:
+detail in Refs~\cite{isamod,isa}. Here we only give a brief summary.
 
 The ISA identifies modules by an iterative procedure. The algorithm starts
-from an input seed (corresponding to some set of genes or samples) which are
+from an input seed (corresponding to some set of genes or samples) which is
 refined at each iteration by adding and/or removing genes and/or samples
@@
 randomly to sample uniformly the immense search space. We also implemented a
-semi-supervised method, to which we refer as "smart seeding", where the
+semi-supervised method, to which we refer as ``smart seeding'', where the
 seeds a biased to start with certain sets of genes or samples for which
@@
 
-In the following we first outline how our software tool is commonly for a
+In the following we first outline how our software tool is commonly used for a
 modular analysis (Section~\ref{sec:methods}) and then we describe in more
@@
   BioConductor \Rclass{ExpressionSet} object. This is normalized to
-  $Z$-scores (1); the ISA iterations are initiated using "smart" or
+  $Z$-scores (1); the ISA iterations are initiated using ``smart'' or
   random seeds (2); and the biclusters are then merged and filtered
   based on the robustness measure (3). The \Rpackage{eisa} package
-  contains tools to perform many GO, KEGG, etc. enrichment calculation
+  contains tools to perform many GO, KEGG, etc. enrichment calculations
   quickly (4), and several bicluster visualization tools (5).
@@
 \emph{Batch correction.}
-In order study the global organization of a transcription program including
+To study the global organization of a transcription program including
 many aspects of transcriptional regulation one often combines several
@@
 see e.g.~\cite{johnson07} for an algorithm that has a GNU R implementation.
-(S: Do we have a standard flag to call their software?)
+% (S: Do we have a standard flag to call their software?)
 
@@
 any information and may reflect ineffective array probes, etc. Since these
-genes are likely to contribute only noise to the analysis, we integrated the
-possibility to remove such genes. (S: Describe how this is done, and if it
-requires setting a threshold!)
+genes are likely to contribute only noise to the analysis, we suggest
+removing them before running the ISA iteration.
+% (S: Describe how this is done, and if it requires setting a threshold!)
 
@@
 In each iteration of the ISA weighted sums of expression levels over either
-genes or samples are computed.  for a number of . Since different genes
+genes or samples are computed. Since different genes
 typically show different levels of base
@@
 
-The \Rpackage{ExpressionView} R package implements an interactive tool
+The \Rpackage{ExpressionView} R package, discussed in detail in
+another application note, implements an interactive tool
 for visualizing ISA (or other) overlapping biclusters.
@@
 packages.
+
+Additional information about the ISA, the \Rpackage{isa2} and
+\Rpackage{eisa} packages and several tutorials are available on the
+ISA homepage.
 

### 2009-12-01 RExpressionView/R/launch.R
RExpressionView: Updated documentation.

@@ -1,5 +1,5 @@
 LaunchEV <- function() {
-	swf <- system.file("ExpressionView.html", package="ExpressionView")
-	url <- URLencode(paste("file://", swf, sep=""))
-	browseURL(url)
+    swf <- system.file("ExpressionView.html", package="ExpressionView")
+    url <- URLencode(paste("file://", swf, sep=""))
+    browseURL(url)
 }

### 2009-12-01 RExpressionView/inst/doc/ExpressionView.tutorial.Rnw

@@
 library(biclust)
-biclusters <- biclust(exprs(ALL), BCPlaid(), fit.model = ~m + a + b, verbose = FALSE)
+biclusters <- biclust(exprs(ALL), BCPlaid(), fit.model=~m+a+b, verbose=FALSE)
 @
@@
 Video tutorials describing how to use the applet can be found on the ExpressionView \\
-\href{http://www2.unil.ch/cbg/index.php?title=ExpressionView}{website}.
+\href{http://www2.unil.ch/cbg/index.php?title=ExpressionView}{website}. The screenshot shown in Fig.~\ref{fig:screenshot} 
+and the description below illustrate the main features of the applet:
+\begin{figure} 
+	\begin{center} 
+		\includegraphics[width=0.95\textwidth]{figures/screenshot.pdf}
+	\end{center} 
+	\caption{Screenshot of the ExpressionView Flash applet.} 
+	\label{fig:screenshot} 
+\end{figure}
+\begin{description}
+	\item[a] Opens an ExpressionView data file. Note that before opening a new data file, you should restart the applet, i.e., refresh your browser window.
+	\item[b] Exports the current view to a pdf file. The file also includes the title (o) of the gene expression data.
+	\item[c] Exports the data of the currently viewed module(=bicluster) to a CSV file, that can be opened as a spreadsheet.
+	\item[d] In inspect mode, you can use the mouse to explore the gene expression data. The information about the data under the mouse pointer is shown in the Info Panel (t).
+	\item[e, f] Zoom and pan modes allow you to restrict the view to a particular part of the gene expression data. They behave like the functions used in graphic programs.
+	\item[e] In zoom mode, you can also use keyboard shortcuts: {\bf a} to auto-zoom onto the modules and {\bf e} to see the whole data.
+In addition to the simple zoom-in feature, you can also use the mouse to select the rectangular area you want to have a closer look at.
+	\item[f] Pan mode.
+	\item[g, h, i, j] Module highlighting and viewing.
+It is in general impossible to present mutually overlapping biclusters as single rectangles.
+They are made up of a collection of rectangles.
+The ordering algorithm used in the R package realigns the gene expression matrix in a way that maximizes the total area of the largest rectangle in every bicluster.
+The outlines of these parts are drawn in a slightly brighter color than the background, making them easily recognizable.
+	\item[g, h] Modules are highlighted as the user moves the moves over the gene expression data.
+The two check boxes allow you to choose between highlighting all the parts of a module (Filling) or alternatively only the largest rectangle (Outline).
+You can also turn it off completely.
+For data sets with many modules, it can be helpful to restrict highlighting to Outline.
+	\item[i, j] Similar to the highlighting, these two check boxes allow you to show either all the parts of the modules (Filling) or only the largest rectangles (Outline).
+By shift-clicking one of the check-boxes you can switch between showing only the modules or only the gene expression data.
+	\item[k] Sets the visibility of the modules layer.
+Moving the slider to the left fades out the gene expression data, focusing on the Biclusters, while towards the opposite direction, the gene expression data moves to the foreground.
+	\item[l] Realigns the windows at their initial positions.
+	\item[m] Puts the program in fullscreen mode.
+Note that due to security reasons, it is impossible to enter text in this mode.
+On Mac OS X, a bug in Flash player prevents you from exporting data in fullscreen mode.
+	\item[n] Opens the ExpressionView website, from where you can download sample files and tutorials.
+	\item[o] Description and dimensions of the data set.
+	\item[p] Modules navigator.
+The Global tab is always available and shows the complete gene expression data.
+Additional tabs appear as you open individual modules.
+To close a module, simply move the mouse over the tab and click the close button that appears.
+	\item[q, r, s] Selected genes, samples and modules. The highlighting reflects the selection in the tables (w). The selection is maintained when switching tabs (p). 
+	\item[q] Selected genes (=probes).
+	\item[r] Selected samples (=conditions).
+	\item[s] Selected modules (=biclusters).
+	\item[t] Info panel showing the data associated with the current mouse position.
+The GO and KEGG list contain the five most significant categories and pathways associated with the modules under the mouse pointer.
+	\item[u] Lists the selected genes, samples and modules, together with the intersecting modules.
+	\item[u1] Opens intersecting modules.
+	\item[u2] Clears the selection.
+	\item[v] Lists the selected GO categories and KEGG pathways
+	\item[w] List navigator.
+Note that depending on the view (p), the lists only show genes and samples contained in the currently viewed module.
+Modules can also be opened by double-clicking on the corresponding row.
+The Experiment tab contains a brief description of the data.
+	\item[x] Searches the tables for a given expression and restricts the view to the matching entries.
+The search function uses Perl-style {\bf regular expressions}.
+By default, the search functions is applied to the whole table.
+To restrict it to a particular column, shift-click the corresponding column header.
+	\item[z] Select a column header to sort the entries according to that column. Shift-click to restrict the search function to that column.
+\end{description}
 

### 2009-12-02 RExpressionView/inst/doc/ExpressionView.tutorial.Rnw
ExpressionView: Bugfix in Makefile. Minor changes.

@@
 The file association is the reason why we do not use the \variable{.xml} extension.
-A description of the XML layout can be found on the  href{http://www2.unil.ch/cbg/index.php?title=ExpressionView}{website}.
+A description of the XML layout can be found on the \href{http://www2.unil.ch/cbg/index.php?title=ExpressionView}{website}.
 Before exporting the data, the
@@
 <<export, eval=FALSE>>=
-OpenEV()
+LaunchEV()
 @
@@
 	\item[b] Exports the current view to a pdf file. The file also includes the title (o) of the gene expression data.
-	\item[c] Exports the data of the currently viewed module(=bicluster) to a CSV file, that can be opened as a spreadsheet.
+	\item[c] Exports the data of the currently viewed module (=bicluster) to a CSV file, that can be opened as a spreadsheet.
 	\item[d] In inspect mode, you can use the mouse to explore the gene expression data. The information about the data under the mouse pointer is shown in the Info Panel (t).
-	\item[e, f] Zoom and pan modes allow you to restrict the view to a particular part of the gene expression data. They behave like the functions used in graphic programs.
+	\item[e, f] Zoom and pan modes allow you to restrict the view to a particular part of the gene expression data.
 	\item[e] In zoom mode, you can also use keyboard shortcuts: {\bf a} to auto-zoom onto the modules and {\bf e} to see the whole data.
@@
 	\item[k] Sets the visibility of the modules layer.
-Moving the slider to the left fades out the gene expression data, focusing on the Biclusters, while towards the opposite direction, the gene expression data moves to the foreground.
+Moving the slider to the left fades out the gene expression data, thus focusing on the Biclusters, while towards the opposite direction, the gene expression data moves to the foreground.
 	\item[l] Realigns the windows at their initial positions.
@@
 
+\section{Using ExpressionView with non-gene expression data}
+While ExpressionView is designed to work with gene expression data available in the form of a \Rpackage{Bioconductor} \Rclass{ExpressionSet}, it can also be used to visualize other data.
+Let us for instance use in-silico data generated by the \Rpackage{isa} package with dimensions 50 $\times$ 500 containing 10 overlapping modules:
+<<in-silico>>=
+library(ExpressionView)
+# generate in-silico data with dimensions m x n containing M overlapping modules
+# and add some noise
+m <- 50
+n <- 500
+M <- 10
+data <- isa.in.silico(num.rows=m, num.cols=n, num.fact=M, noise=0.1, overlap.row=5)[[1]]
+modules <- isa(data)
+@
+
+The \Rfunction{ExportEV} uses the named list provided by the \variable{description} variable to label the data. First, let us annotate the rows and columns of the data set
+<<data set annotation>>=
+v <- vector("character"); for ( i in 1:dim(data)[1] ) { v[i] <- paste("row ", i, sep="") }; rownames(data) <- v
+v <- vector("character"); for ( i in 1:dim(data)[2] ) { v[i] <- paste("column ", i, sep="") }; colnames(data) <- v
+@
+
+Next, we assign the meta data associated with the rows of the data matrix. In this example we use 5 tags labelled ''row tag``:
+<<rowdata annotation>>=
+rowdata <- matrix(nrow=dim(data)[1], ncol=5)
+rownames(rowdata) <- rownames(data)
+v <- vector("character"); for ( i in 1:dim(rowdata)[2] ) { v[i] <- paste("row tag ", i, sep="") }; colnames(rowdata) <- v
+for ( i in 1:dim(rowdata)[1] ) { for ( j in 1:dim(rowdata)[2] ) { rowdata[i,j] <- paste("row description (", i, ", ", j, ")", sep="") }}
+@
+And similarly for the columns, using 10 "column tags``:
+<<coldata annotation>>=
+coldata <- matrix(nrow=dim(data)[2], ncol=10)
+rownames(coldata) <- colnames(data)
+v <- vector("character"); for ( i in 1:dim(coldata)[2] ) { v[i] <- paste("column tag ", i, sep="") }; colnames(coldata) <- v
+for ( i in 1:dim(coldata)[1] ) { for ( j in 1:dim(coldata)[2] ) { coldata[i,j] <- paste("column description (", i, ", ", j, ")", sep="") }}
+@
+
+To finish  the description, we add some general information and merge it with the above tables to get a single named list:
+<<in-silico data annotation>>=
+description <- list(
+experiment=list(
+	title="Title", 
+	xaxislabel="x-Axis Label",
+	yaxislabel="y-Axis Label",
+	name="Author", 
+	lab="Address", 
+	abstract="Abstract", 
+	url="URL", 
+	annotation="Annotation", 
+	organism="Organism"),
+coldata=coldata,
+rowdata=rowdata
+)
+@
+When dealing with gene expression data, the \variable{xaxislabel} is equal to ''genes`` and the  \variable{yaxislabel} is ''samples``. Finally, we export the data set to an ExpressionView file:
+<<in-silico data export, eval=FALSE>>=
+ExportEV(modules, data, filename="file.evf", description=description)
+@
+
+Simply load this file with the Flash applet and check where the various labels appear.
+
 \section{Session information}

### 2009-12-02 RExpressionView/R/export.R

@@ -36,6 +36,9 @@ ExportEV.ISAModules <- function(biclusters, eset, order, filename, norm, cutoff,
     if ( missing(filename) ) {
         con <- file(file.choose(TRUE), open="w")
     } else {
+        if ( filename == "" ) {
+            stop("indicate filename or omit argument to choose it interactively.", call.=FALSE)
+        }
         con <- file(filename, open="w", blocking = TRUE)
     }
     
@@ -378,6 +381,9 @@ ExportEV.list <- function(biclusters, eset, order, filename, norm, cutoff, descr
     if ( missing(filename) ) {
         con <- file(file.choose(TRUE), open="w")
     } else {
+        if ( filename == "" ) {
+            stop("indicate filename or omit argument to choose it interactively.", call.=FALSE)
+        }
         con <- file(filename, open="w", blocking = TRUE)
     }
     

### 2009-12-08 eisa/R/autogen.R
eisa: update HTML generation, because of the changes in the previous commit.

@@ -361,8 +361,8 @@ isa.autogen.module <- function(eset, modules, which, target.dir, template,
     }
     
     ## Create HTML
-    foo <- html.df(df, link=link, display=c("s", "e", "g", "g", "d", extra.display),
-                   digits=c(NA, 3, 4, 4, 4, rep(NA, length(extra.display))))
+    foo <- html.df(df, link=link, display=c("s", "s", "e", "g", "g", "d", extra.display),
+                   digits=c(NA, NA, 3, 4, 4, 4, rep(NA, length(extra.display))))
     foo <- foo[4:(length(foo)-1)]
     foo <- color.table(foo)
     

### 2009-12-08 eisa/inst/doc/EISA_biclust.Rnw
eisa vignettes updated.

@@
 
-The Iterative Signature Algorithm (ISA)~\cite{isa} is a biclustering
-method, that can efficiently find potentially overlapping biclusters
-(modules, according to the ISA terminology) in a matrix. The ISA is
-implemented in the \Rpackage{eisa} package~\cite{eisapackage} it uses
-standard BioConductor classes and includes a number of visualization
-tools as well.
+The Iterative Signature Algorithm (ISA)~\cite{sa,isa,isamod} is a
+biclustering method, that can efficiently find potentially overlapping
+biclusters (modules, according to the ISA terminology) in a
+matrix. The ISA is implemented in the \Rpackage{eisa} package. This
+package uses standard BioConductor classes and includes a number of
+visualization tools as well.
 
@@
 up the computation for our simple examples. We select the genes that are 
-annotated to involved in immune system processes, according to the Gene
+annotated to be involved in immune system processes, according to the Gene
 Ontology database.
@@
 some \Rpackage{eisa} functions, we add the name of the annotation package to 
-the parameters stored in the \Rclass{Biclust} object, this is always adwised.
+the parameters stored in the \Rclass{Biclust} object, this is always advised.
 The procedure makes use of the probe and sample names that are kept and stored 
@@
 
+\subsection{Enrichment analysis}
+
 Now we are able apply the usual \Rclass{ISAModules} methods to 
@@
 
-Doing some enrichment analysis is easy:
+Performing enrichment analysis is easy:
 <<biclust enrichment>>=
@@
 @
+
+\subsection{Heatmaps}
 
@@
   biclustering algorithm. The black squares denote the T-cell samples;
-  all samples in the module belong to T-cell samples.}
+  all samples in the module are from T-cell leukemia patients.}
 \label{fig:heatmap}
@@
 patients with T-cell leukemia.
+
+\subsection{Profile plots}
 
@@
 \end{figure}
+
+\subsection{Gene Ontology tree plots}
 
@@
 
+\subsection{HTML summary of the biclusters}
+
 The \Rfunction{ISAHTML} function creates a HTML overview of all modules.
@@
 @
+
+\subsection{Group-mean plots}
 
@@
 The conversion from \Rclass{ISAModules} to \Rclass{Biclust} can be done
-the usual way:
+the usual way, using the \Rfunction{as} function:
 <<isa to biclust>>=
@@
 @
+
+\subsection{Coherence of biclusters}
 
@@
 <<pairs,eval=FALSE>>=
-pairs( cbind(cV, aV, mV, sV, rob) )
+panel.low <- function(x, y) {
+  usr <- par("usr")
+  m <- c((usr[2]+usr[1])/2, (usr[4]+usr[3])/2)
+  text(m[1], m[2], adj=c(1/2,1/2), cex=1.5,
+       paste(sep="\n", "Correlation:", round(cor(x,y),2)))
+}
+pairs( cbind(cV, aV, mV, sV, rob), lower.panel=panel.low )
 @

### 2009-12-08 eisa/inst/doc/EISA_tutorial.Rnw
eisa vignettes updated.

@@
 
-\title{The Iterative Signature Algorithm for Gene Expression data}
+\title{The Iterative Signature Algorithm for Gene Expression Data}
 \author{G\'abor Cs\'ardi}
@@
 
-The Iterative Signature Algorithm (ISA) is biclustering method. The
-input of a biclustering method a matrix and the output is a set of
-biclusters that fulfill some criteria. A bicluster is a block of the
-potentially reordered input matrix.
-
-Most commonly, this algorithm is used on microarray expression data,
+The Iterative Signature Algorithm (ISA)~\cite{sa,isa,isamod} is a
+biclustering method. The input of a biclustering method is a matrix
+and the output is a set of biclusters that fulfill some criteria. A
+bicluster is a block of the potentially reordered input matrix.
+
+Most commonly, biclustering algorithms are used on microarray expression data,
 to find gene sets that are coexpressed across a subset of the original
@@
 The simplest way to run ISA is to choose the two threshold parameters
-and then call the \Rfunction{isa} function on the
+and then call the \Rfunction{ISA} function on the
 \Rclass{ExpressionSet} object. The threshold parameters tune the size
@@
 Since running ISA might take a couple of minutes and the results 
-depend on the random number generator used, the ISA run is commented out 
-and we just load a precomputed set of modules in this vignette.
-<<simple ISA>>=
+depend on the random number generator used, the ISA run is commented
+out from the next code block, and we just load a precomputed set of
+modules that is distributed with the \Rpackage{eisa} package.
+<<simple ISA,keep.source=TRUE>>=
 thr.gene <- 2.7
@@
 
-The \Rfunction{isa} function returns an \Rclass{ISAModules} object. By
+The \Rfunction{ISA} function returns an \Rclass{ISAModules} object. By
 typing in its name we can get a brief summary of the results:
@@
 \Rfunction{getOrganism} function returns the scientific name of the
-organism, \Rfunction{annotation} the name of the chip. For the former
-the appropriate annotation package must be installed.
+organism under study, \Rfunction{annotation} the name of the chip. For
+the former the appropriate annotation package must be installed.
 <<number of features and samples>>=
@@
 chr <- get(paste(annotation(modules), sep="", "CHR"))
-entrez <- sapply(mget(featureNames(modules), chr), 
+chr1features <- sapply(mget(featureNames(modules), chr), 
                  function(x) "1" %in% x)
-modules[entrez,]
+modules[chr1features,]
 @
@@
 the feature names coming from the \Rclass{ExpressionSet} object) in
-the modules. It returns a list, here we just print the first entry in
-the list.
+the modules. It returns a list, here we just print the first entry.
 <<list genes in modules>>=
@@
 Again, the sample names are taken from the \Rclass{ExpressionSet}
-object that was passed to \Rfunction{isa}:
+object that was passed to \Rfunction{ISA}:
 <<list conditions in modules>>=
@@
 \Rfunction{getFeatureMatrix} and \Rfunction{getSampleMatrix} functions
-are defined for this:
+are defined for this. The probes/samples that are not included in a
+module will have a zero score by definition.
 <<query all scores>>=
@@
 The \Rfunction{runData} function returns additional information about
-the ISA run, seed the \Rfunction{isa} manual page for details.
+the ISA run, see the \Rfunction{ISA} manual page for details.
 <<run data>>=
@@
 
-The \Rfunction{ISAGO} function requires the annotation package on the
+The \Rfunction{ISAGO} function requires the annotation package of the
 chip, e.g. for the ALL data, the \Rpackage{hgu95av2.db} package is
@@
 library(GO.db)
-mget(sigCategories(GO$BP)[[1]], GOTERM)
+mget(sigCategories(GO$BP)[[1]][1:3], GOTERM)
 @
@@
 \Rfunction{pvalues}, \Rfunction{geneCounts}, \Rfunction{oddsRatios},
-\Rfunction{exprectedCounts}, \Rfunction{universeCounts},
+\Rfunction{expectedCounts}, \Rfunction{universeCounts},
 \Rfunction{universeMappedCount}, \Rfunction{geneMappedCount},
@@
 library(KEGG.db)
-mget(sigCategories(KEGG)[[1]], KEGGPATHID2NAME)
-@
-
-The functions mentioned in the Gene ontology enrichment section can be
-used for KEGG, as well.
+mget(sigCategories(KEGG, p=0.001)[[1]], KEGGPATHID2NAME)
+@
+
+The functions mentioned in the Gene ontology enrichment section
+(\Rfunction{summary}, \Rfunction{pvalues}, \Rfunction{htmlReport},
+etc.) can be used for chromosome enrichment, as well.
 
@@
 
-The functions mentioned in the Gene ontology enrichment section can be
-used for chromosome enrichment, as well.
+The functions mentioned in the Gene ontology enrichment section
+(\Rfunction{summary}, \Rfunction{pvalues}, \Rfunction{htmlReport},
+etc.) can be used for chromosome enrichment, as well.
 
@@
 at a time. For some of these we will use the \Rpackage{biclust} R
-package.
+package, \cite{biclust}.
 
@@
 
-The \Rfunction{profile.plot} function has several options to set the
+The \Rfunction{profilePlot} function has several options to set the
 plot colors and styles, please see the manual for the details. This
@@
 and broader categories are subdivided into more specific
-subcategories. But the GO is not exactly a tree, as the same category
+subcategories. But the GO is not a tree graph, as the same category
 can be the subcategory of more than one broader categories: e.g. the
@@
 The \Rpackage{eisa} package provides functions to plot parts of
-the GO graph that is related to a transcription module. The
+the GO graph that are related to a transcription module. The
 \Rfunction{gograph} function creates an object that is a
@@
 device. (The graph attributes of an \Rpackage{igraph} graph can be
-queried with the `\texttt{\$}' selector:
+queried with the `\texttt{\$}' selector.)
 <<gograph object 2>>=
@@
 also abbreviates the names of the GO categories to make them fit on
-the plot. The graph object contains the full name of the category as
+the plot. The graph object contains the full names of the categories as
 well. The full and abbreviated names can be listed by querying the
@@
 It is clear that the \Sexpr{getNoFeatures(modules)[2]} genes included
-in this transcription module can separate B-cell and T-cell leukemia
-samples.
+in this transcription module cannot separate B-cell and T-cell
+leukemia samples. Section~\ref{sec:separators} shows good separator
+modules for this data set.
 
@@
 
-The Rfunction{ISAHTMLModules} generates a HTML page for every single
-module with the following information on it: 
+The \Rfunction{ISAHTMLModules} function generates a HTML page for
+every single module with the following information on it: 
 \begin{itemize}
@@
   separately for the three ontologies., These are produced by
-  calling \Rfunction{gograph}.
+  calling \Rfunction{gograph} and \Rfunction{gographPlot}.
 \item Tables for the enriched Gene Ontology terms, separately for
@@
 \item A table for the enriched KEGG pathways.
-\item A table for the enriched miRNA families.
+\item A table for the enriched miRNA families. (This is optional.)
 \item The list of genes in the module.
@@
 The rows of the summary table generated by \Rfunction{ISAHTMLTable}
-link to the module pages generated by \Rfunction{ISAHTMLModules}.
+link to the module pages generated by \Rfunction{ISAHTMLModules}, if
+the same \Rargument{target.dir} argument was used for both of them.
 
@@
 The previously mentioned visualization techniques focus on a single
-module only. The \Rpackage{ExpressionView} package visualizes a set of
-possibly overlapping biclusters together.
+module at at time. The \Rpackage{ExpressionView} package visualizes a
+set of possibly overlapping biclusters together.
 
@@
 $E_c$. $E_r$ is calculated by transposing $E$ and
-centering and scaling its genes (see the \Rfunction{scale} R function). $E_c$ is
-calculated by centering and scaling the genes of $E$. $E_r$ is
-used to calculate the sample signature of genes and $E_c$ is used
-to calculate the signature of the samples.
+centering and scaling its expression values for each sample (see the
+\Rfunction{scale} R function). $E_c$ is calculated by centering and
+scaling the genes of $E$. $E_r$ is used to calculate the sample
+signature of genes and $E_c$ is used to calculate the gene signature
+of the samples.
 
@@
 
-If the \Rfunction{isa} function was used to find the transcription
+If the \Rfunction{ISA} function was used to find the transcription
 modules, then the robustness measure is used automatically to filter the results.
@@
 
+If you analyze gene expression data, then we suggest using the
+interface provided in the \Rpackage{eisa} package. For other data, use
+the \Rpackage{isa2} package directly.
+
 \section{Finer control over ISA parameters}%
@@
 
-The \Rfunction{ISA} function takes care of all steps performed during
+The \Rfunction{ISA} function takes care of all steps performed in
 a modular study, and for each step it uses parameters, that work reasonably
@@
 filter is based on the inter-quantile ratio of the probesets'
-expression values, a robust measure of variance. Other possible
-filtering techniques include using the AffyMetrix present/absent calls
-produced by the \Rfunction{mas5calls} function of the \Rpackage{affy}
-package, but this requires the raw data, so in this vignette we use a
-simple method based on variance and minimum expression value: only
-probesets that have a variance of at least \variable{varLimit} and
-that have at least \variable{kLimit} samples with expression values
-over \variable{ALimit} are kept.
+expression values, a robust measure of variance.
+
+If other filters are desired, then these can be implemented by using
+the functions of the \Rpackage{genefilter} package directly.
+Possible filtering techniques include using the AffyMetrix
+present/absent calls produced by the \Rfunction{mas5calls} function of
+the \Rpackage{affy} package, but this requires the raw data, so in
+this vignette we use a simple method based on variance and minimum
+expression value: only probesets that have a variance of at least
+\variable{varLimit} and that have at least \variable{kLimit} samples
+with expression values over \variable{ALimit} are kept.
 <<filtering>>=
@@
 
-In this step we match the probesets to Entrez Id and remove the ones
-that don't map to any Entrez genes.
+In this step we match the probesets to Entrez identifiers and remove
+the ones that don't map to any Entrez gene.
 <<Entrez,cache=FALSE>>=
@@
 To reduce ambiguity in the interpretation of the results, we might
-also want to keep only single probeset for each Entrez gene.
+also want to keep only single probeset for each Entrez gene. The
+following code snipplet keeps the probeset with the highest variance.
 <<Entrez unique>>=
@@
 We perform the ISA iterations for our two sets of seeds separately.
+The two threshold parameters we use here were chosen after some
+experimentation; these result modules of the ``right'' size.
 <<iteration>>=
@@
 modules2.unique <- ISAUnique(ALL.normed, modules2)
+length(modules1.unique)
+length(modules2.unique)
 @
@@
 modules2.robust <- ISAFilterRobust(ALL.normed, modules2.unique)
+length(modules1.robust)
+length(modules2.robust)
 @
@@
 
-\subsection{Differentially regulated modules}
+\subsection{Differentially regulated modules}%
+\label{sec:separators}
 
@@
 sign2 <- which(p.adjust(tt2$p.value, "holm") < 0.05)
+sign1
+sign2
 @
@@
 
-As it turns out, one modules found using the smart seeds was also
-found with random seeds.
+As it turns out, the best separator module found using the smart seeds
+was also found with random seeds. This can be seen by calculating the
+correlation between the separator modules found in the two experiments.
+<<dysreg correlation>>=
+cor(getSampleMatrix(modules1.TB), getSampleMatrix(modules2.TB))
+cor(getFeatureMatrix(modules1.TB), getFeatureMatrix(modules2.TB))
+@
 
@@
 <<dysreg enriched categories>>=
-gocats <- unique(unlist(lapply(c(GO.dysreg1, GO.dysreg2), function(x) 
-                               unique(unlist(sigCategories(x))))))
-keggp  <- unique(unlist(lapply(list(KEGG.dysreg1, KEGG.dysreg2), function(x)
-                        unique(unlist(sigCategories(x))))))
+gocats <- unique(unlist(c(sigCategories(GO.dysreg1$BP),
+                          sigCategories(GO.dysreg1$CC),
+                          sigCategories(GO.dysreg1$MF),
+                          sigCategories(GO.dysreg2$BP),
+                          sigCategories(GO.dysreg2$CC),
+                          sigCategories(GO.dysreg2$MF))))
+keggp <- unique(unlist(c(sigCategories(KEGG.dysreg1), 
+                         sigCategories(KEGG.dysreg2))))
 sapply(mget(gocats, GOTERM), Term)
@@
 From the plot it seems that these two modules are essentially the
-same, one is the opposite of the others.
+same.
 <<check B1 same module>>=
@@
 
-Indeed, they are almost the same, their Pearson correlation is around
+Indeed, they are almost the same, their Pearson correlation is
 \Sexpr{round(B1.cor[1],3)} for the sample scores and
@@
 \bibliography{EISA}
-\nocite{*}
 

### 2009-12-09 isanote/modular.tex
Added Sven's changes to the ISA application note.

@@
 \section{Summary:}
-Large sets of data, like expression profile from many samples, require
+Large sets of data, like expression profiles from many samples, require
 analytic tools to reduce their complexity. Classical (bi-)clustering
@@
 other limitations of standard clustering algorithms. It aims to reduce the
-complexity of very large sets of data by decomposing it into so-called
+complexity of a very large set of data by decomposing it into so-called
 ``modules''. In the context of gene expression data these modules consist of
@@
 rely on the computation of correlation matrices (like many other tools), it
-is extremely fast even for very large datsets.
+is extremely fast even for very large datasets.
 
-In this short note, we introduce two BioConductor, \cite{BioC}, software
+In this short note, we introduce two BioConductor \citep{BioC}, software
 packages written in GNU~R: The~\Rpackage{isa2} includes a highly optimized
@@
 from any rectangular matrix of data. To be specific we consider here the
-case of transcriptomics data corresponding to a selection of gene-expression
+case of transcriptomics data corresponding to a set of gene-expression
 profiles from a collection of samples. The method has been described in
-detail in Refs~\cite{isamod,isa}. Here we only give a brief summary.
+detail in Refs~\citep{isamod,isa}. Here we only give a brief summary.
 
@@
   using the acute lymphoblastic leukemia data set, see
-  \cite{chiaretti04} and the \Rpackage{ALL} R package.
+  \citep{chiaretti04} and the \Rpackage{ALL} R package.
   \subfig{(A)} A condition plot for a simple bicluster. Each sample is
@@
 datasets. Several methods address this challenge,
-see e.g.~\cite{johnson07} for an algorithm that has a GNU R implementation.
+see e.g.~\citep{johnson07} for an algorithm that has a GNU R
+implementation. 
 % (S: Do we have a standard flag to call their software?)
@@
 \emph{The ISA iteration.}
-This step is essentially performing the ISA and finding the biclusters
-from the starting seeds.
+This step is the core of the ISA and finds biclusters from the
+starting seeds.
 
@@
 \emph{Merging the modules.}
-It is possible that more seeds converge to the same, or very similar
+It is possible that several seeds converge to the same, or very similar
 biclusters. This step eliminates such duplicates.
@@
 data. The \Rpackage{eisa} package builds on \Rpackage{isa2}. It adds
-support to standard BioConductor gene expression data structures; and
-contains gene expression specific visualization tools, see some of
-these in Fig.~\ref{fig:workflow}.
+support to standard BioConductor data structures; and
+contains gene expression specific visualization tools (see 
+Fig.~\ref{fig:workflow} for examples).
 
@@
 workflows. The \emph{simple workflow} involves a single R function
-call and runs all ISA steps (1-3 on Fig.~\ref{fig:workflow}.) with
+call and runs all ISA steps (1-3 in Fig.~\ref{fig:workflow}.) with
 their default parameters.
@@
 The \Rpackage{eisa} package implements a set of visualization
-techniques for biclusters, some of these are specific to ISA, others
-are not. Please see Fig.~\ref{fig:workflow} for some of these.
+techniques for biclusters, one of which is specific to ISA (see
+Fig.~\ref{fig:workflow} for examples).
 
-The \Rpackage{ExpressionView} R package, discussed in detail in
-another application note, implements an interactive tool
-for visualizing ISA (or other) overlapping biclusters.
+% The \Rpackage{ExpressionView} R package, discussed in detail in
+% another application note, implements an interactive tool
+% for visualizing ISA (or other) overlapping biclusters.
 
@@
 
-The \Rpackage{biclust} package, \cite{biclust}, implements a number of
+The \Rpackage{biclust} package, \citep{biclust}, implements a number of
 biclustering algorithms in a unified framework. The \Rpackage{eisa}
@@
 \paragraph{Funding\textcolon} The authors are greatful to the Swiss
-Institute of Bioinformatics for support.
+Institute of Bioinformatics, the Swiss National Science Foundation
+(3100AO-116323/1), and the European Framework Project 6 (through
+the EuroDia and AnEuploidy projects).
 \paragraph{Conflict of interest\textcolon} None declared.

### 2009-12-11 isanote/modular.tex
isanote: redesign the figure, update caption.

@@
 \includegraphics[width=\textwidth]{isa2workflow3}
-\caption{Work flow of a typical modular analysis study, with the
-  \Rpackage{eisa} package (left); and four bicluster visualization
-  techniques (right). The input of the \Rpackage{eisa} pipeline is a
+\caption{\subfig{(A)} Work flow of a typical modular analysis study, with the
+  \Rpackage{eisa} package. The input of the \Rpackage{eisa} pipeline is a
   BioConductor \Rclass{ExpressionSet} object. This is normalized to
-  $Z$-scores (1); the ISA iterations are initiated using ``smart'' or
-  random seeds (2); and the biclusters are then merged and filtered
-  based on the robustness measure (3). The \Rpackage{eisa} package
-  contains tools to perform many GO, KEGG, etc. enrichment calculations
-  quickly (4), and several bicluster visualization tools (5).
-  Subfigures \subfig{A}, \subfig{B}, \subfig{C} and \subfig{D} were
-generated
+  $Z$-scores (1); the ISA iterations are performed (2); and the
+  biclusters are then merged and filtered based on the robustness
+  measure (3). The \Rpackage{eisa} package contains tools to perform
+  many GO, KEGG, etc. enrichment calculations quickly (4), and several
+  bicluster visualization tools (5).
+  Subfigures \subfig{B} and \subfig{C} were generated
   using the acute lymphoblastic leukemia data set, see
   \citep{chiaretti04} and the \Rpackage{ALL} R package.
-  \subfig{(A)} A condition plot for a simple bicluster. Each sample is
-  represented as a bar, its height is its score in the bicluster.
-  Samples above the red (none in this case) and samples
-  below the green line are part of the biclusters. The coloring
-  indicates cases and controls.
-  \subfig{(B)} A module tree. Each rectangle is a bicluster; see the
+  \subfig{(B)} Heatmap for a single module, showing correlated
+  expression for the genes and samples. The red lines are the gene and
+  sample scores.
+  \subfig{(C)} A module tree. Each rectangle is a bicluster; see the
   definition of the edges in the text. The modules are colored
@@
   the modules is shown on the horizontal axes.
-  \subfig{(C)} Heatmap for a single module, showing correlated
-  expression for the genes and samples. The red lines are the gene and
-  sample scores.
-  \subfig{(D)} Profile plot, for a single module. On the top plot the
-  expression of all samples is plotted, the samples of the module
-  with black. The red and green lines show the mean of the two
-  groups. The bottom plot is the same for the genes of the module.
 }

### 2009-12-11 isanote/modular.tex
Updated isa application note.

@@
 implementation of the ISA and~\Rpackage{eisa} provides a convenient interface
-to run the ISA and visualize its output.
+to run the ISA, visualize its output and puts the biclusters into
+biological context.
 
 \section{Availability:}
-The ISA homepage is at \href{http://www.unil.ch/cbg/homepage/software.html}%
-{\url{http://www.unil.ch/cbg/homepage/software.html}}
+\href{http://www.unil.ch/cbg/ISA}%
+{\url{http://www.unil.ch/cbg/ISA}}
 \section{Contact:} \href{Sven.Bergmann@unil.ch}{Sven.Bergmann@unil.ch}
@@
 until the process converges to a stable set, which is referred to as a
-transcription module. This iterative procedure is typically performed to a
-large number of seeds. In the unsupervised approach these seeds are chosen
-randomly to sample uniformly the immense search space. We also implemented a
-semi-supervised method, to which we refer as ``smart seeding'', where the
-seeds a biased to start with certain sets of genes or samples for which
-there is prior knowledge.
+transcription module.
 
-The output of ISA is a collection of potentially overlapping modules. Each
-gene and each sample is attributed a score that reflects the strength of the
-association with the module.
+The output of ISA is a collection of potentially overlapping
+modules. In every module, each gene and each sample is attributed a
+score that reflects the strength of the association with the module.
 
@@
 \caption{\subfig{(A)} Work flow of a typical modular analysis study, with the
-  \Rpackage{eisa} package. The input of the \Rpackage{eisa} pipeline is a
-  BioConductor \Rclass{ExpressionSet} object. This is normalized to
-  $Z$-scores (1); the ISA iterations are performed (2); and the
-  biclusters are then merged and filtered based on the robustness
-  measure (3). The \Rpackage{eisa} package contains tools to perform
-  many GO, KEGG, etc. enrichment calculations quickly (4), and several
-  bicluster visualization tools (5).
+  \Rpackage{eisa} package. See the text for details.
   Subfigures \subfig{B} and \subfig{C} were generated
@@
 many aspects of transcriptional regulation one often combines several
-microarray experiments into a single dataset. In such a case it is important
-to properly normalize such data and reduce the bias due to constituent
-datasets. Several methods address this challenge,
+microarray experiments into a single dataset. In such a case
+additional data normalization is vital to reduce the bias due to
+constituent datasets. Several methods address this challenge,
 see e.g.~\citep{johnson07} for an algorithm that has a GNU R
@@
 % Normalization.
-\emph{ISA normalization.}
-In each iteration of the ISA weighted sums of expression levels over either
-genes or samples are computed. Since different genes
+\emph{ISA normalization.} (Step 1 on Fig.~\ref{fig:workflow}.)
+In each iteration of the ISA thresholded weighted sums of expression
+levels over either genes or samples are computed. Since different genes
 typically show different levels of base
@@
 % Random and smart seeding.
-\emph{Random and smart seeding.}
-As mentioned previously, ISA can be performed with random or smart
+\emph{Random and smart seeding, ISA iteration} (Step 2.)
+The ISA iteration is typically performed to a
+large number of seeds. In the unsupervised approach these seeds are chosen
+randomly to sample uniformly the immense search space. We also implemented a
+semi-supervised method, to which we refer as ``smart seeding'', where the
+seeds a biased to start with certain sets of genes or samples based on
+prior knowledge.As mentioned previously, ISA can be performed with random or smart
 seeds, based on the application.
 
-% The ISA iteration.
-\emph{The ISA iteration.}
-This step is the core of the ISA and finds biclusters from the
-starting seeds.
-
 % Merging the modules.
-\emph{Merging the modules.}
+\emph{Merging the modules.} (Step 3.)
 It is possible that several seeds converge to the same, or very similar
@@
 % The robustness measure and filtering the modules.
-\emph{Robustness of biclusters.}
+\emph{Robustness of biclusters.} (Step 3.)
 To access the significance of a bicluster, we designed a robustness

### 2009-12-11 isanote/modular.tex
More isa application note changes.

@@
 \includegraphics[width=\textwidth]{isa2workflow3}
-\caption{\subfig{(A)} Work flow of a typical modular analysis study, with the
-  \Rpackage{eisa} package. See the text for details.
+\caption{\subfig{(A)} Work flow of a typical modular analysis with the
+  \Rpackage{eisa} package. See text for details.
   Subfigures \subfig{B} and \subfig{C} were generated
@@
   \citep{chiaretti04} and the \Rpackage{ALL} R package.
-  \subfig{(B)} Heatmap for a single module, showing correlated
-  expression for the genes and samples. The red lines are the gene and
+  \subfig{(B)} Heatmap for a single module, showing coherent
+  expression of the genes across the samples. The red lines are the gene and
   sample scores.
-  \subfig{(C)} A module tree. Each rectangle is a bicluster; see the
-  definition of the edges in the text. The modules are colored
+  \subfig{(C)} Module tree. Each rectangle is a module; see the
+  definition of the edges in the text. Modules are colored
   according to their Gene Ontology enrichment $p$-values, the codes of
@@
   rectangles. The top-right corner shows the number of genes and
-  conditions in the bicluster. The gene thresholds used for finding
+  conditions in the module. The gene thresholds used for finding
   the modules is shown on the horizontal axes.
@@
 microarray experiments into a single dataset. In such a case
-additional data normalization is vital to reduce the bias due to
+additional data normalization is crucial to reduce the bias due to
 constituent datasets. Several methods address this challenge,
@@
 implementation.
-% (S: Do we have a standard flag to call their software?)
 
@@
 any information and may reflect ineffective array probes, etc. Since these
-genes are likely to contribute only noise to the analysis, we suggest
-removing them before running the ISA iteration.
-% (S: Describe how this is done, and if it requires setting a threshold!)
+genes are likely to contribute only noise to the
+analysis~\citep{hackstadt09}, we suggest removing them before running
+the module identification of the ISA.
 
@@
 \emph{ISA normalization.} (Step 1 on Fig.~\ref{fig:workflow}.)
-In each iteration of the ISA thresholded weighted sums of expression
-levels over either genes or samples are computed. Since different genes
+In each iteration the ISA computes thresholded weighted sums of expression
+levels over either genes or samples. Since different genes 
 typically show different levels of base
@@
 \emph{Random and smart seeding, ISA iteration} (Step 2.)
-The ISA iteration is typically performed to a
+The iterative procedure of the module identification is typically applied to a
 large number of seeds. In the unsupervised approach these seeds are chosen
@@
 semi-supervised method, to which we refer as ``smart seeding'', where the
-seeds a biased to start with certain sets of genes or samples based on
-prior knowledge.As mentioned previously, ISA can be performed with random or smart
-seeds, based on the application.
+seeds are biased to start with certain sets of genes or samples based on
+prior knowledge. The ISA can be performed with random or smart
+seeds, depending on the application.
 
 % Merging the modules.
-\emph{Merging the modules.} (Step 3.)
+\emph{Merging and filtering the modules.} (Step 3.)
 It is possible that several seeds converge to the same, or very similar
 biclusters. This step eliminates such duplicates.
-
-% The robustness measure and filtering the modules.
-\emph{Robustness of biclusters.} (Step 3.)
-To access the significance of a bicluster, we designed a robustness
+To access the significance of a module, we designed a robustness
 measure, that can be used to filter out spurious modules. This is done
-based on performing ISA on the scrambled input data.
+by applying the ISA to scrambled input data in order to obtain a
+reference (null) distribution for the significance scores.
 
@@
 threshold and the sample threshold. ISA modules can be organized into
-a directed graph, a module tree, in which there is an edge from module
-$A$ to module $B$, if the ISA converges to module $B$ from module $A$,
+a directed graph, to which we refer as ``module tree''. An edge from module
+$A$ to module $B$ indicates that the ISA converges to module $B$ from module $A$,
 with the same threshold parameters that were used to find module
 $B$. A module tree provides a hierarchical modular description of a
-data set.
+dataset.
 
@@
 workflows. The \emph{simple workflow} involves a single R function
-call and runs all ISA steps (1-3 in Fig.~\ref{fig:workflow}.) with
+call and runs all ISA steps (1-3 in Fig.~\ref{fig:workflow}) with
 their default parameters.
 
-The \emph{detailed workflow} means running every step of the modular
-analysis separately, possibly with non-default parameters. This allows
-the users to tailor the ISA completely according their needs.
+In the \emph{detailed workflow} every step of the modular analysis 
+is executed separately, possibly with non-default parameters. This allows
+the users to tailor the ISA according their needs.
 
@@
 The \Rpackage{eisa} package implements a set of visualization
-techniques for biclusters, one of which is specific to ISA (see
-Fig.~\ref{fig:workflow} for examples).
-
-% The \Rpackage{ExpressionView} R package, discussed in detail in
-% another application note, implements an interactive tool
-% for visualizing ISA (or other) overlapping biclusters.
-
-% \subsubsection{Connection to other software}
+techniques for modules (see Fig.~\ref{fig:workflow} for examples).
 
@@
 
-\paragraph{Funding\textcolon} The authors are greatful to the Swiss
+\paragraph{Funding\textcolon} The authors are grateful to the Swiss
 Institute of Bioinformatics, the Swiss National Science Foundation

### 2010-01-11 isanote/modular.tex
Final version of the ISA application note.

@@
 In this short note, we introduce two BioConductor \citep{BioC}, software
-packages written in GNU~R: The~\Rpackage{isa2} includes a highly optimized
-implementation of the ISA and~\Rpackage{eisa} provides a convenient interface
-to run the ISA, visualize its output and puts the biclusters into
+packages written in GNU~R: The~\Rpackage{isa2} package includes a highly optimized
+implementation of the ISA and the~\Rpackage{eisa} package provides a convenient interface
+to run the ISA, visualize its output and put the biclusters into
 biological context.

### 2010-01-11 expressionviewnote/expressionview.tex
ExpressionView note: Updated screenshot.
Updated ExpressionView application note author list.

@@
 
-\usepackage[pdftex,colorlinks,bookmarks=false,citecolor=blue,linkcolor=red,urlcolor=blue]{hyperref}
+\usepackage[dvipdfm,colorlinks,bookmarks=false,citecolor=blue,linkcolor=red,urlcolor=blue]{hyperref}
 \newcommand{\thefirstpage}{1} \newcommand{\thelastpage}{1}
@@
 \title[ExpressionView]{ExpressionView - an interactive viewer for biclusters identified in gene expression data}
-\author[Andreas L\"uscher, G\'abor Cs\'ardi, Aitana Morton de Lachapelle, Zolt\'an Kutalik, and Sven Bergmann]{Andreas L\"uscher$^1$, G\'abor Cs\'ardi$^{1,2}$, Aitana Morton de Lachapelle$^{1,2}$, Zolt\'an Kutalik$^{1,2}$, and Sven Bergmann$^{1,2}$\footnote{to whom correspondence should be addressed}}
+\author[Andreas L\"uscher, G\'abor Cs\'ardi, Aitana Morton de Lachapelle, Zolt\'an Kutalik, and Sven Bergmann]{Andreas L\"uscher$^1$, G\'abor Cs\'ardi$^{1,2}$, Aitana Morton de Lachapelle$^{1,2}$, Zolt\'an Kutalik$^{1,2}$, Bastian Peter$^{1,2}$ and Sven Bergmann$^{1,2}$\footnote{to whom correspondence should be addressed}}
 \address{
@@
 Cs{\'a}rdi,G., Kutalik,Z. and Bergmann,S. (2009)
-Eisa.
+Modular analysis of gene expression data with R.
 \href{http://dx.doi.org/}{\emph{Bioinformatics}, ...}

### 2010-01-11 expressionviewnote/expressionview.tex
Small corrections in the ExpressionView application note.

@@
 There is a variety of biclustering algorithms described in the literature~\citep{madeira04,prelic06}, several of which have been implemented as R packages.
-ExpressionView can treat results obtained by the iterative signature algorithm (ISA)~\citep{bergmann03,csardi09} and the methods available in the Biclust package~\citep{kaiser08}.
-Since the structure of biclustering results is independent of the algorithm, an extension to other methods is straightforward.
+ExpressionView can treat results obtained by the iterative signature algorithm (ISA)~\citep{bergmann03}%,csardi09
+and the methods available in the Biclust package~\citep{kaiser08}. Since the structure of biclustering results is independent of the algorithm, an extension to other methods is straightforward.
 
@@
 The quality of the order is defined as the sum over the maximal number of neighbouring elements in every cluster and the tasks thus consists in maximizing this quantity by changing the order of the elements.
-Starting from an initial configuration determined by similarity, this is achieved by either shifting well aligned subsequences of a given cluster, herby enlarging the longest contiguous part, or permuting the individual elements of a given subcluster.
+Starting from an initial configuration determined by similarity, this is achieved by either shifting well aligned subsequences of a given cluster, hereby enlarging the longest contiguous part, or permuting the individual elements of a given subcluster.
 
@@
 
-\paragraph{Funding\textcolon} This work was supported by ...
+\paragraph{Funding\textcolon} The authors are grateful to the Swiss
+Institute of Bioinformatics, the Swiss National Science Foundation
+(3100AO-116323/1), and the European Framework Project 6 (through
+the EuroDia and AnEuploidy projects).
 
@@
 
-\bibitem[Cs{\'a}rdi et~al.(2009)]{csardi09}
-Cs{\'a}rdi,G., Kutalik,Z. and Bergmann,S. (2009)
-Modular analysis of gene expression data with R.
-\href{http://dx.doi.org/}{\emph{Bioinformatics}, ...}
+%\bibitem[Cs{\'a}rdi et~al.(2009)]{csardi09}
+%Cs{\'a}rdi,G., Kutalik,Z. and Bergmann,S. (2009)
+%Modular analysis of gene expression data with R.
+%\href{http://dx.doi.org/}{\emph{Bioinformatics}, ...}
 

### 2010-01-11 expressionviewnote/expressionview.tex
More expressionview paper updates, sent to co-authors.

@@
 
-\usepackage[dvipdfm,colorlinks,bookmarks=false,citecolor=blue,linkcolor=red,urlcolor=blue]{hyperref}
+\usepackage[colorlinks,citecolor=blue,linkcolor=red,urlcolor=blue]{hyperref}
 \newcommand{\thefirstpage}{1} \newcommand{\thelastpage}{1}
 
-\copyrightyear{2009}
-\pubyear{2009}
+\copyrightyear{2010}
+\pubyear{2010}
 
@@
 \section{Availability:} 
-\href{http://www2.unil.ch/cbg/index.php?title=ExpressionView}{R package with open source license}
+\href{http://www.unil.ch/cbg/ExpressionView}{R
+  package with open source license, \url{http://www.unil.ch/cbg/ExpressionView}}
 
-\section{Contact:} 
-\href{andreas.luescher@a3.epfl.ch}{Andreas L\"uscher}
+\section{Contact:} \href{Sven.Bergmann@unil.ch}{Sven.Bergmann@unil.ch}
 
 \section{Supplementary information:}
-Screenshots, tutorials and sample data sets can be found on the ExpressionView \href{http://www2.unil.ch/cbg/index.php?title=ExpressionView}{website}.
+Screenshots, tutorials and sample data sets can be found on the ExpressionView \href{http://www.unil.ch/cbg/ExpressionView}{website}.
 
@@
 \end{figure*}
-For users who prefer to install the viewer as a standalone program, we also provide a platform independent Adobe AIR version that can be downloaded from the \href{http://www2.unil.ch/cbg/index.php?title=ExpressionView}{website}.
+For users who prefer to install the viewer as a standalone program, we also provide a platform independent Adobe AIR version that can be downloaded from the \href{http://www.unil.ch/cbg/ExpressionView}{website}.
 
@@
 There is a variety of biclustering algorithms described in the literature~\citep{madeira04,prelic06}, several of which have been implemented as R packages.
-ExpressionView can treat results obtained by the iterative signature algorithm (ISA)~\citep{bergmann03}%,csardi09
+ExpressionView can treat results obtained by the iterative signature algorithm (ISA)~\citep{bergmann03} %,csardi09
 and the methods available in the Biclust package~\citep{kaiser08}. Since the structure of biclustering results is independent of the algorithm, an extension to other methods is straightforward.
@@
 The actual gene expression data is rounded to two digit precision and encoded in Base64.
-The structure of the XML file is described in more detail on the ExpressionView \href{http://www2.unil.ch/cbg/index.php?title=ExpressionView}{website}.
+The structure of the XML file is described in more detail on the ExpressionView \href{http://www.unil.ch/cbg/ExpressionView}{website}.
 In most cases, these files are of the order of a few Megabytes and can thus easily be sent to co-workers.

### 2010-01-14 expressionviewnote/expressionview.tex
ExpressionView paper and docs corrections.

@@
 \section{Summary:}
-ExpressionView is an R package that provides an interactive environment to explore biclusters identified in gene expression data.
+ExpressionView is an R package that provides an interactive graphical environment to explore biclusters identified in gene expression data.
 A sophisticated ordering algorithm is used to present the biclusters in a visually appealing layout that provides an intuitive summary of the results.
@@
 Biclustering is an unsupervised learning technique used to analyse microarray data.
-It consists in identifying co-regulated groups of genes and conditions according to their expression profile.
-While there is a multitude of biclustering software available~\citep{madeira04}, packages with intuitive interfaces that allow the user to interactively explore the results are sparse~\citep{santamaria08}.
-ExpressionView is designed to close this gap, hereby facilitating the collaboration between bioinformaticians and life scientists without consolidated programming experience.
-Implemented as an R package, the user can apply all the powerful microarray analysis tools provided by the Bioconductor~\citep{gentleman04} project before exporting the data to a platform independent Adobe Flash applet that allows him and his co-workers to visualise the biclusters together with the underlying gene expression data.
+It consists of identifying co-regulated groups of genes and conditions according to their expression profile.
+While there is a multitude of biclustering software available~(see in \cite{madeira04}), packages with intuitive interfaces that allow the user to interactively explore the results are sparse~\citep{santamaria08}.
+ExpressionView is designed to close this gap, and facilitate the collaboration between bioinformaticians and life scientists without consolidated programming experience.
+Implemented as an R package, the user can apply all the powerful microarray analysis tools provided by the Bioconductor~\citep{gentleman04} project before exporting the data to a platform independent Adobe Flash applet that allows one to visualise the biclusters together with the underlying gene expression data.
 
 \section{Package design and workflow}
-The ExpressionView package contains two independent parts: An ordering algorithm implemented in C++ and a Adobe Flash applet written in ActionScript and Adobe Flex.
-Fig.~\ref{fig:workflow} schematically summarises the typical workflow that uses the genomic data available as a Bioconductor ExpressionSet and combines it with the biclustering results to produce an ExpressionView data file that can be explored with the Flash applet.
+The ExpressionView package contains two independent parts: An ordering algorithm implemented in C++ and an Adobe Flash applet written in ActionScript and Adobe Flex.
+Fig.~\ref{fig:workflow} schematically summarises the typical workflow that uses the gene expression data available as a Bioconductor ExpressionSet and combines it with the biclustering results to produce an ExpressionView data file that can be explored with the Flash applet.
 In the following, we describe the different steps in more detail.
@@
 In a second step, the rows and columns of the gene expression matrix are rearranged to produce an easily readable overview of the results.
-The last step consist in combining the gene expression data and its associated metadata with the results from the biclustering and produce an ExpressionView data file that can be explored with the interactive Flash applet.}\label{fig:workflow}
+The last step consists of combining the gene expression data and its associated metadata with the results from the biclustering and produce an ExpressionView data file that can be explored with the interactive Flash applet.}\label{fig:workflow}
 \end{figure*}
@@
 \subsection{Order}
-To present the tens of possibly overlapping biclusters in a visually appealing form, it is necessary to reorder the rows and columns of the gene expression matrix in such a way that biclusters form contiguous rectangles.
-Since for more than two mutually overlapping biclusters, it is in general impossible to find such an arrangement, one has to make concessions.
-In contrast to \cite{grothaus06}, who proposed to repeat rows and columns as necessary to achieve this goal, we prefer to optimise the arrangement within the original data by maximising the area of the largest contiguous biclusters.
+To present the dozens of possibly overlapping biclusters in a visually appealing form, it is necessary to reorder the rows and columns of the gene expression matrix in such a way that biclusters form contiguous rectangles.
+Since it is in general impossible to find such an arrangement for more than two mutually overlapping biclusters, one has to make concessions.
+In contrast to \cite{grothaus06}, who proposed to repeat rows and columns as necessary to achieve this goal, we prefer to optimise the arrangement within the original data by maximising the total area of the largest contiguous biclusters.
 
 Since this reordering is an interesting problem on its own, which to the best of our knowledge has not been studied in the literature, we briefly outline our strategy here.
-Noting that the rows and the columns can be ordered separately, the problem reduces to finding the optimal arrangement of a set of $n$ elements that are part of at least one of $m$ clusters.
-The quality of the order is defined as the sum over the maximal number of neighbouring elements in every cluster and the tasks thus consists in maximizing this quantity by changing the order of the elements.
-Starting from an initial configuration determined by similarity, this is achieved by either shifting well aligned subsequences of a given cluster, hereby enlarging the longest contiguous part, or permuting the individual elements of a given subcluster.
+Noting that the rows and the columns can be ordered independently, the problem reduces to finding the optimal arrangement of a set of $n$ elements that are part of at least one of $m$ clusters.
+The quality of the order is defined as the sum of the maximal number of neighbouring elements in every cluster and the aim is to maximize this quantity by changing the order of the elements.
+Starting from an initial configuration determined by hierarchical clustering, this is achieved by either shifting well aligned subsequences of a given cluster, hence enlarging the longest contiguous part, or permuting the individual elements of a given contiguous block.
 
@@
 We find that in almost every situation, the proposed algorithm finds an order that recovers more than 99\% of the score of the optimal solution and in most cases, it actually recovers the correct alignment.
-For random samples, which are more representative for actual gene expression data than orderable situations, execution time increases polynomially with the number of clusters $m$ as ${\mathcal O}(m^\alpha)$, where $\alpha \in [1.6, 2]$, almost independent of the number of elements $n$.
+For random samples, which are more representative for actual gene expression data than orderable situations, execution time increases polynomially with the number of clusters $m$ as ${\mathcal O}(m^\alpha)$, where $\alpha \in [1.6, 2]$, almost independently of the number of elements $n$.
 For a given number of clusters, we find ${\mathcal O}(n^\alpha)$, with $\alpha \in [2.5, 2.7]$.
@@
 The core application of this package is the Flash applet that allows the user to interactively explore the gene expression data and the biclusters.
-Relying on the Flash architecture has the advantage of producing a platform and browser independent applet that can be launched from any computer with even the most restrictive user rights.
+The Flash architecture has the advantage of producing a platform and browser independent applet that can be launched from any computer with even the most restrictive user rights.
 A screenshot is shown in Fig.~\ref{fig:workflow}.
@@
 Institute of Bioinformatics, the Swiss National Science Foundation
-(3100AO-116323/1), and the European Framework Project 6 (through
+(3100AO-116323/1) and the European Framework Project 6 (through
 the EuroDia and AnEuploidy projects).

### 2010-01-14 expressionviewnote/expressionview.tex
More ExpressionView paper updates.

@@
 
-\usepackage[colorlinks,citecolor=blue,linkcolor=red,urlcolor=blue]{hyperref}
+\usepackage{url}
+% \usepackage[colorlinks,citecolor=blue,linkcolor=red,urlcolor=blue]{hyperref}
 \newcommand{\thefirstpage}{1} \newcommand{\thelastpage}{1}
@@
 \application{}
-\title[ExpressionView]{ExpressionView - an interactive viewer for biclusters identified in gene expression data}
-\author[Andreas L\"uscher, G\'abor Cs\'ardi, Aitana Morton de Lachapelle, Zolt\'an Kutalik, and Sven Bergmann]{Andreas L\"uscher$^1$, G\'abor Cs\'ardi$^{1,2}$, Aitana Morton de Lachapelle$^{1,2}$, Zolt\'an Kutalik$^{1,2}$, Bastian Peter$^{1,2}$ and Sven Bergmann$^{1,2}$\footnote{to whom correspondence should be addressed}}
+\title[ExpressionView]{ExpressionView - an interactive viewer for
+  biclusters identified in gene expression data} 
+\author[Andreas L\"uscher, G\'abor Cs\'ardi, Aitana Morton de
+Lachapelle, Zolt\'an Kutalik, and Sven Bergmann]{Andreas
+  L\"uscher$^1$, G\'abor Cs\'ardi$^{1,2}$, Aitana Morton de
+  Lachapelle$^{1,2}$, Zolt\'an Kutalik$^{1,2}$, Bastian Peter$^{1,2}$
+  and Sven Bergmann$^{1,2}$\footnote{to whom correspondence should be
+    addressed}} 
 \address{
   $^{1}$Swiss Institute of Bioinformatics, Lausanne, Switzerland\\
-	$^{2}$Department of Medical Genetics, University of Lausanne, Lausanne, Switzerland
+  $^{2}$Department of Medical Genetics, University of Lausanne,
+  Lausanne, Switzerland 
 }
@@
 \section{Summary:}
-ExpressionView is an R package that provides an interactive graphical environment to explore biclusters identified in gene expression data.
-A sophisticated ordering algorithm is used to present the biclusters in a visually appealing layout that provides an intuitive summary of the results.
-From this overview, the user can select individual biclusters and access all the biologically relevant data associated with it.
-The package is aimed to facilitate the collaboration between bioinformaticians and life scientists who are not familiar with the R language.
+ExpressionView is an R package that provides an interactive graphical
+environment to explore biclusters identified in gene expression
+data. A sophisticated ordering algorithm is used to present the
+biclusters in a visually appealing layout that provides an intuitive
+summary of the results. From this overview, the user can select
+individual biclusters and access all the biologically relevant data
+associated with it. The package is aimed to facilitate the
+collaboration between bioinformaticians and life scientists.
 
 \section{Availability:} 
-\href{http://www.unil.ch/cbg/ExpressionView}{R
-  package with open source license, \url{http://www.unil.ch/cbg/ExpressionView}}
+%\href{http://www.unil.ch/cbg/ExpressionView}{R 
+%  package with open source license,
+\url{http://www.unil.ch/cbg/ExpressionView}
+%}
 
@@
 \section{Supplementary information:}
-Screenshots, tutorials and sample data sets can be found on the ExpressionView \href{http://www.unil.ch/cbg/ExpressionView}{website}.
+Screenshots, tutorials and sample data sets can be found on the
+ExpressionView \href{http://www.unil.ch/cbg/ExpressionView}{website}.
 
@@
 \section{Introduction}
-Biclustering is an unsupervised learning technique used to analyse microarray data.
-It consists of identifying co-regulated groups of genes and conditions according to their expression profile.
-While there is a multitude of biclustering software available~(see in \cite{madeira04}), packages with intuitive interfaces that allow the user to interactively explore the results are sparse~\citep{santamaria08}.
-ExpressionView is designed to close this gap, and facilitate the collaboration between bioinformaticians and life scientists without consolidated programming experience.
-Implemented as an R package, the user can apply all the powerful microarray analysis tools provided by the Bioconductor~\citep{gentleman04} project before exporting the data to a platform independent Adobe Flash applet that allows one to visualise the biclusters together with the underlying gene expression data.
+Biclustering is an unsupervised learning technique used to analyse
+microarray data. It consists of identifying co-regulated groups of
+genes and conditions according to their expression profile. While
+there is a multitude of biclustering software available~(see in
+\cite{madeira04}), packages with intuitive interfaces that allow the
+user to interactively explore the results are
+sparse~\citep{santamaria08}. ExpressionView is designed to close this
+gap, and facilitate the collaboration between bioinformaticians and
+life scientists without consolidated programming
+experience. Implemented as an R package, the user can apply all the
+powerful microarray analysis tools provided by the
+Bioconductor~\citep{gentleman04} project before exporting the data to
+a platform independent Adobe Flash applet that allows one to visualise
+the biclusters together with the underlying gene expression data.
 
 \section{Package design and workflow}
-The ExpressionView package contains two independent parts: An ordering algorithm implemented in C++ and an Adobe Flash applet written in ActionScript and Adobe Flex.
-Fig.~\ref{fig:workflow} schematically summarises the typical workflow that uses the gene expression data available as a Bioconductor ExpressionSet and combines it with the biclustering results to produce an ExpressionView data file that can be explored with the Flash applet.
-In the following, we describe the different steps in more detail.
+The ExpressionView package contains two independent parts: An ordering
+algorithm implemented in C++ and an Adobe Flash applet written in
+ActionScript and Adobe Flex. Fig.~\ref{fig:workflow} schematically
+summarises the typical workflow that uses the gene expression data
+available as a Bioconductor ExpressionSet and combines it with the
+biclustering results to produce an ExpressionView data file that can
+be explored with the Flash applet. In the following, we describe the
+different steps in more detail. 
 \begin{figure*}[!tpb]
 \centerline{\includegraphics[width=0.9\linewidth]{fig1-crop}}
-\caption{How to use ExpressionView: Starting from gene expression data in the form of a Bioconductor ExpressionSet, the user first runs a biclustering algorithm to determine co-regulated groups of genes and samples.
-In a second step, the rows and columns of the gene expression matrix are rearranged to produce an easily readable overview of the results.
-The last step consists of combining the gene expression data and its associated metadata with the results from the biclustering and produce an ExpressionView data file that can be explored with the interactive Flash applet.}\label{fig:workflow}
+\caption{How to use ExpressionView: Starting from gene expression data
+  in the form of a Bioconductor ExpressionSet, the user first runs a
+  biclustering algorithm to determine co-regulated groups of genes and
+  samples. In a second step, the rows and columns of the gene
+  expression matrix are rearranged to produce an easily readable
+  overview of the results. The last step consists of combining the
+  gene expression data and its associated metadata with the results
+  from the biclustering and produce an ExpressionView data file that
+  can be explored with the interactive Flash
+  applet.}\label{fig:workflow} 
 \end{figure*}
-For users who prefer to install the viewer as a standalone program, we also provide a platform independent Adobe AIR version that can be downloaded from the \href{http://www.unil.ch/cbg/ExpressionView}{website}.
+For users who prefer to install the viewer as a standalone program, we
+also provide a platform independent Adobe AIR version that can be
+downloaded from the
+\href{http://www.unil.ch/cbg/ExpressionView}{website}.
 
 \subsection{Gene expression data and biclusters}
-ExpressionView is designed to work with gene expression data in the form of a Bioconductor ExpressionSet.
-This class provides a user-friendly way to access the actual gene expression matrix and its associated metadata.
-There is a variety of biclustering algorithms described in the literature~\citep{madeira04,prelic06}, several of which have been implemented as R packages.
-ExpressionView can treat results obtained by the iterative signature algorithm (ISA)~\citep{bergmann03} %,csardi09
-and the methods available in the Biclust package~\citep{kaiser08}. Since the structure of biclustering results is independent of the algorithm, an extension to other methods is straightforward.
+ExpressionView is designed to work with gene expression data in the
+form of a Bioconductor ExpressionSet. This class provides a
+user-friendly way to access the actual gene expression matrix and its
+associated metadata. There is a variety of biclustering algorithms
+described in the literature~\citep{madeira04,prelic06}, several of
+which have been implemented as R packages. ExpressionView can treat
+results obtained by the iterative signature algorithm
+(ISA)~\citep{bergmann03} %,csardi09 
+and the methods available in the Biclust
+package~\citep{kaiser08}. Since the structure of biclustering results
+is independent of the algorithm, an extension to other methods is
+straightforward.
 
 \subsection{Order}
-To present the dozens of possibly overlapping biclusters in a visually appealing form, it is necessary to reorder the rows and columns of the gene expression matrix in such a way that biclusters form contiguous rectangles.
-Since it is in general impossible to find such an arrangement for more than two mutually overlapping biclusters, one has to make concessions.
-In contrast to \cite{grothaus06}, who proposed to repeat rows and columns as necessary to achieve this goal, we prefer to optimise the arrangement within the original data by maximising the total area of the largest contiguous biclusters.
+To present the dozens of possibly overlapping biclusters in a visually
+appealing form, it is necessary to reorder the rows and columns of the
+gene expression matrix in such a way that biclusters form contiguous
+rectangles. Since it is in general impossible to find such an
+arrangement for more than two mutually overlapping biclusters, one has
+to make concessions. In contrast to \cite{grothaus06}, who proposed to
+repeat rows and columns as necessary, we prefer
+to optimise the arrangement within the original data by maximising the
+total area of the largest contiguous biclusters.
 
-Since this reordering is an interesting problem on its own, which to the best of our knowledge has not been studied in the literature, we briefly outline our strategy here.
-Noting that the rows and the columns can be ordered independently, the problem reduces to finding the optimal arrangement of a set of $n$ elements that are part of at least one of $m$ clusters.
-The quality of the order is defined as the sum of the maximal number of neighbouring elements in every cluster and the aim is to maximize this quantity by changing the order of the elements.
-Starting from an initial configuration determined by hierarchical clustering, this is achieved by either shifting well aligned subsequences of a given cluster, hence enlarging the longest contiguous part, or permuting the individual elements of a given contiguous block.
+Since this reordering is an interesting problem on its own, which to
+the best of our knowledge has not been studied in the literature, we
+briefly outline our strategy here. Noting that the rows and the
+columns can be ordered independently, the problem reduces to finding
+the optimal arrangement of a set of $n$ elements that are part of at
+least one of $m$ clusters. The quality of the order is defined as the
+sum of the maximal number of neighbouring elements in every cluster
+and the aim is to maximize this quantity by changing the order of the
+elements. Starting from an initial configuration determined by
+hierarchical clustering, this is achieved by either shifting well
+aligned subsequences of a given cluster, hence enlarging the longest
+contiguous part, or permuting the individual elements of a given
+contiguous block.
 
-To determine the efficiency of this method, we have studied a large number of perfectly orderable, but initially scrambled, test cases.
-We find that in almost every situation, the proposed algorithm finds an order that recovers more than 99\% of the score of the optimal solution and in most cases, it actually recovers the correct alignment.
-For random samples, which are more representative for actual gene expression data than orderable situations, execution time increases polynomially with the number of clusters $m$ as ${\mathcal O}(m^\alpha)$, where $\alpha \in [1.6, 2]$, almost independently of the number of elements $n$.
-For a given number of clusters, we find ${\mathcal O}(n^\alpha)$, with $\alpha \in [2.5, 2.7]$.
+To determine the efficiency of this method, we have studied a large
+number of perfectly orderable, but initially scrambled, test cases. We
+find that in almost every situation, the proposed algorithm finds an
+order that recovers more than 99\% of the score of the optimal
+solution and in most cases, it recovers the correct
+alignment. For random samples, which are more representative for
+actual gene expression data than orderable situations, execution time
+increases polynomially with the number of clusters $m$ as ${\mathcal
+  O}(m^\alpha)$, where $\alpha \in [1.6, 2]$, almost independently of
+the number of elements $n$. For a given number of clusters, we find
+${\mathcal O}(n^\alpha)$, with $\alpha \in [2.5, 2.7]$.
 
 \subsection{Export}
-Once the optimal order is determined, the program rearranges the gene expression matrix accordingly and exports all the relevant information into an XML file.
-This format has the advantage of being self-explanatory and extensible.
-In addition, the files can be read and edited with any text editor.
-The actual gene expression data is rounded to two digit precision and encoded in Base64.
-The structure of the XML file is described in more detail on the ExpressionView \href{http://www.unil.ch/cbg/ExpressionView}{website}.
-In most cases, these files are of the order of a few Megabytes and can thus easily be sent to co-workers.
+Once the optimal order is determined, the program rearranges the gene
+expression matrix accordingly and exports all the relevant information
+into an XML file. This format has the advantage of being
+self-explanatory and extensible. In addition, the files can be read
+and edited with any text editor. The actual gene expression data is
+rounded to two digit precision and encoded in Base64. The structure of
+the XML file is described in more detail on the ExpressionView
+\href{http://www.unil.ch/cbg/ExpressionView}{website}. In most cases,
+these files are of the order of a few Megabytes and can thus easily be
+sent to co-workers.
 
 \subsection{Visualisation}
-The core application of this package is the Flash applet that allows the user to interactively explore the gene expression data and the biclusters.
-The Flash architecture has the advantage of producing a platform and browser independent applet that can be launched from any computer with even the most restrictive user rights.
-A screenshot is shown in Fig.~\ref{fig:workflow}.
-The interface is divided in two parts: On the left-hand side, the user finds the gene expression data in the common heat map form, on top of which the biclusters are overlaid.
-On the right-hand side, the metadata associated with the gene expression data and the results of the enrichment calculations [GO (Gene Ontology)~\citep{ashburner00} and KEGG (Kyoto Encyclopedia of Genes and Genomes)~\citep{kanehisa04}] are shown.
-Wherever possible, these elements are linked to the corresponding databases.
-The interface essentially behaves as an image viewer, allowing the user to zoom and pan around the expression data, getting instant feedback on the selected item.
-The currently selected region can be exported to a PDF file, helpful to prepare figures for presentations or articles, or a simple text file.\\
+The core application of this package is the Flash applet that allows
+the user to interactively explore the gene expression data and the
+biclusters. The Flash architecture has the advantage of producing a
+platform and browser independent applet that can be launched from any
+computer with even the most restrictive user rights. A screenshot is
+shown in Fig.~\ref{fig:workflow}. The interface is divided in two
+parts: On the left-hand side, the user finds the gene expression data
+in the common heat map form, on top of which the biclusters are
+overlaid. On the right-hand side, the metadata associated with the
+gene expression data and the results of the enrichment calculations
+(Gene Ontology~\citep{ashburner00} and Kyoto Encyclopedia
+of Genes and Genomes ~\citep{kanehisa04}] are shown. Wherever
+possible, these elements are linked to the corresponding
+databases. The interface essentially behaves as an image viewer,
+allowing the user to zoom and pan around the expression data, getting
+instant feedback on the selected item. The currently selected region
+can be exported to a PDF file, or a simple text file.
 
-%\section*{Acknowledgement}
+\section*{Acknowledgement}
 
@@
 Gene Ontology: tool for the unification of biology.
-\href{http://dx.doi.org/10.1038/75556}{\emph{Nature Genetics}, {\bf 25}, 25 -- 29.}
+\href{http://dx.doi.org/10.1038/75556}{\emph{Nature Genetics}, {\bf 25}, 25--29.}
 
@@
 Biclustering algorithms for biological data analysis: A survey.
-\href{http://dx.doi.org/10.1109/TCBB.2004.2}{\emph{IEEE/ACM Transactions on Computational Biology and Bioinformatics}, {\bf 1}, 24 -- 45.}
+\href{http://dx.doi.org/10.1109/TCBB.2004.2}{\emph{IEEE/ACM Transactions on Computational Biology and Bioinformatics}, {\bf 1}, 24--45.}
 
@@
 A systematic comparison and evaluation of biclustering methods for gene expression data 
-\href{http://dx.doi.org/10.1093/bioinformatics/bt1060}{\emph{Bioinformatics}, {\bf 22}(9), 1122 -- 1129.}
+\href{http://dx.doi.org/10.1093/bioinformatics/bt1060}{\emph{Bioinformatics}, {\bf 22}(9), 1122--1129.}
 
@@
 BicOverlapper: A tool for bicluster visualization. 
-\href{http://dx.doi.org/10.1093/bioinformatics/btn076}{\emph{Bioinformatics}, {\bf 24}, 1212 -- 1213.}
+\href{http://dx.doi.org/10.1093/bioinformatics/btn076}{\emph{Bioinformatics}, {\bf 24}, 1212--1213.}
 

### 2010-01-14 expressionviewnote/expressionview.tex
Even more ExpressionView paper updates.

@@
 \bibitem[Ashburner et~al.(2000)]{ashburner00}
-Ashburner,M. \emph{et al.} (2000)
+Ashburner,~M. \emph{et al.} (2000)
 Gene Ontology: tool for the unification of biology.
@@
 \bibitem[Bergmann et~al.(2003)]{bergmann03}
-Bergmann,S., Ihmels,J. and Barkai,N. (2003)
+Bergmann,~S., Ihmels,~J. and Barkai,~N. (2003)
 Iterative signature algorithm for the analysis of large-scale gene expression data.
@@
 \bibitem[Gentleman et~al.(2004)]{gentleman04}
-Gentleman,R.C., \emph{et al.} (2004)
+Gentleman,~R.C., \emph{et al.} (2004)
 Bioconductor: open software development for computational biology and bioinformatics.
@@
 \bibitem[Grothaus et~al.(2006)]{grothaus06}
-Grothaus,G., Mufti,A., and Murali,TM. (2006)
+Grothaus,~G., Mufti,~A., and Murali,~TM. (2006)
 Automatic layout and visualization of biclusters.
@@
 \bibitem[Kaiser and Leisch(2008)]{kaiser08}
-Kaiser,S. and Leisch,F. (2008)
+Kaiser,~S. and Leisch,~F. (2008)
 A toolbox for bicluster analysis in R.
@@
 \bibitem[Kanehisa et~al.(2004)]{kanehisa04}
-Kanehisa,M. \emph{et al.} (2004)
+Kanehisa,~M. \emph{et al.} (2004)
 The KEGG resource for deciphering the genome.
@@
 \bibitem[Madeira and Oliveira(2004)]{madeira04}
-Madeira,S.C. and Oliveira,A.L. (2004)
+Madeira,~S.C. and Oliveira,~A.L. (2004)
 Biclustering algorithms for biological data analysis: A survey.
@@
 \bibitem[Prelic et~al.(2006)]{prelic06}
-Prelic,A., \emph{et~al.} 
+Prelic,~A., \emph{et~al.} 
 A systematic comparison and evaluation of biclustering methods for gene expression data 
@@
 \bibitem[Santamar\'ia et~al.(2008)]{santamaria08}
-Santamar\'ia,R., Ther\'on,R. and Quintales,L. (2008)
+Santamar\'ia,~R., Ther\'on,~R. and Quintales,~L. (2008)
 BicOverlapper: A tool for bicluster visualization. 

### 2010-01-14 expressionviewnote/expressionview.tex
Added Sven's changes to ExpressionView note.

@@
 \title[ExpressionView]{ExpressionView - an interactive viewer for
-  biclusters identified in gene expression data} 
+  modules identified in gene expression data} 
 \author[Andreas L\"uscher, G\'abor Cs\'ardi, Aitana Morton de
 Lachapelle, Zolt\'an Kutalik, and Sven Bergmann]{Andreas
-  L\"uscher$^1$, G\'abor Cs\'ardi$^{1,2}$, Aitana Morton de
-  Lachapelle$^{1,2}$, Zolt\'an Kutalik$^{1,2}$, Bastian Peter$^{1,2}$
+  L\"uscher$^{1\dagger}$, G\'abor Cs\'ardi$^{1,2\dagger}$, Aitana Morton de
+  Lachapelle$^{1,2\dagger}$, Zolt\'an Kutalik$^{1,2\dagger}$, 
+  Bastian Peter$^{1,2}$
   and Sven Bergmann$^{1,2}$\footnote{to whom correspondence should be
@@
   $^{2}$Department of Medical Genetics, University of Lausanne,
-  Lausanne, Switzerland 
+  Lausanne, Switzerland\\
+  +: equal contribution
 }
@@
 ExpressionView is an R package that provides an interactive graphical
-environment to explore biclusters identified in gene expression
+environment to explore transcription modules identified in gene expression
 data. A sophisticated ordering algorithm is used to present the
-biclusters in a visually appealing layout that provides an intuitive
+modules with the expression in a visually appealing layout that provides an intuitive
 summary of the results. From this overview, the user can select
-individual biclusters and access all the biologically relevant data
-associated with it. The package is aimed to facilitate the
-collaboration between bioinformaticians and life scientists.
+individual modules and access all the biologically relevant metadata
+associated with it.
 
@@
 ExpressionView \href{http://www.unil.ch/cbg/ExpressionView}{website}. 
-
 \end{abstract}
 
+\vspace*{-10pt}
 \section{Introduction}
-Biclustering is an unsupervised learning technique used to analyse
-microarray data. It consists of identifying co-regulated groups of
-genes and conditions according to their expression profile. While
+Biclustering is an unsupervised data analysis method which is
+frequently used to explore microarray data. Biclustering algorithms
+identify co-expressed groups of
+genes and conditions (samples) where this co-expression occurs from
+collections of expression profiles. We refer to such groups (or
+biclusters) as modules. While
 there is a multitude of biclustering software available~(see in
-\cite{madeira04}), packages with intuitive interfaces that allow the
-user to interactively explore the results are
+\cite{madeira04,ihmels04}), packages with intuitive interfaces that
+allow for an interactive exploration of the results are
 sparse~\citep{santamaria08}. ExpressionView is designed to close this
@@
 Bioconductor~\citep{gentleman04} project before exporting the data to
-a platform independent Adobe Flash applet that allows one to visualise
-the biclusters together with the underlying gene expression data.
+a platform independent Adobe Flash applet. This stand-alone applet
+provides and integrated, intuitive and interactive visualization of
+the modules together with the underlying gene expression data and
+additional information on the genes and conditions.
 
@@
 \begin{figure*}[!tpb]
-\centerline{\includegraphics[width=0.9\linewidth]{fig1-crop}}
+\centerline{\includegraphics[width=0.7\linewidth]{fig1-crop}}
 \caption{How to use ExpressionView: Starting from gene expression data
   in the form of a Bioconductor ExpressionSet, the user first runs a
-  biclustering algorithm to determine co-regulated groups of genes and
-  samples. In a second step, the rows and columns of the gene
+  biclustering algorithm to determine co-expressed groups of genes and
+  the relevant samples. In a second step, the rows and columns of the gene
   expression matrix are rearranged to produce an easily readable
@@
 
-\subsection{Gene expression data and biclusters}
+\subsection{Gene expression data and modules}
 ExpressionView is designed to work with gene expression data in the
@@
 which have been implemented as R packages. ExpressionView can treat
-results obtained by the iterative signature algorithm
-(ISA)~\citep{bergmann03} %,csardi09 
-and the methods available in the Biclust
+results obtained by the Iterative Signature Algorithm
+(ISA)~\citep{bergmann03} (see \citep{csardi09} for its R
+implementation) and any of the methods available in the Biclust
 package~\citep{kaiser08}. Since the structure of biclustering results
@@
 
-\subsection{Order}
-To present the dozens of possibly overlapping biclusters in a visually
-appealing form, it is necessary to reorder the rows and columns of the
+\subsection{Reordering genes and conditions}
+To present the collection of possibly overlapping modules in a visually
+appealing form, it is necessary to reorder the rows (genes) and
+columns (conditions) of the
 gene expression matrix in such a way that biclusters form contiguous
 rectangles. Since it is in general impossible to find such an
-arrangement for more than two mutually overlapping biclusters, one has
-to make concessions. In contrast to \cite{grothaus06}, who proposed to
-repeat rows and columns as necessary, we prefer
-to optimise the arrangement within the original data by maximising the
-total area of the largest contiguous biclusters.
+arrangement for more than two mutually overlapping modules, we propose
+here an approximate solution that optimizes the arrangement within the
+original data by maximising the total area of the largest contiguous
+module subsets. (An alternative would be to repeat rows and columns as
+necessary~\cite{grothaus06}, but for many modules this will results in
+a very large expression matrix.)
 
-Since this reordering is an interesting problem on its own, which to
-the best of our knowledge has not been studied in the literature, we
-briefly outline our strategy here. Noting that the rows and the
-columns can be ordered independently, the problem reduces to finding
-the optimal arrangement of a set of $n$ elements that are part of at
-least one of $m$ clusters. The quality of the order is defined as the
-sum of the maximal number of neighbouring elements in every cluster
-and the aim is to maximize this quantity by changing the order of the
-elements. Starting from an initial configuration determined by
-hierarchical clustering, this is achieved by either shifting well
-aligned subsequences of a given cluster, hence enlarging the longest
-contiguous part, or permuting the individual elements of a given
-contiguous block.
+This optimization task is an interesting problem on its own, which to
+the best of our knowledge has not been studied in the literature. Thus
+we briefly outline our strategy here (see our website for details). 
+The reordering of the rows and the columns do not depend on each
+other, so the same optimization method can be applied sequentially,
+say first to rows and then to columns. For a given order of the
+elements (either genes or conditions) we compute for each module the
+size of the largest contiguous sequence of elements (i.e. the maximal
+number of neighboring elements $N^{max}_i$). Then as a measure of the
+quality $Q$ of the order we sum this quantity over all modules
+($Q=\sum_i N^{max}_i$). The optimization procedure for $Q$ starts from
+some initial order. Two operations are then applied to this order: (1)
+Permutations that exchange two elements within a module and (2) shifts
+of a sequence of elements within a module to a different position. We
+use a greedy iterative scheme that performs these operations at
+randomly selected positions and keeps the new order if it improves
+$Q$. The algorithm stops if after a given number of operations no
+significant improvement of $Q$ is achieved.
 
@@
 alignment. For random samples, which are more representative for
-actual gene expression data than orderable situations, execution time
-increases polynomially with the number of clusters $m$ as ${\mathcal
-  O}(m^\alpha)$, where $\alpha \in [1.6, 2]$, almost independently of
-the number of elements $n$. For a given number of clusters, we find
-${\mathcal O}(n^\alpha)$, with $\alpha \in [2.5, 2.7]$.
+actual gene expression data, the execution time increases polynomially
+with the number of clusters $m$ as ${\mathcal O}(m^\alpha)$, where
+$\alpha \in [1.6, 2]$, almost independent of the number of elements
+$n$. For a given number of clusters, we find ${\mathcal O}(n^\alpha)$,
+with $\alpha \in [2.5, 2.7]$.
 
@@
 these files are of the order of a few Megabytes and can thus easily be
-sent to co-workers.
+sent by e-mail.
 
@@
 the user to interactively explore the gene expression data and the
-biclusters. The Flash architecture has the advantage of producing a
-platform and browser independent applet that can be launched from any
-computer with even the most restrictive user rights. A screenshot is
-shown in Fig.~\ref{fig:workflow}. The interface is divided in two
-parts: On the left-hand side, the user finds the gene expression data
-in the common heat map form, on top of which the biclusters are
-overlaid. On the right-hand side, the metadata associated with the
-gene expression data and the results of the enrichment calculations
-(Gene Ontology~\citep{ashburner00} and Kyoto Encyclopedia
-of Genes and Genomes ~\citep{kanehisa04}] are shown. Wherever
-possible, these elements are linked to the corresponding
-databases. The interface essentially behaves as an image viewer,
-allowing the user to zoom and pan around the expression data, getting
-instant feedback on the selected item. The currently selected region
-can be exported to a PDF file, or a simple text file.
+modules. The Flash architecture has the advantage of
+producing a platform and browser independent applet that can be
+launched from any computer with even the most restrictive user
+rights. A screenshot is shown in Fig.~\ref{fig:workflow}. The
+interface is divided in two parts: On the left-hand side, the user
+finds the gene expression data in the common heat map form, on top of
+which the modules are overlaid. On the right-hand side, the metadata
+associated with the gene expression data and the results of the
+enrichment calculations, GO (Gene Ontology)~\citep{ashburner00} and
+KEGG (Kyoto Encyclopedia of Genes and Genomes)~\citep{kanehisa04}, are
+shown. Wherever possible, these elements are linked to the
+corresponding databases. The interface essentially behaves as an image
+viewer, allowing the user to zoom and pan around the expression data,
+getting instant feedback on the selected item. The currently selected
+region can be exported to a PDF file, helpful to prepare figures for
+presentations or articles, or a simple text file.
 
@@
 
+\bibitem[Cs{\'a}rdi(2009)]{csardi09}
+G\'abor Cs\'ardi (2009)
+isa2: The Iterative Signature Algorithm. 
+{R package version 0.2.}
+
 \bibitem[Gentleman et~al.(2004)]{gentleman04}
@@
 \href{http://dx.doi.org/10.1186/1748-7188-1-15}{\emph{Algorithms for Molecular Biology}, {\bf 1}, 15.}
+
+\bibitem[Ihmels and Bergmann(2004)]{ihmels04}
+Ihmels JH and Bergmann S. Challenges and prospects in the analysis
+of large-scale gene expression data. (2004)
+{\emph{Brief Bioinform} \bf{5} 313--327.}
 

### 2010-01-14 expressionviewnote/expressionview.tex
EV note updates, getting close.....

@@
 
+\usepackage{amstext}
 \usepackage{url}
@@
   Bastian Peter$^{1,2}$
-  and Sven Bergmann$^{1,2}$\footnote{to whom correspondence should be
-    addressed}}  
+  and Sven Bergmann$^{1,2}$}
 \address{
@@
 \begin{abstract}
-
 \section{Summary:}
@@
 individual modules and access all the biologically relevant metadata
-associated with it.
+associated with them.
 
@@
 
-\vspace*{-10pt}
+\vspace*{-7pt}
 \section{Introduction}
@@
 genes and conditions (samples) where this co-expression occurs from
-collections of expression profiles. We refer to such groups (or
-biclusters) as modules. While
+collections of expression profiles. We refer to such groups as modules. While 
 there is a multitude of biclustering software available~(see in
@@
 Bioconductor~\citep{gentleman04} project before exporting the data to
-a platform independent Adobe Flash applet. This stand-alone applet
-provides and integrated, intuitive and interactive visualization of
+a platform independent Adobe Flash applet. This applet
+provides an integrated, intuitive and interactive visualization of
 the modules together with the underlying gene expression data and
@@
 
+\vspace*{-7pt}
 \section{Package design and workflow}
@@
 For users who prefer to install the viewer as a standalone program, we
-also provide a platform independent Adobe AIR version that can be
-downloaded from the
+also provide an Adobe AIR version that can be downloaded from the
 \href{http://www.unil.ch/cbg/ExpressionView}{website}.
@@
 which have been implemented as R packages. ExpressionView can treat
-results obtained by the Iterative Signature Algorithm
-(ISA)~\citep{bergmann03} (see \citep{csardi09} for its R
-implementation) and any of the methods available in the Biclust
-package~\citep{kaiser08}. Since the structure of biclustering results
-is independent of the algorithm, an extension to other methods is
-straightforward.
+results obtained by the Iterative Signature
+Algorithm~\citep{bergmann03,csardi09} and any of the methods available
+in the Biclust package~\citep{kaiser08}. Since the structure of
+biclustering results is independent of the algorithm, an extension to
+other methods is straightforward.
 
@@
 module subsets. (An alternative would be to repeat rows and columns as
-necessary~\cite{grothaus06}, but for many modules this will results in
+necessary~\citep{grothaus06}, but for many modules this will results in
 a very large expression matrix.)
@@
 size of the largest contiguous sequence of elements (i.e. the maximal
-number of neighboring elements $N^{max}_i$). Then as a measure of the
+number of neighboring elements $N^\text{max}_i$). Then as a measure of the
 quality $Q$ of the order we sum this quantity over all modules
-($Q=\sum_i N^{max}_i$). The optimization procedure for $Q$ starts from
+($Q=\sum_i N^\text{max}_i$). The optimization procedure for $Q$ starts from
 some initial order. Two operations are then applied to this order: (1)
@@
 associated with the gene expression data and the results of the
-enrichment calculations, GO (Gene Ontology)~\citep{ashburner00} and
-KEGG (Kyoto Encyclopedia of Genes and Genomes)~\citep{kanehisa04}, are
-shown. Wherever possible, these elements are linked to the
-corresponding databases. The interface essentially behaves as an image
-viewer, allowing the user to zoom and pan around the expression data,
-getting instant feedback on the selected item. The currently selected
-region can be exported to a PDF file, helpful to prepare figures for
-presentations or articles, or a simple text file.
+enrichment calculations, GO~\citep{ashburner00} and
+KEGG~\citep{kanehisa04}, are shown. Wherever possible, these elements
+are linked to the corresponding databases. The interface essentially
+behaves as an image viewer, allowing the user to zoom and pan around
+the expression data, getting instant feedback on the selected
+item. The currently selected region can be exported to a PDF file,
+helpful to prepare figures for presentations or articles, or a simple
+text file.
 
+\vspace*{-7pt}
 \section*{Acknowledgement}
@@
 
-\begin{thebibliography}{6}
-
-\bibitem[Ashburner et~al.(2000)]{ashburner00}
-Ashburner,~M. \emph{et al.} (2000)
-Gene Ontology: tool for the unification of biology.
-\href{http://dx.doi.org/10.1038/75556}{\emph{Nature Genetics}, {\bf 25}, 25--29.}
-
-\bibitem[Bergmann et~al.(2003)]{bergmann03}
-Bergmann,~S., Ihmels,~J. and Barkai,~N. (2003)
-Iterative signature algorithm for the analysis of large-scale gene expression data.
-\href{http://dx.doi.org/10.1103/PhysRevE.67.031902 }{\emph{Physical Review E}, {\bf 67}, 031902.}
-
-%\bibitem[Cs{\'a}rdi et~al.(2009)]{csardi09}
-%Cs{\'a}rdi,G., Kutalik,Z. and Bergmann,S. (2009)
-%Modular analysis of gene expression data with R.
-%\href{http://dx.doi.org/}{\emph{Bioinformatics}, ...}
-
-\bibitem[Cs{\'a}rdi(2009)]{csardi09}
-G\'abor Cs\'ardi (2009)
-isa2: The Iterative Signature Algorithm. 
-{R package version 0.2.}
-
-\bibitem[Gentleman et~al.(2004)]{gentleman04}
-Gentleman,~R.C., \emph{et al.} (2004)
-Bioconductor: open software development for computational biology and bioinformatics.
-\href{http://dx.doi.org/10.1186/gb-2004-5-10-r80}{\emph{Genome Biology}, {\bf 5}, R80.}
-
-\bibitem[Grothaus et~al.(2006)]{grothaus06}
-Grothaus,~G., Mufti,~A., and Murali,~TM. (2006)
-Automatic layout and visualization of biclusters.
-\href{http://dx.doi.org/10.1186/1748-7188-1-15}{\emph{Algorithms for Molecular Biology}, {\bf 1}, 15.}
-
-\bibitem[Ihmels and Bergmann(2004)]{ihmels04}
-Ihmels JH and Bergmann S. Challenges and prospects in the analysis
-of large-scale gene expression data. (2004)
-{\emph{Brief Bioinform} \bf{5} 313--327.}
-
-\bibitem[Kaiser and Leisch(2008)]{kaiser08}
-Kaiser,~S. and Leisch,~F. (2008)
-A toolbox for bicluster analysis in R.
-\href{http://epub.ub.uni-muenchen.de/3293/}{Technical Report 028, Department of Statistics, University of Munich}.
-
-\bibitem[Kanehisa et~al.(2004)]{kanehisa04}
-Kanehisa,~M. \emph{et al.} (2004)
-The KEGG resource for deciphering the genome.
-\href{http://dx.doi.org/10.1093/nar/gkh063}{\emph{Nucleic Acids Res.}, {\bf 32} (Database special issue), D277.}
-
-\bibitem[Madeira and Oliveira(2004)]{madeira04}
-Madeira,~S.C. and Oliveira,~A.L. (2004)
-Biclustering algorithms for biological data analysis: A survey.
-\href{http://dx.doi.org/10.1109/TCBB.2004.2}{\emph{IEEE/ACM Transactions on Computational Biology and Bioinformatics}, {\bf 1}, 24--45.}
-
-\bibitem[Prelic et~al.(2006)]{prelic06}
-Prelic,~A., \emph{et~al.} 
-A systematic comparison and evaluation of biclustering methods for gene expression data 
-\href{http://dx.doi.org/10.1093/bioinformatics/bt1060}{\emph{Bioinformatics}, {\bf 22}(9), 1122--1129.}
-
-\bibitem[Santamar\'ia et~al.(2008)]{santamaria08}
-Santamar\'ia,~R., Ther\'on,~R. and Quintales,~L. (2008)
-BicOverlapper: A tool for bicluster visualization. 
-\href{http://dx.doi.org/10.1093/bioinformatics/btn076}{\emph{Bioinformatics}, {\bf 24}, 1212--1213.}
-
-\end{thebibliography}
+\vspace*{-7pt}
+\bibliographystyle{natbib}
+\bibliography{expressionview}
 

### 2010-01-14 expressionviewnote/expressionview.tex
Even more ev note updates.....

@@
 Lachapelle, Zolt\'an Kutalik, and Sven Bergmann]{Andreas
-  L\"uscher$^{1\dagger}$, G\'abor Cs\'ardi$^{1,2\dagger}$, Aitana Morton de
-  Lachapelle$^{1,2\dagger}$, Zolt\'an Kutalik$^{1,2\dagger}$, 
+  L\"uscher$^{1*}$, G\'abor Cs\'ardi$^{1,2*}$, Aitana Morton de
+  Lachapelle$^{1,2*}$, Zolt\'an Kutalik$^{1,2*}$,
   Bastian Peter$^{1,2}$
@@
   Lausanne, Switzerland\\
-  +: equal contribution
+  *: equal contribution
 }
@@
 summary of the results. From this overview, the user can select
-individual modules and access all the biologically relevant metadata
+individual modules and access biologically relevant metadata
 associated with them.
@@
 frequently used to explore microarray data. Biclustering algorithms
-identify co-expressed groups of
-genes and conditions (samples) where this co-expression occurs from
-collections of expression profiles. We refer to such groups as modules. While 
-there is a multitude of biclustering software available~(see in
-\cite{madeira04,ihmels04}), packages with intuitive interfaces that
-allow for an interactive exploration of the results are
+process collections of expression profiles to identify co-expressed groups of
+genes and conditions (samples) for which this co-expression occurs. 
+We refer to such groups as modules. While
+there is a multitude of biclustering software available (for a reviews 
+and comparisons see~\cite{madeira04,ihmels04,prelic06}), packages with 
+intuitive interfaces that allow for an interactive exploration of the results are
 sparse~\citep{santamaria08}. ExpressionView is designed to close this
@@
   from the biclustering and produce an ExpressionView data file that
-  can be explored with the interactive Flash
-  applet.}\label{fig:workflow} 
+  can be explored with the interactive Flash applet.}\label{fig:workflow}
 \end{figure*}
@@
 user-friendly way to access the actual gene expression matrix and its
-associated metadata. There is a variety of biclustering algorithms
-described in the literature~\citep{madeira04,prelic06}, several of
-which have been implemented as R packages. ExpressionView can treat
-results obtained by the Iterative Signature
-Algorithm~\citep{bergmann03,csardi09} and any of the methods available
-in the Biclust package~\citep{kaiser08}. Since the structure of
-biclustering results is independent of the algorithm, an extension to
-other methods is straightforward.
+associated metadata. ExpressionView can treat results obtained by the 
+Iterative Signature Algorithm~\citep{bergmann03,csardi09} and any of 
+the methods available in the Biclust package~\citep{kaiser08}. Since 
+the structure of biclustering results is independent of the algorithm, 
+an extension to other methods is straightforward.
 
@@
 To present the collection of possibly overlapping modules in a visually
-appealing form, it is necessary to reorder the rows (genes) and
-columns (conditions) of the
-gene expression matrix in such a way that biclusters form contiguous
+appealing form, it is necessary to reorder the rows (conditions) and
+columns (genes) of the gene expression matrix in such a way that biclusters form contiguous
 rectangles. Since it is in general impossible to find such an
@@
 module subsets. (An alternative would be to repeat rows and columns as
-necessary~\citep{grothaus06}, but for many modules this will results in
+necessary~\citep{grothaus06}, but for many modules this results in
 a very large expression matrix.)
@@
 The reordering of the rows and the columns do not depend on each
-other, so the same optimization method can be applied sequentially,
-say first to rows and then to columns. For a given order of the
-elements (either genes or conditions) we compute for each module the
+other, so the same optimization method can be applied independently
+to rows and columns. For a given order of the
+elements (either genes or conditions) we compute for each module $i$ the
 size of the largest contiguous sequence of elements (i.e. the maximal
-number of neighboring elements $N^\text{max}_i$). Then as a measure of the
-quality $Q$ of the order we sum this quantity over all modules
-($Q=\sum_i N^\text{max}_i$). The optimization procedure for $Q$ starts from
-some initial order. Two operations are then applied to this order: (1)
+number of neighboring elements $N^\text{max}_i$). Then, as a measure of the
+quality ($Q$) of the order we sum this quantity over all modules
+($Q=\sum_i N^\text{max}_i$). To optimize $Q$, an initial sequence is calculated using
+hierarchical clustering. Two operations are then applied to this: (1)
 Permutations that exchange two elements within a module and (2) shifts
-of a sequence of elements within a module to a different position. We
+of a sequence of multiple elements of the same module to a different position. We
 use a greedy iterative scheme that performs these operations at
-randomly selected positions and keeps the new order if it improves
+randomly selected positions and keeps the new sequence if it improves
 $Q$. The algorithm stops if after a given number of operations no
@@
 self-explanatory and extensible. In addition, the files can be read
-and edited with any text editor. The actual gene expression data is
+and edited with any text editor. The actual gene expression values are
 rounded to two digit precision and encoded in Base64. The structure of
@@
 which the modules are overlaid. On the right-hand side, the metadata
-associated with the gene expression data and the results of the
-enrichment calculations, GO~\citep{ashburner00} and
-KEGG~\citep{kanehisa04}, are shown. Wherever possible, these elements
+associated with the expression data and the results of the
+enrichment calculations for GO~\citep{ashburner00} categories and
+KEGG~\citep{kanehisa04} pathways are shown. Wherever possible, these elements
 are linked to the corresponding databases. The interface essentially
@@
 the expression data, getting instant feedback on the selected
-item. The currently selected region can be exported to a PDF file,
-helpful to prepare figures for presentations or articles, or a simple
+item. The currently selected region can be exported to a PDF file (e.g. 
+to prepare figures for presentations or articles) or to a
 text file.

### 2010-01-15 expressionviewnote/expressionview.tex
EV note updates, should be ready to submit.

@@
 \application{}
-\title[ExpressionView]{ExpressionView - an interactive viewer for
+\title[ExpressionView]{ExpressionView -- an interactive viewer for
   modules identified in gene expression data}
@@
 Biclustering is an unsupervised data analysis method which is
-frequently used to explore microarray data. Biclustering algorithms
-process collections of expression profiles to identify co-expressed groups of
-genes and conditions (samples) for which this co-expression occurs. 
+frequently used to explore microarray data. 
+Biclustering algorithms process collections of expression profiles to
+identify groups of genes co-expressed under some conditions (samples).
 We refer to such groups as modules. While
@@
 powerful microarray analysis tools provided by the
-Bioconductor~\citep{gentleman04} project before exporting the data to
+Bioconductor project~\citep{gentleman04} before exporting the data to
 a platform independent Adobe Flash applet. This applet
@@
 here an approximate solution that optimizes the arrangement within the
-original data by maximising the total area of the largest contiguous
+original data, by maximising the total area of the largest contiguous
 module subsets. (An alternative would be to repeat rows and columns as
@@
 hierarchical clustering. Two operations are then applied to this: (1)
-Permutations that exchange two elements within a module and (2) shifts
+permutations that exchange two elements within a module and (2) shifts
 of a sequence of multiple elements of the same module to a different position. We
 use a greedy iterative scheme that performs these operations at
-randomly selected positions and keeps the new sequence if it improves
+all possible positions and keeps the best new sequence if it improves
 $Q$. The algorithm stops if after a given number of operations no

### 2010-02-04 isanote/answers_to_reviewers.tex
ISA application note: added 'response to reviewers' skeleton.

\documentclass[a4paper]{article}
\usepackage{url}

\title{Modular analysis of gene expression data with R\\
  Answers to reviewers}
\author{G\'abor Cs\'ardi, Zolt\'an Kutalik and Sven Bergmann}

\begin{document}

\maketitle

\section{Reviewer 1}

\begin{quote}
Comments to the Author

This Application Note presents a software package that implements the
Iterative Signature Algorithm and provides a convenient visualization
of it’s results. The method is useful and the software may be of help
to interested users. 
\end{quote}

\begin{quote}
I have two problems with the Note in its present form.
The first may be due to my misunderstanding of the method. My
understanding was that some threshold Tg is applied on the genes and
another one, Tc, on the conditions. The aim is to reach a bicluster (a
set of conditions C and a set of genes G) such that G contains only
genes with expression above Tg for all the conditions of C, and
conditions are included in C only if the expression of all genes of G
exceeds Tc (one can replace a lower threshold with an upper one). The
heatmap of the Figure implies that my understanding was incorrect –
the conditions satisfy a double threshold (expressions are either
above or below some thresholds) and the genes have expressions above a
Tg in some conditions or below some other Tg’ in other conditions. 
I urge the authors to define what specifies or defines a bicluster –
do it briefly and sacrifice a few lines from elsewhere. 
\end{quote}

\begin{quote}
My second concern is minor – the abstract implies that the ISA is
special in allowing overlapping cluster assignments. This is incorrect
– there are other biclustering methods that do this. If I am not
mistaken, all three original biclustering methods that came out in
2000 (Cheng, Getz and Califano) allow this. 
\end{quote}

\section{Reviewer 2}

\begin{quote}
Comments to the Author

Summary

This manuscript describes two software packages for GNU R, for
discovering biclusters (here called transcription modules. see [1] for
a survey) in a given gene expression matrix based on the Iterative
Signature Algorithm (ISA) [2]. The first package, called isa2,
contains the implementation of the basic ISA, and can be used to
analyze any tabular data. The second package, called eisa, is built on
isa2. It adds support to standard BioConductor [3] data structures,
and contains gene expression specific tools for visualization of
biclusters. Both packages support execution of ISA with default
parameters, and also execution of ISA step by step with user-specified
parameters. The steps of ISA are normalization, iteration, filtering
of modules and enrichment calculation. In the current note, the
authors present both packages, and describe how to install and run
them step by step, using the acute lymphoblastic leukemia data set
(ALL) [4] as an example. 
\end{quote}

\begin{quote}
Overall evaluation

The manuscript and its accompanying tutorials are well written and
give an adequate theoretical background to ISA. The tutorials include
many detailed and visualized examples that make learning the package
quite easy for an R user. However, the following major points should
be taken into consideration: 
\end{quote}

\begin{quote}
1.      The manuscript and the website lack explanations about how the
authors implemented ISA and how they used the advantages of R. The
authors emphasize that the packages are highly optimized and run
extremely fast even for very large datasets.  Therefore, the
accompanying material should contain a summary describing how this
efficiency was achieved. In addition, a graph showing execution time
as a function of dataset size should be given. 
\end{quote}

\begin{quote}
2.      The tutorials give examples of execution on a single rather
old dataset (ALL) which is quite small in comparison to datasets of
recent years. Here the authors should give results for ISA execution
on more up-to-date datasets, and on heterogeneous datasets containing
profiles from multiple studies. 
\end{quote}

\begin{quote}
To summarize, the software is well motivated and enables users of the
R environment to utilize ISA. I suggest accepting the manuscript
conditional on solving the issues mentioned in this report. 
\end{quote}

\begin{quote}
Additional important remarks:

1.      The manuscript should contain references to additional
biclustering implementations, in R or in other environments, so that
the user could choose the one that best fits his needs in terms of
input/output, species, etc. Examples are SAMBA (via the Expander
software) [5,6] and BicAt [7]. 
\end{quote}

\begin{quote}
2.      The sample score plots example given in the EISA tutorial
(\url{http://www2.unil.ch/cbg/index.php?title=EISA\_tutorial}, section 6.5)
is not sufficiently clear. What are the scores represented by the
horizontal lines and how can one calculate them? Moreover, how can one
separate B-cell and T-cell leukemia samples using this plot? 
\end{quote}

\begin{quote}
3.      It is not quite clear what is the difference between the ISA
implementation included in the isa2 package and the ISA implementation
included in the eisa package. 
\end{quote}

\begin{quote}
4.      It should be clarified who are the potential users for the ISA
implementation for R environment. Assuming these are mainly
biologists, bioinformaticians and biostatisticians (and not only
programmers), more technical details of how to work with the packages
should be given. The following points are important examples: 
a.      Since the ISA implementation in eisa package imports data from
another package (e.g. hgu95av2.db), a tutorial for how to create such
a .db package from a given expression data text file should be given
(or referred to). 
\end{quote}

\begin{quote}
b.      Since both ISA packages use various data structures, a short
explanation of how to construct and use them should be given (or
references to such explanation), and also how to export data from
these structures to external files. 
\end{quote}

\begin{quote}
Minor remarks:

1.      The authors describe methods that perform enrichment tests for
the gene sets corresponding to the ISA modules against various
databases. In particular, they perform enrichment analysis against GO
database, KEGG pathway database, chromosomes and the TargetScan [8]
database. One may also want to investigate the correspondence of
biclusters and protein-protein interaction networks (which have been
derived from other types of data than gene expression data), e.g. the
DIP database [9]. An example for such analysis can be found in [10]. 
\end{quote}

\begin{quote}
2.      The heatmap image of a transcription module obtained by any
one of the two methods presented in the tutorial should include a
color legend. 
\end{quote}

\begin{quote}
3.      Since ISA function accepts an ExpressionSet, one must obtain
annotations for its expression data file in order to run ISA function
on it. Annotations should not be a prerequisite for running ISA. 
\end{quote}

\begin{quote}
4.      In section 10.8 it is mentioned that two significant modules
were found using random seeds and two significant (separating) modules
were found using smart seeds. However, in figure 7 it is noted that
only one module found using smart seeds. At the end of the section it
is mistakenly written "As it turns out, one modules found". 
\end{quote}

\begin{quote}
References:
[1] A. Tanay, R. Sharan, and R. Shamir. Biclustering algorithms: A
survey. In E. by Srinivas Aluru, editor, In Handbook of Computational
Molecular Biology. Chapman \& Hall/CRC, Computer and Information
Science Series, 2005.

[2] Bergmann, S., Ihmels, J., and Barkai, N. (2003). Iterative
signature algorithm for the analysis of large-scale gene expression
data. Phys Rev E Nonlin Soft Matter Phys, page 031902.

[3] Gentleman, R. C., Carey, V. J., Bates, D. M., et
al. (2004). Bioconductor: Open software development for computational
biology and bioinformatics. Genome Biology, 5, R80.

[4] Chiaretti, S., Li, X., Gentleman, R., Vitale, A., Vignetti, M.,
Mandelli, F., Ritz, J., and Foa, R. (2004). Gene expression profile of
adult t-cell acute lymphocytic leukemia identifies distinct subsets of
patients with different response to therapy and survival. Blood,
103(7).

[5] R. Sharan, A. Maron-Katz, N. Arbili, and R. Shamir.EXPANDER:
EXPression ANalyzer and DisplayER, 2002.Software package, Tel-Aviv
University, \url{http://www.cs.tau.ac.il/~rshamir/expander/expander.html}.

[6] A. Tanay, R. Sharan, M. Kupiec, and R. Shamir. Revealing
modularity and organization in the yeast molecular network by
integrated analysis of highly heterogeneous genomewide data. Proc Natl
Acad Sci U S A., 101(9):2981–6, 2004.

[7] Barkow, S., et al. (2006) BicAT: a biclustering analysis
toolbox. Bioinformatics, 22, 1282–1283

[8] Targetscan [\url{http://www.targetscan.org/}]

[9] Salwinski, L., et al. (2004) The Database of Interacting Proteins:
2004 update. Nucleic Acids Res, . 32, D449–D451

[10] Prelic,A. et al. (2006) A systematic comparison and evaluation of
biclustering methods for gene expression data. Bioinformatics, 22,
1122–1129. 
\end{quote}

\end{document}

### 2010-02-04 eisa/R/autogen.R

@@ -646,8 +646,9 @@ isa.autogen.module <- function(eset, modules, which, target.dir, template,
     }
   }
 
-  score <- round(getSampleMatrix(modules, mods=m), 2)
+  score <- getSampleMatrix(modules, mods=m)
   seq <- which(score != 0)
+  score <- round(score, 2)
   ord <- order(getSampleScores(modules, mods=m)[[1]], decreasing=TRUE)
   pd <- pData(modules)[seq,,drop=FALSE][ ord,,drop=FALSE]
   pd <- pd[, cond.to.include,drop=FALSE]
@@ -664,6 +665,7 @@ isa.autogen.module <- function(eset, modules, which, target.dir, template,
   
   # mark under- and over-expression
   under <- score[seq][ord] < 0
+  
   if (any(under)) {
     foo[under] <- gsub('<tr>', '<tr class="under">', foo[under])
   }

### 2010-02-04 eisa/R/viz.R

@@ -138,7 +138,7 @@ gograph <- function(table, colbar.length=30, label.cex=1, GOGRAPHS=NULL,
   ## nonzero is here to ignore an igraph bug in the
   ## Reingold-Tilford implementation, sometime two vertices
   ## are put at the very same coordinates
-  nonzero <- function(x) x[ x != 0 ]
+  nonzero <- function(x) x[ abs(x) > 1e-12 ]
   nodes.y <- tapply(l[,2], l[,1], function(x) {
     if (length(x)==1) Inf else min(nonzero(diff(sort(x)))) })
   nodes.y <- min(nodes.y)

### 2010-03-14 expressionviewnote/expressionview.tex
(R)ExpressionView: BioC modifications.

--- 2010-01-14 expressionviewnote/expressionview.tex
@@
-\documentclass[round]{bioinfo}
+\documentclass{bioinfo}
 
@@
 
-\usepackage{amstext}
-\usepackage{url}
-% \usepackage[colorlinks,citecolor=blue,linkcolor=red,urlcolor=blue]{hyperref}
+\usepackage[dvipdfm,colorlinks,bookmarks=false,citecolor=blue,linkcolor=red,urlcolor=blue]{hyperref}
 \newcommand{\thefirstpage}{1} \newcommand{\thelastpage}{1}
 
-\copyrightyear{2010}
-\pubyear{2010}
+\copyrightyear{2009}
+\pubyear{2009}
 
@@
 \application{}
-\title[ExpressionView]{ExpressionView - an interactive viewer for
-  modules identified in gene expression data} 
-\author[Andreas L\"uscher, G\'abor Cs\'ardi, Aitana Morton de
-Lachapelle, Zolt\'an Kutalik, and Sven Bergmann]{Andreas
-  L\"uscher$^{1\dagger}$, G\'abor Cs\'ardi$^{1,2\dagger}$, Aitana Morton de
-  Lachapelle$^{1,2\dagger}$, Zolt\'an Kutalik$^{1,2\dagger}$, 
-  Bastian Peter$^{1,2}$
-  and Sven Bergmann$^{1,2}$}
+\title[ExpressionView]{ExpressionView - an interactive viewer for biclusters identified in gene expression data}
+\author[Andreas L\"uscher, G\'abor Cs\'ardi, Aitana Morton de Lachapelle, Zolt\'an Kutalik, and Sven Bergmann]{Andreas L\"uscher$^1$, G\'abor Cs\'ardi$^{1,2}$, Aitana Morton de Lachapelle$^{1,2}$, Zolt\'an Kutalik$^{1,2}$, and Sven Bergmann$^{1,2}$\footnote{to whom correspondence should be addressed}}
 \address{
 	$^{1}$Swiss Institute of Bioinformatics, Lausanne, Switzerland\\
-  $^{2}$Department of Medical Genetics, University of Lausanne,
-  Lausanne, Switzerland\\
-  +: equal contribution
+	$^{2}$Department of Medical Genetics, University of Lausanne, Lausanne, Switzerland
 }
@@
 \begin{abstract}
+
 \section{Summary:}
-ExpressionView is an R package that provides an interactive graphical
-environment to explore transcription modules identified in gene expression
-data. A sophisticated ordering algorithm is used to present the
-modules with the expression in a visually appealing layout that provides an intuitive
-summary of the results. From this overview, the user can select
-individual modules and access all the biologically relevant metadata
-associated with them.
+ExpressionView is an R package that provides an interactive environment to explore biclusters identified in gene expression data.
+A sophisticated ordering algorithm is used to present the biclusters in a visually appealing layout that provides an intuitive summary of the results.
+From this overview, the user can select individual biclusters and access all the biologically relevant data associated with it.
+The package is aimed to facilitate the collaboration between bioinformaticians and life scientists who are not familiar with the R language.
 
 \section{Availability:} 
-%\href{http://www.unil.ch/cbg/ExpressionView}{R 
-%  package with open source license,
-\url{http://www.unil.ch/cbg/ExpressionView}
-%}
+\href{http://www2.unil.ch/cbg/index.php?title=ExpressionView}{R package with open source license}
 
-\section{Contact:} \href{Sven.Bergmann@unil.ch}{Sven.Bergmann@unil.ch}
+\section{Contact:} 
+\href{andreas.luescher@a3.epfl.ch}{Andreas L\"uscher}
 
 \section{Supplementary information:}
-Screenshots, tutorials and sample data sets can be found on the
-ExpressionView \href{http://www.unil.ch/cbg/ExpressionView}{website}. 
+Screenshots, tutorials and sample data sets can be found on the ExpressionView \href{http://www2.unil.ch/cbg/index.php?title=ExpressionView}{website}.
+
 \end{abstract}
 
-\vspace*{-7pt}
 \section{Introduction}
-Biclustering is an unsupervised data analysis method which is
-frequently used to explore microarray data. Biclustering algorithms
-identify co-expressed groups of
-genes and conditions (samples) where this co-expression occurs from
-collections of expression profiles. We refer to such groups as modules. While 
-there is a multitude of biclustering software available~(see in
-\cite{madeira04,ihmels04}), packages with intuitive interfaces that
-allow for an interactive exploration of the results are
-sparse~\citep{santamaria08}. ExpressionView is designed to close this
-gap, and facilitate the collaboration between bioinformaticians and
-life scientists without consolidated programming
-experience. Implemented as an R package, the user can apply all the
-powerful microarray analysis tools provided by the
-Bioconductor~\citep{gentleman04} project before exporting the data to
-a platform independent Adobe Flash applet. This applet
-provides an integrated, intuitive and interactive visualization of
-the modules together with the underlying gene expression data and
-additional information on the genes and conditions.
+Biclustering is an unsupervised learning technique used to analyse microarray data.
+It consists in identifying co-regulated groups of genes and conditions according to their expression profile.
+While there is a multitude of biclustering software available~\citep{madeira04}, packages with intuitive interfaces that allow the user to interactively explore the results are sparse~\citep{santamaria08}.
+ExpressionView is designed to close this gap, hereby facilitating the collaboration between bioinformaticians and life scientists without consolidated programming experience.
+Implemented as an R package, the user can apply all the powerful microarray analysis tools provided by the Bioconductor~\citep{gentleman04} project before exporting the data to a platform independent Adobe Flash applet that allows him and his co-workers to visualise the biclusters together with the underlying gene expression data.
 
-\vspace*{-7pt}
 \section{Package design and workflow}
-The ExpressionView package contains two independent parts: An ordering
-algorithm implemented in C++ and an Adobe Flash applet written in
-ActionScript and Adobe Flex. Fig.~\ref{fig:workflow} schematically
-summarises the typical workflow that uses the gene expression data
-available as a Bioconductor ExpressionSet and combines it with the
-biclustering results to produce an ExpressionView data file that can
-be explored with the Flash applet. In the following, we describe the
-different steps in more detail. 
+The ExpressionView package contains two independent parts: An ordering algorithm implemented in C++ and a Adobe Flash applet written in ActionScript and Adobe Flex.
+Fig.~\ref{fig:workflow} schematically summarises the typical workflow that uses the genomic data available as a Bioconductor ExpressionSet and combines it with the biclustering results to produce an ExpressionView data file that can be explored with the Flash applet.
+In the following, we describe the different steps in more detail.
 \begin{figure*}[!tpb]
-\centerline{\includegraphics[width=0.7\linewidth]{fig1-crop}}
-\caption{How to use ExpressionView: Starting from gene expression data
-  in the form of a Bioconductor ExpressionSet, the user first runs a
-  biclustering algorithm to determine co-expressed groups of genes and
-  the relevant samples. In a second step, the rows and columns of the gene
-  expression matrix are rearranged to produce an easily readable
-  overview of the results. The last step consists of combining the
-  gene expression data and its associated metadata with the results
-  from the biclustering and produce an ExpressionView data file that
-  can be explored with the interactive Flash
-  applet.}\label{fig:workflow} 
+\centerline{\includegraphics[width=0.9\linewidth]{fig1-crop}}
+\caption{How to use ExpressionView: Starting from gene expression data in the form of a Bioconductor ExpressionSet, the user first runs a biclustering algorithm to determine co-regulated groups of genes and samples.
+In a second step, the rows and columns of the gene expression matrix are rearranged to produce an easily readable overview of the results.
+The last step consist in combining the gene expression data and its associated metadata with the results from the biclustering and produce an ExpressionView data file that can be explored with the interactive Flash applet.}\label{fig:workflow}
 \end{figure*}
-For users who prefer to install the viewer as a standalone program, we
-also provide an Adobe AIR version that can be downloaded from the
-\href{http://www.unil.ch/cbg/ExpressionView}{website}.
+For users who prefer to install the viewer as a standalone program, we also provide a platform independent Adobe AIR version that can be downloaded from the \href{http://www2.unil.ch/cbg/index.php?title=ExpressionView}{website}.
 
-\subsection{Gene expression data and modules}
-ExpressionView is designed to work with gene expression data in the
-form of a Bioconductor ExpressionSet. This class provides a
-user-friendly way to access the actual gene expression matrix and its
-associated metadata. There is a variety of biclustering algorithms
-described in the literature~\citep{madeira04,prelic06}, several of
-which have been implemented as R packages. ExpressionView can treat
-results obtained by the Iterative Signature
-Algorithm~\citep{bergmann03,csardi09} and any of the methods available
-in the Biclust package~\citep{kaiser08}. Since the structure of
-biclustering results is independent of the algorithm, an extension to
-other methods is straightforward.
+\subsection{Gene expression data and biclusters}
+ExpressionView is designed to work with gene expression data in the form of a Bioconductor ExpressionSet.
+This class provides a user-friendly way to access the actual gene expression matrix and its associated metadata.
+There is a variety of biclustering algorithms described in the literature~\citep{madeira04,prelic06}, several of which have been implemented as R packages.
+ExpressionView can treat results obtained by the iterative signature algorithm (ISA)~\citep{bergmann03,csardi09} and the methods available in the Biclust package~\citep{kaiser08}.
+Since the structure of biclustering results is independent of the algorithm, an extension to other methods is straightforward.
 
-\subsection{Reordering genes and conditions}
-To present the collection of possibly overlapping modules in a visually
-appealing form, it is necessary to reorder the rows (genes) and
-columns (conditions) of the
-gene expression matrix in such a way that biclusters form contiguous
-rectangles. Since it is in general impossible to find such an
-arrangement for more than two mutually overlapping modules, we propose
-here an approximate solution that optimizes the arrangement within the
-original data by maximising the total area of the largest contiguous
-module subsets. (An alternative would be to repeat rows and columns as
-necessary~\citep{grothaus06}, but for many modules this will results in
-a very large expression matrix.)
+\subsection{Order}
+To present the tens of possibly overlapping biclusters in a visually appealing form, it is necessary to reorder the rows and columns of the gene expression matrix in such a way that biclusters form contiguous rectangles.
+Since for more than two mutually overlapping biclusters, it is in general impossible to find such an arrangement, one has to make concessions.
+In contrast to \cite{grothaus06}, who proposed to repeat rows and columns as necessary to achieve this goal, we prefer to optimise the arrangement within the original data by maximising the area of the largest contiguous biclusters.
 
-This optimization task is an interesting problem on its own, which to
-the best of our knowledge has not been studied in the literature. Thus
-we briefly outline our strategy here (see our website for details). 
-The reordering of the rows and the columns do not depend on each
-other, so the same optimization method can be applied sequentially,
-say first to rows and then to columns. For a given order of the
-elements (either genes or conditions) we compute for each module the
-size of the largest contiguous sequence of elements (i.e. the maximal
-number of neighboring elements $N^\text{max}_i$). Then as a measure of the
-quality $Q$ of the order we sum this quantity over all modules
-($Q=\sum_i N^\text{max}_i$). The optimization procedure for $Q$ starts from
-some initial order. Two operations are then applied to this order: (1)
-Permutations that exchange two elements within a module and (2) shifts
-of a sequence of elements within a module to a different position. We
-use a greedy iterative scheme that performs these operations at
-randomly selected positions and keeps the new order if it improves
-$Q$. The algorithm stops if after a given number of operations no
-significant improvement of $Q$ is achieved.
+Since this reordering is an interesting problem on its own, which to the best of our knowledge has not been studied in the literature, we briefly outline our strategy here.
+Noting that the rows and the columns can be ordered separately, the problem reduces to finding the optimal arrangement of a set of $n$ elements that are part of at least one of $m$ clusters.
+The quality of the order is defined as the sum over the maximal number of neighbouring elements in every cluster and the tasks thus consists in maximizing this quantity by changing the order of the elements.
+Starting from an initial configuration determined by similarity, this is achieved by either shifting well aligned subsequences of a given cluster, herby enlarging the longest contiguous part, or permuting the individual elements of a given subcluster.
 
-To determine the efficiency of this method, we have studied a large
-number of perfectly orderable, but initially scrambled, test cases. We
-find that in almost every situation, the proposed algorithm finds an
-order that recovers more than 99\% of the score of the optimal
-solution and in most cases, it recovers the correct
-alignment. For random samples, which are more representative for
-actual gene expression data, the execution time increases polynomially
-with the number of clusters $m$ as ${\mathcal O}(m^\alpha)$, where
-$\alpha \in [1.6, 2]$, almost independent of the number of elements
-$n$. For a given number of clusters, we find ${\mathcal O}(n^\alpha)$,
-with $\alpha \in [2.5, 2.7]$.
+To determine the efficiency of this method, we have studied a large number of perfectly orderable, but initially scrambled, test cases.
+We find that in almost every situation, the proposed algorithm finds an order that recovers more than 99\% of the score of the optimal solution and in most cases, it actually recovers the correct alignment.
+For random samples, which are more representative for actual gene expression data than orderable situations, execution time increases polynomially with the number of clusters $m$ as ${\mathcal O}(m^\alpha)$, where $\alpha \in [1.6, 2]$, almost independent of the number of elements $n$.
+For a given number of clusters, we find ${\mathcal O}(n^\alpha)$, with $\alpha \in [2.5, 2.7]$.
 
 \subsection{Export}
-Once the optimal order is determined, the program rearranges the gene
-expression matrix accordingly and exports all the relevant information
-into an XML file. This format has the advantage of being
-self-explanatory and extensible. In addition, the files can be read
-and edited with any text editor. The actual gene expression data is
-rounded to two digit precision and encoded in Base64. The structure of
-the XML file is described in more detail on the ExpressionView
-\href{http://www.unil.ch/cbg/ExpressionView}{website}. In most cases,
-these files are of the order of a few Megabytes and can thus easily be
-sent by e-mail.
+Once the optimal order is determined, the program rearranges the gene expression matrix accordingly and exports all the relevant information into an XML file.
+This format has the advantage of being self-explanatory and extensible.
+In addition, the files can be read and edited with any text editor.
+The actual gene expression data is rounded to two digit precision and encoded in Base64.
+The structure of the XML file is described in more detail on the ExpressionView \href{http://www2.unil.ch/cbg/index.php?title=ExpressionView}{website}.
+In most cases, these files are of the order of a few Megabytes and can thus easily be sent to co-workers.
 
 \subsection{Visualisation}
-The core application of this package is the Flash applet that allows
-the user to interactively explore the gene expression data and the
-modules. The Flash architecture has the advantage of
-producing a platform and browser independent applet that can be
-launched from any computer with even the most restrictive user
-rights. A screenshot is shown in Fig.~\ref{fig:workflow}. The
-interface is divided in two parts: On the left-hand side, the user
-finds the gene expression data in the common heat map form, on top of
-which the modules are overlaid. On the right-hand side, the metadata
-associated with the gene expression data and the results of the
-enrichment calculations, GO~\citep{ashburner00} and
-KEGG~\citep{kanehisa04}, are shown. Wherever possible, these elements
-are linked to the corresponding databases. The interface essentially
-behaves as an image viewer, allowing the user to zoom and pan around
-the expression data, getting instant feedback on the selected
-item. The currently selected region can be exported to a PDF file,
-helpful to prepare figures for presentations or articles, or a simple
-text file.
+The core application of this package is the Flash applet that allows the user to interactively explore the gene expression data and the biclusters.
+Relying on the Flash architecture has the advantage of producing a platform and browser independent applet that can be launched from any computer with even the most restrictive user rights.
+A screenshot is shown in Fig.~\ref{fig:workflow}.
+The interface is divided in two parts: On the left-hand side, the user finds the gene expression data in the common heat map form, on top of which the biclusters are overlaid.
+On the right-hand side, the metadata associated with the gene expression data and the results of the enrichment calculations [GO (Gene Ontology)~\citep{ashburner00} and KEGG (Kyoto Encyclopedia of Genes and Genomes)~\citep{kanehisa04}] are shown.
+Wherever possible, these elements are linked to the corresponding databases.
+The interface essentially behaves as an image viewer, allowing the user to zoom and pan around the expression data, getting instant feedback on the selected item.
+The currently selected region can be exported to a PDF file, helpful to prepare figures for presentations or articles, or a simple text file.\\
 
-\vspace*{-7pt}
-\section*{Acknowledgement}
+%\section*{Acknowledgement}
 
-\paragraph{Funding\textcolon} The authors are grateful to the Swiss
-Institute of Bioinformatics, the Swiss National Science Foundation
-(3100AO-116323/1) and the European Framework Project 6 (through
-the EuroDia and AnEuploidy projects).
+\paragraph{Funding\textcolon} This work was supported by ...
 
@@
 
-\vspace*{-7pt}
-\bibliographystyle{natbib}
-\bibliography{expressionview}
+\begin{thebibliography}{6}
+
+\bibitem[Ashburner et~al.(2000)]{ashburner00}
+Ashburner,M. \emph{et al.} (2000)
+Gene Ontology: tool for the unification of biology.
+\href{http://dx.doi.org/10.1038/75556}{\emph{Nature Genetics}, {\bf 25}, 25 -- 29.}
+
+\bibitem[Bergmann et~al.(2003)]{bergmann03}
+Bergmann,S., Ihmels,J. and Barkai,N. (2003)
+Iterative signature algorithm for the analysis of large-scale gene expression data.
+\href{http://dx.doi.org/10.1103/PhysRevE.67.031902 }{\emph{Physical Review E}, {\bf 67}, 031902.}
+
+\bibitem[Cs{\'a}rdi et~al.(2009)]{csardi09}
+Cs{\'a}rdi,G., Kutalik,Z. and Bergmann,S. (2009)
+Modular analysis of gene expression data with R.
+\href{http://dx.doi.org/}{\emph{Bioinformatics}, ...}
+
+\bibitem[Gentleman et~al.(2004)]{gentleman04}
+Gentleman,R.C., \emph{et al.} (2004)
+Bioconductor: open software development for computational biology and bioinformatics.
+\href{http://dx.doi.org/10.1186/gb-2004-5-10-r80}{\emph{Genome Biology}, {\bf 5}, R80.}
+
+\bibitem[Grothaus et~al.(2006)]{grothaus06}
+Grothaus,G., Mufti,A., and Murali,TM. (2006)
+Automatic layout and visualization of biclusters.
+\href{http://dx.doi.org/10.1186/1748-7188-1-15}{\emph{Algorithms for Molecular Biology}, {\bf 1}, 15.}
+
+\bibitem[Kaiser and Leisch(2008)]{kaiser08}
+Kaiser,S. and Leisch,F. (2008)
+A toolbox for bicluster analysis in R.
+\href{http://epub.ub.uni-muenchen.de/3293/}{Technical Report 028, Department of Statistics, University of Munich}.
+
+\bibitem[Kanehisa et~al.(2004)]{kanehisa04}
+Kanehisa,M. \emph{et al.} (2004)
+The KEGG resource for deciphering the genome.
+\href{http://dx.doi.org/10.1093/nar/gkh063}{\emph{Nucleic Acids Res.}, {\bf 32} (Database special issue), D277.}
+
+\bibitem[Madeira and Oliveira(2004)]{madeira04}
+Madeira,S.C. and Oliveira,A.L. (2004)
+Biclustering algorithms for biological data analysis: A survey.
+\href{http://dx.doi.org/10.1109/TCBB.2004.2}{\emph{IEEE/ACM Transactions on Computational Biology and Bioinformatics}, {\bf 1}, 24 -- 45.}
+
+\bibitem[Prelic et~al.(2006)]{prelic06}
+Prelic,A., \emph{et~al.} 
+A systematic comparison and evaluation of biclustering methods for gene expression data 
+\href{http://dx.doi.org/10.1093/bioinformatics/bt1060}{\emph{Bioinformatics}, {\bf 22}(9), 1122 -- 1129.}
+
+\bibitem[Santamar\'ia et~al.(2008)]{santamaria08}
+Santamar\'ia,R., Ther\'on,R. and Quintales,L. (2008)
+BicOverlapper: A tool for bicluster visualization. 
+\href{http://dx.doi.org/10.1093/bioinformatics/btn076}{\emph{Bioinformatics}, {\bf 24}, 1212 -- 1213.}
+
+\end{thebibliography}
 

### 2010-03-14 RExpressionView/inst/doc/ExpressionView.ordering.Rnw
ExpressionView:  Removed old images.

@@
 
-The \code{C++} code contains the functions \code{reposition} and \code{exchange} that 
-execute these two basic moves.
+Internally, the data is represented as a binary matrix (i.e., elements are either 0 or 1) with dimensions $m$ $\times$ $n$, where the number of rows $m$ correspond to the number of genes or conditions and the number of columns $n$ is determined by the number of clusters.
 
-Before optimizing the arrangement, the matrix is simplified (\code{simplify}) to get rid of duplicate slots.
-The multiplicity of each slot is taken into account when calculating the score $Q$. 
-At the end of the ordering the initial dimensions of the data are restored with the \code{complexify} 
-function.
+The ordering is done in four steps:
+\begin{enumerate}
+	\item{\code{simplify}: Reduces the dimensions of the problem by excluding duplicate rows. The multiplicity of each slot is taken into account when calculating the score $Q$.}
+	\item{\code{prearrange}: Tries to find a good starting point for the optimization by creating a new order, starting from the first row and adding the consecutive rows one after the other, placing them at the optimal position.
+For smaller samples, the fitness function is used, while for larger samples only the similarity (scalar product) between rows is taken into account.}
+	\item{\code{arrange}: Uses the brute-force approach to further increase the alignment score.}
+	\item{\code{complexify}: Reinstates the original dimensions of the problem by inserting the omitted rows at the newly found position.}
+\end{enumerate}
+
+If the user specifies a time limit, the algorithm stops the alignment after the elapsed interval and returns the best alignment.
+Otherwise, it continues until it can no longer improve the alignment score.
+
+The ordering iterates over the clusters and first calls the function \code{reposition}, taking \code{cluster} as argument.
+It goes through all the rows contained in a given cluster and tries to shift them to a better position.
+A contiguous group of rows can be repositioned at once or split up in subgroups.
+Empirically, it turned out that the alignment is fastest by starting the repositioning with a single row, increasing the length of the group only if no better position could be found.
+The opposite approach can be used by replacing the \emph{start forward} section on line 545 by the \emph{start backwards} code on line 556.
+
+After repositioning, the main loop calls the exchange function, taking \code{cluster} and \code{what} as argument.
+It exchanges rows contained (\code{what = 1}) / not contained (\code{what = 0}) in a given cluster.
+It randomly picks \code{maxslots} number of rows (currently six), generates all possible permutations, calculates the corresponding alignment scores, and retains the best order.
+This realignment is repeated \code{maxiter} times, so that on average every slot is part of one optimization.
+
+The loop over clusters calls the \code{reposition} function first for all the rows not contained in a given cluster and afterwards for the remaining ones.
+The main loop is repeated until the maximal computation time is reached or no improved alignment could be found anymore.
 
@@
 
-\bibliographystyle{unsrt}
-\bibliography{ExpressionView}
+%\bibliographystyle{unsrt}
+%\bibliography{ExpressionView}
 

### 2010-03-15 isanote/answers_to_reviewers.tex
ISA paper updated for the resubmission.

@@
 \usepackage{url}
+\usepackage{natbib}
+
+\newcommand{\Rpackage}[1]{`\texttt{#1}'}
+\newcommand{\Rfunction}[1]{`\texttt{#1}'}
+\newcommand{\Rclass}[1]{\textsl{#1}}
+
+\setlength{\parskip}{12pt}
+\setlength{\parindent}{0pt}
 
@@
 \maketitle
+
+First of all, we want to thank the reviewers for their thorough work
+and useful suggestions. In the following, the reviewers' comments are
+indented as a block.
 
@@
 
+Indeed, by default the software uses a double threshold. This can be
+changed, as explained in the reference manual of the ISA functions,
+and the ISA tutorial titled `The Iterative Signature Algorithm for
+Gene Expression Data'.
+
+To  clarify things better, we have added the following to the main
+text:
+\textsl{The output of ISA is a collection of potentially overlapping
+modules. Every module contains genes, that are up- and/or
+down-regulated together, exactly in the samples of the module.
+In every module, each gene and each sample is attributed a
+score between minus one and one, that reflects the strength of the
+association with the module. Moreover, if the scores of two genes of a
+module have the same sign, then they are correlated (in the samples of
+the module), opposite signs mean anti-correlation. If two sample
+scores have the same sign, then these samples are correlated (along
+the genes of the module), opposite signs signal anti-correlation.}
+
+The tutorial on the ISA homepage explains the ISA iteration in detail:
+\textsl{The thresholding is an important step of the ISA, without
+  thresholding ISA would be equivalent to a (not too eﬀective)
+  numerical singular value decomposition (SVD) algorithm. Currently
+  thresholding is done by calculating the mean and standard deviation
+  of the vector and keeping only elements that are further than a
+  given number of standard deviations from the mean. Based on the
+  direction parameter, this means keeping values that are signiﬁcantly
+  higher than the mean (``up''), or keeping the ones that are
+  signiﬁcantly lower than the mean (``down''); or keeping both
+  (``updown'').}
+
 \begin{quote}
@@
 \end{quote}
+
+We have removed the emphasis on the overlapping biclusters of the
+ISA and added references to these methods.
 
@@
 This manuscript describes two software packages for GNU R, for
-discovering biclusters (here called transcription modules. see [1] for
+discovering biclusters (here called transcription modules. see \citet{tanay05} for
 a survey) in a given gene expression matrix based on the Iterative
-Signature Algorithm (ISA) [2]. The first package, called isa2,
+Signature Algorithm (ISA) \citep{isa}. The first package, called isa2,
 contains the implementation of the basic ISA, and can be used to
 analyze any tabular data. The second package, called eisa, is built on
-isa2. It adds support to standard BioConductor [3] data structures,
+isa2. It adds support to standard BioConductor \citep{BioC} data structures,
 and contains gene expression specific tools for visualization of
@@
 them step by step, using the acute lymphoblastic leukemia data set
-(ALL) [4] as an example. 
+(ALL) \citep{chiaretti04} as an example. 
 \end{quote}
@@
 
+We have added a new document to the ISA homepage at
+\url{http://www.unil/ch/cbg/ISA}, titled `ISA internals'. This
+document contains the details of our implementations, and also
+execution time analysis.
+
 \begin{quote}
@@
 \end{quote}
+
+We believe that the tutorials are best if they can be run on an
+average desktop computer, interactively, without spending much
+effort on collecting and assembling the data. This is why we chose the
+ALL dataset. It is included in BioConductor, and used in the tutorials
+of many BioConductor packages. There is a high chance that it is
+already installed on the user's machine, and if it is not, then it can
+be downloaded easily and quickly.
+
+We wrote a second tutorial that uses a bigger, more recent data set, 
+titled `Tissue specific expression with the Iterative Signature
+Algorithm', please see the ISA homepage at
+\url{http://www.unil.ch/cbg/ISA}. The `ISA internals' document uses a
+new, larger data set, as well.
+
+Of course, the ISA can be used for much larger data sets (see
+e.g. \cite{bergmann04}), but then the analysis takes at least hours
+instead of minutes.
 
@@
 input/output, species, etc. Examples are SAMBA (via the Expander
-software) [5,6] and BicAt [7]. 
+software) \citep{sharan02,tanay04} and BicAt \citep{barkow06}. 
 \end{quote}
+
+We have added these references.
 
@@
 
+The reviewer is right, that the module plotted in section 6.5 cannot
+be used to separate B-cell and T-cell samples, unfortunately the wrong
+module was plotted. We corrected this.
+
+We have also explained sample score plots better in the tutorial. 
+Please note, that the tutorials do not aim to explain all details 
+about the ISA functions; the reference manual, that is included in the
+package, is the place for that.
+
 \begin{quote}
@@
 
+We have explained the difference between the two packages in the newly
+added `ISA internals' document, section `Why two packages?'.
+
 \begin{quote}
@@
 programmers), more technical details of how to work with the packages
-should be given. The following points are important examples: 
+should be given.
+\end{quote}
+
+We have added the following to the abstract:
+\textsl{
+Potential users for these packages are all R and
+BioConductor users dealing with tabular (e.g. gene expression) data.
+}
+
+\begin{quote}
+The following points are important examples: 
 a.      Since the ISA implementation in eisa package imports data from
@@
 \end{quote}
+
+Users analyzing gene expression data almost never need to build
+these annotation packages, they are included in BioConductor, for all
+standard chips. Only for custom chips might be useful to create such
+packages by hand; but even in this case, they are not required, the
+user can create ExpressionSet objects that do not contain chip
+annotation, or use the \Rpackage{isa2} package.
 
@@
 
+For the \Rpackage{isa2} package, this is included in the reference
+manual. For the \Rpackage{eisa} package, the \Rclass{ISAModules} class
+is an opaque data type, the representation of it might change in the
+future, and the user should not rely on its current
+definition. Instead, various functions are provided that deal with
+these kind of objects, they are documented on the manual page of the
+\Rclass{ISAModules} class, in the reference manual.
+
 \begin{quote}
@@
 databases. In particular, they perform enrichment analysis against GO
-database, KEGG pathway database, chromosomes and the TargetScan [8]
+database, KEGG pathway database, chromosomes and the TargetScan \citep{targetscan}
 database. One may also want to investigate the correspondence of
@@
 derived from other types of data than gene expression data), e.g. the
-DIP database [9]. An example for such analysis can be found in [10]. 
+DIP database \citep{salwinski04}. An example for such analysis can be
+found in \citet{prelic06}.
 \end{quote}
+
+We have added new classes and functions to perform enrichment analysis
+on the ISA modules, against any gene label or database that the user
+provides. These functions are explained in the manual page of the
+\Rfunction{ISAEnrichment} function.
 
@@
 
+We have added the color legends.
+
 \begin{quote}
@@
 \end{quote}
+
+Annotations are not required to run ISA. First, the purpose of the
+\Rpackage{isa2} package is to run ISA on any tabular data set,
+biological or not. As an alternative, the user can construct an
+\Rclass{ExpressionSet} object without annotation, and this can be used
+with the function of the \Rpackage{eisa} package.
 
@@
 
-\begin{quote}
-References:
-[1] A. Tanay, R. Sharan, and R. Shamir. Biclustering algorithms: A
-survey. In E. by Srinivas Aluru, editor, In Handbook of Computational
-Molecular Biology. Chapman \& Hall/CRC, Computer and Information
-Science Series, 2005.
+We have corrected this.
 
-[2] Bergmann, S., Ihmels, J., and Barkai, N. (2003). Iterative
-signature algorithm for the analysis of large-scale gene expression
-data. Phys Rev E Nonlin Soft Matter Phys, page 031902.
-
-[3] Gentleman, R. C., Carey, V. J., Bates, D. M., et
-al. (2004). Bioconductor: Open software development for computational
-biology and bioinformatics. Genome Biology, 5, R80.
-
-[4] Chiaretti, S., Li, X., Gentleman, R., Vitale, A., Vignetti, M.,
-Mandelli, F., Ritz, J., and Foa, R. (2004). Gene expression profile of
-adult t-cell acute lymphocytic leukemia identifies distinct subsets of
-patients with different response to therapy and survival. Blood,
-103(7).
-
-[5] R. Sharan, A. Maron-Katz, N. Arbili, and R. Shamir.EXPANDER:
-EXPression ANalyzer and DisplayER, 2002.Software package, Tel-Aviv
-University, \url{http://www.cs.tau.ac.il/~rshamir/expander/expander.html}.
-
-[6] A. Tanay, R. Sharan, M. Kupiec, and R. Shamir. Revealing
-modularity and organization in the yeast molecular network by
-integrated analysis of highly heterogeneous genomewide data. Proc Natl
-Acad Sci U S A., 101(9):2981–6, 2004.
-
-[7] Barkow, S., et al. (2006) BicAT: a biclustering analysis
-toolbox. Bioinformatics, 22, 1282–1283
-
-[8] Targetscan [\url{http://www.targetscan.org/}]
-
-[9] Salwinski, L., et al. (2004) The Database of Interacting Proteins:
-2004 update. Nucleic Acids Res, . 32, D449–D451
-
-[10] Prelic,A. et al. (2006) A systematic comparison and evaluation of
-biclustering methods for gene expression data. Bioinformatics, 22,
-1122–1129. 
-\end{quote}
+\bibliographystyle{natbib}
+\bibliography{isa}
 

### 2010-03-15 isanote/modular.tex
ISA paper updated for the resubmission.

--- 2009-12-11 isanote/modular.tex
@@
 Large sets of data, like expression profiles from many samples, require
-analytic tools to reduce their complexity. Classical (bi-)clustering
-algorithms typically attribute elements (genes, arrays) to disjoint groups
-(``clusters''). Yet, in some cases overlapping cluster assignments would suit
-the biological reality much better.
-
-The Iterative Signature Algorithm (ISA) was designed to overcome this and
-other limitations of standard clustering algorithms. It aims to reduce the
-complexity of a very large set of data by decomposing it into so-called
+analytic tools to reduce their complexity. The Iterative Signature
+Algorithm (ISA) is a biclustering algorithm. 
+It was designed to decompose a large set of data into so-called
 ``modules''. In the context of gene expression data these modules consist of
@@
 multiple modules and the level of required coherence can be varied resulting
-in different ``resolutions'' of the modular mapping. Since the ISA does not
-rely on the computation of correlation matrices (like many other tools), it
-is extremely fast even for very large datasets.
+in different ``resolutions'' of the modular mapping.
 
-In this short note, we introduce two BioConductor \citep{BioC}, software
-packages written in GNU~R: The~\Rpackage{isa2} includes a highly optimized
-implementation of the ISA and~\Rpackage{eisa} provides a convenient interface
-to run the ISA, visualize its output and puts the biclusters into
-biological context.
+In this short note, we introduce two BioConductor \citep{BioC} software
+packages written in GNU~R: The~\Rpackage{isa2} package includes an optimized
+implementation of the ISA and the~\Rpackage{eisa} package provides a
+convenient interface 
+to run the ISA, visualize its output and put the biclusters into
+biological context. Potential users for these packages are all R and
+BioConductor users dealing with tabular (e.g. gene expression) data.
 
@@
 
+\vspace{-1cm}
 \section{Introduction}
@@
 The output of ISA is a collection of potentially overlapping
-modules. In every module, each gene and each sample is attributed a
-score that reflects the strength of the association with the module.
+modules. Every module contains genes, that are up- and/or
+down-regulated together, exactly in the samples of the module.
+In every module, each gene and each sample is attributed a
+score between minus one and one, that reflects the strength of the
+association with the module. Moreover, if the scores of two genes of a
+module have the same sign, then they are correlated (in the samples of
+the module), opposite signs mean anti-correlation. If two sample
+scores have the same sign, then these samples are correlated (along
+the genes of the module), opposite signs signal anti-correlation.
 
-In the following we first outline how our software tool is commonly used for a
-modular analysis (Section~\ref{sec:methods}) and then we describe in more
-detail the R packages that implement the ISA and visualize its output
-(Section~\ref{sec:implementation}).
+For other biclustering algorithms, please see
+\citet{cheng00,getz00,califano00,sharan02,tanay04,barkow06}.
 
@@
 \centering
-\includegraphics[width=\textwidth]{isa2workflow3}
+\includegraphics[width=0.8\textwidth]{isa2workflow3}
 \caption{\subfig{(A)} Work flow of a typical modular analysis with the
@@
 
-For users that prefer Matlab over R, we created a Matlab package for
-ISA, please see the ISA homepage for more information.
-
 % \subsubsection{Visualization}
@@
 
-Additional information about the ISA, the \Rpackage{isa2} and
-\Rpackage{eisa} packages and several tutorials are available on the
-ISA homepage.
+Additional information and a Matlab implementation of ISA are
+available on the ISA homepage.
 
+\vspace*{-0.4cm}
 \section*{Acknowledgement}

### 2010-03-15 isanote/answers_to_reviewers.tex
ISA paper: added Zoltan's changes.

@@
 modules. Every module contains genes, that are up- and/or
-down-regulated together, exactly in the samples of the module.
+down-regulated, in samples that belong to the module.
 In every module, each gene and each sample is attributed a
@@
 association with the module. Moreover, if the scores of two genes of a
-module have the same sign, then they are correlated (in the samples of
-the module), opposite signs mean anti-correlation. If two sample
-scores have the same sign, then these samples are correlated (along
-the genes of the module), opposite signs signal anti-correlation.}
+module have the same sign, then they are correlated (across the
+samples of the module), opposite signs mean
+anti-correlation. Similarly, if two sample
+scores have the same sign, then these samples are correlated (across
+the genes of the module), opposite signs indicate anti-correlation.}
 
@@
 \textsl{The thresholding is an important step of the ISA, without
-  thresholding ISA would be equivalent to a (not too eﬀective)
+  thresholding ISA would be equivalent to a (not too effective)
   numerical singular value decomposition (SVD) algorithm. Currently
@@
   of the vector and keeping only elements that are further than a
-  given number of standard deviations from the mean. Based on the
-  direction parameter, this means keeping values that are signiﬁcantly
-  higher than the mean (``up''), or keeping the ones that are
-  signiﬁcantly lower than the mean (``down''); or keeping both
-  (``updown'').}
+  given number of standard deviations from the mean. Using the
+  ``direction'' parameter, one can keep values that are
+  (a) significantly higher (``up''); (b) lower (``down'') than the
+  mean; or (c) both (``updown'').}
 

### 2010-03-15 eisa/inst/doc/EISA_tutorial.Rnw
ISA paper updated for the resubmission.
ISA paper: added Zoltan's changes.

@@
 @
-\caption{Condition plot for the ALL data and the second module. Every
-  sample is represented as a bar, and its ISA score is plotted. The
-  scores are also printed at the top or bottom of the bars. The scores
-  corresponding to the ISA thresholds are represented as horizontal
-  lines. The dashed line is the mean of the condition scores. B-cell
-  samples are dark green, T-cell samples are orange.
+\caption{Condition (or sample score) plot for the ALL data and the
+  second module. Every sample is represented as a bar, and its ISA
+  score is plotted. The scores are also printed at the top or bottom
+  of the bars. The scores corresponding to the ISA thresholds are
+  represented as horizontal lines. The dashed line is the mean of the
+  condition scores. B-cell samples are dark green, T-cell samples are
+  orange. The green and red horizontal lines show the sample
+  thresholds. Samples above the red line, and samples below the green
+  line are included in the module.
 }
@@
 
-The thresholding is an important step of the ISA, without thresholding
-ISA would be equivalent to a (not too effective) numerical singular
-value decomposition (SVD) algorithm. Currently thresholding is done by
-calculating the mean and standard deviation of the vector and keeping
-only elements that are further than a given number of standard
-deviations from the mean. Based on the \Rargument{direction} parameter,
-this means keeping values that are significantly higher than the
-mean (``\emph{up}''), or keeping the ones that are significantly lower
-than the mean (``\emph{down}''); or keeping both
-(``\emph{updown}'').
+The thresholding is an important step of the ISA, without
+thresholding ISA would be equivalent to a (not too effective)
+numerical singular value decomposition (SVD) algorithm. Currently
+thresholding is done by calculating the mean and standard deviation
+of the vector and keeping only elements that are further than a
+given number of standard deviations from the mean. Using the
+``direction'' parameter, one can keep values that are
+(a) significantly higher (``up''); (b) lower (``down'') than the
+mean; or (c) both (``updown'').
 

### 2010-03-15 isanote/modular.tex
ISA note: make figure a bit bigger.
ISA paper: added Zoltan's changes.

@@
 to run the ISA, visualize its output and put the biclusters into
-biological context. Potential users for these packages are all R and
+biological context. Potential users of these packages are all R and
 BioConductor users dealing with tabular (e.g. gene expression) data.
@@
 association with the module. Moreover, if the scores of two genes of a
-module have the same sign, then they are correlated (in the samples of
-the module), opposite signs mean anti-correlation. If two sample
-scores have the same sign, then these samples are correlated (along
-the genes of the module), opposite signs signal anti-correlation.
+module have the same sign, then they are correlated (across the
+samples of the module), opposite signs mean
+anti-correlation. Similarly, if two sample
+scores have the same sign, then these samples are correlated (across
+the genes of the module), opposite signs indicate anti-correlation.
 
@@
 \centering
-\includegraphics[width=0.8\textwidth]{isa2workflow3}
+\includegraphics[width=0.842\textwidth]{isa2workflow3}
 \caption{\subfig{(A)} Work flow of a typical modular analysis with the

### 2010-03-15 eisa/inst/doc/EISA_tutorial.Rnw
Update EISA tutorial.

@@
 \Rargument{norm} argument to \Rfunction{ISA2heatmap}, see the manual
-for details. The result is in Fig.~\ref{fig:heatmap}.
-<<heatmap,results=hide,include=FALSE,eval=FALSE>>=
-ISA2heatmap(modules, 1, ALL, norm="sample", scale="none")
+for details. 
+<<heatmap,results=hide,eval=FALSE>>=
+ISA2heatmap(modules, 1, ALL, norm="sample", scale="none", 
+            col=heat.colors(12))
+@
+
+We also create a color bar.
+<<heatmap-cb,results=hide,eval=FALSE>>=
+sc <- function(x, ab) {
+  x <- x-min(x)
+  r <- range(x) 
+  x <- x/(r[2]-r[1])
+  x * (ab[2]-ab[1]) + ab[1]
+}
+cb <- sc(1:12, range(exprs(ALL)[getFeatureNames(modules, 1)[[1]],
+                                      getSampleNames(modules, 2)[[1]] ]))
+par(mar=c(3,4,5,2)+0.1)
+image(rbind(cb), axes=FALSE)
+axis(2, at=sc(1:12, 0:1), cb, label=format(cb, digits=2))
+@
+
+The result is in Fig.~\ref{fig:heatmap}.
+
+<<save heatmap files,echo=FALSE,print=FALSE,results=hide>>=
+png("plot-heatmap.png", 400, 400)
+<<heatmap>>
+dev.off()
+png("plot-heatmap-colorbar.png", 100, 400)
+<<heatmap-cb>>
+dev.off()
 @
@@
 \centering
-<<heatmap-real,fig=TRUE,echo=FALSE,results=hide>>=
-<<heatmap>>
-@
+\includegraphics[width=0.8\textwidth]{plot-heatmap.png}%
+\includegraphics[width=0.2\textwidth]{plot-heatmap-colorbar.png}
 \caption{Heatmap of the first ISA module, plotted via the
-\Rfunction{ISA2heatmap} function. The expression values are
-sample-wised normalized.}
+\Rfunction{ISA2heatmap} function.}
 \label{fig:heatmap}
@@
 \Rfunction{heatmap} for details.
-
-You can also use the \Rpackage{biclust} package to create image
-plots, by calling the \Rfunction{drawHeatmap} function. The result is
-in Fig.~\ref{fig:drawHeatmap}.
-<<heatmap biclust pre,echo=FALSE,results=hide>>=
-if (require(Cairo)) {
-  CairoPNG("plot-heatmap.png", 400, 400 )  
-} else {
-  png("plot-heatmap.png", 400, 400 )
-}
-@ 
-<<heatmap>>=
-drawHeatmap(exprs(ALL[featureNames(modules),]), Bc, number=1) 
-@
-<<heatmap post,echo=FALSE,results=hide>>=
-dev.off() 
-@ 
-\begin{figure}
-  \centering
-  \includegraphics[width=0.7\textwidth]{plot-heatmap.png}
-  \caption{Heatmap of the first transcription module, plotted with the
-  \Rfunction{drawHeatmap} function from the \Rpackage{biclust}
-  package.}
-  \label{fig:drawHeatmap}
-\end{figure}
 
@@
 
-The \Rpackage{ExpressionView} package~\cite{ev} tries to reorder the
+The \Rpackage{ExpressionView} package tries to reorder the
 rows and columns of the expression matrix in such a way that the

### 2010-03-15 isanote/answers_to_reviewers.tex
Refine answers to reviewers.

@@
 
-The reviewer is right, that the module plotted in section 6.5 cannot
-be used to separate B-cell and T-cell samples, unfortunately the wrong
-module was plotted. We corrected this.
-
 We have also explained sample score plots better in the tutorial. 
@@
 package, is the place for that.
+
+The module plotted in section 6.5 indeed cannot be used to separate
+B-cell and T-cell samples, we point this out in the text.
 
@@
 
-We have added the color legends.
+We have added the color legend to the first heatmap. We removed the 
+second heatmap from the tutorial entirely, as adding a heatmap to 
+it would be too cumbersome for the user.
 

### 2010-03-15 isanote/modular.tex
ISA paper: added Sven's changes.

@@
 The output of ISA is a collection of potentially overlapping
-modules. Every module contains genes, that are up- and/or
-down-regulated, in samples that belong to the module.
+modules. Every module contains genes, that are over- and/or
+under-expressed, in samples that belong to the module.
 In every module, each gene and each sample is attributed a
@@
 
-For other biclustering algorithms, please see
-\citet{cheng00,getz00,califano00,sharan02,tanay04,barkow06}.
+For other biclustering algorithms, please see e.g.
+\citet{cheng00,getz00,califano00,sharan02,tanay04,barkow06} and
+\citet{ihmels04} for a review.
 
@@
   sample scores.
-  \subfig{(C)} Module tree. Each rectangle is a module; see the
-  definition of the edges in the text. Modules are colored
+  \subfig{(C)} Module tree. Each rectangle is a module; its numeric id
+  is shown on top of it. See the definition of the edges in the
+  text. Modules are colored 
   according to their Gene Ontology enrichment $p$-values, the codes of
@@
 additional data normalization is crucial to reduce the bias due to
-constituent datasets. Several methods address this challenge,
+the constituent datasets. Several methods address this challenge,
 see e.g.~\citep{johnson07} for an algorithm that has a GNU R
@@
 any information and may reflect ineffective array probes, etc. Since these
-genes are likely to contribute only noise to the
+genes are likely to contribute mostly noise to the
 analysis~\citep{hackstadt09}, we suggest removing them before running
@@
 % Normalization.
-\emph{ISA normalization.} (Step 1 on Fig.~\ref{fig:workflow}.)
+\emph{ISA normalization.} (Step 1 in Fig.~\ref{fig:workflow}.)
 In each iteration the ISA computes thresholded weighted sums of expression

### 2010-03-15 isanote/answers_to_reviewers.tex
ISA paper: added Sven's changes.
ISA paper: added Sven's final comments, ready for (re)submission.

@@
 \usepackage{natbib}
+\usepackage{color}
 
@@
 \newcommand{\Rclass}[1]{\textsl{#1}}
+
+\newenvironment{myquote}{\begin{quote}\color{blue}}{\end{quote}}
 
@@
 and useful suggestions. In the following, the reviewers' comments are
-indented as a block.
+typeset in blue, and indented as a block.
 
-\section{Reviewer 1}
+\section*{Reviewer 1}
 
-\begin{quote}
+\begin{myquote}
 Comments to the Author
@@
 to interested users. 
-\end{quote}
+\end{myquote}
 
-\begin{quote}
+\begin{myquote}
 I have two problems with the Note in its present form.
@@
 do it briefly and sacrifice a few lines from elsewhere. 
-\end{quote}
+\end{myquote}
 
@@
 \textsl{The output of ISA is a collection of potentially overlapping
-modules. Every module contains genes, that are up- and/or
-down-regulated, in samples that belong to the module.
+modules. Every module contains genes, that are over- and/or
+under-expressed, in samples that belong to the module.
 In every module, each gene and each sample is attributed a
@@
 
-\begin{quote}
+\begin{myquote}
 My second concern is minor – the abstract implies that the ISA is
@@
 2000 (Cheng, Getz and Califano) allow this. 
-\end{quote}
+\end{myquote}
 
@@
 
-\section{Reviewer 2}
+\section*{Reviewer 2}
 
-\begin{quote}
+\begin{myquote}
 Comments to the Author
@@
 (ALL) \citep{chiaretti04} as an example. 
-\end{quote}
+\end{myquote}
 
-\begin{quote}
+\begin{myquote}
 Overall evaluation
@@
 be taken into consideration: 
-\end{quote}
+\end{myquote}
 
-\begin{quote}
+\begin{myquote}
 1.      The manuscript and the website lack explanations about how the
@@
 as a function of dataset size should be given. 
-\end{quote}
+\end{myquote}
 
@@
 
-\begin{quote}
+\begin{myquote}
 2.      The tutorials give examples of execution on a single rather
@@
 profiles from multiple studies. 
-\end{quote}
+\end{myquote}
 
@@
 
-\begin{quote}
+\begin{myquote}
 To summarize, the software is well motivated and enables users of the
@@
 conditional on solving the issues mentioned in this report. 
-\end{quote}
+\end{myquote}
 
-\begin{quote}
+\begin{myquote}
 Additional important remarks:
@@
 software) \citep{sharan02,tanay04} and BicAt \citep{barkow06}. 
-\end{quote}
+\end{myquote}
 
@@
 
-\begin{quote}
+\begin{myquote}
 2.      The sample score plots example given in the EISA tutorial
@@
 separate B-cell and T-cell leukemia samples using this plot? 
-\end{quote}
+\end{myquote}
 
-We have also explained sample score plots better in the tutorial. 
+We have explained sample score plots better in the tutorial. 
 Please note, that the tutorials do not aim to explain all details 
@@
 
-\begin{quote}
+\begin{myquote}
 3.      It is not quite clear what is the difference between the ISA
@@
 included in the eisa package. 
-\end{quote}
+\end{myquote}
 
@@
 
-\begin{quote}
+In brief, the \Rpackage{isa2} package implements the ISA iteration and
+provides and interface to people dealing with data, that is not gene
+expression. The \Rpackage{eisa} package provides an interface that
+uses the standard BioConductor data structures. It does not
+reimplement the ISA iteration, but simply calls the implementation in
+the \Rpackage{isa2} package.
+
+The two packages allow ease of installation and use: users dealing
+with gene expression data install the \Rpackage{eisa} package, and
+this automatically installs the \Rpackage{isa2} package as well. Users
+analyzing other data install the \Rpackage{isa2} package only, this
+does not need any BioConductor packages.
+
+\begin{myquote}
 4.      It should be clarified who are the potential users for the ISA
@@
 should be given.
-\end{quote}
+\end{myquote}
 
@@
 
-\begin{quote}
+\begin{myquote}
 The following points are important examples: 
@@
 (or referred to). 
-\end{quote}
+\end{myquote}
 
@@
 
-\begin{quote}
+\begin{myquote}
 b.      Since both ISA packages use various data structures, a short
@@
 these structures to external files. 
-\end{quote}
+\end{myquote}
 
@@
 manual. For the \Rpackage{eisa} package, the \Rclass{ISAModules} class
-is an opaque data type, the representation of it might change in the
+is an opaque data type, i.e. the representation of it might change in the
 future, and the user should not rely on its current
@@
 
-\begin{quote}
+\begin{myquote}
 Minor remarks:
@@
 found in \citet{prelic06}.
-\end{quote}
+\end{myquote}
 
 We have added new classes and functions to perform enrichment analysis
-on the ISA modules, against any gene label or database that the user
+on the ISA modules, against any gene labels or database that the user
 provides. These functions are explained in the manual page of the
@@
 
-\begin{quote}
+\begin{myquote}
 2.      The heatmap image of a transcription module obtained by any
@@
 color legend. 
-\end{quote}
+\end{myquote}
 
@@
 
-\begin{quote}
+\begin{myquote}
 3.      Since ISA function accepts an ExpressionSet, one must obtain
@@
 on it. Annotations should not be a prerequisite for running ISA. 
-\end{quote}
+\end{myquote}
 
@@
 
-\begin{quote}
+\begin{myquote}
 4.      In section 10.8 it is mentioned that two significant modules
@@
 is mistakenly written "As it turns out, one modules found". 
-\end{quote}
+\end{myquote}
 

### 2010-04-06 eisa/inst/doc/EISA_biclust.Rnw
eisa vignette: do not open web browser.

@@
         GO=GO, KEGG=KEGG, CHR=CHR, condPlot=FALSE)
+@
+<<html-show,eval=FALSE>>=
 if (interactive()) {

### 2010-04-12 eisa/inst/doc/EISA_tutorial.Rnw
eisa: tutorials use 'xlib' graphics, if available.

@@
 options(continue=" ")
+try(X11.options(type="xlib"), silent=TRUE)
 @
@@
 <<parallel.pre,echo=FALSE,results=hide>>=
-if (require(Cairo)) {
-  CairoPNG("plot-parallel.png", 400, 400 )
-} else {
 png("plot-parallel.png", 400, 400 )
-}
 @ 

### 2010-04-16 isanote/modular.tex
ISA paper: update to make it fit into two pages.
ISA paper: added Sven's final comments, ready for (re)submission.
ISA application note, final version.

@@
 
-\vspace{-1cm}
+\vspace{-0.7cm}
 \section{Introduction}
@@
   sample scores.
-  \subfig{(C)} Module tree. Each rectangle is a module; its numeric id
-  is shown on top of it. See the definition of the edges in the
+  \subfig{(C)} Module tree. Each module is represented by a rectangle
+  with its numeric id in the center. See the definition of the edges in the
   text. Modules are colored 
@@
   conditions in the module. The gene thresholds used for finding
-  the modules is shown on the horizontal axes.
+  the modules are shown on the horizontal axes.
 }
@@
 biclustering algorithms in a unified framework. The \Rpackage{eisa}
-package include tools to convert between \Rpackage{biclust} and ISA
+package includes tools to convert between \Rpackage{biclust} and ISA
 biclusters. This allows the cross-talk of the functions in the two
@@
 
-\vspace*{-0.4cm}
+\vspace*{-0.5cm}
 \section*{Acknowledgement}

### 2010-04-16 RExpressionView/R/export.R
EV package: GO and KEGG results can be supplied now, then they are not recalculated.

@@ -27,7 +27,9 @@ if (require(eisa)) {
     )
 }
 
-ExportEV.ISAModules <- function(biclusters, eset, order, filename, norm, cutoff, description) {
+ExportEV.ISAModules <- function(biclusters, eset, order, filename, norm, cutoff,
+                                description, GO=ISAGO(biclusters),
+                                KEGG=ISAKEGG(biclusters)) {
 
     if ( missing(order) ) {
         order <- OrderEV(biclusters)
@@ -81,10 +83,7 @@ ExportEV.ISAModules <- function(biclusters, eset, order, filename, norm, cutoff,
     writeLines("\t</summary>", con)
 
     go.table <- toTable(GOTERM)
-    gos <- ISAGO(biclusters)
-
     kegg.table <- toTable(KEGGPATHID2NAME)
-    keggs <- ISAKEGG(biclusters)
 
     ## get gene info
     ann <- annotation(eset)
@@ -279,7 +278,7 @@ ExportEV.ISAModules <- function(biclusters, eset, order, filename, norm, cutoff,
             writeLines("\t\t\t<gos>", con)
             k <- 1
             for ( i in 1:3 ) {
-                s <- summary(gos[[i]])[[module]]
+                s <- summary(GO[[i]])[[module]]
                 if ( dim(s)[1] > 0 ) {
                     temp <- match(rownames(s), go.table[,1])
                     
@@ -305,7 +304,7 @@ ExportEV.ISAModules <- function(biclusters, eset, order, filename, norm, cutoff,
             writeLines("", con)
 
             writeLines("\t\t\t<keggs>", con)
-            s <- summary(keggs)[[module]]
+            s <- summary(KEGG)[[module]]
             if ( dim(s)[1] > 0 ) {
                 temp <- match(rownames(s), kegg.table[,1])
                 for ( j in 1:dim(s)[1] ) {

### 2010-04-21 RExpressionView/inst/doc/ExpressionView.ordering.Rnw
EV: improved vignettes.

@@
 \SweaveOpts{cache=TRUE}
+
+\setlength{\parindent}{2em}
 
@@
 The problem can be formulated very generally:
-Given a matrix in which each element is part of at least one bicluster, arrange the rows and columns
+given a matrix in which each element is part of at least one bicluster, arrange the rows and columns
 in such a way that the biclusters are easily recognizable on a two-dimensional plot.
@@
 Optimizing the order thus consists in maximizing $Q$ by changing the order of the slots.
-To achieve this goal we use a brute-force approach that 
+To achieve this goal we use a greedy approach that 
 \begin{itemize}
@@
 For smaller samples, the fitness function is used, while for larger samples only the similarity (scalar product) between rows is taken into account.}
-	\item{\code{arrange}: Uses the brute-force approach to further increase the alignment score.}
+	\item{\code{arrange}: Uses the greedy approach to further increase the alignment score.}
 	\item{\code{complexify}: Reinstates the original dimensions of the problem by inserting the omitted rows at the newly found position.}
@@
 
-The ordering iterates over the clusters and first calls the function \code{reposition}, taking \code{cluster} as argument.
-It goes through all the rows contained in a given cluster and tries to shift them to a better position.
-A contiguous group of rows can be repositioned at once or split up in subgroups.
-Empirically, it turned out that the alignment is fastest by starting the repositioning with a single row, increasing the length of the group only if no better position could be found.
-The opposite approach can be used by replacing the \emph{start forward} section on line 545 by the \emph{start backwards} code on line 556.
+The ordering iterates over the clusters and first tries to shift the
+rows of them to a better position. A contiguous group of rows can be
+repositioned at once or split up in subgroups. Empirically, it turned
+out that the alignment is fastest by starting the repositioning with a
+single row, increasing the length of the group only if no better
+position could be found.
 
-After repositioning, the main loop calls the exchange function, taking \code{cluster} and \code{what} as argument.
-It exchanges rows contained (\code{what = 1}) / not contained (\code{what = 0}) in a given cluster.
-It randomly picks \code{maxslots} number of rows (currently six), generates all possible permutations, calculates the corresponding alignment scores, and retains the best order.
-This realignment is repeated \code{maxiter} times, so that on average every slot is part of one optimization.
+After repositioning, the algorithm tries to exchanges rows
+contained or not contained in a given cluster. It randomly picks
+a number of rows (up to six right now) and generates all their
+possible permutations, calculates the corresponding alignment scores,
+and retains the best order. This realignment is repeated
+a number of times, so that on average every slot is part of one
+optimization.
 
-The loop over clusters calls the \code{reposition} function first for all the rows not contained in a given cluster and afterwards for the remaining ones.
-The main loop is repeated until the maximal computation time is reached or no improved alignment could be found anymore.
+These two steps are repeated until the maximal computation time is
+reached or no improved alignment could be found anymore.
 
-The \code{C++} program can also be compiled to a standalone executable by defining \code{-DSTANDALONE}
-to exclude R-specific functions. You can also define \code{-DTIMING} to measure the time spent in the function
-that calculates $Q$ (\code{getfitness}).
+The \code{C++} program can also be compiled to a standalone executable
+by defining \code{-DSTANDALONE} to exclude R-specific functions. You
+can also define \code{-DTIMING} to measure the time spent in the
+function that calculates $Q$ (\code{getfitness}).
 
@@
 \section{Possible improvements}
-More than 90\% of the calculation time is spent to recalculate the length of the longest contiguous subclusters. Optimizing the \code{getfitness} routine
-thus has a direct impact on the overall performance of the algorithm. Since the different clusters can be treated independently, it is
-straightforward to parallelize the calculation, maybe even on the graphic chip.
+More than 90\% of the calculation time is spent to recalculate the
+length of the longest contiguous subclusters. Optimizing this
+algorithm thus has a direct impact on the overall performance of the
+algorithm. Since the different clusters can be treated independently,
+it is straightforward to parallelize the calculation, maybe even on
+the graphic chip.
 

### 2010-04-21 RExpressionView/inst/doc/ExpressionView.tutorial.Rnw
ExpressionView paper and docs corrections.
EV: improved vignettes.

@@
 %\SweaveOpts{prefix.string=graphics/plot}
+
+\setlength{\parindent}{2em}
 
@@
 an R package designed to interactively explore biclusters identified in gene 
-expression data in conjunction with the Iterative Signature Algorithm (ISA)~\cite{bergmann03} 
+expression data, in conjunction with the Iterative Signature Algorithm (ISA)~\cite{bergmann03} 
 and the biclustering methods available in the \Rpackage{Biclust} package~\cite{kaiser08}.
@@
 The \Rpackage{ExpressionView} package requires the gene expression data to be available in the
-form of a \Rpackage{Bioconductor} \Rclass{ExpressionSet}. In this tutorial we will use the 
-\Rpackage{Bioconductor} sample data from a 
+form of a BioConductor \Rclass{ExpressionSet}. In this tutorial we will use the 
+BioConductor sample data from a 
 clinical trial in acute lymphoblastic leukemia provided by the \Rpackage{ALL} package.
@@
 All of them aim to reduce the complexity of the gene expression data by identifying 
-suitable groups of genes and conditions that are co-regulated. In what follows we
-show how to use two of the freely available biclustering methods.
+suitable groups of genes and conditions that are co-expressed. In this tutorial
+show how to use \Rpackage{ExpressionView} with some of the available biclustering algorithms.
 
@@
 
-The ISA biclustering is not binary, every gene (and
-similarly, every sample or condition) has a score between -1 and 1; the further the
-score is from zero, the stronger the association between the gene
-(or sample) and the module. If two features both have scores with the
-same sign, they are correlated, otherwise, they are anti-correlated.
+This object can be directly used with the functions of the
+\Rpackage{ExpressionView} package. See Section~\ref{sec:ordering} for
+details.
 
@@
 
-The standard R function \Rfunction{as} 
-can be used to convert \Rclass{ISAModules} to \Rclass{Biclust} objects. This
-requires the binarization of the modules, i.e., the ISA scores are
-lost since they are converted to zeros and ones:
-<<biclust>>=
-as(modules, "Biclust")
-@
-
-The conversion also works in the other direction. Starting from a \Rclass{Biclust} object, just use
+\Rclass{Biclust} objects can be directly used with the
+\Rpackage{ExpressionView} functions. Alternatively, they can be
+converted to \Rclass{ISAModules} objects, using the standard
+\Rfunction{as} R function:
 <<isa>>=
@@
 @
-to get an \Rclass{ISAModules} object.
+results an \Rclass{ISAModules} object.
 
@@
 
-\section{Order}
+\section{Order}%
+\label{sec:ordering}
 To present the tens of possibly overlapping biclusters in a visually appealing form, 
-it is necessary to regroup the rows and columns of the gene expression matrix in 
+it is necessary to reorder the rows and columns of the gene expression matrix in 
 such a way that biclusters form contiguous rectangles. Since for more than two 
 mutually overlapping biclusters, it is in general impossible to find such an arrangement,
- one has to make concessions. In contrast to~\cite{grothaus06}, who proposed to repeat 
- rows and columns as necessary to achieve this goal, we prefer to optimize the 
+ one has to make concessions. In contrast methods that propose to repeat 
+ rows and columns as necessary to achieve this goal~\cite{grothaus06}, we prefer to optimize the 
  arrangement within the original data by maximizing the area of the largest contiguous biclusters.
@@
 The \Rfunction{OrderEV} function implemented in the \Rpackage{ExpressionView} package
-determines the optimal order of the gene expression matrix for a given set of biclusters:
+determines the optimal order of the gene expression matrix for a given
+set of biclusters. It can be called with \Rclass{ISAModules} or
+\Rclass{Biclust} objects as the first argument:
 <<expressionview, results=HIDE>>=
@@
 @
-Here, \variable{i=0}, returns the map of the complete data set. By default, the \Rfunction{OrderEV}
+The first elements of the lists contain the optimal ordering of the
+complete matrix. By default, the \Rfunction{OrderEV} function
 runs for roughly one minute, this might not be sufficient to find an appropriate order for data 
@@
 If the status is set to \variable{1}, the algorithm has found the optimal solution. A \variable{0} indicates that the
-the calculation could not be terminated within the given timeframe. The \Rfunction{OrderEV} 
+the calculation could not be terminated within the given timeframe. The \Rfunction{OrderEV} function
 accepts two additional parameters to circumvent the problem of partial alignment:
@@
 The file association is the reason why we do not use the \variable{.xml} extension.
-A description of the XML layout can be found on the \href{http://www2.unil.ch/cbg/index.php?title=ExpressionView}{website}.
+A description of the XML layout can be found on the \Rpackage{ExpressionView} website at \url{http://www.unil.ch/cbg/ExpressionView}.
 Before exporting the data, the
@@
 The outlines of these parts are drawn in a slightly brighter color than the background, making them easily recognizable.
-	\item[g, h] Modules are highlighted as the user moves the moves over the gene expression data.
+	\item[g, h] Modules are highlighted as the user moves the mouse over the gene expression data.
 The two check boxes allow you to choose between highlighting all the parts of a module (Filling) or alternatively only the largest rectangle (Outline).
@@
 <<data set annotation>>=
-v <- vector("character"); for ( i in 1:dim(data)[1] ) { v[i] <- paste("row ", i, sep="") }; rownames(data) <- v
-v <- vector("character"); for ( i in 1:dim(data)[2] ) { v[i] <- paste("column ", i, sep="") }; colnames(data) <- v
+rownames(data) <- paste("row", seq_len(nrow(data)))
+colnames(data) <- paste("column", seq_len(ncol(data)))
 @
@@
 <<rowdata annotation>>=
-rowdata <- matrix(nrow=dim(data)[1], ncol=5)
+rowdata <- outer(1:nrow(data), 1:5, function(x, y) {
+  paste("row description (", x, ", ", y, ")", sep="")
+})             
 rownames(rowdata) <- rownames(data)
-v <- vector("character"); for ( i in 1:dim(rowdata)[2] ) { v[i] <- paste("row tag ", i, sep="") }; colnames(rowdata) <- v
-for ( i in 1:dim(rowdata)[1] ) { for ( j in 1:dim(rowdata)[2] ) { rowdata[i,j] <- paste("row description (", i, ", ", j, ")", sep="") }}
+colnames(rowdata) <- paste("row tag", seq_len(ncol(rowdata)))
 @
-And similarly for the columns, using 10 "column tags``:
+And similarly for the columns, using 10 ``column tags'':
 <<coldata annotation>>=
-coldata <- matrix(nrow=dim(data)[2], ncol=10)
+coldata <- outer(1:ncol(data), 1:10, function(x, y) {
+  paste("column description (", x, ", ", y, ")", sep="")
+})
 rownames(coldata) <- colnames(data)
-v <- vector("character"); for ( i in 1:dim(coldata)[2] ) { v[i] <- paste("column tag ", i, sep="") }; colnames(coldata) <- v
-for ( i in 1:dim(coldata)[1] ) { for ( j in 1:dim(coldata)[2] ) { coldata[i,j] <- paste("column description (", i, ", ", j, ")", sep="") }}
+colnames(coldata) <- paste("column tag", seq_len(ncol(coldata)))
 @

### 2010-04-21 RExpressionView/inst/doc/ExpressionView.ordering.Rnw
EV: reindented vignettes.

@@
 \section{Introduction}
-Clustering genes according to their expression profiles is an important task 
-in analyzing microarray data. To present the mutually overlapping biclusters
-in a visually appealing layout, the gene expression matrix has to be arranged
-in such a way that biclusters form contiguous rectangles. For more than two biclusters,
-this is in general impossible. Here, we present an algorithm that finds the arrangement
-that maximizes the areas of the largest contiguous parts of the biclusters.
+Clustering genes according to their expression profiles is an
+important task in analyzing microarray data. To present the mutually
+overlapping biclusters in a visually appealing layout, the gene
+expression matrix has to be arranged in such a way that biclusters
+form contiguous rectangles. For more than two biclusters, this is in
+general impossible. Here, we present an algorithm that finds the
+arrangement that maximizes the areas of the largest contiguous parts
+of the biclusters.
 
@@
 The problem can be formulated very generally:
-given a matrix in which each element is part of at least one bicluster, arrange the rows and columns
-in such a way that the biclusters are easily recognizable on a two-dimensional plot.
-Since biclusters are rectangular, rows and columns can be ordered separately.
-The problem thus reduces to finding the optimal arrangement of a set of $n$ elements 
-that are part of at least one of $m$ clusters. In what follows, we refer to the $n$ elements as
-slots.
+given a matrix in which each element is part of at least one
+bicluster, arrange the rows and columns 
+in such a way that the biclusters are easily recognizable on a
+two-dimensional plot. Since biclusters are rectangular, rows and
+columns can be ordered separately. The problem thus reduces to finding
+the optimal arrangement of a set of $n$ elements that are part of at
+least one of $m$ clusters. In what follows, we refer to the $n$
+elements as slots.
 
 \section{Quantifying the order}
-There are several ways of quantifying a ``visually appealing layout''. We have chosen to
-define the quality of the order $Q$ as the sum over the maximal number of neighboring elements 
-in every cluster, as illustrated in Fig.~\ref{fig:ordering}.
+There are several ways of quantifying a ``visually appealing
+layout''. We have chosen to define the quality of the order $Q$ as the
+sum over the maximal number of neighboring elements in every cluster,
+as illustrated in Fig.~\ref{fig:ordering}.
 \begin{figure} 
@@
   \end{center} 
-	\caption{The initial configuration (a) with a score $Q$=58 can be rearranged by modifying the order of the slots. The optimal 
+  \caption{The initial configuration (a) with a score $Q$=58 can be
+    rearranged by modifying the order of the slots. The optimal
     solution (b) maximizes $Q$.} 
@@
 \section{Optimizing the order}
-Optimizing the order thus consists in maximizing $Q$ by changing the order of the slots.
-To achieve this goal we use a greedy approach that 
+Optimizing the order thus consists in maximizing $Q$ by changing the
+order of the slots. To achieve this goal we use a greedy approach that 
 \begin{itemize}
@@
 
-Internally, the data is represented as a binary matrix (i.e., elements are either 0 or 1) with dimensions $m$ $\times$ $n$, where the number of rows $m$ correspond to the number of genes or conditions and the number of columns $n$ is determined by the number of clusters.
+Internally, the data is represented as a binary matrix (i.e., elements
+are either 0 or 1) with dimensions $m$ $\times$ $n$, where the number
+of rows $m$ correspond to the number of genes or conditions and the
+number of columns $n$ is determined by the number of clusters.
 
@@
 \begin{enumerate}
-	\item{\code{simplify}: Reduces the dimensions of the problem by excluding duplicate rows. The multiplicity of each slot is taken into account when calculating the score $Q$.}
-	\item{\code{prearrange}: Tries to find a good starting point for the optimization by creating a new order, starting from the first row and adding the consecutive rows one after the other, placing them at the optimal position.
-For smaller samples, the fitness function is used, while for larger samples only the similarity (scalar product) between rows is taken into account.}
-	\item{\code{arrange}: Uses the greedy approach to further increase the alignment score.}
-	\item{\code{complexify}: Reinstates the original dimensions of the problem by inserting the omitted rows at the newly found position.}
+\item \code{simplify}: Reduces the dimensions of the problem by
+  excluding duplicate rows. The multiplicity of each slot is taken
+  into account when calculating the score $Q$.
+\item \code{prearrange}: Tries to find a good starting point for the
+  optimization by creating a new order, starting from the first row
+  and adding the consecutive rows one after the other, placing them at
+  the optimal position. For smaller samples, the fitness function is
+  used, while for larger samples only the similarity (scalar product)
+  between rows is taken into account.
+\item \code{arrange}: Uses the greedy approach to further increase the
+  alignment score.
+\item \code{complexify}: Reinstates the original dimensions of the
+  problem by inserting the omitted rows at the newly found position.
 \end{enumerate}
 
-If the user specifies a time limit, the algorithm stops the alignment after the elapsed interval and returns the best alignment.
-Otherwise, it continues until it can no longer improve the alignment score.
+If the user specifies a time limit, the algorithm stops the alignment
+after the elapsed interval and returns the best alignment. Otherwise,
+it continues until it can no longer improve the alignment score.
 
@@
 \section{Calculation time}
-To get an idea of the complexity of the algorithm, we determined the scaling of the calculation time 
-with the dimensions of the data, i.e., the number of clusters $m$ and the slots $n$. The results, obtained as
-the average over at least 40 runs, are summarized in Figs.~\ref{fig:clusters} and \ref{fig:slots}. 
+To get an idea of the complexity of the algorithm, we determined the
+scaling of the calculation time with the dimensions of the data, i.e.,
+the number of clusters $m$ and the slots $n$. The results, obtained as
+the average over at least 40 runs, are summarized in
+Figs.~\ref{fig:clusters} and \ref{fig:slots}.
 \begin{figure} 
@@
   \end{center} 
-	\caption{Scaling of the execution time as a function of the number of clusters. Data points connected
-	by dashed lines represent results obtained from orderable samples. The second set is obtained
-	from random samples and is thus more representative of gene expression data. Both sets scale polynomially
-	with the number of clusters $m$ as $\mathcal{O}(m^\alpha)$, with $\alpha \in [1.6, 2]$ for random samples and
-	$\alpha \in [2.3, 4.2]$ in the case of orderable samples.} 
+  \caption{Scaling of the execution time as a function of the number
+    of clusters. Data points connected by dashed lines represent
+    results obtained from orderable samples. The second set is
+    obtained from random samples and is thus more representative of
+    gene expression data. Both sets scale polynomially 	with the
+    number of clusters $m$ as $\mathcal{O}(m^\alpha)$, with $\alpha
+    \in [1.6, 2]$ for random samples and $\alpha \in [2.3, 4.2]$ in the
+    case of orderable samples.}
   \label{fig:clusters} 
 \end{figure}
-We consider orderable and completely random samples. In both cases, the time increases polynomially with the number
-of clusters as $\mathcal{O}(m^\alpha)$, but with different exponents, see Fig.~\ref{fig:clusters}: For random samples, we find 
-$\alpha \in [1.6, 2]$, almost independent of the number of elements $n$, while for orderable samples, 
-$\alpha \in [2.3, 4.3]$. We note that despite of having a slower increase of computation time with $m$, 
-it takes considerably more time to order random samples.
+We consider orderable and completely random samples. In both cases,
+the time increases polynomially with the number of clusters as
+$\mathcal{O}(m^\alpha)$, but with different exponents, see
+Fig.~\ref{fig:clusters}: For random samples, we find $\alpha \in [1.6,
+2]$, almost independent of the number of elements $n$, while for
+orderable samples, $\alpha \in [2.3, 4.3]$. We note that despite of
+having a slower increase of computation time with $m$, it takes
+considerably more time to order random samples.
 
-The scaling of the computation time with the number of slots $n$ is shown in Fig.~\ref{fig:slots}. 
+The scaling of the computation time with the number of slots $n$ is
+shown in Fig.~\ref{fig:slots}.
 \begin{figure} 
@@
   \end{center} 
-	\caption{Scaling of the execution time as a function of the number of slots. The second set is obtained
-	from random samples and is thus more representative of gene expression data. The random set scales polynomially
-	with the number of slots $n$ as $\mathcal{O}(n^\beta)$, with $\beta \in [2.5, 2.7]$. The orderable data set shows an initially
-	polynomial increase and then saturates for larger $n$.} 
+  \caption{Scaling of the execution time as a function of the number
+    of slots. The second set is obtained from random samples and is
+    thus more representative of gene expression data. The random set
+    scales polynomially with the number of slots $n$ as
+    $\mathcal{O}(n^\beta)$, with $\beta \in [2.5, 2.7]$. The orderable
+    data set shows an initially polynomial increase and then saturates
+    for larger $n$.}
   \label{fig:slots} 
 \end{figure}
-Here, only the random samples exhibit a polynomial dependence $\mathcal{O}(n^\beta)$, with $\beta \in [2.5, 2.7]$. 
-The orderable batch saturates at larger $n$. This is due to the fact that by increasing the number of slots $n$ while keeping
-the number of clusters constant, the complexity of the problem stays roughly the same because most of the additional slots are 
-equal to those already contained in the data.
+Here, only the random samples exhibit a polynomial dependence
+$\mathcal{O}(n^\beta)$, with $\beta \in [2.5, 2.7]$. The orderable
+batch saturates at larger $n$. This is due to the fact that by
+increasing the number of slots $n$ while keeping the number of
+clusters constant, the complexity of the problem stays roughly the
+same because most of the additional slots are equal to those already
+contained in the data.
 

### 2010-04-21 RExpressionView/inst/doc/ExpressionView.tutorial.Rnw
EV: reindented vignettes.

@@
 \section{Introduction}
-Clustering genes according to their expression profiles is an important task 
-in analyzing microarray data. In this tutorial, we explain how to use ExpressionView, 
-an R package designed to interactively explore biclusters identified in gene 
-expression data, in conjunction with the Iterative Signature Algorithm (ISA)~\cite{bergmann03} 
-and the biclustering methods available in the \Rpackage{Biclust} package~\cite{kaiser08}.
+Clustering genes according to their expression profiles is an
+important task in analyzing microarray data. In this tutorial, we
+explain how to use ExpressionView, an R package designed to
+interactively explore biclusters identified in gene expression data,
+in conjunction with the Iterative Signature Algorithm
+(ISA)~\cite{bergmann03} and the biclustering methods available in the
+\Rpackage{Biclust} package~\cite{kaiser08}.
 
 \section{Loading the gene expression data}
-The \Rpackage{ExpressionView} package requires the gene expression data to be available in the
-form of a BioConductor \Rclass{ExpressionSet}. In this tutorial we will use the 
-BioConductor sample data from a 
-clinical trial in acute lymphoblastic leukemia provided by the \Rpackage{ALL} package.
+The \Rpackage{ExpressionView} package requires the gene expression
+data to be available in the form of a BioConductor
+\Rclass{ExpressionSet}. In this tutorial we will use the BioConductor
+sample data from a clinical trial in acute lymphoblastic leukemia
+provided by the \Rpackage{ALL} package.
 <<load the data>>=
@@
 @
-The data set contains \Sexpr{ncol(ALL)} samples and \Sexpr{nrow(ALL)} features.
+The data set contains \Sexpr{ncol(ALL)} samples and \Sexpr{nrow(ALL)}
+features.
 
 \section{Find biclusters}
-There are many biclustering algorithms described in the literature~\cite{madeira04}.
-All of them aim to reduce the complexity of the gene expression data by identifying 
-suitable groups of genes and conditions that are co-expressed. In this tutorial
-show how to use \Rpackage{ExpressionView} with some of the available biclustering algorithms.
+There are many biclustering algorithms described in the
+literature~\cite{madeira04}. All of them aim to reduce the complexity
+of the gene expression data by identifying suitable groups of genes
+and conditions that are co-expressed. In this tutorial show how to use
+\Rpackage{ExpressionView} with some of the available biclustering
+algorithms.
 
 \subsection{Iterative Signature Algorithm (ISA)}
-The ISA~\cite{bergmann03} for gene expression data is implemented in the \Rpackage{eisa} package:
+The ISA~\cite{bergmann03} for gene expression data is implemented in
+the \Rpackage{eisa} package:
 <<load the packages>>=
@@
 
-To run the ISA for the given data set, simply call the \Rfunction{ISA} function on the
+To run the ISA for the given data set, simply call the \Rfunction{ISA}
+function on the
 \Rclass{ExpressionSet} object:
@@
 @ 
-Depending on your computing resources, this should take roughly two minutes. If you do not
-want to wait that long, you can shorten the calculation by selecting the thresholds for genes and
-conditions:
+Depending on your computing resources, this should take roughly two
+minutes. If you do not want to wait that long, you can shorten the
+calculation by selecting the thresholds for genes and conditions:
 <<fast ISA, cache=FALSE>>=
@@
 @
-If you leave the thresholds undefined, as in the first example, the ISA runs with the 
-default values, i.e., \variable{thr.gene=c(2,2.5,3,3.5,4)} and \variable{thr.cond=c(1,1.5,2,2.5,3)}.
-In both cases, the random number generator is initialized manually using \Rfunction{set.seed(5)}, 
-to give reproducible results. The \Rfunction{isa} function returns an \Rclass{ISAModules} object. 
-Typing its name returns a brief summary of the results:
+If you leave the thresholds undefined, as in the first example, the
+ISA runs with the default values, i.e.,
+\variable{thr.gene=c(2,2.5,3,3.5,4)} and
+\variable{thr.cond=c(1,1.5,2,2.5,3)}. In both cases, the random number
+generator is initialized manually using \Rfunction{set.seed(5)}, to
+give reproducible results. The \Rfunction{isa} function returns an
+\Rclass{ISAModules} object. Typing its name returns a brief summary of
+the results:
 <<ISAModules summary>>=
@@
 algorithms in a unified framework. It uses the \Rclass{Biclust} class 
-to store a set of biclusters. Let us use the Plaid Model Bicluster Algorithm~\cite{turner04}
-on the \Rpackage{ALL} data set
+to store a set of biclusters. Let us use the Plaid Model Bicluster
+Algorithm~\cite{turner04} on the \Rpackage{ALL} data set
 <<biclust bcplaid,cache=FALSE>>=
@@
 \subsection{External clustering programs}
-Since the structure of biclustering results is independent of the applied method, it is straightforward to
-import results obtained from external clustering programs and convert them to \Rclass{ISAModules}.
-To illustrate the conversion, let us consider the sample data and {\bf randomly} assign the  \Sexpr{nrow(ALL)} genes 
-and  \Sexpr{ncol(ALL)} samples to  \Sexpr{length(modules)} modules. The resulting modules can be described by two
-binary matrices
+Since the structure of biclustering results is independent of the
+applied method, it is straightforward to import results obtained from
+external clustering programs and convert them to
+\Rclass{ISAModules}. To illustrate the conversion, let us consider the
+sample data and {\bf randomly} assign the  \Sexpr{nrow(ALL)} genes and
+\Sexpr{ncol(ALL)} samples to  \Sexpr{length(modules)} modules. The
+resulting modules can be described by two binary matrices
 <<random-modules>>=
-modules.genes <- matrix(as.integer(runif(dim(ALL)[1]*length(modules),0,1)>0.8),nrow=dim(ALL)[1])
-modules.conditions <- matrix(as.integer(runif(dim(ALL)[2]*length(modules),0,1)>0.8),nrow=dim(ALL)[2])
+modules.genes <- matrix(as.integer(runif(nrow(ALL) * length(modules)) > 0.8), 
+                        nrow=nrow(ALL))
+modules.conditions <- matrix(as.integer(runif(ncol(ALL) *
+                                        length(modules))>0.8),
+                        nrow=ncol(ALL))
 @
-indicating if a given gene \variable{i} is contained in module \variable{j} if \variable{modules.genes[i,j]$\ne$0}.
-Using these matrices, it is straightforward to create an \Rclass{ISAModules} object:
+indicating if a given gene \variable{i} is contained in module
+\variable{j} if \variable{modules.genes[i,j]$\ne$0}. Using these
+matrices, it is straightforward to create an \Rclass{ISAModules}
+object: 
 <<toisa>>=
 new("ISAModules",
-genes=modules.genes, conditions=modules.conditions,rundata=data.frame(),seeddata=data.frame())
+    genes=modules.genes, conditions=modules.conditions,
+    rundata=data.frame(), seeddata=data.frame())
 @
@@
 \label{sec:ordering}
-To present the tens of possibly overlapping biclusters in a visually appealing form, 
-it is necessary to reorder the rows and columns of the gene expression matrix in 
-such a way that biclusters form contiguous rectangles. Since for more than two 
-mutually overlapping biclusters, it is in general impossible to find such an arrangement,
- one has to make concessions. In contrast methods that propose to repeat 
- rows and columns as necessary to achieve this goal~\cite{grothaus06}, we prefer to optimize the 
- arrangement within the original data by maximizing the area of the largest contiguous biclusters.
+To present the tens of possibly overlapping biclusters in a visually
+appealing form, it is necessary to reorder the rows and columns of the
+gene expression matrix in such a way that biclusters form contiguous
+rectangles. Since for more than two mutually overlapping biclusters,
+it is in general impossible to find such an arrangement, one has to
+make concessions. In contrast methods that propose to repeat rows and
+columns as necessary to achieve this goal~\cite{grothaus06}, we prefer
+to optimize the arrangement within the original data by maximizing the
+area of the largest contiguous biclusters.
 
-The \Rfunction{OrderEV} function implemented in the \Rpackage{ExpressionView} package
-determines the optimal order of the gene expression matrix for a given
-set of biclusters. It can be called with \Rclass{ISAModules} or
-\Rclass{Biclust} objects as the first argument:
+The \Rfunction{OrderEV} function implemented in the
+\Rpackage{ExpressionView} package determines the optimal order of the
+gene expression matrix for a given set of biclusters. It can be called
+with \Rclass{ISAModules} or \Rclass{Biclust} objects as the first argument:
 <<expressionview, results=HIDE>>=
@@
 
-The result is a list containing various mappings between the original data and the optimal arrangement.
-Note that the genes and the samples can be ordered separately. Apart form reordering the 
-full gene expression matrix, the algorithm also determines the best arrangement of individual biclusters.
-The mapping of the genes and the samples contained in bicluster \variable{i} can be accessed by 
+The result is a list containing various mappings between the original
+data and the optimal arrangement. Note that the genes and the samples
+can be ordered separately. Apart form reordering the full gene
+expression matrix, the algorithm also determines the best arrangement
+of individual biclusters. The mapping of the genes and the samples
+contained in bicluster \variable{i} can be accessed by
 <<gene maps, eval=FALSE>>=
@@
 complete matrix. By default, the \Rfunction{OrderEV} function
-runs for roughly one minute, this might not be sufficient to find an appropriate order for data 
-containing many overlapping biclusters. The status of the ordering is stored in 
+runs for roughly one minute, this might not be sufficient to find an
+appropriate order for data containing many overlapping biclusters. The
+status of the ordering is stored in
 <<status ordering>>=
@@
 @
-If the status is set to \variable{1}, the algorithm has found the optimal solution. A \variable{0} indicates that the
-the calculation could not be terminated within the given timeframe. The \Rfunction{OrderEV} function
-accepts two additional parameters to circumvent the problem of partial alignment:
-One can start the ordering from a given initial configuration, i.e., the result of a previous arrangement
-by defining the \Rargument{initialorder} argument
+% $
+If the status is set to \variable{1}, the algorithm has found the
+optimal solution. A \variable{0} indicates that the the calculation
+could not be terminated within the given timeframe. The
+\Rfunction{OrderEV} function accepts two additional parameters to
+circumvent the problem of partial alignment: One can start the
+ordering from a given initial configuration, i.e., the result of a
+previous arrangement by defining the \Rargument{initialorder} argument
 <<status ordering, eval=FALSE>>=
@@
 @
-and one can increase the time limit by specifying \Rargument{maxtime}. Note that the time is indicated in seconds
-and cannot be  smaller than 1.
+and one can increase the time limit by specifying
+\Rargument{maxtime}. Note that the time is indicated in seconds and
+cannot be  smaller than 1.
 
 \section{Export}
-The \Rfunction{ExportEV} function allows the user to combine the available data and export it to an XML file that can be read by the Flash applet:
+The \Rfunction{ExportEV} function allows the user to combine the
+available data and export it to an XML file that can be read by the
+Flash applet:
 <<export, eval=FALSE>>=
@@
 @
-The function gathers the data contained in the \Rclass{ExpressionSet} \variable{ALL}, orders it according to the optimal arrangement 
-\variable{optimalorder} and adds the biclusters defined in \variable{modules}.
-The output is an uncompressed XML file that can be opened with any text viewer.
-We have chosen to use the extension \variable{.evf} (for ExpressionView file) for the data files.
-This extension is associated with the stand-alone version of the viewer, so that one can simply double-click on such a file to launch the program and load the data.
-The file association is the reason why we do not use the \variable{.xml} extension.
-A description of the XML layout can be found on the \Rpackage{ExpressionView} website at \url{http://www.unil.ch/cbg/ExpressionView}.
-Before exporting the data, the
-\Rfunction{ExportEV} function automatically calculates GO~\cite{ashburner00} and KEGG~\cite{kanehisa04} enrichments for the given biclusters.
+The function gathers the data contained in the \Rclass{ExpressionSet}
+\variable{ALL}, orders it according to the optimal arrangement
+\variable{optimalorder} and adds the biclusters defined in
+\variable{modules}. The output is an uncompressed XML file that can be
+opened with any text viewer. We have chosen to use the extension
+\variable{.evf} (for ExpressionView file) for the data files. This
+extension is associated with the stand-alone version of the viewer, so
+that one can simply double-click on such a file to launch the program
+and load the data. The file association is the reason why we do not
+use the \variable{.xml} extension. A description of the XML layout can
+be found on the \Rpackage{ExpressionView} website at
+\url{http://www.unil.ch/cbg/ExpressionView}. Before exporting the
+data, the \Rfunction{ExportEV} function automatically calculates
+GO~\cite{ashburner00} and KEGG~\cite{kanehisa04} enrichments for the
+given biclusters.
 
 \section{Visualize}
-The ExpressionView Flash applet can be launched from the R environment:
+The ExpressionView Flash applet can be launched from the R
+environment:
 <<export, eval=FALSE>>=
@@
 @
-Video tutorials describing how to use the applet can be found on the ExpressionView \\
-\href{http://www2.unil.ch/cbg/index.php?title=ExpressionView}{website}. The screenshot shown in Fig.~\ref{fig:screenshot} 
-and the description below illustrate the main features of the applet:
+Video tutorials describing how to use the applet can be found on the
+ExpressionView website at
+\url{http://www.unil.ch/cbg/ExpressionView}. The screenshot shown in
+Fig.~\ref{fig:screenshot} and the description below illustrate the
+main features of the applet:
 \begin{figure} 
@@
 \begin{description}
-	\item[a] Opens an ExpressionView data file. Note that before opening a new data file, you should restart the applet, i.e., refresh your browser window.
-	\item[b] Exports the current view to a pdf file. The file also includes the title (o) of the gene expression data.
-	\item[c] Exports the data of the currently viewed module (=bicluster) to a CSV file, that can be opened as a spreadsheet.
-	\item[d] In inspect mode, you can use the mouse to explore the gene expression data. The information about the data under the mouse pointer is shown in the Info Panel (t).
-	\item[e, f] Zoom and pan modes allow you to restrict the view to a particular part of the gene expression data.
-	\item[e] In zoom mode, you can also use keyboard shortcuts: {\bf a} to auto-zoom onto the modules and {\bf e} to see the whole data.
-In addition to the simple zoom-in feature, you can also use the mouse to select the rectangular area you want to have a closer look at.
+\item[a] Opens an ExpressionView data file. Note that before opening a
+  new data file, you should restart the applet, i.e., refresh your
+  browser window.
+\item[b] Exports the current view to a pdf file. The file also
+  includes the title (o) of the gene expression data.
+\item[c] Exports the data of the currently viewed module (=bicluster)
+  to a CSV file, that can be opened as a spreadsheet.
+\item[d] In inspect mode, you can use the mouse to explore the gene
+  expression data. The information about the data under the mouse
+  pointer is shown in the Info Panel (t).
+\item[e, f] Zoom and pan modes allow you to restrict the view to a
+  particular part of the gene expression data.
+\item[e] In zoom mode, you can also use keyboard shortcuts: {\bf a} to
+  auto-zoom onto the modules and {\bf e} to see the whole data. In
+  addition to the simple zoom-in feature, you can also use the mouse
+  to select the rectangular area you want to have a closer look at.
 \item[f] Pan mode.
-	\item[g, h, i, j] Module highlighting and viewing.
-It is in general impossible to present mutually overlapping biclusters as single rectangles.
-They are made up of a collection of rectangles.
-The ordering algorithm used in the R package realigns the gene expression matrix in a way that maximizes the total area of the largest rectangle in every bicluster.
-The outlines of these parts are drawn in a slightly brighter color than the background, making them easily recognizable.
-	\item[g, h] Modules are highlighted as the user moves the mouse over the gene expression data.
-The two check boxes allow you to choose between highlighting all the parts of a module (Filling) or alternatively only the largest rectangle (Outline).
-You can also turn it off completely.
-For data sets with many modules, it can be helpful to restrict highlighting to Outline.
-	\item[i, j] Similar to the highlighting, these two check boxes allow you to show either all the parts of the modules (Filling) or only the largest rectangles (Outline).
-By shift-clicking one of the check-boxes you can switch between showing only the modules or only the gene expression data.
-	\item[k] Sets the visibility of the modules layer.
-Moving the slider to the left fades out the gene expression data, thus focusing on the Biclusters, while towards the opposite direction, the gene expression data moves to the foreground.
+\item[g, h, i, j] Module highlighting and viewing. It is in general
+  impossible to present mutually overlapping biclusters as single
+  rectangles. They are made up of a collection of rectangles. The
+  ordering algorithm used in the R package realigns the gene
+  expression matrix in a way that maximizes the total area of the
+  largest rectangle in every bicluster. The outlines of these parts
+  are drawn in a slightly brighter color than the background, making
+  them easily recognizable.
+\item[g, h] Modules are highlighted as the user moves the mouse over
+  the gene expression data. The two check boxes allow you to choose
+  between highlighting all the parts of a module (Filling) or
+  alternatively only the largest rectangle (Outline). You can also
+  turn it off completely. For data sets with many modules, it can be
+  helpful to restrict highlighting to Outline.
+\item[i, j] Similar to the highlighting, these two check boxes allow
+  you to show either all the parts of the modules (Filling) or only
+  the largest rectangles (Outline). By shift-clicking one of the
+  check-boxes you can switch between showing only the modules or only
+  the gene expression data.
+\item[k] Sets the visibility of the modules layer. Moving the slider
+  to the left fades out the gene expression data, thus focusing on the
+  Biclusters, while towards the opposite direction, the gene
+  expression data moves to the foreground.
 \item[l] Realigns the windows at their initial positions.
-	\item[m] Puts the program in fullscreen mode.
-Note that due to security reasons, it is impossible to enter text in this mode.
-On Mac OS X, a bug in Flash player prevents you from exporting data in fullscreen mode.
-	\item[n] Opens the ExpressionView website, from where you can download sample files and tutorials.
+\item[m] Puts the program in fullscreen mode. Note that due to
+  security reasons, it is impossible to enter text in this mode. On
+  Mac OS X, a bug in Flash player prevents you from exporting data in
+  fullscreen mode.
+\item[n] Opens the ExpressionView website, from where you can download
+  sample files and tutorials.
 \item[o] Description and dimensions of the data set.
-	\item[p] Modules navigator.
-The Global tab is always available and shows the complete gene expression data.
-Additional tabs appear as you open individual modules.
-To close a module, simply move the mouse over the tab and click the close button that appears.
-	\item[q, r, s] Selected genes, samples and modules. The highlighting reflects the selection in the tables (w). The selection is maintained when switching tabs (p). 
+\item[p] Modules navigator. The Global tab is always available and
+  shows the complete gene expression data. Additional tabs appear as
+  you open individual modules. To close a module, simply move the
+  mouse over the tab and click the close button that appears. 
+\item[q, r, s] Selected genes, samples and modules. The highlighting
+  reflects the selection in the tables (w). The selection is
+  maintained when switching tabs (p).
 \item[q] Selected genes (=probes). 
@@
 \item[s] Selected modules (=biclusters).
-	\item[t] Info panel showing the data associated with the current mouse position.
-The GO and KEGG list contain the five most significant categories and pathways associated with the modules under the mouse pointer.
-	\item[u] Lists the selected genes, samples and modules, together with the intersecting modules.
+\item[t] Info panel showing the data associated with the current mouse
+  position. The GO and KEGG list contain the five most significant
+  categories and pathways associated with the modules under the mouse
+  pointer.
+\item[u] Lists the selected genes, samples and modules, together with
+  the intersecting modules.
 \item[u1] Opens intersecting modules.
@@
 \item[v] Lists the selected GO categories and KEGG pathways
-	\item[w] List navigator.
-Note that depending on the view (p), the lists only show genes and samples contained in the currently viewed module.
-Modules can also be opened by double-clicking on the corresponding row.
-The Experiment tab contains a brief description of the data.
-	\item[x] Searches the tables for a given expression and restricts the view to the matching entries.
-The search function uses Perl-style {\bf regular expressions}.
-By default, the search functions is applied to the whole table.
-To restrict it to a particular column, shift-click the corresponding column header.
-	\item[z] Select a column header to sort the entries according to that column. Shift-click to restrict the search function to that column.
+\item[w] List navigator. Note that depending on the view (p), the
+  lists only show genes and samples contained in the currently viewed
+  module. Modules can also be opened by double-clicking on the
+  corresponding row. The Experiment tab contains a brief description
+  of the data.
+\item[x] Searches the tables for a given expression and restricts the
+  view to the matching entries. The search function uses Perl-style
+  {\bf regular expressions}. By default, the search functions is
+  applied to the whole table. To restrict it to a particular column,
+  shift-click the corresponding column header.
+\item[z] Select a column header to sort the entries according to that
+  column. Shift-click to restrict the search function to that column.
 \end{description}
@@
 \section{Using ExpressionView with non-gene expression data}
-While ExpressionView is designed to work with gene expression data available in the form of a \Rpackage{Bioconductor} \Rclass{ExpressionSet}, it can also be used to visualize other data.
-Let us for instance use in-silico data generated by the \Rpackage{isa} package with dimensions 50 $\times$ 500 containing 10 overlapping modules:
+While ExpressionView is designed to work with gene expression data
+available in the form of a \Rpackage{Bioconductor}
+\Rclass{ExpressionSet}, it can also be used to visualize other
+data. Let us for instance use in-silico data generated by the
+\Rpackage{isa} package with dimensions 50 $\times$ 500 containing 10
+overlapping modules:
 <<in-silico>>=
 library(ExpressionView)
-# generate in-silico data with dimensions m x n containing M overlapping modules
+# generate in-silico data with dimensions m x n 
+# containing M overlapping modules
 # and add some noise
@@
 M <- 10
-data <- isa.in.silico(num.rows=m, num.cols=n, num.fact=M, noise=0.1, overlap.row=5)[[1]]
+data <- isa.in.silico(num.rows=m, num.cols=n, 
+                      num.fact=M, noise=0.1, overlap.row=5)[[1]]
 modules <- isa(data)
@@
 
-The \Rfunction{ExportEV} uses the named list provided by the \variable{description} variable to label the data. First, let us annotate the rows and columns of the data set
+The \Rfunction{ExportEV} uses the named list provided by the
+\variable{description} variable to label the data. First, let us
+annotate the rows and columns of the data set
 <<data set annotation>>=
@@
 
-Next, we assign the meta data associated with the rows of the data matrix. In this example we use 5 tags labelled ''row tag``:
+Next, we assign the meta data associated with the rows of the data
+matrix. In this example we use 5 tags labelled ``row tag'':
 <<rowdata annotation>>=
@@
 
-To finish  the description, we add some general information and merge it with the above tables to get a single named list:
+To finish  the description, we add some general information and merge
+it with the above tables to get a single named list:
 <<in-silico data annotation>>=
@@
 @
-When dealing with gene expression data, the \variable{xaxislabel} is equal to ''genes`` and the  \variable{yaxislabel} is ''samples``. Finally, we export the data set to an ExpressionView file:
+When dealing with gene expression data, the \variable{xaxislabel} is
+equal to ``genes'' and the  \variable{yaxislabel} is
+``samples''. Finally, we export the data set to an ExpressionView
+file: 
 <<in-silico data export, eval=FALSE>>=
-ExportEV(modules, data, filename="file.evf", description=description)
+ExportEV(modules, data, filename="file.evf", 
+         description=description)
 @
 
-Simply load this file with the Flash applet and check where the various labels appear.
+Simply load this file with the Flash applet and check where the
+various labels appear.
 

### 2010-05-04 RExpressionView/inst/doc/ExpressionView.format.Rnw
EV: added file format description.

% \VignetteIndexEntry{ExpressionView file format}
\documentclass{article}
\usepackage{ragged2e}
\usepackage{url}
\usepackage{fancyvrb}
\usepackage{color}

\newcommand{\Rfunction}[1]{\texttt{#1}}
\newcommand{\Rpackage}[1]{\texttt{#1}}
\newcommand{\Rclass}[1]{\texttt{#1}}
\newcommand{\Rargument}[1]{\textsl{#1}}
\newcommand{\filename}[1]{\texttt{#1}}
\newcommand{\variable}[1]{\texttt{#1}}

\DefineVerbatimEnvironment%
    {XML}{Verbatim}{fontsize=\small,frame=single,rulecolor=\color{green}}

\DefineVerbatimEnvironment%
    {schema}{Verbatim}{fontsize=\small,frame=single,rulecolor=\color{blue}}

\begin{document}

\title{ExpressionView file format}
\author{G\'abor Cs\'ardi}
\maketitle

\tableofcontents

\RaggedRight

\setlength{\parindent}{2em}

\section{Introduction}

ExpressionView is an interactive visualization tool for biclusters in
gene expression data. The software has two parts. The first part is an
ordering algorithm, written in GNU R, thay shuffles the rows and
columns of the gene expression matrix to make (potentially
overlapping) biclusters more visible. The second part is the
interactive tool, written in Adobe Flex. This runs in an Adobe Flash
enabled web browser. The user can export the ordered gene expression
matrix, with additional meta-data to a data file, that can be openened
by the Adobe Flash application. In this document we briefly discuss
the format of this data file.

\section{The file format}

The EVF data file, used by ExpressionView, is a standard XML file. The
R package contains an XML Schema file that describes the exact
format. In the following we will show this schema file and explain its
parts step by step, while also showing samples from an example EVF
data file. Parts of the schema file appear in blue boxes, EVF file
code snipplets are in green boxes.

\subsection{Header and main parts}

The schema file starts with a standard header:
\begin{schema}
<xsd:schema xmlns:xsd="http://www.w3.org/2001/XMLSchema">

  <xsd:annotation>
    <xsd:documentation xml:lang="en">
     ExpressionView file format schema, version 1.1.
     ExpressionView is a tool to visualize modules (biclusters) in
     gene expression data. Please see http://www.unil.ch/cbg for
     details. Copyright 2010 UNIL DGM Computational Biology Group
    </xsd:documentation>
  </xsd:annotation>
\end{schema}
This is the 1.1 version of the EVF file. ExpressionView can also read
the older 1.0 versions.

\begin{schema}
<xsd:element name="evf" type="evfType" />
\end{schema}
An EVF file contains a single \texttt{<evf>} tag, which has a following
parts:
\begin{schema}
<xsd:complexType name="evfType">
  <xsd:sequence>
    <xsd:element name="summary" type="SummaryType" />
    <xsd:element name="experimentdata" type="ExperimentDataType" />
    <xsd:element name="genes" type="GenesType" />
    <xsd:element name="samples" type="SamplesType" />
    <xsd:element name="modules" type="ModulesType" />
    <xsd:element name="data" type="xsd:base64Binary" />
  </xsd:sequence>
</xsd:complexType>
\end{schema}

\subsection{Summary}

The \texttt{<summary>} tag contains information about the data set, such
as the number of genes, samples, modules, etc. \texttt{<experimentdata>}
is for experiment meta-data, i.e. the lab where it was performed, 
the abstract of the related publication and possibly
more. \texttt{<genes>} and \texttt{<samples>} have the gene and sample
meta data. \texttt{<modules>} defines the biclusters. Finally
\texttt{<data>} contains the expression values themselves. Let us see
all of these tags in detail now.

\begin{schema}  
<xsd:complexType name="SummaryType">
  <xsd:sequence>
    <xsd:element name="description" minOccurs="0" type="xsd:string" />
    <xsd:element name="version" type="xsd:string" />
    <xsd:element name="dataorigin" minOccurs="0" type="xsd:string" />
    <xsd:element name="xaxislabel" minOccurs="0" type="xsd:string" />
    <xsd:element name="yaxislabel" minOccurs="0" type="xsd:string" />
    <xsd:element name="nmodules" type="xsd:nonNegativeInteger" />
    <xsd:element name="ngenes" type="xsd:positiveInteger" />
    <xsd:element name="nsamples" type="xsd:positiveInteger" />
  </xsd:sequence>
</xsd:complexType>
\end{schema}
The \texttt{<summary>} tag optionally contains the description of the
data file (\texttt{<description>} tag), this is displayed above the main
gene expression window. The \texttt{<version>} tag is required must be
\texttt{1.1} for files of the format we are discussing here. The
\texttt{<dataorigin>} tag is optional and has the value \texttt{eisa} for
modules generated with the ISA algorithm and exported using the
ExpressionView R package. \texttt{<xaxislabel>} and \texttt{<yaxislabel>}
are optional axis labels for the gene expression plot. The last three
tags (\texttt{<nmodules>}, \texttt{<ngenes>} and \texttt{<nsamples>}) are
required and give the number of modules, number of genes and number of
samples in the data set.

A sample EVF file header, together with the \texttt{<summary>} tag looks
like this:
\begin{XML}
<?xml version="1.0" encoding="UTF-8"?>
<evf>
  <summary>
    <description>ExpressionView data file</description>
    <version>1.1</version>
    <dataorigin>eisa</dataorigin>
    <nmodules>8</nmodules>
    <ngenes>3522</ngenes>
    <nsamples>128</nsamples>
  </summary>
\end{XML}

\subsection{Experiment meta-data}

The \texttt{<experimentdata>} tag is next, this contains the
experiment meta-data.
\begin{schema}
<xsd:complexType name="ExperimentDataType">
  <xsd:all>
    <xsd:element name="title" minOccurs="0" type="xsd:string" />
    <xsd:element name="name" minOccurs="0" type="xsd:string" />
    <xsd:element name="lab" minOccurs="0" type="xsd:string" />
    <xsd:element name="abstract" minOccurs="0" type="xsd:string" />
    <xsd:element name="url" minOccurs="0" type="xsd:anyURI" />
    <xsd:element name="annotation" minOccurs="0" type="xsd:string" />
    <xsd:element name="organism" minOccurs="0" type="xsd:string" />
  </xsd:all>
</xsd:complexType>
\end{schema}
All fields are optional here: \texttt{<title>}, \texttt{<name>},
\texttt{<lab>}, \texttt{<abstract>}, \texttt{<url>},
\texttt{<annotation>}, \texttt{<organism>}. They are collected and
shown in the \emph{Experiment} tab in the interactive ExpressionView
viewer. The \texttt{<annotation>} tag shoudl contain the name of the
chip on which the experiment was performed. If the \texttt{<organism>}
tag contains \emph{Homo sapiens}, then ExpressionView links genes with
the Gene Cards homepage, for other organism is uses Entrez Gene. The
tags within \texttt{experimentdata} can be specified in arbitrary
order.

Here is an example for the \texttt{<experimentdata>} tag:
\begin{XML}
<experimentdata>
  <title>Gene expression profile of adult T-cell acute
    lymphocytic leukemia identifies distinct subsets of
    patients with different response to therapy and
    survival.
  </title> 
  <name>Chiaretti et al.</name>
  <lab>Department of Medical Oncology, ...</lab>
  <abstract>Gene expression profiles were examined ...</abstact>
  <url>http://...</url>
  <annotation>hgu95av2</annotation>
  <organism>Homo sapiens</organism>
</experimentdata>
\end{XML}

\subsection{Genes}

The next tag within \texttt{<evf>} is \texttt{<genes>}:
\begin{schema}  
<xsd:complexType name="GenesType">
  <xsd:sequence>
    <xsd:element name="genetags" type="GeneTagsType" />
    <xsd:element name="gene" maxOccurs="unbounded" type="GeneType" />
  </xsd:sequence>
</xsd:complexType>
\end{schema}
It has two main parts, a \texttt{<genetags>} tag first, and then a
number of \texttt{<gene>} tags, one for each gene in the data set.

Before we continue with the gene tags, we define a new tag type, that
is required to include a \texttt{name} attribute. We call this type
\texttt{ExtraType}. This tag type can be used to add any kind of meta
data to the table in the \emph{Genes} tab in ExpressionView.
\begin{schema}  
<xsd:complexType name="ExtraType">
  <xsd:simpleContent>
    <xsd:extension base="xsd:string">
      <xsd:attribute name="name" type="xsd:string" use="required" />
    </xsd:extension>
  </xsd:simpleContent>
</xsd:complexType>
\end{schema}

The \texttt{<genetags>} tag has one required and many optional
subtags:
\begin{schema}  
<xsd:complexType name="GeneTagsType">
  <xsd:sequence>
    <xsd:element name="id" type="xsd:string" />
    <xsd:element name="name" minOccurs="0" type="xsd:string" />
    <xsd:element name="symbol" minOccurs="0" type="xsd:string" />
    <xsd:element name="entrezid" minOccurs="0" type="xsd:string" />
    <xsd:element name="x" minOccurs="0" maxOccurs="unbounded"
                 type="ExtraType" />
  </xsd:sequence>
</xsd:complexType>
\end{schema}
The \texttt{<id>} must be an integer number, that starts with one and
goes up to the number of genes. \texttt{<name>} is typically the name
of the probe-set, but be used for other purposes as well,
\texttt{<symbol>} is typically the canonical gene
symbol. \texttt{<entrezid>} is the Entrez gene id. If a given
probeset does not map to a known gene, then \texttt{<symbol>} and
\texttt{<entrezid>} can be set to \texttt{NA}. Finally, the gene tags
might contain any number number of \texttt{<x>} tags, each defining an
additional column in the \emph{Genes} tab in ExpressionView. These
tags must have a \texttt{name} attribute, which is refered by the
individual genes. Here is an example gene tags section:
\begin{XML}
<genetags>
  <id>#</id>
  <name>Name</name>
  <symbol>Symbol</symbol>
  <entrezid>EntrezID</entrezid>
  <x name="chr">Chromosome</x>
</genetags>
\end{XML}
Here we define an extra column for the gene table, this will give the
chromosome on which the gene is located.

After the gene tags, we have a typically large number of genes in the file:
\begin{schema}
<xsd:complexType name="GeneType">
  <xsd:sequence>
    <xsd:element name="id" type="xsd:string" />
    <xsd:element name="name" minOccurs="0" type="xsd:string" />
    <xsd:element name="symbol" minOccurs="0" type="xsd:string" />
    <xsd:element name="entrezid" minOccurs="0" type="xsd:string" />
    <xsd:element name="x" minOccurs="0" maxOccurs="unbounded"
                 type="ExtraType"/>
  </xsd:sequence>
</xsd:complexType>
\end{schema}
These must have exactly the same subtags as the ones given in the gene
tags section. For example:
\begin{XML}
<gene>
  <id>1</id>
  <name>33500_i_at</name>
  <symbol>NA</symbol>
  <entrezid>NA</entrezid>
  <x name="chr">NA</x>
</gene>
<gene>
  <id>2</id>
  <name>40990_at</name>
  <symbol>TSPAN5</symbol>
  <entrezid>10098</entrezid>
  <x name="chr">4</x>
</gene>
...
\end{XML}
Observe, how the extra tag for the chromosomes is referenced here.

\subsection{Samples}

The samples and their associated data have a similar format to the
genes:
\begin{schema}  
<xsd:complexType name="SamplesType">
  <xsd:sequence>
    <xsd:element name="sampletags" type="SampleTagsType" />
    <xsd:element name="sample" maxOccurs="unbounded" 
                 type="SampleType" />
  </xsd:sequence>
</xsd:complexType>
\end{schema}
The \texttt{<samples>} tag has two parts, the first defines the sample
meta data entries and the second part contains one \texttt{<sample>}
tag for each sample in the data set.

\begin{schema}  
<xsd:complexType name="SampleTagsType">
  <xsd:sequence>
    <xsd:element name="id" type="xsd:string" />
    <xsd:element name="name" minOccurs="0" type="xsd:string" />
    <xsd:element name="x" minOccurs="0" maxOccurs="unbounded"
                 type="ExtraType" />
  </xsd:sequence>
</xsd:complexType>
\end{schema}
The \texttt{<id>} and \texttt{<name>} sample tags are
required. \texttt{<id>} contains numeric ids, from one, up to the
number of samples in the data set. \texttt{<name>} is an arbitrary
string, it is typically the sample name in the R ExpressionSet object,
that was used to find the biclusters. Similarly to the genes, any
number of additional \texttt{<x>} tags can be added, each with a
unique \texttt{name} attribute, to define additional information about
the samples.

Here is an example \texttt{<sampletags>} tag:
\begin{XML}
<sampletags>
  <id>#</id>
  <name>Name</name>
  <x name="diagnosis"> Date of diagnosis</x>
  <x name="sex"> Gender of the patient</x>
  <x name="age"> Age of the patient at entry</x>
  <x name="BT"> does the patient have B-cell or T-cell ALL</x>
</sampletags>
\end{XML}

Then the samples follow:
\begin{schema}  
<xsd:complexType name="SampleType">
  <xsd:sequence>
    <xsd:element name="id" type="xsd:string" />
    <xsd:element name="name" minOccurs="0" type="xsd:string" />
    <xsd:element name="x" minOccurs="0" maxOccurs="unbounded"
                 type="ExtraType" />
  </xsd:sequence>
</xsd:complexType>
\end{schema}
Here is an example, that corresponds to the \texttt{<sampletags>}
entry given above:
\begin{XML}
<sample>
  <id>1</id>
  <name>01010</name>
  <x name="diagnosis">3/29/2000</x>
  <x name="sex">2</x>
  <x name="age">19</x>
  <x name="BT">3</x>
</sample>
<sample>
  <id>2</id>
  <name>04010</name>
  <x name="diagnosis">10/30/1997</x>
  <x name="sex">1</x>
  <x name="age">18</x>
  <x name="BT">2</x>
</sample>
...
\end{XML}

\subsection{Modules}

Again, the \texttt{<modules>} tag has a syntax that is similar to the
syntax of the genes and the samples:
\begin{schema}
<xsd:complexType name="ModulesType">
  <xsd:sequence>
    <xsd:element name="moduletags" type="ModuleTagsType" />
    <xsd:element name="gotags" minOccurs="0" type="GoTagsType" />
    <xsd:element name="keggtags" minOccurs="0" type="KeggTagsType" />
    <xsd:element name="module" maxOccurs="unbounded" 
                 type="ModuleType" />
  </xsd:sequence>      
</xsd:complexType>
\end{schema}
There are two required and two optional parts. The first part is
\texttt{<moduletags>}, this defines the meta-data associated with the
modules. The \texttt{<gotags>} and \texttt{<keggtags>} tags are
optional, they are only present if the file contains Gene Ontology
and/or KEGG pathway enrichment calculation results for the
biclusters. Finally, there is a list of \texttt{<module>} tags, one
for each bicluster.

\begin{schema}  
<xsd:complexType name="ModuleTagsType">
  <xsd:sequence>
    <xsd:element name="id" type="xsd:string" />
    <xsd:element name="name" minOccurs="0" type="xsd:string" />
    <xsd:element name="iterations" minOccurs="0" type="xsd:string" />
    <xsd:element name="oscillation" minOccurs="0" type="xsd:string" />
    <xsd:element name="thr_row" minOccurs="0" type="xsd:string" />
    <xsd:element name="thr_col" minOccurs="0" type="xsd:string" />
    <xsd:element name="freq" minOccurs="0" type="xsd:string" />
    <xsd:element name="rob" minOccurs="0" type="xsd:string" />
    <xsd:element name="rob_limit" minOccurs="0" type="xsd:string" />
    <xsd:element name="x" minOccurs="0" maxOccurs="unbounded"
		 type="ExtraType" />
  </xsd:sequence>
</xsd:complexType>
\end{schema}
There is only one tag required in \texttt{<moduletags>}, the
\texttt{<id>}, which is a numeric id, going from 1 to the number of
modules. The rest is optional, and it is also possible to create any
kind of extra tags with the \texttt{<x>} tag. \texttt{<name>} is an
arbitrary character string. The other optional tags are
mainly defined to modules coming from the ISA algorithm:
\texttt{<iterations>} gives the number of ISA iterations needed to
find the module, \texttt{<oscillation>} specifies the oscillation
cycle length for oscillating modules. It is mostly zero, meaning that
the module does not oscillate. \texttt{<thr\_row>} is the ISA gene
threshold, \texttt{<thr\_col>} is the ISA condition
threshold. \texttt{<freq>} is the number of ISA seeds that converged
to the module, \texttt{<rob>} is its robustness
score. \texttt{<rob\_limit>} is the robustness threshold that was used
to filter the module.

Here is an example \texttt{<moduletags>} tag:
\begin{XML}
<moduletags>
  <id>#</id>
  <name>Name</name>
  <iterations>iterations</iterations>
  <oscillation>oscillation</oscillation>
  <thr_row>thr.row</thr_row>
  <thr_col>thr.col</thr_col>
  <freq>freq</freq>
  <rob>rob</rob>
  <rob_limit>rob.limit</rob_limit>
</moduletags>
\end{XML}

The optional \texttt{<gotags>} and \texttt{<keggtags>} tags are
discussed in the next subsection, let us now see how the data for a
single module looks like:
\begin{schema}
<xsd:complexType name="ModuleType">
  <xsd:sequence>
    <xsd:element name="id" type="xsd:string" />
    <xsd:element name="name" minOccurs="0" type="xsd:string" />
    <xsd:element name="iterations" minOccurs="0" type="xsd:string" />
    <xsd:element name="oscillation" minOccurs="0" type="xsd:string" />
    <xsd:element name="thr_row" minOccurs="0" type="xsd:string" />
    <xsd:element name="thr_col" minOccurs="0" type="xsd:string" />
    <xsd:element name="freq" minOccurs="0" type="xsd:string" />
    <xsd:element name="rob" minOccurs="0" type="xsd:string" />
    <xsd:element name="rob_limit" minOccurs="0" type="xsd:string" />
    <xsd:element name="x" minOccurs="0" maxOccurs="unbounded"
		 type="ExtraType" />
    <xsd:element name="containedgenes" type="integerList" />
    <xsd:element name="genescores" type="doubleList" />
    <xsd:element name="containedsamples" type="integerList" />
    <xsd:element name="samplescores" type="doubleList" />
    <xsd:element name="intersectingmodules" type="integerList" />
    <xsd:element name="gos" minOccurs="0" type="GosType" />
    <xsd:element name="keggs" minOccurs="0" type="KeggsType" />
  </xsd:sequence>
</xsd:complexType>  
\end{schema}
The first couple of fields refer to the ones defined in the
\texttt{<moduletags>} section above. The others are:
\texttt{<containedgenes>}, a space separated list of gene ids, for the
genes that are included in the module; \texttt{<genescores>}, the gene
scores of these genes, in the order of the list in
\texttt{<containedgenes>}; \texttt{<containedsamples>}, the ids of the
samples in the module; \texttt{<samplescores>}, the scores for these
samples; \texttt{<intersectingmodules>}, the ids of the modules that
have an overlap with the given module. \texttt{<gos>} and
\texttt{<keggs>} are optional and describe the Gene Ontology and KEGG
pathway enrichment of the given module, see their details in the next
subsection.

Here is an example \texttt{<module>} tag:
\begin{XML}
<module>
  <id>1</id>
  <name>module 1</name>
  <iterations>22</iterations>
  <oscillation>0</oscillation>
  <thr_row>2.7</thr_row>
  <thr_col>1.4</thr_col>
  <freq>1</freq>
  <rob>21.98</rob>
  <rob_limit>21.98</rob_limit>
  <containedgenes>214 215 216 217 218 219 220 221 222
  </containedgenes>
  <genescores>-0.94 -0.88 0.74 -0.76 -1.00 -0.84 -0.74 -0.76 -0.85
  </genescores>
  <containedsamples>63 64 65 54 66</containedsamples>
  <samplescores>-0.62 -1.00 -0.77 -0.40 -0.28</samplescores>
  <intersectingmodules>7</intersectingmodules>
  <gos>
    ... see below ...
  </gos>
  <keggs>
    ... see below ...
  </keggs>
</module>
\end{XML}

We refered to the integer list and double list data types above, now
we define them:
\begin{schema}  
<xsd:simpleType name="integerList">
  <xsd:list itemType="xsd:positiveInteger"/>
</xsd:simpleType>
\end{schema}

\begin{schema}
<xsd:simpleType name="doubleList">
  <xsd:list itemType="xsd:double"/>
</xsd:simpleType>
\end{schema}

\subsection{Gene Ontology and KEGG pathway enrichment}

EVF files can optionally contain Gene Ontology (GO) and/or KEGG
pathway enrichment calculation results. In this case
\texttt{<moduletags>} is followed by a \texttt{<gotags>} and/or a
\texttt{keggtags} tag:
\begin{schema}
<xsd:complexType name="GoTagsType">
  <xsd:sequence>
    <xsd:element name="id" type="xsd:string" />
    <xsd:element name="go" type="xsd:string" />
    <xsd:element name="term" type="xsd:string" />
    <xsd:element name="ontology" type="xsd:string" />
    <xsd:element name="pvalue" type="xsd:string" />
    <xsd:element name="oddsratio" type="xsd:string" />
    <xsd:element name="expcount" type="xsd:string" />
    <xsd:element name="count" type="xsd:string" />
    <xsd:element name="size" type="xsd:string" />
  </xsd:sequence>
</xsd:complexType>
\end{schema}
Here \texttt{<id>} is a numeric id, starting from one, up to the total
number of enriched GO categories; \texttt{<go>} is the GO id;
\texttt{term} is the GO term; \texttt{ontology} is the GO ontology of
the term, possible values: \texttt{BP}, \texttt{CC}, \texttt{MF};
\texttt{pvalue} is the enrichment $p$-value; \texttt{oddsratio} is the
odds ratio; \texttt{expcount} is the expected number of genes in the
intersection, by chance; \texttt{count} is the number of genes in the
intersection; \texttt{size} is the size of the GO term, i.e. the
number of genes (in the current gene universe) that are annotated with
the term.

\begin{schema}
<xsd:complexType name="KeggTagsType">
  <xsd:sequence>
    <xsd:element name="id" type="xsd:string" />
    <xsd:element name="kegg" type="xsd:string" />
    <xsd:element name="pathname" type="xsd:string" />
    <xsd:element name="pvalue" type="xsd:string" />
    <xsd:element name="oddsratio" type="xsd:string" />
    <xsd:element name="expcount" type="xsd:string" />
    <xsd:element name="count" type="xsd:string" />
    <xsd:element name="size" type="xsd:string" />    
  </xsd:sequence>
</xsd:complexType>  
\end{schema}
The tags for \texttt{<keggtags>} is almost the same as for
\texttt{<gotags>}, but here \texttt{<kegg>} id the KEGG pathway ID,
and \texttt{<pathname>} is the name of the pathway. The rest is the
same.

Here is part of an EVF file with the \texttt{<gotags>} and
\texttt{<keggtags>} tags:
\begin{XML}
<gotags>
  <id>#</id>
  <go>GO</go>
  <term>Term</term>
  <ontology>Ontology</ontology>
  <pvalue>PValue</pvalue>
  <oddsratio>OddsRatio</oddsratio>
  <expcount>ExpCount</expcount>
  <count>Count</count>
  <size>Size</size>
</gotags>
<keggtags>
  <id>#</id>
  <kegg>KEGG</kegg>
  <pathname>Path Name</pathname>
  <pvalue>PValue</pvalue>
  <oddsratio>OddsRatio</oddsratio>
  <expcount>ExpCount</expcount>
  <count>Count</count>
  <size>Size</size>
</keggtags>
\end{XML}

The actual enrichment data is given in the \texttt{<module>} tags,
these can contain a \texttt{<gos>} and/or a \texttt{<keggs>} tag with
the enrichment $p$-values and other statistics.

\begin{schema}
<xsd:complexType name="GosType">
  <xsd:sequence>
    <xsd:element name="go" minOccurs="0" maxOccurs="unbounded" />
  </xsd:sequence>
</xsd:complexType>
\end{schema}
\texttt{<gos>} is a list of \texttt{<go>} tags.

\begin{schema}
<xsd:complexType name="KeggsType">
  <xsd:sequence>
    <xsd:element name="kegg" minOccurs="0" maxOccurs="unbounded" />
  </xsd:sequence>
</xsd:complexType>
\end{schema}
\texttt{<keggs>} is a list of \texttt{<kegg>} tags.

The subtags within a \texttt{<go>} or a \texttt{<kegg>} tag refer to
the tags already listed in the \texttt{<gotags>} and
\texttt{<keggtags>} tags:
\begin{schema}
<xsd:complexType name="go">
  <xsd:sequence>
    <xsd:element name="id" type="xsd:positiveInteger" />
    <xsd:element name="go" type="xsd:string" />
    <xsd:element name="ontology" type="xsd:string" />
    <xsd:element name="pvalue" type="xsd:double" />
    <xsd:element name="oddsratio" type="xsd:double" />
    <xsd:element name="expcount" type="xsd:double" />
    <xsd:element name="count" type="xsd:nonNegativeInteger" />
    <xsd:element name="size" type="xsd:nonNegativeInteger" />
  </xsd:sequence>
</xsd:complexType>
\end{schema}

\begin{schema}
<xsd:complexType name="kegg">
  <xsd:sequence>
    <xsd:element name="id" type="xsd:positiveInteger" />
    <xsd:element name="kegg" type="xsd:string" />
    <xsd:element name="pathname" type="xsd:string" />
    <xsd:element name="pvalue" type="xsd:double" />
    <xsd:element name="oddsratio" type="xsd:double" />
    <xsd:element name="expcount" type="xsd:double" />
    <xsd:element name="count" type="xsd:nonNegativeInteger" />
    <xsd:element name="size" type="xsd:nonNegativeInteger" />
  </xsd:sequence>
</xsd:complexType>
\end{schema}

Here is an example, this is part of a \texttt{<module>} tag:
\begin{XML}
<gos>
  <go>
    <id>1</id>
    <go>GO:0002376</go>
    <term>immune system process</term>
    <ontology>BP</ontology>
    <pvalue>1.64e-04</pvalue>
    <oddsratio>7.03</oddsratio>
    <expcount>3.95</expcount>
    <count>16</count>
    <size>353</size>
  </go>
  <go>
    <id>2</id>
    <go>GO:0002504</go>
    <term>antigen processing and presentation of peptide or 
      polysaccharide antigen via MHC class II</term>
    <ontology>BP</ontology>
    <pvalue>1.06e-03</pvalue>
    <oddsratio>79.95</oddsratio>
    <expcount>0.10</expcount>
    <count>4</count>
    <size>9</size>
  </go>
</gos>
<keggs>
  <kegg>
    <id>1</id>
    <kegg>05310</kegg>
    <pathname>Asthma</pathname>
    <pvalue>0.02</pvalue>
    <oddsratio>31.60</oddsratio>
    <expcount>0.15</expcount>
    <count>3</count>
    <size>10</size>
  </kegg>
  <kegg>
    <id>2</id>
    <kegg>05320</kegg>
    <pathname>Autoimmune thyroid disease</pathname>
    <pvalue>0.03</pvalue>
    <oddsratio>24.54</oddsratio>
    <expcount>0.18</expcount>
    <count>3</count>
    <size>12</size>
  </kegg>
</keggs>
\end{XML}

\subsection{The expression data}

Finally, the expression data in included in the \texttt{<data>}
tag. This is a Base64 encoded string, generated by representing each
expression value with a single unsigned byte, and then concatenating
and Base64 encoding these bytes. This data looks like this:
\begin{XML}
<data>
/Acs1hZT+FDeFGH/29zL+3rK4+sTCvzZzxVgvNrM5Bw8Kfjw2vb6Of8qti943/0QIF8xJy7g
zMoOqyYA4/ne1KsZ1bH69u3c28zewtQM9G8T5zGK7P0szwIM/ub2z9zwHxclIdz3ywC6xtfO
DewMklghAfGqzsQtDeHZ3wolAC1GSN+9N+/i5Oz99bTzzRb1QvzHE+Qlr1Ej1U697+AFvtyn
...
</data>
\end{XML}

\begin{schema}  
</xsd:schema>
\end{schema}

\section{Additional information}

Please see the ExpressionView homepage at
\url{http://www.unil.ch/cbg/ExpressionView} for more infomation, the
schema file and example data file.

\end{document}

### 2010-05-05 RExpressionView/inst/doc/ExpressionView.format.Rnw
EV: XML format vignette updates.

@@
 
-\setlength{\parindent}{2em}
+\setlength{\parindent}{12pt}
 
@@
 gene expression data. The software has two parts. The first part is an
-ordering algorithm, written in GNU R, thay shuffles the rows and
+ordering algorithm, written in GNU R, that reorders the rows and
 columns of the gene expression matrix to make (potentially
 overlapping) biclusters more visible. The second part is the
-interactive tool, written in Adobe Flex. This runs in an Adobe Flash
+interactive tool, written in Adobe Flex. It runs in an Adobe Flash
 enabled web browser. The user can export the ordered gene expression
-matrix, with additional meta-data to a data file, that can be openened
-by the Adobe Flash application. In this document we briefly discuss
-the format of this data file.
+matrix, with additional meta-data from R to a data file, that can be
+openened by the Adobe Flash application. In this document we briefly
+discuss the format of this data file.
 
@@
 data file. Parts of the schema file appear in blue boxes, EVF file
-code snipplets are in green boxes.
+code snipplets are in boxes with green frames.
 
@@
 This is the 1.1 version of the EVF file. ExpressionView can also read
-the older 1.0 versions.
+the older 1.0 version.
 
@@
 \end{schema}
-An EVF file contains a single \texttt{<evf>} tag, which has a following
+An EVF file contains a single \texttt{<evf>} tag. It has the following
 parts:
@@
 
-\subsection{Summary}
-
 The \texttt{<summary>} tag contains information about the data set, such
@@
 
+\subsection{Summary}
+
+This is the type of the \texttt{<summary>} tag:
 \begin{schema}  
@@
 data file (\texttt{<description>} tag), this is displayed above the main
-gene expression window. The \texttt{<version>} tag is required must be
-\texttt{1.1} for files of the format we are discussing here. The
-\texttt{<dataorigin>} tag is optional and has the value \texttt{eisa} for
-modules generated with the ISA algorithm and exported using the
-ExpressionView R package. \texttt{<xaxislabel>} and \texttt{<yaxislabel>}
-are optional axis labels for the gene expression plot. The last three
-tags (\texttt{<nmodules>}, \texttt{<ngenes>} and \texttt{<nsamples>}) are
-required and give the number of modules, number of genes and number of
-samples in the data set.
+gene expression window in ExpressionView. The \texttt{<version>} tag
+must be \texttt{1.1} for files for the format we are discussing here. The
+\texttt{<dataorigin>} tag is optional and has the value `\texttt{eisa}' for
+modules generated with the ISA algorithm~\cite{bergmann03} and
+exported using the ExpressionView R package. \texttt{<xaxislabel>} and
+\texttt{<yaxislabel>} are optional axis labels for the gene expression
+plot. The last three tags (\texttt{<nmodules>}, \texttt{<ngenes>} and
+\texttt{<nsamples>}) are required and give the number of modules,
+number of genes and number of samples in the data set.
 
@@
 The \texttt{<experimentdata>} tag is next, this contains the
-experiment meta-data.
+experiment meta-data. It has the following parts:
 \begin{schema}
@@
 shown in the \emph{Experiment} tab in the interactive ExpressionView
-viewer. The \texttt{<annotation>} tag shoudl contain the name of the
+viewer. The \texttt{<annotation>} tag should contain the name of the
 chip on which the experiment was performed. If the \texttt{<organism>}
-tag contains \emph{Homo sapiens}, then ExpressionView links genes with
+tag contains `\texttt{Homo sapiens}', then ExpressionView links genes with
 the Gene Cards homepage, for other organism is uses Entrez Gene. The
@@
 
-The next tag within \texttt{<evf>} is \texttt{<genes>}:
+The next tag within \texttt{<evf>} is \texttt{<genes>}, its type os
+defined as
 \begin{schema}  
@@
 Before we continue with the gene tags, we define a new tag type, that
-is required to include a \texttt{name} attribute. We call this type
+is required to contain a \texttt{name} attribute. We call this type
 \texttt{ExtraType}. This tag type can be used to add any kind of meta
@@
 The \texttt{<id>} must be an integer number, that starts with one and
-goes up to the number of genes. \texttt{<name>} is typically the name
-of the probe-set, but be used for other purposes as well,
+goes up to the number of genes. It is used as an id that is refered
+from other tags in the document. \texttt{<name>} is typically the name
+of the probe-set, but it can be used for other purposes as well,
 \texttt{<symbol>} is typically the canonical gene
@@
 \texttt{<entrezid>} can be set to \texttt{NA}. Finally, the gene tags
-might contain any number number of \texttt{<x>} tags, each defining an
+might contain any number of \texttt{<x>} tags, each defining an
 additional column in the \emph{Genes} tab in ExpressionView. These
 tags must have a \texttt{name} attribute, which is refered by the
-individual genes. Here is an example gene tags section:
+tags of the individual genes. Here is an example gene tags section:
 \begin{XML}
@@
 \end{XML}
-Here we define an extra column for the gene table, this will give the
+We define an extra column for the gene table, this will give the
 chromosome on which the gene is located.
 
-After the gene tags, we have a typically large number of genes in the file:
+After the gene tags, we have a typically large number of genes in the
+file. The definition of the \texttt{<gene>} tag:
 \begin{schema}
@@
 \end{schema}
-These must have exactly the same subtags as the ones given in the gene
-tags section. For example:
+The \texttt{<gene>} tags must have exactly the same subtags as the
+ones given in the gene tags section. For example, assuming the
+\texttt{<genetags>} example above we can have:
 \begin{XML}
@@
 
-The samples and their associated data have a similar format to the
+The samples and their associated data have a format similar to the
 genes:
@@
 required. \texttt{<id>} contains numeric ids, from one, up to the
-number of samples in the data set. \texttt{<name>} is an arbitrary
+number of samples in the data set. These are referenced by other tags,
+e.g. the ones that define the modules. \texttt{<name>} is an arbitrary
 string, it is typically the sample name in the R ExpressionSet object,
@@
 \end{XML}
+We defined four extra tags for the samples in this example.
 
@@
 
+The type of \texttt{<moduletags>}:
 \begin{schema}  
@@
   <name>Name</name>
-  <iterations>iterations</iterations>
-  <oscillation>oscillation</oscillation>
-  <thr_row>thr.row</thr_row>
-  <thr_col>thr.col</thr_col>
-  <freq>freq</freq>
-  <rob>rob</rob>
-  <rob_limit>rob.limit</rob_limit>
+  <iterations>Iterations</iterations>
+  <oscillation>Oscillation cycle</oscillation>
+  <thr_row>Gene threshold</thr_row>
+  <thr_col>Column threshold</thr_col>
+  <freq>Frequency</freq>
+  <rob>Robustness</rob>
+  <rob_limit>Robustness limit</rob_limit>
 </moduletags>
@@
 The optional \texttt{<gotags>} and \texttt{<keggtags>} tags are
-discussed in the next subsection, let us now see how the data for a
+discussed in Section~\ref{sec:gokegg}, let us now see how the data for a
 single module looks like:
@@
 \texttt{<keggs>} are optional and describe the Gene Ontology and KEGG
-pathway enrichment of the given module, see their details in the next
-subsection.
+pathway enrichment of the given module, see their details in Section~\ref{sec:gokegg}.
 
@@
 
-\subsection{Gene Ontology and KEGG pathway enrichment}
+\subsection{Gene Ontology and KEGG pathway enrichment}%
+\label{sec:gokegg}
 
@@
 \texttt{<moduletags>} is followed by a \texttt{<gotags>} and/or a
-\texttt{keggtags} tag:
+\texttt{<keggtags>} tag. The definition of \texttt{<gotags>}:
 \begin{schema}
@@
 \end{schema}
-Here \texttt{<id>} is a numeric id, starting from one, up to the total
+\texttt{<id>} is a numeric id, starting from one, up to the total
 number of enriched GO categories; \texttt{<go>} is the GO id;
-\texttt{term} is the GO term; \texttt{ontology} is the GO ontology of
-the term, possible values: \texttt{BP}, \texttt{CC}, \texttt{MF};
-\texttt{pvalue} is the enrichment $p$-value; \texttt{oddsratio} is the
-odds ratio; \texttt{expcount} is the expected number of genes in the
-intersection, by chance; \texttt{count} is the number of genes in the
-intersection; \texttt{size} is the size of the GO term, i.e. the
+\texttt{<term>} is the GO term; \texttt{<ontology>} is the GO ontology of
+the term, its possible values are: \texttt{BP}, \texttt{CC},
+\texttt{MF}, standing for \emph{Biological process}, \emph{Cellular
+  component} and \emph{Molecular function};
+\texttt{<pvalue>} is the enrichment $p$-value; \texttt{<oddsratio>} is the
+odds ratio; \texttt{<expcount>} is the expected number of genes in the
+intersection, by chance; \texttt{<count>} is the number of genes in the
+intersection; \texttt{<size>} is the size of the GO term, i.e. the
 number of genes (in the current gene universe) that are annotated with
-the term.
+the enriched term.
 
@@
 
-Here is part of an EVF file with the \texttt{<gotags>} and
+Part of an EVF file with the \texttt{<gotags>} and
 \texttt{<keggtags>} tags:
@@
 The subtags within a \texttt{<go>} or a \texttt{<kegg>} tag refer to
-the tags already listed in the \texttt{<gotags>} and
+the tags already listed above in the \texttt{<gotags>} and
 \texttt{<keggtags>} tags:
@@
 \url{http://www.unil.ch/cbg/ExpressionView} for more infomation, the
-schema file and example data file.
+schema file and example data files.
+
+\bibliographystyle{unsrt}
+\bibliography{ExpressionView}
 

### 2010-05-05 RExpressionView/inst/doc/ExpressionView.format.Rnw
EV file format vignette: use 'listings' instead of 'fancyvrb', better results.

@@
 \usepackage{url}
-\usepackage{fancyvrb}
+\usepackage{listings}
 \usepackage{color}
+\usepackage{courier}
 
@@
 
-\DefineVerbatimEnvironment%
-    {XML}{Verbatim}{fontsize=\small,frame=single,rulecolor=\color{green}}
-
-\DefineVerbatimEnvironment%
-    {schema}{Verbatim}{fontsize=\small,frame=single,rulecolor=\color{blue}}
+\definecolor{green}{rgb}{0.8,1,0.8}
+\definecolor{blue}{rgb}{0.8,0.8,1}
+\lstset{basicstyle=\footnotesize\ttfamily,%
+        keywordstyle=\footnotesize\ttfamily,%
+        language=XML,frame=single,columns=fullflexible}
+\lstnewenvironment{schema}{\lstset{backgroundcolor=\color{green}}}{}
+\lstnewenvironment{XML}{\lstset{backgroundcolor=\color{blue}}}{}
 
@@
 
-\setlength{\parindent}{12pt}
+\setlength{\parskip}{12pt}
 
@@
 parts step by step, while also showing samples from an example EVF
-data file. Parts of the schema file appear in blue boxes, EVF file
-code snipplets are in boxes with green frames.
+data file. Parts of the schema file appear in green boxes, EVF file
+code snipplets are in blue boxes.
 
@@
 <data>
-/Acs1hZT+FDeFGH/29zL+3rK4+sTCvzZzxVgvNrM5Bw8Kfjw2vb6Of8qti943/0QIF8xJy7g
-zMoOqyYA4/ne1KsZ1bH69u3c28zewtQM9G8T5zGK7P0szwIM/ub2z9zwHxclIdz3ywC6xtfO
-DewMklghAfGqzsQtDeHZ3wolAC1GSN+9N+/i5Oz99bTzzRb1QvzHE+Qlr1Ej1U697+AFvtyn
+/Acs1hZT+FDeFGH/29zL+3rK4+sTCvzZzxVgvNrM5Bw8Kfjw2vb6Of8qti943/0QIF8xJy7g2ufB
+zMoOqyYA4/ne1KsZ1bH69u3c28zewtQM9G8T5zGK7P0szwIM/ub2z9zwHxclIdz3ywC6xtfOySje
+DewMklghAfGqzsQtDeHZ3wolAC1GSN+9N+/i5Oz99bTzzRb1QvzHE+Qlr1Ej1U697+AFvtynDeHd
 ...

### 2010-05-05 RExpressionView/inst/doc/ExpressionView.format.Rnw
EV file format vignette: better spacing.
EV format vignette: use a macro to typeset the XML tags.

@@
         keywordstyle=\footnotesize\ttfamily,%
-        language=XML,frame=single,columns=fullflexible}
+        language=XML,frame=single,columns=fullflexible,%
+        aboveskip=8pt,belowskip=-5pt}
 \lstnewenvironment{schema}{\lstset{backgroundcolor=\color{green}}}{}
-\lstnewenvironment{XML}{\lstset{backgroundcolor=\color{blue}}}{}
+\lstnewenvironment{example}{\lstset{backgroundcolor=\color{blue}}}{}
+
+\newcommand{\xmltag}[1]{%
+  \csname a:xmltag\endcsname\texttt{<#1>}\csname b:xmltag\endcsname}
 
@@
 \end{schema}
-An EVF file contains a single \texttt{<evf>} tag. It has the following
+An EVF file contains a single \xmltag{evf} tag. It has the following
 parts:
@@
 
-The \texttt{<summary>} tag contains information about the data set, such
-as the number of genes, samples, modules, etc. \texttt{<experimentdata>}
+The \xmltag{summary} tag contains information about the data set, such
+as the number of genes, samples, modules, etc. \xmltag{experimentdata}
 is for experiment meta-data, i.e. the lab where it was performed, 
 the abstract of the related publication and possibly
-more. \texttt{<genes>} and \texttt{<samples>} have the gene and sample
-meta data. \texttt{<modules>} defines the biclusters. Finally
-\texttt{<data>} contains the expression values themselves. Let us see
+more. \xmltag{genes} and \xmltag{samples} have the gene and sample
+meta data. \xmltag{modules} defines the biclusters. Finally
+\xmltag{data} contains the expression values themselves. Let us see
 all of these tags in detail now.
@@
 
-This is the type of the \texttt{<summary>} tag:
+This is the type of the \xmltag{summary} tag:
 \begin{schema}  
@@
 \end{schema}
-The \texttt{<summary>} tag optionally contains the description of the
-data file (\texttt{<description>} tag), this is displayed above the main
-gene expression window in ExpressionView. The \texttt{<version>} tag
+The \xmltag{summary} tag optionally contains the description of the
+data file (\xmltag{description} tag), this is displayed above the main
+gene expression window in ExpressionView. The \xmltag{version} tag
 must be \texttt{1.1} for files for the format we are discussing here. The
-\texttt{<dataorigin>} tag is optional and has the value `\texttt{eisa}' for
+\xmltag{dataorigin} tag is optional and has the value `\texttt{eisa}' for
 modules generated with the ISA algorithm~\cite{bergmann03} and
-exported using the ExpressionView R package. \texttt{<xaxislabel>} and
-\texttt{<yaxislabel>} are optional axis labels for the gene expression
-plot. The last three tags (\texttt{<nmodules>}, \texttt{<ngenes>} and
-\texttt{<nsamples>}) are required and give the number of modules,
+exported using the ExpressionView R package. \xmltag{xaxislabel} and
+\xmltag{yaxislabel} are optional axis labels for the gene expression
+plot. The last three tags (\xmltag{nmodules}, \xmltag{ngenes} and
+\xmltag{nsamples}) are required and give the number of modules,
 number of genes and number of samples in the data set.
 
-A sample EVF file header, together with the \texttt{<summary>} tag looks
+A sample EVF file header, together with the \xmltag{summary} tag looks
 like this:
-\begin{XML}
+\begin{example}
 <?xml version="1.0" encoding="UTF-8"?>
@@
   </summary>
-\end{XML}
+\end{example}
 
@@
 
-The \texttt{<experimentdata>} tag is next, this contains the
+The \xmltag{experimentdata} tag is next, this contains the
 experiment meta-data. It has the following parts:
@@
 \end{schema}
-All fields are optional here: \texttt{<title>}, \texttt{<name>},
-\texttt{<lab>}, \texttt{<abstract>}, \texttt{<url>},
-\texttt{<annotation>}, \texttt{<organism>}. They are collected and
+All fields are optional here: \xmltag{title}, \xmltag{name},
+\xmltag{lab}, \xmltag{abstract}, \xmltag{url},
+\xmltag{annotation}, \xmltag{organism}. They are collected and
 shown in the \emph{Experiment} tab in the interactive ExpressionView
-viewer. The \texttt{<annotation>} tag should contain the name of the
-chip on which the experiment was performed. If the \texttt{<organism>}
+viewer. The \xmltag{annotation} tag should contain the name of the
+chip on which the experiment was performed. If the \xmltag{organism}
 tag contains `\texttt{Homo sapiens}', then ExpressionView links genes with
@@
 
-Here is an example for the \texttt{<experimentdata>} tag:
-\begin{XML}
+Here is an example for the \xmltag{experimentdata} tag:
+\begin{example}
 <experimentdata>
@@
 </experimentdata>
-\end{XML}
+\end{example}
 
@@
 
-The next tag within \texttt{<evf>} is \texttt{<genes>}, its type os
+The next tag within \xmltag{evf} is \xmltag{genes}, its type os
 defined as
@@
 \end{schema}
-It has two main parts, a \texttt{<genetags>} tag first, and then a
-number of \texttt{<gene>} tags, one for each gene in the data set.
+It has two main parts, a \xmltag{genetags} tag first, and then a
+number of \xmltag{gene} tags, one for each gene in the data set.
 
@@
 
-The \texttt{<genetags>} tag has one required and many optional
+The \xmltag{genetags} tag has one required and many optional
 subtags:
@@
 \end{schema}
-The \texttt{<id>} must be an integer number, that starts with one and
+The \xmltag{id} must be an integer number, that starts with one and
 goes up to the number of genes. It is used as an id that is refered
-from other tags in the document. \texttt{<name>} is typically the name
+from other tags in the document. \xmltag{name} is typically the name
 of the probe-set, but it can be used for other purposes as well,
-\texttt{<symbol>} is typically the canonical gene
-symbol. \texttt{<entrezid>} is the Entrez gene id. If a given
-probeset does not map to a known gene, then \texttt{<symbol>} and
-\texttt{<entrezid>} can be set to \texttt{NA}. Finally, the gene tags
-might contain any number of \texttt{<x>} tags, each defining an
+\xmltag{symbol} is typically the canonical gene
+symbol. \xmltag{entrezid} is the Entrez gene id. If a given
+probeset does not map to a known gene, then \xmltag{symbol} and
+\xmltag{entrezid} can be set to \texttt{NA}. Finally, the gene tags
+might contain any number of \xmltag{x} tags, each defining an
 additional column in the \emph{Genes} tab in ExpressionView. These
@@
 tags of the individual genes. Here is an example gene tags section:
-\begin{XML}
+\begin{example}
 <genetags>
@@
 </genetags>
-\end{XML}
+\end{example}
 We define an extra column for the gene table, this will give the
@@
 After the gene tags, we have a typically large number of genes in the
-file. The definition of the \texttt{<gene>} tag:
+file. The definition of the \xmltag{gene} tag:
 \begin{schema}
@@
 \end{schema}
-The \texttt{<gene>} tags must have exactly the same subtags as the
+The \xmltag{gene} tags must have exactly the same subtags as the
 ones given in the gene tags section. For example, assuming the
-\texttt{<genetags>} example above we can have:
-\begin{XML}
+\xmltag{genetags} example above we can have:
+\begin{example}
 <gene>
@@
 ...
-\end{XML}
+\end{example}
 Observe, how the extra tag for the chromosomes is referenced here.
@@
 \end{schema}
-The \texttt{<samples>} tag has two parts, the first defines the sample
-meta data entries and the second part contains one \texttt{<sample>}
+The \xmltag{samples} tag has two parts, the first defines the sample
+meta data entries and the second part contains one \xmltag{sample}
 tag for each sample in the data set.
@@
 \end{schema}
-The \texttt{<id>} and \texttt{<name>} sample tags are
-required. \texttt{<id>} contains numeric ids, from one, up to the
+The \xmltag{id} and \xmltag{name} sample tags are
+required. \xmltag{id} contains numeric ids, from one, up to the
 number of samples in the data set. These are referenced by other tags,
-e.g. the ones that define the modules. \texttt{<name>} is an arbitrary
+e.g. the ones that define the modules. \xmltag{name} is an arbitrary
 string, it is typically the sample name in the R ExpressionSet object,
 that was used to find the biclusters. Similarly to the genes, any
-number of additional \texttt{<x>} tags can be added, each with a
+number of additional \xmltag{x} tags can be added, each with a
 unique \texttt{name} attribute, to define additional information about
@@
 
-Here is an example \texttt{<sampletags>} tag:
-\begin{XML}
+Here is an example \xmltag{sampletags} tag:
+\begin{example}
 <sampletags>
@@
 </sampletags>
-\end{XML}
+\end{example}
 We defined four extra tags for the samples in this example.
@@
 \end{schema}
-Here is an example, that corresponds to the \texttt{<sampletags>}
+Here is an example, that corresponds to the \xmltag{sampletags}
 entry given above:
-\begin{XML}
+\begin{example}
 <sample>
@@
 ...
-\end{XML}
+\end{example}
 
@@
 
-Again, the \texttt{<modules>} tag has a syntax that is similar to the
+Again, the \xmltag{modules} tag has a syntax that is similar to the
 syntax of the genes and the samples:
@@
 There are two required and two optional parts. The first part is
-\texttt{<moduletags>}, this defines the meta-data associated with the
-modules. The \texttt{<gotags>} and \texttt{<keggtags>} tags are
+\xmltag{moduletags}, this defines the meta-data associated with the
+modules. The \xmltag{gotags} and \xmltag{keggtags} tags are
 optional, they are only present if the file contains Gene Ontology
 and/or KEGG pathway enrichment calculation results for the
-biclusters. Finally, there is a list of \texttt{<module>} tags, one
+biclusters. Finally, there is a list of \xmltag{module} tags, one
 for each bicluster.
 
-The type of \texttt{<moduletags>}:
+The type of \xmltag{moduletags}:
 \begin{schema}  
@@
 \end{schema}
-There is only one tag required in \texttt{<moduletags>}, the
-\texttt{<id>}, which is a numeric id, going from 1 to the number of
+
+There is only one tag required in \xmltag{moduletags}, the
+\xmltag{id}, which is a numeric id, going from 1 to the number of
 modules. The rest is optional, and it is also possible to create any
-kind of extra tags with the \texttt{<x>} tag. \texttt{<name>} is an
+kind of extra tags with the \xmltag{x} tag. \xmltag{name} is an
 arbitrary character string. The other optional tags are
 mainly defined to modules coming from the ISA algorithm:
-\texttt{<iterations>} gives the number of ISA iterations needed to
-find the module, \texttt{<oscillation>} specifies the oscillation
+\xmltag{iterations} gives the number of ISA iterations needed to
+find the module, \xmltag{oscillation} specifies the oscillation
 cycle length for oscillating modules. It is mostly zero, meaning that
-the module does not oscillate. \texttt{<thr\_row>} is the ISA gene
-threshold, \texttt{<thr\_col>} is the ISA condition
-threshold. \texttt{<freq>} is the number of ISA seeds that converged
-to the module, \texttt{<rob>} is its robustness
-score. \texttt{<rob\_limit>} is the robustness threshold that was used
+the module does not oscillate. \xmltag{thr\_row} is the ISA gene
+threshold, \xmltag{thr\_col} is the ISA condition
+threshold. \xmltag{freq} is the number of ISA seeds that converged
+to the module, \xmltag{rob} is its robustness
+score. \xmltag{rob\_limit} is the robustness threshold that was used
 to filter the module.
 
-Here is an example \texttt{<moduletags>} tag:
-\begin{XML}
+Here is an example \xmltag{moduletags} tag:
+\begin{example}
 <moduletags>
@@
 </moduletags>
-\end{XML}
+\end{example}
 
-The optional \texttt{<gotags>} and \texttt{<keggtags>} tags are
+The optional \xmltag{gotags} and \xmltag{keggtags} tags are
 discussed in Section~\ref{sec:gokegg}, let us now see how the data for a
@@
 The first couple of fields refer to the ones defined in the
-\texttt{<moduletags>} section above. The others are:
-\texttt{<containedgenes>}, a space separated list of gene ids, for the
-genes that are included in the module; \texttt{<genescores>}, the gene
+\xmltag{moduletags} section above. The others are:
+\xmltag{containedgenes}, a space separated list of gene ids, for the
+genes that are included in the module; \xmltag{genescores}, the gene
 scores of these genes, in the order of the list in
-\texttt{<containedgenes>}; \texttt{<containedsamples>}, the ids of the
-samples in the module; \texttt{<samplescores>}, the scores for these
-samples; \texttt{<intersectingmodules>}, the ids of the modules that
-have an overlap with the given module. \texttt{<gos>} and
-\texttt{<keggs>} are optional and describe the Gene Ontology and KEGG
+\xmltag{containedgenes}; \xmltag{containedsamples}, the ids of the
+samples in the module; \xmltag{samplescores}, the scores for these
+samples; \xmltag{intersectingmodules}, the ids of the modules that
+have an overlap with the given module. \xmltag{gos} and
+\xmltag{keggs} are optional and describe the Gene Ontology and KEGG
 pathway enrichment of the given module, see their details in Section~\ref{sec:gokegg}.
 
-Here is an example \texttt{<module>} tag:
-\begin{XML}
+Here is an example \xmltag{module} tag:
+\begin{example}
 <module>
@@
 </module>
-\end{XML}
+\end{example}
 
 We refered to the integer list and double list data types above, now
-we define them:
+we define them, this is the integer list:
 \begin{schema}  
@@
 \end{schema}
+
+and similarly for the double list:
 
@@
 pathway enrichment calculation results. In this case
-\texttt{<moduletags>} is followed by a \texttt{<gotags>} and/or a
-\texttt{<keggtags>} tag. The definition of \texttt{<gotags>}:
+\xmltag{moduletags} is followed by a \xmltag{gotags} and/or a
+\xmltag{keggtags} tag. The definition of \xmltag{gotags}:
 \begin{schema}
@@
 \end{schema}
-\texttt{<id>} is a numeric id, starting from one, up to the total
-number of enriched GO categories; \texttt{<go>} is the GO id;
-\texttt{<term>} is the GO term; \texttt{<ontology>} is the GO ontology of
+\xmltag{id} is a numeric id, starting from one, up to the total
+number of enriched GO categories; \xmltag{go} is the GO id;
+\xmltag{term} is the GO term; \xmltag{ontology} is the GO ontology of
 the term, its possible values are: \texttt{BP}, \texttt{CC},
@@
   component} and \emph{Molecular function};
-\texttt{<pvalue>} is the enrichment $p$-value; \texttt{<oddsratio>} is the
-odds ratio; \texttt{<expcount>} is the expected number of genes in the
-intersection, by chance; \texttt{<count>} is the number of genes in the
-intersection; \texttt{<size>} is the size of the GO term, i.e. the
+\xmltag{pvalue} is the enrichment $p$-value; \xmltag{oddsratio} is the
+odds ratio; \xmltag{expcount} is the expected number of genes in the
+intersection, by chance; \xmltag{count} is the number of genes in the
+intersection; \xmltag{size} is the size of the GO term, i.e. the
 number of genes (in the current gene universe) that are annotated with
@@
 \end{schema}
-The tags for \texttt{<keggtags>} is almost the same as for
-\texttt{<gotags>}, but here \texttt{<kegg>} id the KEGG pathway ID,
-and \texttt{<pathname>} is the name of the pathway. The rest is the
+The tags for \xmltag{keggtags} is almost the same as for
+\xmltag{gotags}, but here \xmltag{kegg} id the KEGG pathway ID,
+and \xmltag{pathname} is the name of the pathway. The rest is the
 same.
 
-Part of an EVF file with the \texttt{<gotags>} and
-\texttt{<keggtags>} tags:
-\begin{XML}
+Part of an EVF file with the \xmltag{gotags} and
+\xmltag{keggtags} tags:
+\begin{example}
 <gotags>
@@
 </keggtags>
-\end{XML}
+\end{example}
 
-The actual enrichment data is given in the \texttt{<module>} tags,
-these can contain a \texttt{<gos>} and/or a \texttt{<keggs>} tag with
+The actual enrichment data is given in the \xmltag{module} tags,
+these can contain a \xmltag{gos} and/or a \xmltag{keggs} tag with
 the enrichment $p$-values and other statistics.
@@
 \end{schema}
-\texttt{<gos>} is a list of \texttt{<go>} tags.
+\xmltag{gos} is a list of \xmltag{go} tags.
 
@@
 \end{schema}
-\texttt{<keggs>} is a list of \texttt{<kegg>} tags.
+\xmltag{keggs} is a list of \xmltag{kegg} tags.
 
-The subtags within a \texttt{<go>} or a \texttt{<kegg>} tag refer to
-the tags already listed above in the \texttt{<gotags>} and
-\texttt{<keggtags>} tags:
+The subtags within a \xmltag{go} or a \xmltag{kegg} tag refer to
+the tags already listed above in the \xmltag{gotags} and
+\xmltag{keggtags} tags:
 \begin{schema}
@@
 
-Here is an example, this is part of a \texttt{<module>} tag:
-\begin{XML}
+Here is an example, this is part of a \xmltag{module} tag:
+\begin{example}
 <gos>
@@
 </keggs>
-\end{XML}
+\end{example}
 
@@
 
-Finally, the expression data in included in the \texttt{<data>}
+Finally, the expression data in included in the \xmltag{data}
 tag. This is a Base64 encoded string, generated by representing each
@@
 and Base64 encoding these bytes. This data looks like this:
-\begin{XML}
+\begin{example}
 <data>
@@
 </data>
-\end{XML}
+\end{example}
 

### 2010-05-26 expressionviewnote/answers-to-reviewers.tex
EV note: added answers to reviewers.

\documentclass[a4paper]{article}
\usepackage{url}
\usepackage{natbib}
\usepackage{ragged2e}
\usepackage{color}

\newcommand{\Rpackage}[1]{`\texttt{#1}'}
\newcommand{\Rfunction}[1]{`\texttt{#1}'}
\newcommand{\Rclass}[1]{\textsl{#1}}

\newenvironment{myquote}{\begin{quote}\color{blue}}{\end{quote}}

\setlength{\parskip}{12pt}
\setlength{\parindent}{0pt}

\title{ExpressionView -- an interactive viewer for
  modules identified in gene expression data\\[10pt]
  Answers to reviewers}
\author{Andreas L\"uscher, G\'abor Cs\'ardi, \\ Aitana Morton de
  Lachapelle, Zolt\'an Kutalik, \\ and Sven Bergmann}

\begin{document}

\maketitle

\RaggedRight

First of all, we want to thank the reviewers for their thorough work
and useful suggestions. In the following, the reviewers' comments are
typeset in blue, and indented as a block.

\section*{Reviewer 1}

\begin{myquote}
All comments are for minor revisions:

1) BiVisu by Cheng et al shall be cited:
Bioinformatics. 2007;23(17):2342-4. Please briefly elaborate the major
difference(s) of ExpressionView from BiVisu and BicOverlapper.
\end{myquote}

We have added a brief comparison to Bivisu~\citep{cheng07} and
BicOverlapper\citep{santamaria08}.

\begin{myquote}
2) Page 2, line 30: please re-write the sentence "We find that in
almost every situation, the proposed algorithm finds an order that
recovers more than 99\% of the score of the optimal solution and in
most cases, it recovers the correct alignment". 
\end{myquote}

We have rewritten this part.

\begin{myquote}
3) Page 2, line 55, change "any computer" to "a computer".
\end{myquote}

We have corrected this.

\section*{Reviewer 2}

\begin{myquote}
Comments to the Author

This is a very useful tool which users might  benefit if there is
clearly stated in the website, what sort of input format the
application accepts. There does not seem to be a complete and explicit
description of the various types of input format beyond just a
expression matrix.
\end{myquote}

For the matrix reordering, ExpressionView accepts modules in the
format defined in the eisa R package~\citep{csardi10}
(\textit{ISAModules} object) and the
Biclust R package~\citep{kaiser08} (\textit{Biclust} object). These
cover the existing R implementations of biclustering algorithms (to our
knowledge). Moreover, the Biclust package defines a very simple class
for storing biclustering results, so the output of any other algorithm
(implemented in R or not) can be easily transformed into this
format. This is documented in the Biclust package.

For the expression data itself, ExpressionView accepts
\textit{ExpressionSet} objects as the input. This is the standard data
type in BioConductor for storing expression data. E.g. the GCRMA and
other normalization methods all output this format and it is also easy
to create an \textit{ExpressionSet} object from a simple tabular text
file.

Alternatively, ExpressionView can visualize non-expression data,
inputted as a simple matrix, this is explained in the R manual that
comes with the package.

\begin{myquote}
For example, how do we incorporate GO terms and
KEGG pathways into the raw input dataset, or are these also computed
by the application? 
\end{myquote}

These are computed by the R part of the application, using functions
in the eisa R package. There are examples for this in the manual that
comes with the package.

\begin{myquote}
It is not entirely clear to the casual reader,
whether the Flash application takes raw data or the intermediate XML
result from the workflow, which means user may need to use command
line R anyway to generate the data. If it does so, the format of this
needs to be clearly stated for a non-bioinformatics biologist to be
able to load and run the application. 
\end{myquote}

The Flash application takes the intermediate XML file, we made this
clear in Fig.~1. The idea is that the user performing the analysis in
R is usually not the same, as the user browsing the results. So the
(bioinformatician) user generates the data in R, exports it into an
XML file, and the other (life scientist) user explores it using an
ordinary web browser. We have made this distinction clear in the
manuscript.

The file format of the XML file is not relevant in most cases, as the
file is produced by the ExpressionView package, and the user does not
have to produce it by hand. It is only relevant, if the user wants to
import data that was produced by another reordering algorithm, into
the Flash applet. We have created an XML Schema document that precisely
defines the XML data file's format, and also written a document that
discusses the format. These are available on the ExpressionView website.

\begin{myquote}
The tutorials are impressive, but the audio volume could be increased
or spoken by a person with a higher pitch voice. Very difficult to
hear.  
\end{myquote}

We have increased the audio volume in the video tutorials.

\begin{myquote}
ExpressionView source code (for the Flash applet) in winrar
format contains a lot of empty folders and tmp files some of which may
not be necessary. 
\end{myquote}

We have simplified the directory structure of the source code.

\begin{myquote}
The paper could do with at least a few words in the body of the paper
or in the website, to explain what improvement over features of other
existing Expression data biclustering visualisers currently
available. 
\end{myquote}

We have included a brief comparison to Bivisu~\citep{cheng07} and
BiocOverlapper~\citep{santamaria08}, two existing bicluster
visualization software packages.

\begin{myquote}
Moreover, the paper must a least emphasize justification
for why biologists should find this application/algorithm more useful
over other existing applications/algorithms besides just visual
appeal. 
\end{myquote}

We emphasized the advantages of ExpressionView more in the paper:
\begin{itemize}
\item visualization of all the modules and the expression values together.
\item ease of use for the (life scientist) end user, only a web browser is
  needed, no extra programs needed to be installed.
\item the inclusion of meta-data and enrichment data into the same
  environment, with interactive links to corresponding databases.
\end{itemize}

\section*{Reviewer 3}

\begin{myquote}
Comments to the Author

General comments:

Luescher et al. describe software for visualizing modules in gene
expression data using an efficient implementation of matrix
reordering. Such a package is of high practical value for biologists
to analyze and visualize gene expression data, and I therefore
recommend publication as an Applications Note. In particular, I
appreciate the connection to the Bioconductor package, it is well
done.

Specific comments:

Minor:

I would recommend a different name for the package, ExpressionView is
too general.
\end{myquote}

While this is just criticism, we cannot easily change the name of the
package, as it is part of BioConductor now, and the name change would
cause a lot problems for the users and the BioConductor developers.

\begin{myquote}
Biologists should have no hurdles to access and apply the software
\ldots I could not easily install it on my LINUX/Ubuntu computer after
having installed the isa2 package, as recommended by the authors. 
I assume the download and installation of the package will be as
straightforward as the isa2 package by the authors.
\end{myquote}

The ExpressionView package is an official BioConductor package since
April 2010, so its installation should be easy on any platforms
supported by BioConductor (Linux, MS Windows, Apple OSX). We have
updated the installation instructions on the ExpressionView homepage
accordingly.

\begin{myquote}
The authors should consider that some computers might not play the
Flash applets, and should provide short and clear instructions for
installation on their website.
\end{myquote}

We have added a note about this, and a link to the relevant Adobe
website, to the ExpressionView homepage.

\bibliographystyle{natbib}
\bibliography{expressionview}

\end{document}

### 2010-05-26 expressionviewnote/expressionview.tex
EV note: updated.

--- 2010-01-15 expressionviewnote/expressionview.tex
@@
 
-\vspace*{-7pt}
+\vspace*{-9pt}
 \section{Introduction}
 Biclustering is an unsupervised data analysis method which is
-frequently used to explore microarray data. 
-Biclustering algorithms process collections of expression profiles to
-identify groups of genes co-expressed under some conditions (samples).
-We refer to such groups as modules. While
-there is a multitude of biclustering software available (for a reviews 
-and comparisons see~\cite{madeira04,ihmels04,prelic06}), packages with 
-intuitive interfaces that allow for an interactive exploration of the results are
-sparse~\citep{santamaria08}. ExpressionView is designed to close this
-gap, and facilitate the collaboration between bioinformaticians and
-life scientists without consolidated programming
-experience. Implemented as an R package, the user can apply all the
-powerful microarray analysis tools provided by the
-Bioconductor project~\citep{gentleman04} before exporting the data to
-a platform independent Adobe Flash applet. This applet
-provides an integrated, intuitive and interactive visualization of
-the modules together with the underlying gene expression data and
-additional information on the genes and conditions.
+frequently used to explore microarray data. Biclustering algorithms
+process collections of expression profiles to identify groups of genes
+co-expressed under some conditions (samples). We refer to such groups
+as modules. While there is a multitude of biclustering software
+available (for reviews  and comparisons
+see~\cite{madeira04,ihmels04,prelic06}), packages with intuitive
+interfaces that allow for an interactive exploration of the results
+are sparse.
 
-\vspace*{-7pt}
+Two existing approaches are Bivisu~\citep{cheng07}, an interactive
+biclustering program, it plots modules individually, thus the
+relationship between the overlapping modules is hard to identify; and
+BicOverlapper~\citep{santamaria08}, a novel tool for plotting
+overlapping modules in an abstract space. Our approach in
+ExpressionView is different, as we use the usual gene-sample
+space and visualize all modules together, on top of the reordered
+expression matrix. The reordering ensures that, the genes and samples
+that appear in the same module, are kept together.
+
+\vspace*{-9pt}
 \section{Package design and workflow}
-The ExpressionView package contains two independent parts: An ordering
-algorithm implemented in C++ and an Adobe Flash applet written in
-ActionScript and Adobe Flex. Fig.~\ref{fig:workflow} schematically
-summarises the typical workflow that uses the gene expression data
-available as a Bioconductor ExpressionSet and combines it with the
-biclustering results to produce an ExpressionView data file that can
-be explored with the Flash applet. In the following, we describe the
-different steps in more detail.
-\begin{figure*}[!tpb]
-\centerline{\includegraphics[width=0.7\linewidth]{fig1-crop}}
-\caption{How to use ExpressionView: Starting from gene expression data
-  in the form of a Bioconductor ExpressionSet, the user first runs a
-  biclustering algorithm to determine co-expressed groups of genes and
-  the relevant samples. In a second step, the rows and columns of the gene
+With the ExpressionView package, bicluster analysis can be separated
+into two parts. The first part involves finding the
+modules in the data set with some algorithm, possibly running
+enrichment analysis for the modules, and reordering the rows and
+columns of the expression matrix according to the modules. This
+part is typically done by a bioinformatician. The second part of the
+analysis involves the visualization and interactive exploration of the
+results. This part is typically done by researchers without extensive
+programming knowledge.
+
+ExpressionView has thus two parts. The first part is written in GNU
+R~\citep{R} and contains an implementation of the matrix reordering
+algorithm. The second part is an interactive visualization tool, in
+the form of an Adobe Flash applet, the user of this part only needs a
+Flash-enabled web browser.
+
+This dual implementation has the advantage that all the power of the
+GNU R environment and the BioConductor~\citep{gentleman04} packages
+can be used for the analysis itself, e.g. all organisms that are (and
+will be) supported by BioConductor are automatically supported by
+ExpressionView. On the other hand, the exploration of the results does
+not need any GNU R knowledge, in most cases no extra software needs to
+be installed. Please see Fig.~\ref{fig:workflow} for a typical
+ExpressionView workflow.
+
+\begin{figure*}
+\centering
+\includegraphics[width=0.7\linewidth]{fig1}
+\caption{
+  Workflow of ExpressionView, showing the two parts of the analysis.
+  \textbf{(A)} These steps are performed by a bioinformatician,
+  using GNU R. Starting from gene expression data in the form of a
+  Bioconductor \textit{ExpressionSet}, the first step is finding the
+  modules. In a second step, the rows and columns of the gene
   expression matrix are rearranged to produce an easily readable
   overview of the results. The last step consists of combining the
-  gene expression data and its associated metadata with the results
-  from the biclustering and produce an ExpressionView data file that
-  can be explored with the interactive Flash applet.}\label{fig:workflow}
+  gene expression data and its associated metadata, possibly including
+  results of enrichment analysis, with the results
+  from the biclustering and produce an ExpressionView data file.
+  \textbf{(B)}~This file can be distributed and finally explored with
+  the interactive Flash applet by the end-user. Please see the
+  website for details on the data file format.}
+\label{fig:workflow}
 \end{figure*}
-For users who prefer to install the viewer as a standalone program, we
-also provide an Adobe AIR version that can be downloaded from the
-\href{http://www.unil.ch/cbg/ExpressionView}{website}.
 
-\subsection{Gene expression data and modules}
+\vspace*{-9pt}
+\subsection{Reordering genes and conditions}
+
+% \subsection{Gene expression data and modules}
 ExpressionView is designed to work with gene expression data in the
-form of a Bioconductor ExpressionSet. This class provides a
+form of a Bioconductor \textit{ExpressionSet}. This class provides a
 user-friendly way to access the actual gene expression matrix and its
 associated metadata. ExpressionView can treat results obtained by the 
-Iterative Signature Algorithm~\citep{bergmann03,csardi09} and any of 
+Iterative Signature Algorithm~\citep{bergmann03,csardi10} and any of 
 the methods available in the Biclust package~\citep{kaiser08}. Since 
@@
 
-\subsection{Reordering genes and conditions}
 To present the collection of possibly overlapping modules in a visually
 appealing form, it is necessary to reorder the rows (conditions) and
-columns (genes) of the gene expression matrix in such a way that biclusters form contiguous
+columns (genes) of the gene expression matrix in such a way that
+biclusters form contiguous
 rectangles. Since it is in general impossible to find such an
@@
 here an approximate solution that optimizes the arrangement within the
-original data, by maximising the total area of the largest contiguous
+original data, by maximizing the total area of the largest contiguous
 module subsets. (An alternative would be to repeat rows and columns as
@@
 
-To determine the efficiency of this method, we have studied a large
-number of perfectly orderable, but initially scrambled, test cases. We
-find that in almost every situation, the proposed algorithm finds an
+We have studied a large number of perfectly orderable, but initially
+scrambled, test cases. We find that the proposed algorithm finds an
 order that recovers more than 99\% of the score of the optimal
@@
 with the number of clusters $m$ as ${\mathcal O}(m^\alpha)$, where
-$\alpha \in [1.6, 2]$, almost independent of the number of elements
+$\alpha \in [1.6, 2]$, almost independently of the number of elements
 $n$. For a given number of clusters, we find ${\mathcal O}(n^\alpha)$,
@@
 
-\subsection{Export}
 Once the optimal order is determined, the program rearranges the gene
 expression matrix accordingly and exports all the relevant information
-into an XML file. This format has the advantage of being
-self-explanatory and extensible. In addition, the files can be read
-and edited with any text editor. The actual gene expression values are
-rounded to two digit precision and encoded in Base64. The structure of
-the XML file is described in more detail on the ExpressionView
-\href{http://www.unil.ch/cbg/ExpressionView}{website}. In most cases,
-these files are of the order of a few Megabytes and can thus easily be
-sent by e-mail.
+into an XML file, that can be placed on a web-server or distributed by
+email and then imported by the interactive viewer.
 
 \subsection{Visualisation}
-The core application of this package is the Flash applet that allows
-the user to interactively explore the gene expression data and the
-modules. The Flash architecture has the advantage of
-producing a platform and browser independent applet that can be
-launched from any computer with even the most restrictive user
-rights. A screenshot is shown in Fig.~\ref{fig:workflow}. The
+
+A screenshot of the viewer is shown in Fig.~\ref{fig:workflow}. The
 interface is divided in two parts: On the left-hand side, the user
@@
 the expression data, getting instant feedback on the selected
-item. The currently selected region can be exported to a PDF file (e.g. 
-to prepare figures for presentations or articles) or to a
-text file.
+item.
 
-\vspace*{-7pt}
+\vspace*{-9pt}
 \section*{Acknowledgement}
@@
 
-\vspace*{-7pt}
+\vspace*{-9pt}
 \bibliographystyle{natbib}

### 2010-05-26 expressionviewnote/answers-to-reviewers.tex
EV note: finishing touches in the answers to reviewers.

@@
 \author{Andreas L\"uscher, G\'abor Cs\'ardi, \\ Aitana Morton de
-  Lachapelle, Zolt\'an Kutalik, \\ and Sven Bergmann}
+  Lachapelle, Zolt\'an Kutalik, \\ Bastian Peter and Sven Bergmann}
 
@@
 
-We have added a brief comparison to Bivisu~\citep{cheng07} and
-BicOverlapper\citep{santamaria08}.
+We have added a citation to Bivisu and a brief comparison to
+Bivisu~\citep{cheng07} and BicOverlapper\citep{santamaria08}.
 
@@
 We have included a brief comparison to Bivisu~\citep{cheng07} and
-BiocOverlapper~\citep{santamaria08}, two existing bicluster
+BicOverlapper~\citep{santamaria08}, two existing bicluster
 visualization software packages.
@@
 While this is just criticism, we cannot easily change the name of the
-package, as it is part of BioConductor now, and the name change would
-cause a lot problems for the users and the BioConductor developers.
+package, as it is part of BioConductor, and the name change would
+cause a lot of problems for the users and the BioConductor developers.
 

### 2010-05-31 expressionviewnote/answers-to-reviewers.tex
EV note: added changes from the co-authors.

@@
 
+\hyphenation{Bic-Over-lap-per}
+
 \begin{document}
@@
 and useful suggestions. In the following, the reviewers' comments are
-typeset in blue, and indented as a block.
+typeset in blue, and indented as a block. Quotations from the
+manuscript itself appear in a slanted font.
 
@@
 We have added a citation to Bivisu and a brief comparison to
-Bivisu~\citep{cheng07} and BicOverlapper\citep{santamaria08}.
+Bivisu~\citep{cheng07} and BicOverlapper\citep{santamaria08}:
+
+\textsl{Existing approaches include Bivisu~\citep{cheng07} and
+BicOverlapper~\citep{santamaria08}. The former is an interactive
+biclustering program that plots modules individually, making it
+difficult to identify the relationship between the overlapping
+modules. The latter is a novel tool for plotting overlapping modules,
+yet in an abstract space. Our approach in ExpressionView is different,
+as we use the usual gene-sample space and visualize all modules
+together, on top of the reordered expression matrix. The reordering
+ensures that the genes and samples that appear in the same module are
+kept together.}
 
@@
 
-We have rewritten this part.
+We have cut the ambiguous part of the sentence, it now reads:
+
+\textsl{We find that the proposed algorithm finds an
+order that recovers more than 99\% of the score of the optimal
+solution and in most cases, it recovers the correct
+alignment.}
 
@@
 Alternatively, ExpressionView can visualize non-expression data,
-inputted as a simple matrix, this is explained in the R manual that
-comes with the package.
+provided as a simple matrix, this is explained in the R manual that
+comes with the package, including examples. This manual is also
+available on the ExpressionView website now, under ``Tutorials''.
+
+For the Flash application itself, we have extensively documented the
+ExpressionView XML file format. This is documentation is available on
+the website, under ``Tutorials''.
 
@@
 These are computed by the R part of the application, using functions
-in the eisa R package. There are examples for this in the manual that
-comes with the package.
+of the eisa R package. There are examples for this in the manual that
+comes with the package. This manual is also available now on the
+ExpressionView website.
 
@@
 
-The Flash application takes the intermediate XML file, we made this
-clear in Fig.~1. The idea is that the user performing the analysis in
-R is usually not the same, as the user browsing the results. So the
+The Flash application takes the intermediate XML file, as shown
+in Fig.~1. The idea is that the user performing the analysis in
+R is usually not the same as the user browsing the results. So the
 (bioinformatician) user generates the data in R, exports it into an
-XML file, and the other (life scientist) user explores it using an
+XML file, and the other (e.g. life scientist) user explores it using an
 ordinary web browser. We have made this distinction clear in the
-manuscript.
+manuscript:
+
+{\slshape
+With the ExpressionView package, bicluster analysis can be separated
+into two parts. The first part involves finding the
+modules in the data set with some algorithm, possibly running
+enrichment analysis for the modules, and reordering the rows and
+columns of the expression matrix according to the modules. This
+part is typically done by a bioinformatician. The second part of the
+analysis involves the visualization and interactive exploration of the
+results. This part is typically done by researchers without extensive
+programming knowledge.
+
+The first part of is written in GNU
+R~\citep{R} and contains an implementation of the matrix reordering
+algorithm. The second part is an interactive visualization tool in
+the form of an Adobe Flash applet, for which the user only needs a
+Flash-enabled web browser.
+
+This dual implementation has the advantage that all the power of the
+GNU R environment and the BioConductor~\citep{gentleman04} packages
+can be used for the analysis itself, e.g. all organisms that are (and
+will be) supported by BioConductor are automatically supported by
+ExpressionView. On the other hand, the exploration of the results does
+not need any GNU R knowledge and in most cases no extra software needs to
+be installed. See Fig.~1 for a typical
+ExpressionView workflow.
+}
 
@@
 
-We have simplified the directory structure of the source code.
+We have simplified the directory structure of the source code and 
+removed all empty folders.
 
@@
 BicOverlapper~\citep{santamaria08}, two existing bicluster
-visualization software packages.
+visualization software packages:
+
+\textsl{Existing approaches include Bivisu~\citep{cheng07} and
+BicOverlapper~\citep{santamaria08}. The former is an interactive
+biclustering program that plots modules individually, making it
+difficult to identify the relationship between the overlapping
+modules. The latter is a novel tool for plotting overlapping modules,
+yet in an abstract space. Our approach in ExpressionView is different,
+as we use the usual gene-sample space and visualize all modules
+together, on top of the reordered expression matrix. The reordering
+ensures that the genes and samples that appear in the same module are
+kept together.
+}
 
@@
 \begin{itemize}
-\item visualization of all the modules and the expression values together.
+\item visualization of all the modules and the expression values
+  together:
+
+  \textsl{\noindent
+    Our approach in ExpressionView is different,
+    as we use the usual gene-sample space and visualize all modules
+    together, on top of the reordered expression matrix. The reordering
+    ensures that the genes and samples that appear in the same module are
+    kept together.
+  }
+\item implementation as a BioConductor package:
+
+  \textsl{\noindent
+    This dual implementation has the advantage that all the power of the
+    GNU R environment and the BioConductor~\citep{gentleman04} packages
+    can be used for the analysis itself, e.g. all organisms that are (and
+    will be) supported by BioConductor are automatically supported by
+    ExpressionView. 
+  }
 \item ease of use for the (life scientist) end user, only a web browser is
-  needed, no extra programs needed to be installed.
+  needed, no extra programs need to be installed.
+
+  \textsl{\noindent
+    On the other hand, the exploration of the results does
+    not need any GNU R knowledge and in most cases no extra software needs to
+    be installed.    
+  }
 \item the inclusion of meta-data and enrichment data into the same
-  environment, with interactive links to corresponding databases.
+  environment, with interactive links to corresponding databases:
+
+  \textsl{\noindent
+    On the right-hand side, the metadata
+    associated with the expression data and the results of the
+    enrichment calculations for GO~\citep{ashburner00} categories and
+    KEGG~\citep{kanehisa04} pathways are shown. Wherever possible, these elements
+    are linked to the corresponding databases.    
+  }
 \end{itemize}
@@
 
-While this is just criticism, we cannot easily change the name of the
+While this is a just criticism, we cannot easily change the name of the
 package, as it is part of BioConductor, and the name change would
-cause a lot of problems for the users and the BioConductor developers.
+cause problems for the users and the BioConductor developers.
 
@@
 The ExpressionView package is an official BioConductor package since
-April 2010, so its installation should be easy on any platforms
+April 2010, so its installation should be easy on any platform
 supported by BioConductor (Linux, MS Windows, Apple OSX). We have

### 2010-05-31 expressionviewnote/expressionview.tex
EV note: added changes from the co-authors.

@@
 \pubyear{2010}
+
+\hyphenation{Bic-Over-lap-per}
 
@@
 
-Two existing approaches are Bivisu~\citep{cheng07}, an interactive
-biclustering program, it plots modules individually, thus the
-relationship between the overlapping modules is hard to identify; and
-BicOverlapper~\citep{santamaria08}, a novel tool for plotting
-overlapping modules in an abstract space. Our approach in
-ExpressionView is different, as we use the usual gene-sample
-space and visualize all modules together, on top of the reordered
-expression matrix. The reordering ensures that, the genes and samples
-that appear in the same module, are kept together.
+Existing approaches include Bivisu~\citep{cheng07} and
+BicOverlapper~\citep{santamaria08}. The former is an interactive
+biclustering program that plots modules individually, making it
+difficult to identify the relationship between the overlapping
+modules. The latter is a novel tool for plotting overlapping modules,
+yet in an abstract space. Our approach in ExpressionView is different,
+as we use the usual gene-sample space and visualize all modules
+together, on top of the reordered expression matrix. The reordering
+ensures that the genes and samples that appear in the same module are
+kept together.
 
@@
 
-ExpressionView has thus two parts. The first part is written in GNU
+The first part of ExpressionView is written in GNU
 R~\citep{R} and contains an implementation of the matrix reordering
-algorithm. The second part is an interactive visualization tool, in
-the form of an Adobe Flash applet, the user of this part only needs a
+algorithm. The second part is an interactive visualization tool in
+the form of an Adobe Flash applet, for which the user only needs a
 Flash-enabled web browser.
@@
 ExpressionView. On the other hand, the exploration of the results does
-not need any GNU R knowledge, in most cases no extra software needs to
-be installed. Please see Fig.~\ref{fig:workflow} for a typical
+not need any GNU R knowledge and in most cases no extra software needs to
+be installed. See Fig.~\ref{fig:workflow} for a typical
 ExpressionView workflow.
@@
 user-friendly way to access the actual gene expression matrix and its
-associated metadata. ExpressionView can treat results obtained by the 
+associated metadata. ExpressionView can treat biclustering results obtained by the 
 Iterative Signature Algorithm~\citep{bergmann03,csardi10} and any of 
@@
 This optimization task is an interesting problem on its own, which to
-the best of our knowledge has not been studied in the literature. Thus
-we briefly outline our strategy here (see our website for details).
-The reordering of the rows and the columns do not depend on each
-other, so the same optimization method can be applied independently
+the best of our knowledge has not been studied in the literature.
+We briefly outline our strategy here (see our website for details).
+The reordering of the rows is independent from that of the columns,
+so the same optimization method can be applied separately
 to rows and columns. For a given order of the
-elements (either genes or conditions) we compute for each module $i$ the
+elements (either genes or conditions), we compute for each module $i$ the
 size of the largest contiguous sequence of elements (i.e. the maximal
 number of neighboring elements $N^\text{max}_i$). Then, as a measure of the
-quality ($Q$) of the order we sum this quantity over all modules
+quality ($Q$) of the order, we sum this quantity over all modules
 ($Q=\sum_i N^\text{max}_i$). To optimize $Q$, an initial sequence is calculated using
@@
 
-\vspace*{-9pt}
 \section*{Acknowledgement}

### 2011-10-18 eisa/inst/doc/EISA_biclust.Rnw
eisa: tutorials use 'xlib' graphics, if available.
Updated a vignette because of the new biclust package.

@@
 the parameters stored in the \Rclass{Biclust} object, this is always advised.
-The procedure makes use of the probe and sample names that are kept and stored 
-in the \Rclass{Biclust} object, this information will be used later, e.g. for 
-the enrichment analysis. The conversion itself can be performed with
-the usual \Rfunction{as} function. 
+We also need to add the input data to the \Rclass{Biclust} object by
+hand, as this is not included in the \Rclass{Biclust} object, when
+using a recent version of the \Rpackage{biclust} package. The column
+and row names of the data will be used later, e.g. for the enrichment
+analysis. The conversion itself can be performed with the usual
+\Rfunction{as} function.  
 <<convert bc to isa>>=
 Bc@Parameters$annotation <- annotation(ALL.filtered)
+Bc@Parameters$Data <- exprs(ALL.filtered)
 modules <- as(Bc, "ISAModules")

### 2011-10-18 eisa/inst/doc/EISA_biclust.Rnw
eisa: added 'annotate' function, from the BioC SVN. This is a better way to convert Biclust objects.

--- 2010-04-06 eisa/inst/doc/EISA_biclust.Rnw
@@
 object, that is used in the \Rpackage{eisa} package. To help
-some \Rpackage{eisa} functions, we add the name of the annotation package to 
+some \Rpackage{eisa} functions, we add some biological annotation to
+the \Rclass{Biclust} object.  of the annotation package to 
 the parameters stored in the \Rclass{Biclust} object, this is always advised.
@@
 <<convert bc to isa>>=
-Bc@Parameters$annotation <- annotation(ALL.filtered)
+Bc <- annotate(Bc, ALL.filtered)
 modules <- as(Bc, "ISAModules")

### 2011-10-18 eisa/inst/doc/EISA_tutorial.Rnw
eisa: tutorial updated, it failed on R 2.12.0 (beta).
eisa: added 'annotate' function, from the BioC SVN. This is a better way to convert Biclust objects.

@@
 library(GO.db)
-mget(sigCategories(GO$BP)[[1]][1:3], GOTERM)
+mget(na.omit(sigCategories(GO$BP)[[1]][1:3]), GOTERM)
 @
@@
 
+<<conditions-plots-dysreg-real,width=10,height=8,echo=FALSE,results=hide>>=
+png("cond-dysreg.png", width=800, height=400)
+<<conditions-plots-dysreg>>
+dev.off()
+@
+
 \begin{figure}
-<<conditions-plots-dysreg-real,fig=TRUE,width=10,height=8,echo=FALSE,results=hide>>=
-<<conditions-plots-dysreg>>
-@
+\includegraphics[width=\textwidth]{cond-dysreg.png}
 \caption{Condition plots for the best separator module found with
@@
 
+<<B1-cond-plots-real,echo=FALSE,results=hide,width=10,height=8>>=
+png("cond-dysreg2.png", width=800, height=400)
+<<B1-cond-plots>>
+dev.off()
+@ 
 \begin{figure}
 \centering
-<<B1-cond-plots-real,fig=TRUE,echo=FALSE,results=hide,width=10,height=8>>=
-<<B1-cond-plots>>
-@ 
+\includegraphics[width=\textwidth]{cond-dysreg2.png}
 \caption{Condition plots for the modules that best separate the B1-type 

### 2011-10-18 eisa/R/biclust.R

@@ -1,23 +1,29 @@
 
+annotate <- function(biclusters, data) {
+
+  if (!inherits(biclusters, "Biclust")) {
+    stop("`biclusters' must be a Biclust")
+  }
+  
+  if (!inherits(ALL, "ExpressionSet")) {
+    stop("`data' must be an ExpressionSet")
+  }
+
+  par <- biclusters@Parameters
+  par$annotation <- annotation(data)
+  par$featureNames <- featureNames(data)
+  par$sampleNames <- sampleNames(data)
+  par$pData <- pData(data)
+  biclusters@Parameters <- par
+
+  biclusters
+}
+
 setAs(from="Biclust", to="ISAModules", def=function(from) {
 
-  feat <- rownames(from@Parameters$Data)
-  samp <- colnames(from@Parameters$Data)
-  anno <- from@Parameters$annotation
-  
-  if (is.null(feat)) {
-    warning("No feature names in `Biclust' object")
-  }
-  if (is.null(samp)) {
-    warning("No sample names in `Biclust' object")
-  }
-  if (is.null(anno)) {
-    warning("No `annotation' in `Biclust' object")
-  }
-  if (is.null(from@Parameters$pData) && !is.null(samp)) {
-    from@Parameters$pData <- data.frame(row.names=samp)
-  }
-  
+  feat <- from@Parameters$featureNames
+  samp <- from@Parameters$sampleNames
+
   new("ISAModules",
       genes=structure(from@RowxNumber, dimnames=list(feat, NULL)),
       conditions=structure(t(from@NumberxCol), dimnames=list(samp, NULL)),

### 2011-10-18 RExpressionView/inst/doc/ExpressionView.tutorial.Rnw
EV updates from SVN, version number increased.

@@
 with \Rclass{ISAModules} or \Rclass{Biclust} objects as the first argument:
-<<expressionview, results=HIDE>>=
+<<expressionview, results=hide>>=
 library(ExpressionView)

### 2011-10-18 RExpressionView/R/zzz.R

@@ -1,5 +1,5 @@
-.onLoad <- function(dir, package) {
-    library.dynam("ExpressionView", package, dir, local=FALSE);
+.onLoad <- function(libname, pkgname) {
+    library.dynam("ExpressionView", pkgname, libname, local=FALSE);
 }
 
 .onUnload <- function(libpath) {

### 2012-09-22 isa2/R/isa3.R
Fixed a bug in isa2 sweep, when there are no modules left at some level.

@@ -469,7 +469,7 @@ isa.sweep.default <- function(data, normed.data, isaresult, method=c("cor"),
       tmpures$columns <- tmpures$columns[,keep,drop=FALSE]
       tmpures$seeddata <- tmpures$seeddata[keep,,drop=FALSE]
       tmpures$seeddata$rob <- rob[keep]
-      tmpures$seeddata$rob.limit <- max(isalist.seed[[i]]$rob.limit)
+      tmpures$seeddata$rob.limit[] <- max(isalist.seed[[i]]$rob.limit)
     }
 
     if (neg.cor) { ABS <- abs } else { ABS <- function(x) x }

### 2012-09-22 eisa/R/biclust.R
Fixed an eisa bug in 'Biclust' conversion.

@@ -5,7 +5,7 @@ annotate <- function(biclusters, data) {
     stop("`biclusters' must be a Biclust")
   }
   
-  if (!inherits(ALL, "ExpressionSet")) {
+  if (!inherits(data, "ExpressionSet")) {
     stop("`data' must be an ExpressionSet")
   }
 

### 2012-09-22 eisa/R/eisa.R
Fixed an eisa sweep graph plot bug, when layout coordinates only differ by eps, because of numerical inaccuracies in the layout algorithm.

@@ -207,7 +207,7 @@ ISASweepGraphPlot <- function(graph, vertex.label=V(graph)$id,
   
   if (thresholds) {
     la <- sort(unique(V(graph)$thr))
-    at <- sort(unique(graph$layout[,1]))
+    at <- sort(unique(round(graph$layout[,1], 3)))
     axis(1, at=at, labels=la)
     axis(3, at=at, labels=la)
     par(xpd=FALSE)

### 2012-09-22 eisa/inst/doc/EISA_tutorial.Rnw
Removed the unneeded loading of the 'affy' package in an eisa vignette.

@@
 library(hgu95av2.db)
-library(affy)
 data(ALL)

### 2012-09-22 eisa/R/autogen.R
Eliminated some eisa check NOTEs, coming from incomplete argument names.

@@ -449,7 +449,7 @@ isa.autogen.module <- function(eset, modules, which, target.dir, template,
   m <- which
 
   ep <- expPlotCreate(eset, modules, m, norm="sample")
-  png(file=paste(sep="", target.dir, "/expression-", m, ".png"),
+  png(filename=paste(sep="", target.dir, "/expression-", m, ".png"),
       width=ep$width, height=ep$height)
   ## returns the box coordinates of the expression image
   bbox <- expPlot(ep)
@@ -458,7 +458,7 @@ isa.autogen.module <- function(eset, modules, which, target.dir, template,
   dev.off() 
 
   ## Color bar
-  png(file=paste(sep="", target.dir, "/expcolbar-", m, ".png"),
+  png(filename=paste(sep="", target.dir, "/expcolbar-", m, ".png"),
       width=840, height=56)
   expPlotColbar(ep)
   dev.off()
@@ -564,7 +564,7 @@ isa.autogen.module <- function(eset, modules, which, target.dir, template,
     gop <- gograph(data.frame(pval),
                    colbar.length=20, label.cex=1,
                    go.terms=go.terms, GOGRAPHS=GOGRAPHS)
-    png(file=filename, width=gop$width*4, height=gop$height*4)
+    png(filename=filename, width=gop$width*4, height=gop$height*4)
     co <- gographPlot(gop, coords=TRUE)
     dev.off()
     list(graph=gop, coords=co)
@@ -720,7 +720,7 @@ isa.autogen.module <- function(eset, modules, which, target.dir, template,
   ## Create the condition plot
 
   if (condPlot) {
-    png(file=paste(sep="", target.dir, "/condplot-", m, ".png"),
+    png(filename=paste(sep="", target.dir, "/condplot-", m, ".png"),
         width=1200, height=400)
     condPlot(modules, number=m, eset=eset,
              col=cond.col, sep=sep)

### 2012-09-22 eisa/R/viz.R

@@ -396,10 +396,10 @@ print.ISAexpPlot <- function(x, ...) {
 
 expPlotColbar <- function(epo) {
   par(mar=c(1,1,2,1))
-  image( matrix((-30):30, nc=1), col=epo$colbar, axes=FALSE)
+  image( matrix((-30):30, ncol=1), col=epo$colbar, axes=FALSE)
   at <- seq(0, 60, length=13)/60
   label <- round(seq(epo$zlim[1], epo$zlim[2], length=13), 2)
-  axis(3, at=at, label=label, tick=FALSE, line=-1, cex.axis=1.5)
+  axis(3, at=at, labels=label, tick=FALSE, line=-1, cex.axis=1.5)
   abline(v=(0:61-0.5)/60)
   abline(h=c(-1,1))
   invisible(NULL)
@@ -409,7 +409,7 @@ expPlot <- function(epo, scores=TRUE) {
 
   if (scores) {
   
-    layout( matrix(c(5,2,4, 6,1,3, 7,8,9), nc=3, byrow=TRUE),
+    layout( matrix(c(5,2,4, 6,1,3, 7,8,9), ncol=3, byrow=TRUE),
            c( 50/epo$width, epo$exp.width/epo$width, 100/epo$width ),
            c( 70/epo$height, epo$exp.height/epo$height, 50/epo$height) )
 
@@ -426,7 +426,7 @@ expPlot <- function(epo, scores=TRUE) {
   label1 <- unique(round(seq(1, no.genes, length=10)))
   at1 <- (label1-1) / (no.genes-1)
   if (length(at1) != 1) {
-    axis(1, at=at1, label=label1, cex.axis=2)
+    axis(1, at=at1, labels=label1, cex.axis=2)
   }
   if (epo$cond.height.px >= 10) {
     label2 <- seq(1, no.conds, by=2)
@@ -435,7 +435,7 @@ expPlot <- function(epo, scores=TRUE) {
   }
   at2 <- (no.conds-label2) / (no.conds-1)
   if (any(is.finite(at2))) {
-    axis(2, at=at2, label=label2, cex.axis=2, tick=FALSE)
+    axis(2, at=at2, labels=label2, cex.axis=2, tick=FALSE)
   }
 
   usr <- par("usr")
@@ -448,13 +448,13 @@ expPlot <- function(epo, scores=TRUE) {
     if (!is.finite(pp)) pp <- 0.5
     
     par(mar=c(1,0,1,0))
[…]

### 2013-09-30 eisa/inst/doc/EISA_tutorial.Rnw
eisa: work out imports and depends better.

@@
 <<filtering>>=
+library(genefilter)
 varLimit <- 0.5

### 2013-09-30 eisa/R/KEGG.R

@@ -77,7 +77,6 @@ ISAKEGG <- function(modules,
   isa2:::isa.status("Calculating KEGG enrichment", "in")
   
   library(paste(sep="", ann, ".db"), character.only=TRUE)
-  library(Category)
   library(KEGG.db)
 
   ENTREZ <- get(paste(sep="", ann, "ENTREZID"))

### 2013-09-30 eisa/R/autogen.R

@@ -235,7 +235,6 @@ ISAHTMLModules <- function(eset, modules, which=NULL,
 
   isa2:::isa.status("Generating module pages", "in")
 
-  library(Biobase)
   library(GO.db)
   library(KEGG.db)
   library(igraph)

### 2013-09-30 eisa/R/eisa.R

@@ -57,7 +57,6 @@ ISA <- function(data,
   }
 
   if (is.function(flist)) {
-    library(genefilter)
     selected <- genefilter(data, flist)
     data <- data[selected,]
   } else if (! is.na(flist)) {

### 2013-09-30 eisa/R/generalenrichment.R

@@ -54,7 +54,6 @@ ISAEnrichment <- function(modules, categories,
   }
   
   library(paste(sep="", ann, ".db"), character.only=TRUE)
-  library(Category)
 
   ENTREZ <- get(paste(sep="", ann, "ENTREZID"))
 

### 2013-09-30 eisa/R/miRNA.R

@@ -76,7 +76,6 @@ ISAmiRNA <- function(modules,
   library(paste(sep=".", "targetscan", short.organism, "eg.db"),
                 character.only=TRUE)
   library(paste(sep="", ann, ".db"), character.only=TRUE)
-  library(Category)
 
   ENTREZ <- get(paste(sep="", ann, "ENTREZID"))
 

### 2014-03-26 eisa/inst/doc/EISA_biclust.Rnw
Fix EISA biclust vignette for new biclust
Because it creates different biclusters and we
ended up with an empty set.

@@
 GO <- ISAGO(modules)
-gog <- gograph(summary(GO$CC)[[2]])
+gog <- gograph(summary(GO$CC)[[3]])
 summary(gog)

### 2015-04-08 isa2/R/biclust.R
Various fixes for isa2 R CMD check

@@ -1,7 +1,7 @@
 
 isa.biclust <- function(modules) {
 
-  if (!require(biclust)) {
+  if (!requireNamespace("biclust")) {
     stop("The `biclust' package is required for this")
   }
 

### 2015-04-08 isa2/R/isa3.R

@@ -537,7 +537,9 @@ sweep.graph.default <- function(sweep.result) {
     stop("Not a sweep result")
   }
 
-  require(igraph)
+  if (!requireNamespace("igraph")) {
+    stop("The igraph package is required for sweeping")
+  }
 
   nnodes <- nrow(sweep.result$seeddata)
 
@@ -548,25 +550,26 @@ sweep.graph.default <- function(sweep.result) {
   from <- from[valid]
   to <- to[valid]
 
-  G <- graph( rbind(from, to), n=nnodes )
+  G <- igraph::graph( rbind(from, to), n=nnodes )
 
   if (length(unique(sweep.result$seeddata$thr.row)) == 1) {
-    V(G)$thr <- sweep.result$seeddata$thr.col
+    igraph::V(G)$thr <- sweep.result$seeddata$thr.col
   } else {
-    V(G)$thr <- sweep.result$seeddata$thr.row
+    igraph::V(G)$thr <- sweep.result$seeddata$thr.row
   }
 
-  V(G)$id <- seq(vcount(G))
-  graphs <- decompose.graph(G)
+  igraph::V(G)$id <- seq(igraph::vcount(G))
+  graphs <- igraph::decompose.graph(G)
 
   layouts <- lapply(graphs, function(g) {
-    l <- layout.reingold.tilford(g, root=tail(topological.sort(g, mode="out"), 1))
+    l <- igraph::layout.reingold.tilford(
+      g, root=tail(igraph::topological.sort(g, mode="out"), 1))
     r <- sqrt(l[,1]^2 + l[,2]^2)
     phi <- atan2(l[,2], l[,1]) - pi/2
     l[,1] <- r * cos(phi)
     l[,2] <- r * sin(phi)
-    labels <- sort(unlist(V(g)$thr))
-    l <- layout.norm(l, labels[1]-2, labels[length(labels)]-2, NULL, NULL)
+    labels <- sort(unlist(igraph::V(g)$thr))
+    l <- igraph::layout.norm(l, labels[1]-2, labels[length(labels)]-2, NULL, NULL)
     l[,2] <- l[,2] - min(l[,2])
     l})
 
@@ -578,20 +581,23 @@ sweep.graph.default <- function(sweep.result) {
     offs <- offs + r[2] - r[1]
   }
   offs <- offs-0.5
-  
-  G$layout <- do.call(rbind, layouts)
[…]

### 2015-04-08 isa2/R/viz.R

@@ -32,8 +32,6 @@ plotModules.default <- function(modules, to.plot=seq_len(ncol(modules$rows)),
                                 xlab="", ylab="",
                                 ...) {
 
-  require(lattice)
-  
   no.mods <- length(to.plot)
   no.rows <- nrow(modules$rows)
   no.cols <- nrow(modules$columns)
